---
title: "Stochastic Gradient Descent"
---

{{< include _common.qmd >}}

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(mvtnorm)

set.seed(1422)

source(here::here("R", "sgd.R"))
```

## Last Time

### A Deeper Look Into the EM Algorithm

We looked at convergence properties and the generalized EM algorithm.

. . .

### Fisher Information

We talked about computing the Fisher information and how it relates to the EM
algorithm.

## Today

### Stochastic Gradient Descent

How can we optimize a loss when each full gradient is expensive?

\medskip\pause

We will see why a noisy gradient can be useful, when its noise matters, and how
the learning rate controls it.

# Stochastic Gradient Descent

## Minimizing Average Loss

Many problems in statistics and machine learning have an objective that is an
average of observation-level losses:
$$
  H_n(\theta) = \frac{1}{n} \sum_{i=1}^n L(y_i, x_i, \theta),
$$
where $(y_i, x_i)$, $i = 1,\ldots,n$, are the observed data.

\pause

### Examples

- Least squares:
  $L(y_i, x_i, \beta) = \frac{1}{2} (y_i - \beta^\intercal x_i)^2$
- Logistic regression:
  $L(y_i, x_i, \beta) = -y_i \beta^\intercal x_i + \log\left( 1 + e^{\beta^\intercal x_i} \right)$
- Neural networks: $L$ is the loss of the network on observation $i$

## Gradient Descent, and Its Price

::: {.columns align="center"}
::::: {.column width="47%"}
Because $H_n$ is an average of observation-level losses, its gradient is also an
average:
$$
  \nabla H_n(\theta_k)
  = \frac{1}{n}\sum_{i=1}^n \nabla_\theta L(y_i,x_i,\theta_k).
$$

\medskip\pause

Gradient descent updates
$$
  \theta_{k+1} = \theta_k - \gamma_k \nabla H_n(\theta_k).
$$
:::::

. . .

::::: {.column width="47%"}
\begin{algorithm}[H]
  \caption{Gradient descent}
  \KwData{$\theta_0$, learning rates $\{\gamma_k\}$}
  \For{$k = 0,1,\ldots$}{
    \colorbox{mLightBrown!20}{%
      $\displaystyle \rho_k \gets \frac{1}{n}\sum_{i=1}^n \nabla_\theta L(y_i,x_i,\theta_k)$%
    }\;
    $\theta_{k+1} \gets \theta_k - \gamma_k\rho_k$\;
  }
\end{algorithm}

\medskip\pause

### The Problem {.alert}

Computing $\nabla H_n$ means touching **every** observation. Each contribution
costs $O(p)$, so one iteration costs $O(np)$.

\medskip\pause

*Can we estimate this gradient without computing every term?*
:::::
:::

## Stochastic Gradient Descent

::: {.columns}
::::: {.column width="47%"}
Sample $J_k$ uniformly from $\{1,\ldots,n\}$, compute
$$
  \rho_k = \nabla_\theta L(y_{J_k},x_{J_k},\theta_k),
$$
and update $\theta_{k+1} = \theta_k - \gamma_k\rho_k$.

\bigskip\pause

\begin{algorithm}[H]
  \caption{SGD}
  \KwData{$\theta_0$, learning rates $\{\gamma_k\}$}
  \For{$k = 0,1,\ldots$}{
    \colorbox{mLightGreen!20}{%
      $\displaystyle J_k \sim \operatorname{Uniform}\{1,\ldots,n\}$%
    }\;
    \colorbox{mLightGreen!20}{%
      $\displaystyle \rho_k \gets \nabla_\theta L(y_{J_k},x_{J_k},\theta_k)$%
    }\;
    $\theta_{k+1} \gets \theta_k - \gamma_k\rho_k$\;
  }
\end{algorithm}
:::::

::::: {.column width="47%"}
\pause

### Unbiasedness

Each step follows the true gradient *on average*:
$$
  \E(\rho_k \mid \theta_k) = \nabla H_n(\theta_k).
$$

\medskip\pause

### Iteration cost

The cost falls from $O(np)$ to $O(p)$.
:::::
:::

## Basic Implementation of SGD

```{r}
#| label: sgd-basic

sgd_basic <- function(
  par,
  grad,
  N,                                      # number of data points
  gamma,                                  # learning rate
  sampler = function(n) sample.int(n, 1), # sample one index
  maxit = 100,
  ...
) {
  for (k in seq_len(maxit)) {
    for (i in seq_len(N)) {
      ind <- sampler(N)
      par <- par - gamma * grad(par, ind, ...)
    }
  }
  par
}
```

## Example: Ordinary Least Squares

The empirical risk is
$$
  H_N(\beta) = \frac{1}{N} \sum_{i=1}^N L(y_i,x_i,\beta),
$$
where
$$
  L(y_i,x_i,\beta) = \frac{1}{2} (y_i - \beta^T x_i)^2.
$$

And the gradient is
$$
  \nabla_\beta L(y_i,x_i,\beta) = -x_i(y_i - \beta^T x_i).
$$

## Example: Logistic Regression

::: {.columns align="center"}
::::: {.column width="47%"}
With $y_i \in \{0, 1\}$, the logistic regression loss is
$$
  L(y_i,x_i,\beta) = -y_i \beta^T x_i + \log\left( 1 + e^{\beta^T x_i} \right)
$$
and the gradient is
$$
  \nabla_\beta L(y_i,x_i,\beta)
  = x_i \left( \frac{e^{\beta^T x_i}}{1 + e^{\beta^T x_i}} - y_i \right).
$$

\medskip\pause

Gradient descent (GD) follows a smooth path to the minimizer. How do you think
SGD will behave?
:::::

::::: {.column width="47%"}
```{r}
#| label: logistic-regression-paths
#| echo: false
#| fig-show: animate
#| fig-height: 3.4
#| fig-width: 3.0
#| dev.args: { bg: white, pointsize: 9 }

set.seed(123)

N <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(N, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(N, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())

loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

beta1 <- seq(-1, 3, length = 100)
beta2 <- seq(-1, 3, length = 100)

loss_2d_vectorized <- Vectorize(function(b1, b2) {
  loss_2d(b1, b2, X, y)
})

z <- outer(beta1, beta2, loss_2d_vectorized)

res_sgd <- logreg_sgd(X, y, max_epochs = 100, batch_size = 1)
res_sgd2 <- logreg_sgd(
  X,
  y,
  max_epochs = 100,
  batch_size = 1,
  K = 5,
  a = 0.5
)
res_gd <- logreg_sgd(X, y, max_epochs = 100, batch_size = N)

loss_optim <- loss_2d(
  res_glm$coefficients[1],
  res_glm$coefficients[2],
  X,
  y
)

res_gd <- logreg_sgd(X, y, max_epochs = 100, batch_size = N)
loss1 <- loss_2d_vectorized(
  res_gd$beta_history[1, ],
  res_gd$beta_history[2, ]
)
loss2 <- loss_2d_vectorized(
  res_sgd$beta_history[1, ],
  res_sgd$beta_history[2, ]
)
loss3 <- loss_2d_vectorized(
  res_sgd2$beta_history[1, ],
  res_sgd2$beta_history[2, ]
)

n1 <- length(loss1)
n2 <- length(loss2)
n3 <- length(loss3)

pal <- palette.colors(palette = "Okabe-Ito")

sgd_reveals <- c(0L, 1L, 5L, 20L, 100L, 1000, 10000)

for (steps in sgd_reveals) {
  contour(
    beta1,
    beta2,
    z,
    col = "dark grey",
    drawlabels = FALSE,
    asp = 1,
    main = sprintf("SGD updates: %d", steps)
  )

  lines(
    res_gd$beta_history[1, ],
    res_gd$beta_history[2, ],
    col = pal[1]
  )
  points(
    res_gd$beta_history[1, ],
    res_gd$beta_history[2, ],
    col = pal[1],
    pch = 19,
    cex = 0.5
  )

  if (steps > 0L) {
    path <- seq_len(steps + 1L)
    lines(
      res_sgd$beta_history[1, path],
      res_sgd$beta_history[2, path],
      col = pal[3]
    )
    points(
      res_sgd$beta_history[1, steps + 1L],
      res_sgd$beta_history[2, steps + 1L],
      col = pal[3],
      pch = 19,
      cex = 0.9
    )
  }

  points(
    res_glm$coefficients[1],
    res_glm$coefficients[2],
    col = pal[2],
    pch = 19,
    cex = 1
  )

  legend(
    "bottomright",
    legend = if (steps > 0L) c("SGD", "GD") else "GD",
    col = if (steps > 0L) pal[c(3, 1)] else pal[1],
    pch = 19,
    cex = 0.8
  )
}
```
:::::
:::

## The Same Runs, Counted by Iterations and Epochs

::: {.columns}
::::: {.column width="47%"}
Over the first 100 iterations, GD reduces the loss much faster.

```{r}
#| label: convergence-by-updates
#| echo: false
#| fig-cap: "The first 100 updates of the same GD and SGD runs."

conv_data <- tibble(
  method = rep(c("GD", "SGD"), times = c(n1, n2)),
  iteration = c(seq_len(n1) - 1L, seq_len(n2) - 1L),
  observations = c((seq_len(n1) - 1L) * N, seq_len(n2) - 1L),
  loss = c(loss1 - loss_optim, loss2 - loss_optim)
)

ggplot(
  dplyr::filter(conv_data, iteration <= 100),
  aes(iteration, loss, color = method)
) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Iterations", y = "Suboptimality", color = NULL)
```
:::::

. . .

::::: {.column width="47%"}
One GD iteration or $N$ single-observation SGD iterations: one **epoch** for
either method.

```{r}
#| label: convergence-by-epochs
#| echo: false
#| fig-cap: "The same GD and SGD runs, measured in epochs."

ggplot(
  dplyr::mutate(conv_data, epoch = observations / N),
  aes(epoch, loss, color = method)
) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Epochs", y = "Suboptimality", color = NULL)
```
:::::
:::

## What Does an Epoch See?

For single-observation SGD, one epoch means $N$ updates. It does not necessarily
mean that every observation was used once.

\medskip\pause

```{r}
#| label: sgd-epoch-sampling
#| echo: false
#| fig-width: 5
#| fig-height: 1.6
#| fig-cap: "Two ways to select eight observations over one epoch."

draws <- strsplit(c("ACCHDAFB", "DAGBHECF"), "")
names(draws) <- c("Independent draws", "Shuffle once")
colors <- setNames(
  palette.colors(8, palette = "Okabe-Ito"),
  LETTERS[1:8]
)

old_par <- par(mar = c(0, 0, 0, 0))
plot.new()
plot.window(xlim = c(-2.3, 8.5), ylim = c(0.4, 2.6))

for (row in seq_along(draws)) {
  y <- 3 - row
  text(-2.2, y, names(draws)[row], adj = 0, cex = 0.8)

  for (i in seq_along(draws[[row]])) {
    observation <- draws[[row]][i]
    rect(
      i - 0.45,
      y - 0.4,
      i + 0.35,
      y + 0.4,
      col = colors[observation],
      border = NA
    )
    text(
      i - 0.05,
      y,
      observation,
      col = ifelse(
        observation %in% c("A", "F"),
        "white",
        "black"
      )
    )
  }
}

par(old_par)
```

\medskip\pause

Independent draws may revisit some observations and miss others. Shuffling the
data once per epoch uses each observation exactly once.

## Unbiased Does Not Mean Noiseless

Suppose $\theta_k = \theta^*$, so the **full** gradient is zero. Gradient
descent would stop there.

\bigskip\pause

What about SGD?

\bigskip\pause

No, because although the stochastic gradient is unbiased, it is still noisy:
$$
  \E(\rho_k) = 0, \qquad \E\bigl(\lVert\rho_k\rVert^2\bigr) = \sigma_\star^2
                                                            > 0.
$$

. . .

With a fixed learning rate $\gamma > 0$,
$$
  \theta_{k+1} - \theta^\star = -\gamma\rho_k,
$$
so
$$
  \E\bigl(\lVert\theta_{k+1} - \theta^*\rVert^2 \mid \theta_k = \theta^*\bigr)
  = \gamma^2\sigma_\star^2 > 0.
$$

\medskip\pause

The next iterate is correct on average, but it still moves away from the
minimizer.

## Learning Rate

::: {.columns align = "center"}
::::: {.column width="47%"}
With a suitable fixed step length, GD can converge to the minimizer.

\medskip\pause

When observation gradients remain noisy at the minimizer, fixed-rate SGD keeps
fluctuating around it.

\medskip

We **can** improve this, but first we need to introduce the concept of **strong
convexity**.
:::::

::::: {.column width="47%"}
```{r}
#| label: fixed-learning-rate
#| fig-cap: "Convergence of SGD for logistic regression with a fixed learning rate."
#| fig-height: 2.6
#| echo: false

L <- 0.25 * norm(crossprod(X), "2") / N
gamma_fixed <- 0.05 / L
fixed_updates <- 30L * N
record_every <- 100L
fixed_beta <- numeric(p)
fixed_loss <- numeric(fixed_updates / record_every + 1L)
fixed_loss[1] <- loss_2d(fixed_beta[1], fixed_beta[2], X, y)

set.seed(481)
for (k in seq_len(fixed_updates)) {
  ind <- sample.int(N, 1L)
  x_ind <- X[ind, ]
  gradient <- x_ind * (plogis(sum(x_ind * fixed_beta)) - y[ind])
  fixed_beta <- fixed_beta - gamma_fixed * gradient

  if (k %% record_every == 0L) {
    fixed_loss[k / record_every + 1L] <- loss_2d(
      fixed_beta[1],
      fixed_beta[2],
      X,
      y
    )
  }
}

fixed_data <- dplyr::bind_rows(
  tibble(
    method = "Fixed-rate SGD",
    epoch = seq(0, fixed_updates, by = record_every) / N,
    loss = pmax(fixed_loss - loss_optim, 1e-12)
  ),
  tibble(
    method = "GD",
    epoch = seq_len(n1) - 1L,
    loss = pmax(loss1 - loss_optim, 1e-12)
  )
)

ggplot(fixed_data, aes(epoch, loss, color = method)) +
  geom_line() +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 30)) +
  labs(x = "Epochs", y = "Suboptimality", color = NULL) +
  theme(legend.position = c(0.6, 0.9))
```
:::::
:::

## Strong Convexity

A differentiable function $H: \mathbb{R}^p \to \mathbb{R}$ is **strongly
convex** if
$$
  H(\theta) \geq H(\theta') + \nabla H(\theta')^T (\theta - \theta')
                 + \frac{l}{2} \| \theta - \theta'\|_2^2
$$
for all $\theta,\theta' \in \mathbb{R}^p$ and some $l > 0$.

\medskip\pause

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        width=7cm, height=4.3cm,
        axis lines=none,
        xlabel={$\theta$},
        ylabel={$H(\theta)$},
        ymin=-0.5, ymax=4,
        xmin=-1, xmax=2.2,
        samples=200,
        domain=-0.8:2.2,
        clip=false
      ]
      % f(x) = x^2
      \addplot[thick] {x^2};
      % Tangent at x0=1: 2*(x-1) + 1
      \addplot[DarkOrange, dashed, thick] {2*(x-1) + 1};
      % Strong convexity lower bound: tangent + 0.5*(x-1)^2
      \addplot[RoyalBlue, dotted, thick] {2*(x-1) + 1 + 0.5*(x-1)^2};
      % Point at (1,1)
      \addplot[only marks, mark=*] coordinates {(1,1)};
      % Annotations
      \node[anchor=west] at (axis cs:2.2,5) {$H(\theta)$};
      \node[anchor=west, RoyalBlue] at (axis cs:2.2,4.3) {$H(\theta') + \nabla H(\theta')^T(\theta-\theta') + \frac{l}{2}\|\theta-\theta' \|^2$};
      \node[anchor=west, DarkOrange] at (axis cs:2.2,3.5) {$H(\theta') + \nabla H(\theta')^T(\theta-\theta')$};
    \end{axis}
  \end{tikzpicture}
\end{figure}

## Example: Strong Convexity

The function
$$
  H(\theta) = \theta^2
$$
\pause is strongly convex with $l = 2$ since $$
\begin{aligned}
H(\theta) - H(\theta') - \nabla H(\theta')^T(\theta - \theta') & = \theta^2 - (\theta')^2 - 2\theta'(\theta - \theta') \\ \pause
                                         & = (\theta - \theta')^2 \\ \pause
                                         & \geq \frac{l}{2}(\theta - \theta')^2.
\end{aligned}
$$

\medskip\pause

The inequality holds for $l \leq 2$, and in particular for $l = 2$.

## Relationship to Lipschitz Gradient

If $H$ is strongly convex with parameter $l$ and has Lipschitz gradient with
constant $L$, then
$$
  l I \preceq D^2H(\theta) \preceq L I.
$$

\medskip\pause

- Strong convexity provides a **lower** bound on the curvature of $H$\pause
- Lipschitz gradient provides an **upper** bound on the curvature of $H$

. . .

### Example

The function $H(\theta) = \theta^2$ is strongly convex with $l = 2$ and has
Lipschitz gradient with $L = 2$, since $D^2H(\theta) = 2$.

\medskip\pause

This is *identical* to the strong convexity parameter.

## Example of Strong Convexity--Lipschitz Gap

::: {.columns}
::::: {.column width="47%"}
Let's take the function
$$
  H(\theta) = \theta^2 + \sin(\theta).
$$

\medskip\pause

This is twice differentiable with
$$
  D^2H(\theta) = 2 - \sin(\theta).
$$

\medskip\pause

And since
$$
  1 \leq 2 - \sin(x) \leq 3,
$$
$H$ is strongly convex with $l = 1$ and has Lipschitz gradient with $L = 3$.

\medskip\pause
:::::

::::: {.column width="47%"}
\begin{figure}
  \begin{tikzpicture}
    \pgfmathsetmacro{\a}{1}
    \pgfmathsetmacro{\xzero}{4.712389} % 3*pi/2
    \pgfmathsetmacro{\mu}{2 - \a}      % =1
    \pgfmathsetmacro{\Lips}{2 + \a}    % =3
    \pgfmathsetmacro{\fz}{\xzero*\xzero + sin(deg(\xzero))}
    \pgfmathsetmacro{\dfz}{2*\xzero + cos(deg(\xzero))}
    \begin{axis}[
        width=5cm,
        height=6.2cm,
        domain=-0.5:12.5,
        clip=false,
        samples=200,
        axis lines=none,
        xlabel={$\theta$},
        ylabel={$H(\theta)$},
        legend style={draw=none,at={(0.02,0.98)},anchor=north west}
      ]
      % f
      \addplot[thick, line label={$H(\theta)$}] {x^2 + sin(deg(x))};
      % lower quadratic
      \addplot[
        DarkOrange,
        dashed,
        thick,
        line label={Strong convexity bound},
      ] {
        \fz + \dfz*(x-\xzero) + 0.5*\mu*(x-\xzero)^2
      };
      % upper quadratic
      \addplot[
        RoyalBlue,
        dashed,
        thick,
        line label={Lipschitz bound}
      ] {
        \fz + \dfz*(x-\xzero) + 0.5*\Lips*(x-\xzero)^2
      };
      % point
      \addplot[only marks,mark=*] coordinates {(\xzero,\fz)};
    \end{axis}
  \end{tikzpicture}
  \caption{The function $H(\theta) = \theta^2 + \sin(\theta)$ is strongly convex with $l = 1$
    and has Lipschitz gradient with $L = 3$.}
\end{figure}
:::::
:::

## Conditioning

The **condition number** of a strongly convex function with Lipschitz gradient
is
$$
  \kappa = \frac{L}{l} \geq 1.
$$

. . .

The closer $\kappa$ is to $1$, the better conditioned the problem is.

. . .

### Example

For $H(\theta) = \theta^2$, we have $\kappa = 2/2 = 1$, so the problem is
perfectly conditioned.

\medskip\pause

We reach the minimum in one step with gradient descent and step length
$t = 1/L = 1/2$.

## Why Scale Predictors Before SGD?

When predictors have very different scales, the loss can be steep in one
direction and flat in another. One learning rate must work for both directions.

\medskip\pause

```{r}
#| label: sgd-predictor-scaling
#| echo: false
#| fig-width: 5
#| fig-height: 2

grid <- seq(-2, 2, length.out = 101)
contour_levels <- c(0.25, 0.5, 1, 2, 3)
old_par <- par(mfrow = c(1, 2), mar = c(2.2, 2.2, 1.5, 0.5))

contour(
  grid,
  grid,
  outer(grid, grid, function(x, y) ((3 * x)^2 + y^2) / 2),
  levels = contour_levels,
  drawlabels = FALSE,
  col = "royalblue",
  asp = 1,
  xlab = "",
  ylab = "",
  main = "Original predictors"
)
contour(
  grid,
  grid,
  outer(grid, grid, function(x, y) (x^2 + y^2) / 2),
  levels = contour_levels,
  drawlabels = FALSE,
  col = "royalblue",
  asp = 1,
  xlab = "",
  ylab = "",
  main = "Scaled predictors"
)

par(old_par)
```

\medskip\pause

Scaling can make the contours more nearly round and the learning rate easier to
choose. Use training-data scales for new observations too.

## The Robbins--Monro Convergence Theorem

Let $H(\theta) = \E L(Y,\theta)$ be the population risk. Suppose $H$ is strongly
convex and
$$
  \E\left( \|\nabla_\theta L(Y,\theta)\|_2^2 \right) \leq A + B \| \theta \|_2^2
$$
for some constants $A$ and $B$.

\medskip\pause

If $\theta^*$ is the global minimizer of $H$, then $\theta_n$ converges almost
surely toward $\theta^*$ if
$$
  \sum_{n=1}^{\infty} \gamma_n^2
  < \infty \quad \text{and} \quad \sum_{n=1}^{\infty} \gamma_n = \infty.
$$

. . .

The theorem applies to learning rates satisfying
$$
  \gamma_n \propto \frac{1}{n}.
$$

## What Do the Two Sums Mean?

The Robbins--Monro conditions ask the learning rates to do two jobs:
$$
  \sum_n \gamma_n = \infty \quad\text{and}\quad \sum_n \gamma_n^2 < \infty.
$$

\medskip\pause

The first sum keeps the algorithm moving far enough to reach the minimum. The
second makes the accumulated gradient noise finite.

\medskip\pause

For $\gamma_n \propto n^{-a}$, both hold when $1/2 < a \leq 1$. That still
leaves a wide range of finite-run behavior.

## Learning Rate

The learning rate $\gamma_n$ is crucial for SGD, but hard to set.

\medskip\pause

Convergence theorem only gives a few hints.

. . .

### A Class of Decay Schedules

$$
  \gamma_n = \frac{\gamma_0 K}{K + n^a} = \frac{\gamma_0}{1 + K^{-1} n^a}
$$
with initial learning rate $\gamma_0 > 0$ and constants $K, a > 0$.

\medskip\pause

Convergence is ensured by Robbins--Monro if $a \in (0.5, 1]$.

\medskip\pause

Fixing the exponent $a$ and picking a target rate, $\gamma_1$, to be reached
after $n_1$ steps, we can solve for $K$ and get
$$
  K = \frac{n_1^a \gamma_1}{\gamma_0 - \gamma_1}.
$$

##

```{r}
#| label: decay
#| cache: true

decay_scheduler <- function(
  gamma0 = 1,
  a = 1,
  K = 1,
  gamma1 = NULL,
  n1 = NULL,
  t0 = gamma0
) {
  force(a)
  gamma0 <- t0

  if (!is.null(gamma1) && !is.null(n1)) {
    K <- n1^a * gamma1 / (gamma0 - gamma1)
  }

  b <- gamma0 * K

  function(n) b / (K + n^a)
}
```

##

```{r, dependson="decay"}
#| label: decay-fig
#| echo: false
#| fig-height: 3
#| fig-width: 4
#| fig-cap: "Examples of learning rate schedules,
#|   using $\\gamma_n = \\frac{\\gamma_0}{1 + K^{-1} n^{a}}$."

grid_par <- expand.grid(
  n = seq(0, 1000, 2),
  K = c(25, 50, 100),
  a = c(0.6, 1),
  gamma0 = 1
)

grid_par <- dplyr::mutate(
  grid_par,
  rate = gamma0 * K / (K + n^a)
)

ggplot(
  grid_par,
  aes(n, rate, color = factor(K), linetype = factor(a))
) +
  geom_line() +
  scale_color_discrete("K") +
  scale_linetype_discrete("a") +
  guides(
    color = guide_legend(order = 1),
    linetype = guide_legend(order = 2)
  )
```

## So, Is The Problem of Setting Learning Rate Solved?

No, not really, because it's still hard to pick $\gamma_0$, $K$, and $a$.

```{r}
#| label: step-size-illustration
#| echo: false

set.seed(123)

N <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(N, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(N, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())

loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

beta1 <- seq(-1, 3, length = 100)
beta2 <- seq(-1, 3, length = 100)

loss_2d_vectorized <- Vectorize(function(b1, b2) {
  loss_2d(b1, b2, X, y)
})

z <- outer(beta1, beta2, loss_2d_vectorized)

step_size_illustration <- function(a) {
  old_par <- par(mar = c(2.5, 2.5, 0.5, 0.5))
  on.exit(par(old_par))

  res_sgd <- logreg_sgd(
    X,
    y,
    max_epochs = 100,
    batch_size = 1,
    a = a,
    t0 = 0.5
  )

  pal <- palette.colors(palette = "Okabe-Ito")

  contour(
    beta1,
    beta2,
    z,
    col = "dark grey",
    drawlabels = FALSE,
    asp = 1
  )

  lines(
    res_sgd$beta_history[1, ],
    res_sgd$beta_history[2, ],
    col = pal[3]
  )
  points(
    res_sgd$beta_history[1, ],
    res_sgd$beta_history[2, ],
    col = pal[3],
    pch = 19,
    cex = 0.5
  )

  points(
    res_glm$coefficients[1],
    res_glm$coefficients[2],
    col = pal[2],
    pch = 19,
    cex = 1
  )
}
```

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: step-size-illustration-06
#| echo: false
#| dev: png
#| dev.args: { pointsize: 9 }
#| dpi: 300
#| crop: false
#| fig-height: 2.4
#| fig-cap: SGD on the logistic regression example with $a = 0.6$ (slow decay).

step_size_illustration(0.6)
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: step-size-illustration-09
#| echo: false
#| dev: png
#| dev.args: { pointsize: 9 }
#| dpi: 300
#| crop: false
#| fig-height: 2.4
#| fig-cap: SGD on the logistic regression example with $a = 0.9$ (fast decay).

step_size_illustration(0.9)
```
:::::
:::

## Learning Rates in Practice

::: {.columns align="center"}
::::: {.column width="47%"}
The inverse-power schedule is useful for the Robbins--Monro theorem. A finite
training run need not use it.

\medskip\pause

In practice, with a fixed training budget, common choices include:

- A constant rate for a moderate accuracy target.\pause
- Warmup, then linear or cosine decay over a planned run.\pause
- Reduce the rate when validation performance stalls.
:::::

::::: {.column width="47%"}
```{r}
#| label: practical-learning-rate-schedules
#| echo: false
#| fig-height: 2.7
#| fig-cap: "Illustrative planned schedules."

progress <- seq(0, 1, length.out = 201)
warmup <- 0.1
after_warmup <- pmax(0, (progress - warmup) / (1 - warmup))

schedule_data <- tibble(
  progress,
  Constant = 1,
  `Warmup + linear` = pmin(progress / warmup, 1 - after_warmup),
  `Warmup + cosine` = pmin(
    progress / warmup,
    (1 + cos(pi * after_warmup)) / 2
  )
) |>
  pivot_longer(
    -progress,
    names_to = "schedule",
    values_to = "rate"
  )

ggplot(schedule_data, aes(progress, rate, color = schedule)) +
  geom_line() +
  scale_color_manual(
    values = c(
      Constant = "darkorange",
      `Warmup + linear` = "royalblue",
      `Warmup + cosine` = "forestgreen"
    )
  ) +
  theme(legend.position = c(0.75, 0.8)) +
  labs(
    x = "Fraction of training run",
    y = expression(gamma / gamma[max]),
    color = NULL
  )
```
:::::
:::

## Convergence Rates

### Gradient Descent

For convex $H_N$, GD with diminishing step lengths converges at rate
$$
  H_N(\theta_n) - H_N^* = O\left( \frac{1}{\sqrt{n}} \right).
$$

. . .

When $H_N$ is differentiable with Lipschitz gradient, we get
$$
  H_N(\theta_n) - H_N^* = O\left( \frac{1}{n} \right).
$$

. . .

### Stochastic Gradient Descent

For convex $H_N$ with diminishing learning rates, SGD converges at rate
$$
  \E\big(H_N(\theta_n) - H_N^*\big) = O\left( \frac{1}{\sqrt{n}} \right).
$$

. . .

But this **does not improve** when $H_N$ is differentiable with Lipschitz
gradient.

## Rates for Strongly Convex Functions

::: {.columns}
::::: {.column width="47%"}
When $H_N$ is strongly convex and has Lipschitz gradient, GD satisfies
$$
  H_N(\theta_n) - H_N^* = O(\delta^n)
$$
with $\delta \in (0, 1)$ and $\delta = 1 - l/L$.

\medskip\pause

Under the **same conditions**, SGD gives us
$$
  \E\big(H_N(\theta_n) - H_N^*\big) = O\left( \frac{1}{n} \right).
$$

. . .

So SGD does **not** enjoy linear convergence rates under strong convexity.
:::::

::::: {.column width="47%"}
\begin{figure}
\begin{tikzpicture}
  \begin{axis}[
      width=6.5cm, height=6.7cm,
      xlabel={$n$},
      ylabel={Suboptimality},
      legend pos=north east,
      ymode=log,
      xmax=36,
      domain=1:20
    ]
    % O(1/sqrt(n))
    \addplot[
      thick,
      line label={$O(1/\sqrt{n})$}
    ] {1/sqrt(x)};

    % O(1/n)
    \addplot[
      thick,
      dashed,
      line label={$O(1/n)$}
    ] {1/x};

    % O(delta^n) with delta = 0.5
    \addplot[
      thick,
      dotted,
      samples=50,
      line label={$O(\delta^n)$}
    ] {0.5^x};
  \end{axis}
\end{tikzpicture}
\caption{Illustration of convergence rates. Here, $\delta = 0.5$.}
\end{figure}
:::::
:::

## Another Way to Control Noise

::: {.columns}
::::: {.column width="47%"}
Decay controls SGD's noise, but choosing the learning-rate schedule is
difficult.

\medskip\pause

Could averaging several observation gradients make this choice less delicate?

\medskip\pause

Larger batches lower the noise floor, but process more observations per update.
:::::

::::: {.column width="47%"}
```{r}
#| label: fixed-rate-batch-noise
#| echo: false
#| fig-cap: "Mean excess loss (100 runs, $N = 1000$, fixed $\\gamma = 0.2$)."
#| fig-height: 2.7

set.seed(820)

batch_demo_y <- rnorm(1000)
batch_demo_target <- mean(batch_demo_y)
batch_demo_sizes <- c(1L, 10L, 100L)
batch_demo_updates <- 120L
batch_demo_runs <- 100L
batch_demo_rate <- 0.2

batch_demo_data <- purrr::map_dfr(batch_demo_sizes, function(m) {
  mean_gap <- numeric(batch_demo_updates + 1L)

  for (run in seq_len(batch_demo_runs)) {
    theta <- batch_demo_target + 3
    mean_gap[1] <- mean_gap[1] +
      (theta - batch_demo_target)^2 / 2

    for (k in seq_len(batch_demo_updates)) {
      gradient <- theta - mean(sample(batch_demo_y, size = m))
      theta <- theta - batch_demo_rate * gradient
      mean_gap[k + 1L] <- mean_gap[k + 1L] +
        (theta - batch_demo_target)^2 / 2
    }
  }

  tibble(
    batch = factor(m, levels = batch_demo_sizes),
    update = 0:batch_demo_updates,
    gap = mean_gap / batch_demo_runs
  )
})

ggplot(batch_demo_data, aes(update, gap, color = batch)) +
  geom_line() +
  scale_y_log10() +
  theme(legend.position = c(0.5, 0.8)) +
  labs(x = "Updates", y = "Mean loss gap", color = "Batch size")
```
:::::
:::

## Mini-Batch Gradient Descent

::: {.columns}
::::: {.column width="47%"}
Take a mini-batch $I_n$ of size $m$ and compute
$$
  \rho_n = \frac{1}{m}\sum_{i \in I_n} \nabla_\theta L(y_i,x_i,\theta_n),
$$
then update $\theta_{n+1} = \theta_n - \gamma_n\rho_n$.

\medskip\pause

Estimates are still unbiased,
$$
  \E(\rho_n \mid \theta_n) = \nabla H_N(\theta_n),
$$
but variance falls roughly as $1/m$ when $m$ is small relative to $N$.
:::::

::::: {.column width="47%"}
\begin{algorithm}[H]
  \caption{Stochastic gradient descent (with mini-batches)}
  \KwData{Initial parameter $\theta_0$ and learning-rate schedule $(\gamma_n)$}
  \For{$n = 0,1,\dots$}{
    $I_n \gets$ mini-batch of $m$ indices\;

    $\rho_n \gets \frac{1}{m}\sum_{i\in I_n}\nabla_\theta L(y_i,x_i,\theta_n)$\;

    $\theta_{n+1} \gets \theta_n - \gamma_n\rho_n$\;
    \If{converged}{
      \Return{$\theta_{n+1}$}
    }
  }
\end{algorithm}
:::::
:::

--------------------------------------------------------------------------------

## Mini-Batch Convergence Rates and Sizes

::: {.columns}
::::: {.column width="47%"}
### Rates

Under Lipschitz gradient, the rate goes from $O(1/\sqrt{n})$ to
$O(1/\sqrt{mn}+1/n)$.

\medskip\pause

After $E$ epochs, $n = EN/m$, so the bound becomes
$$
  O\big(1/\sqrt{EN}+m/(EN)\big).
$$
Its leading term does not improve with batch size.
:::::

. . .

::::: {.column width="47%"}
### How Large Should a Mini-Batch Be?

There is no universal best $m$: useful sizes depend on gradient noise, the
learning-rate schedule, and the hardware.

\medskip\pause

Try different sizes and tune the learning rate for each.

\medskip\pause

Choose the size that reaches a useful loss (or validation error) fastest in
**wall-clock time**.
:::::
:::

\pdfpcnote{
Connect to law of large numbers: averaging reduces variance.
}

--------------------------------------------------------------------------------

```{r}
#| label: batch-size-convergence
#| echo: false
#| fig-width: 5
#| fig-height: 2.8
#| fig-cap: "The best tested rate for each batch size minimizes mean loss at
#|   epoch 100 over three pilot runs. Mini-batch rates decay per update, while
#|   the full-batch rate stays fixed."

set.seed(123)

N <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(N, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(N, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())
loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

loss_optim <- loss_2d(
  res_glm$coefficients[1],
  res_glm$coefficients[2],
  X,
  y
)

batch_sizes <- c(1L, 10L, 100L, N)
rate_grid <- c(0.3, 1, 3, 5, 10, 15)
tuning_seeds <- 101:103
plot_seeds <- 201:205

tune_rate <- function(b) {
  scores <- vapply(
    rate_grid,
    function(rate) {
      gaps <- vapply(
        tuning_seeds,
        function(seed) {
          set.seed(seed)
          result <- logreg_sgd(
            X,
            y,
            max_epochs = 25,
            batch_size = b,
            a = 1,
            t0 = rate
          )
          tail(result$loss, 1) - loss_optim
        },
        numeric(1)
      )
      mean(pmax(gaps, 0))
    },
    numeric(1)
  )

  rate_grid[which.min(scores)]
}

best_rates <- vapply(batch_sizes, tune_rate, numeric(1))

conv_data <- purrr::map2_dfr(
  batch_sizes,
  best_rates,
  function(b, rate) {
    purrr::map_dfr(plot_seeds, function(seed) {
      set.seed(seed)
      result <- logreg_sgd(
        X,
        y,
        max_epochs = 25,
        batch_size = b,
        a = 1,
        t0 = rate,
        record_time = TRUE
      )

      tibble(
        b = b,
        seed = seed,
        epoch = seq_len(25),
        seconds = result$elapsed,
        loss = pmax(result$loss - loss_optim, 1e-10)
      )
    })
  }
)

plot_data <- conv_data |>
  group_by(b, epoch) |>
  summarize(
    seconds = median(seconds),
    loss = median(loss),
    lower = quantile(loss, 0.25),
    upper = quantile(loss, 0.75),
    .groups = "drop"
  ) |>
  mutate(
    batch = factor(
      b,
      levels = batch_sizes,
      labels = paste0(batch_sizes, " (", best_rates, ")")
    )
  )

ggplot(
  plot_data,
  aes(epoch, loss, color = batch, fill = batch)
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.15,
    color = NA
  ) +
  geom_line() +
  scale_y_log10() +
  labs(
    color = expression("Batch size (" * gamma[0] * ")"),
    fill = expression("Batch size (" * gamma[0] * ")"),
    y = "Suboptimality"
  )
```

##

```{r}
#| label: batch-size-wall-time
#| echo: false
#| fig-width: 5
#| fig-height: 2.8
#| fig-cap: "The same tuned runs for $N = 1000$ and $p = 2$, measured in this R
#|   implementation. Timings include updates and full-loss evaluation each
#|   epoch, but exclude rate tuning."

ggplot(
  plot_data,
  aes(seconds, loss, color = batch, fill = batch)
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.15,
    color = NA
  ) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Elapsed time (s)",
    color = expression("Batch size (" * gamma[0] * ")"),
    fill = expression("Batch size (" * gamma[0] * ")"),
    y = "Suboptimality"
  )
```

## So, Why Use SGD?

We have seen that convergence rates of SGD are worse than those of GD.

\medskip\pause

So why even care about SGD?

. . .

### Reasons to Use SGD

- On large data, mini-batches can reach a useful loss before GD completes many
  full-gradient steps. \pause
- In many applications (e.g. deep learning), we don't *care* about optimizing to
  high accuracy. \pause
- Sometimes, the stochasticity in SGD helps it escape saddle points and local
  minima.

## When SGD Wins: A Moderate Accuracy Target

```{r}
#| label: large-data-sgd-vs-gd
#| echo: false
#| fig-width: 5
#| fig-height: 2.8
#| fig-cap: "Logistic regression with $N = 10^5$ and $p = 20$."

set.seed(1729)

N_large <- 100000L
p_large <- 20L
X_large <- matrix(rnorm(N_large * p_large), nrow = N_large)
beta_true_large <- seq(0.8, 0.1, length.out = p_large)
y_large <- rbinom(
  N_large,
  1,
  plogis(drop(X_large %*% beta_true_large))
)

full_loss_large <- function(theta) {
  z <- drop(X_large %*% theta)
  mean(log1p(exp(z)) - y_large * z)
}

beta_opt_large <- glm.fit(
  X_large,
  y_large,
  family = binomial()
)$coefficients
loss_opt_large <- full_loss_large(beta_opt_large)
L_large <- 0.25 * norm(crossprod(X_large), "2") / N_large
target_gap <- 0.01

run_large_logreg <- function(batch_size, seed) {
  set.seed(seed)
  theta <- numeric(p_large)
  elapsed <- 0
  max_updates <- if (batch_size == N_large) 15L else 150L
  history <- vector("list", max_updates)

  for (k in seq_len(max_updates)) {
    start <- as.numeric(Sys.time())

    if (batch_size == N_large) {
      X_batch <- X_large
      y_batch <- y_large
    } else {
      ind <- sample.int(N_large, batch_size, replace = TRUE)
      X_batch <- X_large[ind, , drop = FALSE]
      y_batch <- y_large[ind]
    }

    gradient <- drop(crossprod(
      X_batch,
      plogis(drop(X_batch %*% theta)) - y_batch
    )) /
      batch_size
    theta <- theta - gradient / L_large
    elapsed <- elapsed + as.numeric(Sys.time()) - start

    if (batch_size == N_large || k <= 20L || k %% 5L == 0L) {
      history[[k]] <- tibble(
        update = k,
        milliseconds = 1000 * elapsed,
        gap = pmax(
          full_loss_large(theta) - loss_opt_large,
          1e-10
        )
      )
    }
  }

  dplyr::bind_rows(history)
}

comparison_large <- purrr::map_dfr(
  c(N_large, 1000L),
  function(batch_size) {
    purrr::map_dfr(201:205, function(seed) {
      run_large_logreg(batch_size, seed) |>
        mutate(
          method = if (batch_size == N_large) {
            "GD"
          } else {
            "SGD (m = 1000)"
          }
        )
    })
  }
)

comparison_large <- comparison_large |>
  group_by(method, update) |>
  summarize(
    milliseconds = median(milliseconds),
    gap = median(gap),
    lower = quantile(gap, 0.25),
    upper = quantile(gap, 0.75),
    .groups = "drop"
  )

ggplot(
  comparison_large,
  aes(milliseconds, gap, color = method, fill = method)
) +
  geom_hline(
    yintercept = target_gap,
    linetype = "dashed",
    color = "gray50"
  ) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.15,
    color = NA
  ) +
  geom_line() +
  scale_y_log10() +
  coord_cartesian(xlim = c(0.1, 20), ylim = c(1e-3, 0.3)) +
  labs(
    x = "Optimizer time (ms)",
    y = "Suboptimality",
    color = NULL,
    fill = NULL
  )
```

##

```{r}
#| label: escape-minima-setup
#| echo: false

set.seed(1)

f <- function(x) x^4 - 2 * x^2 + 1 * x
df <- function(x) 4 * x^3 - 4 * x + 1

# Parameters
eta <- 0.07
steps <- 25
x_gd <- numeric(steps)
x_sgd <- numeric(steps)
x_gd[1] <- x_sgd[1] <- 1.4

for (k in 1:(steps - 1)) {
  # Exact gradient descent
  x_gd[k + 1] <- x_gd[k] - eta * df(x_gd[k])

  # Noisy gradient descent (SGD-like)
  noise <- rnorm(1, sd = 4) # stochasticity
  x_sgd[k + 1] <- x_sgd[k] - eta * (df(x_sgd[k]) + noise)
}

plot_curves <- function(main) {
  curve(
    f,
    from = -1.6,
    to = 1.7,
    n = 200,
    lwd = 2,
    col = "gray60",
    ylab = "",
    xlab = "",
    axes = TRUE,
    yaxt = if (main == "GD") "s" else "n",
    main = main
  )
}

plot_iteration <- function(i, x_gd, x_sgd) {
  x_gd_i <- x_gd[1:i]
  x_sgd_i <- x_sgd[1:i]

  opar <- par(no.readonly = TRUE)

  par(
    mfrow = c(1, 2),
    oma = c(4.1, 4.1, 0.1, 0.3),
    mai = c(0.1, 0.2, 0.5, 0.0)
  )

  plot_curves("GD")
  lines(x_gd_i, f(x_gd_i), col = "royalblue", lwd = 1)
  points(
    x_gd_i,
    f(x_gd_i),
    col = "royalblue",
    pch = 19,
    cex = 0.8
  )

  plot_curves("SGD")
  lines(x_sgd_i, f(x_sgd_i), col = "darkorange", lwd = 1)
  points(
    x_sgd_i,
    f(x_sgd_i),
    col = "darkorange",
    pch = 19,
    cex = 0.8
  )

  mtext(expression(theta), side = 1, outer = TRUE, line = 2)
  mtext(expression(H(theta)), side = 2, outer = TRUE, line = 2)

  par(opar)
}
```

```{r}
#| label: sgd-escape-minima
#| echo: false
#| fig-show: animate
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: >-
#|   The stochasticity in SGD sometimes helps it escape local minima and saddle
#|   points.

for (i in c(1, 2, 3, 5, 10, 15, 20, steps)) {
  plot_iteration(i, x_gd, x_sgd)
}
```

##

## Implementation Details

### Nested Loops

Implementing an SGD algorithm typically means implementing a nested loop, and we
know this isn't R's strongest suite.

\medskip\pause

Next lecture, when we turn to Rcpp, you will have the tools to solve this
problem.

. . .

### Data Storage

SGD algorithms typically subset rows, or blocks of rows.

\medskip\pause

In a column-major order matrix, this is not optimal. Transposing the matrix
first can help.

## Exercise: Estimate a Mean with SGD

::: {.columns}
::::: {.column width="47%"}
For observations $y_1,\ldots,y_N$, minimize
$$
  H_N(\theta) = \frac{1}{N}\sum_{i=1}^N \frac{1}{2}(\theta - y_i)^2.
$$

\medskip\pause

1. Find the minimizer $\theta^*$ and the sampled gradient. \pause
2. Complete the SGD update and plot its path. \pause
3. Compare $\gamma_k = 0.1$ with $\gamma_k = 1/k$. What changes?
:::::

::::: {.column width="47%"}
\footnotesize

```{r}
#| label: sgd-mean-exercise
#| eval: false

set.seed(1)

N <- 100
y <- rnorm(N, mean = 2)
steps <- 200
theta <- numeric(steps + 1)
theta[1] <- 0

for (k in seq_len(steps)) {
  i <- sample.int(N, 1)
  gamma <- 0.1

  theta[k + 1] <- NA # <1>
}

plot(0:steps, theta, type = "l")
abline(h = mean(y), lty = 2)
```

1. Complete the update.
:::::
:::

## Summary

Stochastic gradient descent is a popular method for large-scale optimization.

\medskip\pause

Convergence results are relatively weak, but for many applications this does not
matter.

\medskip\pause

In practice, most implementations use mini-batches.

## Next TIme

### Rcpp

We will look at how to use Rcpp to speed up R code.

\medskip\pause

Because SGD typically involves nested loops, these algorithms benefit greatly
from relying on C++.
