---
title: "Stochastic Gradient Descent"
---

{{< include _common.qmd >}}

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(mvtnorm)

set.seed(1422)

source(here::here("R", "sgd.R"))
```

## Last Time

### The Expectation-Maximization (EM) Algorithm

A general approach to maximum likelihood estimation with latent variables.

. . .

### Fisher Information

We saw three different ways to define the Fisher information and how they relate
to the EM algorithm.

## Recap: The EM Algorithm

The EM algorithm iteratively optimizes
$$
  \theta_{n + 1} = \operatorname{argmax}_{\theta \in \Theta} Q(\theta \mid \theta_n),
$$
with
$$
  Q(\theta \mid \theta_n) = \E_{\theta_n}\big(\log p(X, Z \mid \theta) \mid X\big).
$$

\medskip\pause

Convergence is **linear** near a local maximum.

## Recap: Fisher Information

The Fisher information is defined as
$$
  \mathcal{I}(\theta) = -\E_\theta\left(D_\theta^2 \ell(\theta)\right),
$$
where $\ell(\theta) = \log p(X \mid \theta)$.

\medskip\pause

We mentioned three ways to compute the observed Fisher information:

1. Differentiate the score function (defined through $Q$).\pause
2. Compute complete and missing information and using the information identity.
   \pause
3. Differentiate the EM mapping.

## Today

### Gaussian Mixture EM Example

We pick up where we left off last time, focusing on the Gaussian mixture
example.

. . .

### Stochastic Gradient Descent

Useful stochastic method for optimization

\medskip\pause

Can be used in a **mini-batch** version.

# EM Gaussian Mixture Example

## A Mixture of Two Gaussians

::: {.columns align="center"}
::::: {.column width="47%"}
### Goal

Fit a two-component Gaussian mixture model to data.

. . .

### Model

We assume that the data $X_i \sim \operatorname{\mathcal{N}}(\mu_k, \sigma_k^2)$
with probability $\pi_k$, $k = 1, 2$.
:::::

::::: {.column width="47%"}
```{r}
#| echo: false
#| fig-cap: "A mixture of two Gaussian distributions."

gausdens <- function(x, mu1, mu2, sigma1, sigma2, p) {
  (p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
    (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2) /
    sqrt(2 * pi)
}

xx <- seq(-3, 11, 0.1)

sigma1 <- 1
sigma2 <- 2
mu1 <- -0.5
mu2 <- 4
p <- 0.5

il_df <- data.frame(
  x = xx,
  y1 = dnorm(xx, mu1, sigma1) * p,
  y2 = dnorm(xx, mu2, sigma2) * (1 - p),
  ymix = gausdens(xx, mu1, mu2, sigma1, sigma2, p)
)

pl <- ggplot(il_df, aes(x, ymix)) +
  geom_line() +
  geom_line(aes(y = y1), linetype = "dashed", color = "steelblue") +
  geom_line(aes(y = y2), linetype = "dashed", color = "darkorange") +
  labs(y = "Density", x = "x")

pl
```
:::::
:::

## Finite Mixtures

In a finite mixture, the marginal density of $X$ is
$$
  p(x \mid \psi) = \sum_{k = 1}^K \pi_k f_k(x \mid \psi_k)
$$
where $\pi_k \geq 0$, $\sum_k \pi_k = 1$, and $f_k$ are densities.

\medskip\pause

### Latent Variable

Assume each $X_i$ comes from one of the $K$ components, but we don't know which
one.

\medskip\pause

We introduce a latent variable $Z_i \in \{1, \ldots, K\}$ such that
$$
  \Pr(Z_i = k) = \pi_k.
$$

. . .

### Complete-Data Model

Then the complete-data model is
$$
  p(x_i, z_i \mid \psi) = \pi_{z_i} f_{z_i}(x_i \mid \psi_{z_i}).
$$

## Gaussian Mixture

For Gaussian mixtures, the components are Gaussian:
$$
  f_k(x \mid \mu_k, \sigma_k^2) = \frac{1}{\sqrt{2 \pi \sigma_k^2}} e^{-\frac{(x - \mu_k)^2}{2 \sigma_k^2}}.
$$
with $\psi_k = (\mu_k, \sigma_k^2)$.

\medskip\pause

### Complete-Data Log-Likelihood

Letting $\theta = (\pi_1, \ldots, \pi_K, \psi_1, \ldots, \psi_K)$, the
complete-data log-likelihood is
$$
\begin{aligned}
  \ell_c (\theta) & = \sum_{i = 1}^n \log p(x_i, z_i \mid \theta)                                                                       \\
  \pause
                  & = \sum_{i = 1}^n \log \left( f(x_i \mid z_i, \theta) f(z_i \mid \theta) \right)                                     \\
  \pause
                  & = \sum_{i = 1}^n \sum_{k = 1}^K \symbf{1}(Z_i = k) \left( \log \pi_k + \log f_k(x_i \mid \mu_k, \sigma_k^2) \right)
\end{aligned}
$$

## Simple Case: Fixed Component Variances, $K = 2$

::: {.columns align="center"}
::::: {.column width="47%"}
Let's simplify and assume $K = 2$ and that $\sigma_1$ and $\sigma_2$ are known.

\medskip\pause

Then the model has three unknown parameters:
$$
  \theta = (\pi, \mu_1, \mu_2).
$$
:::::

::::: {.column width="47%"}
\begin{tikzpicture}[
    node distance=1.6cm,
    latent/.style={circle,draw,minimum size=10mm,inner sep=0pt},
    observed/.style={circle,draw,fill=gray!20,minimum size=10mm,inner sep=0pt},
    param/.style={rectangle,draw,minimum size=7mm,inner sep=2pt},
  ]

  % Parameters for components
  \node[param] (pi) {$\pi$};
  \node[param, right=2.5cm of pi, yshift=1cm] (mu1) {$\mu_1$};
  \node[param, right=2.5cm of pi, yshift=-1cm] (mu2) {$\mu_2$};

  % Latent and observed variables
  \node[latent, below=1.8cm of pi] (z) {$Z_i$};
  \node[observed, below=1.8cm of z] (x) {$X_i$};

  % Arrows
  \draw[->] (pi) -- (z);
  \draw[->] (z) -- (x);
  \draw[->] (mu1) -- (x);
  \draw[->] (mu2) -- (x);

  % Plate for data points
  \node[plate={below:{$i=1,\dots,n$}}, fit=(z)(x)] (plateN) {};
\end{tikzpicture}
:::::
:::

## The E-Step

The complete-data log-likelihood is now
$$
\begin{aligned}
  \ell_c(\theta) & = \sum_{i = 1}^n \symbf{1}(Z_i = 1) \left( \log(\pi) - \frac{(x_i - \mu_1)^2}{2 \sigma_1^2} \right)                     \\
                 & \phantom{ = {}} + \symbf{1}(Z_i = 2)\left( \log(1 - \pi) - \frac{(x_i - \mu_2)^2}{2 \sigma_2^2} \right) + \text{const}.
\end{aligned}
$$

. . .

Taking expectations gives
$$
  Q(\theta \mid \theta') = \sum_{i = 1}^n \tau_{i} \left(\log(\pi) - \frac{(x_i - \mu_1)^2}{2 \sigma_1^2} \right)
                           + (1 - \tau_{i})\left( \log(1 - \pi) - \frac{(x_i - \mu_2)^2}{2 \sigma_2^2} \right)
$$
where $\tau_i = \Pr(Z_i = 1 \mid X_i = x_i, \theta')$ is the **responsibility**
of component 1 for observation $i$.

\medskip\pause

The responsibility tells us how responsible component 1 is for generating $x_i$.

## The Responsibilities

The conditional probability in a mixture model is generally
$$
  \Pr(Z = z \mid X = x) = \frac{\pi_z f_z(x \mid \psi_z)}{\sum_{k = 1}^K \pi_k f_k(x \mid \psi_k)}.
$$

\medskip\pause

For the $K = 2$ Gaussian case, this gives us
$$
  \tau_i = \Pr(Z_i = 1 \mid X = x_i, \theta')
         = \frac{\pi'_1 e^{-\frac{(x_i - \mu_1')^2}{2 \sigma_1^2}}}{\left( \pi'_1 e^{-\frac{(x_i - \mu_1')^2}{2 \sigma_1^2}} + \frac{\sigma_1}{\sigma_2} (1 - \pi'_1) e^{-\frac{(x_i - \mu_2')^2}{2 \sigma_2^2}}\right)}.
$$

## The Q-Step

The MLEs of $\theta$ given $Q(\theta \mid \theta')$ are
$$
  \hat{\theta} = \left(\frac{1}{n} \sum_{i = 1}^n \tau_i, \quad \frac{1}{\sum_{i = 1}^n \tau_i} \sum_{i = 1}^n \tau_i x_i, \quad \frac{1}{\sum_{i = 1}^n (1 - \tau_i)} \sum_{i = 1}^n (1 - \tau_i) x_i \right).
$$

. . .

::: {.columns}
::::: {.column width="47%"}
### Updating $\pi$

We update $\pi$ as
$$
  \hat{\pi} = \frac{1}{n} \sum_{i = 1}^n \tau_i,
$$
which is the fraction of all data points "assigned" to component 1.
:::::

. . .

::::: {.column width="47%"}
### Updating $\mu_1$ and $\mu_2$

We update $\mu_1$ as
$$
  \hat{\mu}_1 = \frac{\sum_{i = 1}^n \tau_i x_i}{\sum_{i = 1}^n \tau_i},
$$
and similarly for $\mu_2$.

\medskip\pause

This is a weighted mean, weighted by how much each component is responsible for
each observation.
:::::
:::

## Example: Simulation

We start by setting the parameters.

```{r}
#| label: mixture-gaus-sim

sigma1 <- 1
sigma2 <- 2
mu1 <- -0.5
mu2 <- 4
p <- 0.5
```

. . .

Then we simulate data.

```{r}
n <- 5000
z <- sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(p, 1 - p))
x <- numeric(n)
n1 <- sum(z)
x[z] <- rnorm(n1, mu1, sigma1)
x[!z] <- rnorm(n - n1, mu2, sigma2)
```

##

```{r}
#| echo: false
#| fig-cap: "Histogram of simulated data from a two-component Gaussian mixture model."
#| fig-width: 4
#| fig-height: 2.7

pl <- ggplot(data.frame(x), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "lightgray") +
  geom_line(
    aes(x, gausdens(x, mu1, mu2, sigma1, sigma2, p)),
    color = "black",
    linewidth = 1
  ) +
  labs(y = "Density", x = "x")

pl
```

```{r}
#| label: gaus-mix-loglik
#| echo: false

neg_loglik <- function(par, x) {
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]

  if (p < 0 || p > 1) {
    return(Inf)
  }

  -sum(log(
    p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
      (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2
  ))
}
```

## General-Purpose Optimization

```{r, dependson=c("gaus-mix-loglik")}
#| label: gaus-mix-example

optim_par <- optim(c(0.5, -0.5, 4), neg_loglik, x = x)
optim_par[1:2]
```

. . .

But the starting value matters!

```{r, dependson=c("gaus-mix-loglik")}
#| label: gaus-mix-example-bad

optim(c(0.9, 3, 1), neg_loglik, x = x)[1:2]
```

## The EM Algorithm

```{r}
#| label: em-alg

em <- function(par, em_step, eps = 1e-6, maxit = 20, cb = NULL) {
  for (i in seq_len(maxit)) {
    par0 <- par
    par <- em_step(par)

    if (!is.null(cb)) {
      cb() # Callback function (tracer)
    }

    if (sum((par - par0)^2) <= eps * (sum(par^2) + eps)) {
      break
    }
  }
  par
}
```

## Implementation

Next, we implement the E and M steps. Implementations are not shown here, but
can be found in the [source code](./slides/lecture11.qmd)

```{r}
#| label: source-em-mix
#| echo: false

library(numDeriv)

e_step_gauss <- function(par, x, sigma1 = 1, sigma2 = sigma1) {
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  a <- p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1
  b <- (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2
  a / (a + b)
}

m_step_gauss <- function(p_hat, x) {
  n <- length(x)
  N1 <- sum(p_hat)
  N2 <- n - N1
  c(N1 / n, sum(p_hat * x) / N1, sum((1 - p_hat) * x) / N2)
}

create_em_gauss_step <- function(x, sigma1 = 1, sigma2 = sigma1) {
  force(x)
  force(sigma1)
  force(sigma2)

  function(par) m_step_gauss(e_step_gauss(par, x, sigma1, sigma2), x)
}

Q <- function(par, par_prime) {
  p_hat <- e_step_gauss(par_prime, x, sigma1, sigma2)
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  sum(
    p_hat * (log(p) - (x - mu1)^2 / (2 * sigma1^2)) +
      (1 - p_hat) * (log(1 - p) - (x - mu2)^2 / (2 * sigma2^2))
  )
}

grad1 <- function(par) grad(function(par) neg_loglik(par, x), par)
grad2 <- function(par) -grad(Q, par, par_prime = par)

em_gauss_step <- create_em_gauss_step(x, sigma1, sigma2)
```

## Testing

```{r}
#| label: test-em-gauss-mix

em_par <- em(c(0.5, -0.5, 4), em_gauss_step)
```

. . .

```{r, dependson=c("gaus-mix-loglik")}
#| label: gaus-mix-example2

testthat::test_that("EM and optim give similar results", {
  testthat::expect_equal(em_par, optim_par$par, tolerance = 1e-3)
})
```

. . .

```{r}
#| label: run-em-gauss-mix-again

em(c(0.9, 3, 1), em_gauss_step) # Starting value still matters
```

## Tracing Convergence

The `tracer` object from the **CSwR** package lets us evaluate arbitrary
expressions at each iteration of the EM algorithm via `expr`.

```{r}
#| label: tracing-convergence

library(CSwR)
em_tracer <- tracer(
  c("par0", "par", "nll"),
  Delta = 0, # No printing
  expr = quote({
    nll <- neg_loglik(par, x)
  })
)
```

. . .

```{r}
par_hat <- em(c(0.2, 2, 2), em_gauss_step, cb = em_tracer$tracer)

em_trace <- summary(em_tracer)
```

##

::: {.columns}
::::: {.column width="47%"}
**CSwR** contains an `autoplot` S3 method for `trace` objects.

. . .

```{r}
#| label: gaussmix-convergence-plot
#| eval: false

autoplot(em_trace, y = nll - min(nll))
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: convergence-plots-2
#| warning: false
#| echo: false
#| ref-label: gaussmix-convergence-plot
#| fig-cap: "Convergence of the EM algorithm for Gaussian mixture model."
```
:::::
:::

##

```{r}
#| label: mixture-convergence
#| echo: false

gaussdens2 <- function(x, p, mu1, mu2) {
  (p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
    (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2) /
    sqrt(2 * pi)
}

em_gauss_illustration <- function(i, em_trace) {
  p_i <- em_trace$par0.1[i]
  mu1_i <- em_trace$par0.2[i]
  mu2_i <- em_trace$par0.3[i]

  pl +
    geom_line(
      aes(y = gaussdens2(x, p_i, mu1_i, mu2_i)),
      color = "darkorange",
      size = 1
    ) +
    lims(y = c(0, 0.25)) +
    labs(title = paste0("k = ", i))
}
```

```{r}
#| fig-width: 4.2
#| fig-height: 2.8
#| echo: false
#| fig-cap: "Iterations of the EM algorithm"

em_gauss_illustration(1, em_trace)
```

##

```{r}
#| fig-width: 4.2
#| fig-height: 2.8
#| echo: false
#| fig-cap: "Iterations of the EM algorithm"

em_gauss_illustration(4, em_trace)
```

##

```{r}
#| fig-width: 4.2
#| fig-height: 2.8
#| echo: false
#| fig-cap: "Iterations of the EM algorithm"

em_gauss_illustration(8, em_trace)
```

##

```{r}
#| fig-width: 4.2
#| fig-height: 2.8
#| echo: false
#| fig-cap: "Iterations of the EM algorithm"

em_gauss_illustration(12, em_trace)
```

##

```{r}
#| fig-width: 4.2
#| fig-height: 2.8
#| echo: false
#| fig-cap: "Iterations of the EM algorithm"

em_gauss_illustration(16, em_trace)
```

##

```{r}
#| fig-width: 4.2
#| fig-height: 2.8
#| echo: false
#| fig-cap: "Iterations of the EM algorithm"

em_gauss_illustration(20, em_trace)
```

## Benchmarking

You might be inclined to wonder which method is faster: EM or `optim()`?

\medskip\pause

```{r}
#| label: benchmark-em-gauss-mix

gauss_bm <- bench::mark(
  em = em(c(0.5, -0.5, 4), em_gauss_step),
  optim = optim(c(0.5, -0.5, 4), neg_loglik, x = x),
  check = FALSE
)
```

##

```{r}
#| fig-cap: "Benchmarking results for EM vs. optim() for Gaussian mixture model."
#| fig-width: 4.5

plot(gauss_bm)
```

\medskip\pause

But this comparison is flawed: **why?**

## Label Switching

If we set $\sigma_1 = \sigma_2$, then the likelihood is invariant under
switching the labels of the components, i.e., $(p, \mu_1, \mu_2)$ and
$(1 - p, \mu_2, \mu_1)$ give the same likelihood.

\medskip\pause

So the EM algorithm may converge to either of the two symmetric modes.

```{r}
#| label: mixture-gaus-sim2
#| echo: false

sigma1 <- 1
sigma2 <- 1
mu1 <- -0.5
mu2 <- 4
p <- 0.8
n <- 5000
z <- sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(p, 1 - p))
x <- numeric(n)
n1 <- sum(z)
x[z] <- rnorm(n1, mu1, sigma1)
x[!z] <- rnorm(n - n1, mu2, sigma2)

em_gauss_step_alt <- create_em_gauss_step(x, sigma1)
```

. . .

```{r}
em(c(0.2, 3, -0.5), em_gauss_step_alt)
```

. . .

```{r}
em(c(0.8, -3, -0.5), em_gauss_step_alt)
```

. . .

In this example, we generated data with $p = 0.8$ and $\sigma_1 = \sigma_2$.

# Stochastic Gradient Descent

## Empirical Risk Minimization

For data $(y_i,x_i)$, $i = 1,\ldots,N$, the empirical risk is
$$
  H_N(\theta) = \frac{1}{N} \sum_{i = 1}^N L(y_i,x_i,\theta).
$$

. . .

::: {.columns}
::::: {.column width="47%"}
### Gradient Descent

Gradient descent uses the full gradient:
$$
  \theta_{n + 1} = \theta_n - \gamma_n \nabla H_N(\theta_n).
$$
:::::

. . .

::::: {.column width="47%"}
### Stochastic Gradient Descent

Sample $J_n$ uniformly from $\{1,\ldots,N\}$, compute
$$
  \rho_n = \nabla_\theta L(y_{J_n},x_{J_n},\theta_n),
$$
and update $\theta_{n + 1} = \theta_n - \gamma_n\rho_n$.
:::::
:::

## Example: Ordinary Least Squares

The empirical risk is
$$
  H_N(\beta) = \frac{1}{N} \sum_{i = 1}^N L(y_i,x_i,\beta),
$$
where
$$
  L(y_i,x_i,\beta) = \frac{1}{2} (y_i - \beta^T x_i)^2.
$$

And the gradient is
$$
  \nabla_\beta L(y_i,x_i,\beta) = -x_i(y_i - \beta^T x_i).
$$

## Basic Implementation of SGD

```{r}
sgd_basic <- function(
  par,
  grad,
  n,
  # number of data points
  gamma,
  # learning rate
  sampler = sample,
  # function that samples indices
  maxit = 100,
  ...
) {
  for (k in seq_len(maxit)) {
    for (i in seq_len(n)) {
      ind <- sampler(n)
      par <- par - gamma * grad(par, ind, ...)
    }
  }
  par
}
```

## Unbiasedness

If $J_n$ is drawn uniformly from $\{1, \ldots, N\}$, then
$$
  \E(\rho_n \mid \theta_n) = \nabla H_N(\theta_n).
$$

So SGD gradients are **unbiased** estimates of the full gradient.

. . .

### Iteration Cost

The main benefit of this is that the cost of each iteration is much lower:
$O(p)$ vs. $O(Np)$ (for full gradient descent with $p$ variables).

. . .

### Epoch

When discussing SGD, we often refer to an **epoch** as $N$ iterations of SGD: a
full pass over the data.

## Example: Logistic Regression

::: {.columns}
::::: {.column width="47%"}
With $y_i \in \{0, 1\}$, the logistic regression loss is
$$
  L(y_i,x_i,\beta) = -y_i \beta^T x_i + \log\left(1 + e^{\beta^T x_i}\right).
$$
and
$$
  \nabla_\beta L(y_i,x_i,\beta) = x_i \left( \frac{e^{\beta^T x_i}}{1 + e^{\beta^T x_i}} - y_i \right).
$$

. . .

SGD typically converges quickly at the start but slows down as it approaches the
minimum.
:::::

::::: {.column width="47%"}
```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 3.2

set.seed(123)

n <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(n, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(n, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())

loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

beta1 <- seq(-1, 3, length = 100)
beta2 <- seq(-1, 3, length = 100)

loss_2d_vectorized <- Vectorize(function(b1, b2) loss_2d(b1, b2, X, y))

z <- outer(beta1, beta2, loss_2d_vectorized)

res_sgd <- logreg_sgd(X, y, max_epochs = 100, batch_size = 1)
res_sgd2 <- logreg_sgd(X, y, max_epochs = 100, batch_size = 1, K = 5, a = 0.5)
res_gd <- logreg_sgd(X, y, max_epochs = 100, batch_size = n)

loss_optim <- loss_2d(res_glm$coefficients[1], res_glm$coefficients[2], X, y)

res_gd <- logreg_sgd(X, y, max_epochs = 100, batch_size = n)
loss1 <- loss_2d_vectorized(res_gd$beta_history[1, ], res_gd$beta_history[2, ])
loss2 <- loss_2d_vectorized(
  res_sgd$beta_history[1, ],
  res_sgd$beta_history[2, ]
)
loss3 <- loss_2d_vectorized(
  res_sgd2$beta_history[1, ],
  res_sgd2$beta_history[2, ]
)

n1 <- length(loss1)
n2 <- length(loss2)
n3 <- length(loss3)

pal <- palette.colors(palette = "Okabe-Ito")

contour(beta1, beta2, z, col = "dark grey", drawlabels = FALSE, asp = 1)

lines(res_sgd$beta_history[1, ], res_sgd$beta_history[2, ], col = pal[3])
points(
  res_sgd$beta_history[1, ],
  res_sgd$beta_history[2, ],
  col = pal[3],
  pch = 19,
  cex = 0.5
)

lines(res_gd$beta_history[1, ], res_gd$beta_history[2, ], col = pal[1])
points(
  res_gd$beta_history[1, ],
  res_gd$beta_history[2, ],
  col = pal[1],
  pch = 19,
  cex = 0.5
)

points(
  res_glm$coefficients[1],
  res_glm$coefficients[2],
  col = pal[2],
  pch = 19,
  cex = 1
)

legend(
  "bottomright",
  legend = c("SGD", "GD"),
  col = pal[c(3, 1)],
  pch = 19,
  cex = 0.8
)
```
:::::
:::

## Learning Rate

::: {.columns align = "center"}
::::: {.column width="47%"}
For gradient descent, we can use a fixed step length and still guarantee
convergence.

\medskip\pause

But this is **not** the case for SGD.
:::::

::::: {.column width="47%"}
```{r}
#| fig-cap: "Convergence of SGD for logistic regression with a fixed learning rate."
#| echo: false

res_sgd <- logreg_sgd(X, y, max_epochs = 200, batch_size = 1, t0 = 5)
pal <- palette.colors(palette = "Okabe-Ito")

loss <- loss_2d_vectorized(res_sgd$beta_history[1, ], res_sgd$beta_history[2, ])

ggplot(
  tibble(iteration = seq_along(loss), loss = loss - loss_optim),
  aes(iteration, loss)
) +
  geom_line() +
  scale_y_log10() +
  labs(y = "Suboptimality")
```
:::::
:::

\medskip\pause

Can we find conditions on the learning rates $\gamma_n$ that ensure convergence
of SGD? **Yes!**

\medskip

But first, we need to introduce the concept of **strong convexity**.

## Strong Convexity

A differentiable function $H: \mathbb{R}^p \to \mathbb{R}$ is **strongly
convex** if
$$
  H(\theta) \geq H(\theta')
                 + \nabla H(\theta')^T (\theta - \theta')
                 + \frac{l}{2} \lVert \theta
                 - \theta'\rVert_2^2
$$
for all $\theta,\theta' \in \mathbb{R}^p$ and some $l > 0$.

\medskip\pause

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        width=7cm, height=4.3cm,
        axis lines=none,
        xlabel={$\theta$},
        ylabel={$H(\theta)$},
        ymin=-0.5, ymax=4,
        xmin=-1, xmax=2.2,
        samples=200,
        domain=-0.8:2.2,
        clip=false
      ]
      % f(x) = x^2
      \addplot[thick] {x^2};
      % Tangent at x0=1: 2*(x-1) + 1
      \addplot[DarkOrange, dashed, thick] {2*(x-1) + 1};
      % Strong convexity lower bound: tangent + 0.5*(x-1)^2
      \addplot[RoyalBlue, dotted, thick] {2*(x-1) + 1 + 0.5*(x-1)^2};
      % Point at (1,1)
      \addplot[only marks, mark=*] coordinates {(1,1)};
      % Annotations
      \node[anchor=west] at (axis cs:2.2,5) {$H(\theta)$};
      \node[anchor=west, RoyalBlue] at (axis cs:2.2,4.3) {$H(\theta') + \nabla H(\theta')^T(\theta-\theta') + \frac{l}{2}\lVert\theta-\theta' \rVert^2$};
      \node[anchor=west, DarkOrange] at (axis cs:2.2,3.5) {$H(\theta') + \nabla H(\theta')^T(\theta-\theta')$};
    \end{axis}
  \end{tikzpicture}
\end{figure}

## Example: Strong Convexity

The function
$$
  H(\theta) = \theta^2
$$
\pause is strongly convex with $l = 2$ since $$
\begin{aligned}
H(\theta) - H(\theta') - \nabla H(\theta')^T(\theta - \theta') & = \theta^2 - (\theta')^2 - 2\theta'(\theta - \theta') \\ \pause
                                         & = (\theta - \theta')^2 \\ \pause
                                         & \geq \frac{l}{2}(\theta - \theta')^2.
\end{aligned}
$$

\medskip\pause

The inequality holds for $l \leq 2$, and in particular for $l = 2$.

## Relationship to Lipschitz Gradient

If $H$ is strongly convex with parameter $l$ and has Lipschitz gradient with
constant $L$, then
$$
  l I \preceq D^2H(\theta) \preceq L I.
$$

\medskip\pause

- Strong convexity provides a **lower** bound on the curvature of $H$\pause
- Lipschitz gradient provides an **upper** bound on the curvature of $H$

. . .

### Example

The function $H(\theta) = \theta^2$ is strongly convex with $l = 2$ and has
Lipschitz gradient with $L = 2$, since $D^2H(\theta) = 2$.

\medskip\pause

This is *identical* to the strong convexity parameter.

## Example of Strong Convexity--Lipschitz Gap

::: {.columns}
::::: {.column width="47%"}
Let's take the function
$$
  H(\theta) = \theta^2 + \sin(\theta).
$$

\medskip\pause

This is twice differentiable with
$$
  D^2H(\theta) = 2 - \sin(\theta).
$$

\medskip\pause

And since
$$
  1 \leq 2 - \sin(x) \leq 3,
$$
$H$ is strongly convex with $l = 1$ and has Lipschitz gradient with $L = 3$.

\medskip\pause
:::::

::::: {.column width="47%"}
\begin{figure}
  \begin{tikzpicture}
    \pgfmathsetmacro{\a}{1}
    \pgfmathsetmacro{\xzero}{4.712389} % 3*pi/2
    \pgfmathsetmacro{\mu}{2 - \a}      % =1
    \pgfmathsetmacro{\Lips}{2 + \a}    % =3
    \pgfmathsetmacro{\fz}{\xzero*\xzero + sin(deg(\xzero))}
    \pgfmathsetmacro{\dfz}{2*\xzero + cos(deg(\xzero))}
    \begin{axis}[
        width=5cm,
        height=6.2cm,
        domain=-0.5:12.5,
        clip=false,
        samples=200,
        axis lines=none,
        xlabel={$\theta$},
        ylabel={$H(\theta)$},
        legend style={draw=none,at={(0.02,0.98)},anchor=north west}
      ]
      % f
      \addplot[thick, line label={$H(\theta)$}] {x^2 + sin(deg(x))};
      % lower quadratic
      \addplot[
        DarkOrange,
        dashed,
        thick,
        line label={Strong convexity bound},
      ] {
        \fz + \dfz*(x-\xzero) + 0.5*\mu*(x-\xzero)^2
      };
      % upper quadratic
      \addplot[
        RoyalBlue,
        dashed,
        thick,
        line label={Lipschitz bound}
      ] {
        \fz + \dfz*(x-\xzero) + 0.5*\Lips*(x-\xzero)^2
      };
      % point
      \addplot[only marks,mark=*] coordinates {(\xzero,\fz)};
    \end{axis}
  \end{tikzpicture}
  \caption{The function $H(\theta) = \theta^2 + \sin(\theta)$ is strongly convex with $l = 1$
    and has Lipschitz gradient with $L = 3$.}
\end{figure}
:::::
:::

## Conditioning

The **condition number** of a strongly convex function with Lipschitz gradient
is
$$
  \kappa = \frac{L}{l} \geq 1.
$$

. . .

The closer $\kappa$ is to $1$, the better conditioned the problem is.

### Example

For $H(\theta) = \theta^2$, we have $\kappa = 2/2 = 1$, so the problem is
perfectly conditioned.

\medskip\pause

We reach the minimum in one step with gradient descent and step length
$t = 1/L = 1/2$.

## The Robbins--Monro Convergence Theorem

Let $H(\theta) = \E L(Y,\theta)$ be the population risk. Suppose $H$ is strongly
convex and
$$
  \E\left(\lVert\nabla_\theta L(Y,\theta)\rVert_2^2\right) \leq A
                                                                + B \lVert \theta \rVert_2^2
$$
for some constants $A$ and $B$.

\medskip\pause

If $\theta^*$ is the global minimizer of $H$, then $\theta_n$ converges almost
surely toward $\theta^*$ if
$$
  \sum_{n = 1}^{\infty} \gamma_n^2 < \infty \quad \text{and} \quad \sum_{n = 1}^{\infty} \gamma_n
                                   = \infty.
$$

. . .

The theorem applies to learning rates satisfying
$$
  \gamma_n \propto \frac{1}{n}.
$$

## Diminishing Learning Rates

Assume a cyclic rule and set $\gamma_n = \gamma$ for $N$ updates. We get
$$
  \theta_{n + N} = \theta_n
                   - \gamma \sum_{i = 1}^N \nabla_\theta L(y_i,x_i,\theta_{n + i - 1}).
$$

. . .

Meanwhile, **full** gradient descent with step length $N\gamma$ would give
$$
  \theta_{n + 1} = \theta_n
                   - \gamma \sum_{i = 1}^N \nabla_\theta L(y_i,x_i,\theta_n).
$$

. . .

Difference between the two is
$$
  \gamma \sum_{i = 1}^N \left( \nabla_\theta L(y_i,x_i,\theta_{n + i - 1}) - \nabla_\theta L(y_i,x_i,\theta_n) \right)
$$
which does not tend to zero for constant $\gamma$. So the noise in the gradient
estimates does not vanish as we approach the minimum.

## Convergence Rates

### Gradient Descent

For convex $H_N$, GD with diminishing step lengths converges at rate
$$
  H_N(\theta_n) - H_N^* = O\left(\frac{1}{\sqrt{n}}\right).
$$

. . .

When $H_N$ is differentiable with Lipschitz gradient, we get
$$
  H_N(\theta_n) - H_N^* = O\left(\frac{1}{n}\right).
$$

. . .

### Stochastic Gradient Descent

For convex $H_N$ with diminishing learning rates, SGD converges at rate
$$
  \E\big(H_N(\theta_n) - H_N^*\big) = O\left(\frac{1}{\sqrt{n}}\right).
$$

. . .

But this **does not improve** when $H_N$ is differentiable with Lipschitz
gradient.

## Rates for Strongly Convex Functions

When $H_N$ is strongly convex and has Lipschitz gradient, GD satisfies
$$
  H_N(\theta_n) - H_N^* = O(\delta^n), \qquad \delta \in (0, 1)
$$
and $\delta = 1 - l/L$.

. . .

Under same conditions, SGD gives us
$$
  \E\big(H_N(\theta_n) - H_N^*\big) = O\left(\frac{1}{n}\right).
$$

. . .

So SGD does **not** enjoy linear convergence rates under strong convexity.

##

\begin{figure}
\begin{tikzpicture}
  \begin{axis}[
      width=9cm, height=8cm,
      xlabel={$k$},
      ylabel={Suboptimality},
      legend pos=north east,
      ymode=log,
      xmax=36,
      domain=1:20
    ]
    % O(1/sqrt(k))
    \addplot[
      thick,
      line label={$O(1/\sqrt{k})$}
    ] {1/sqrt(x)};

    % O(1/k)
    \addplot[
      thick,
      dashed,
      line label={$O(1/k)$}
    ] {1/x};

    % O(delta^k) with delta = 0.5
    \addplot[
      thick,
      dotted,
      samples=50,
      line label={$O(\delta^k)$}
    ] {0.5^x};
  \end{axis}
\end{tikzpicture}
\caption{Illustration of convergence rates. Here, $\delta = 0.5$.}
\end{figure}

## Mini-Batch Gradient Descent

Take a mini-batch $I_n$ of size $m$ and compute
$$
  \rho_n = \frac{1}{m}\sum_{i \in I_n} \nabla_\theta L(y_i,x_i,\theta_n),
$$
then update $\theta_{n + 1} = \theta_n - \gamma_n\rho_n$.

. . .

Estimates are still unbiased,
$$
  \E(\rho_n \mid \theta_n) = \nabla H_N(\theta_n),
$$
but variance is reduced by a factor of $1/m$.

\medskip\pause

Under Lipschitz gradient, the rate goes from $O(1/\sqrt{n})$ to
$O(1/\sqrt{mn} + 1/n)$.

\medskip\pause

Typically more efficient than standard SGD for various computational reasons,
but somewhat **less** robust to local minima.

\pdfpcnote{
Connect to law of large numbers: averaging reduces variance.
}

##

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 2.8
#| fig-cap: "Convergence of SGD on the logistic regression problem, with
#|   different batch sizes. Note that batch size 1000 is standard gradient
#|   descent."

set.seed(123)

n <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(n, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(n, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())
loss_optim <- loss_2d(res_glm$coefficients[1], res_glm$coefficients[2], X, y)

loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

conv_data <- data.frame(b = NULL, epoch = NULL, loss = NULL)

for (b in c(1, 10, 100, n)) {
  res_sgd <- logreg_sgd(X, y, max_epochs = 100, batch_size = b, a = 1, t0 = 3)

  conv_data <- rbind(
    conv_data,
    data.frame(b = b, epoch = seq_len(100), loss = res_sgd$loss - loss_optim)
  )
}

ggplot(conv_data, aes(epoch, loss, col = as.factor(b))) +
  geom_line() +
  scale_y_log10() +
  labs(color = "Batch size", y = "Suboptimality")
```

## So, Why Use SGD?

We have seen that convergence rates of SGD are worse than those of GD.

\medskip\pause

So why even care about SGD?

. . .

### Reasons to Use SGD

- If $n \gg p$, then we can pick a batch size that is reasonably accurate but
  still much smaller than $n$. \pause
- In many applications (e.g. deep learning), we don't *care* about optimizing to
  high accuracy. \pause
- Sometimes, the stochasticity in SGD helps it escape local minima.

## Convergence in Epochs and Iterations

In terms of iterations, gradient descent converges much more quickly than SGD.
But in terms of epochs, SGD can be faster.

\medskip\pause

```{r}
#| fig-width: 5
#| fig-height: 2.3
#| fig-cap: "Convergence of gradient descent (GD) and stochastic gradient descent
#|   (SGD) for logistic regression in terms of iterations and epochs."
#| echo: false

conv_data <- tibble(
  method = rep(c("GD", "SGD"), times = c(length(loss1), length(loss2))),
  epoch = c(seq_len(n1), seq_len(n2) / n),
  loss = c(loss1 - loss_optim, loss2 - loss_optim),
  iteration = c(seq_len(n1), seq_len(n2))
)
epochs_plot <- ggplot(conv_data, aes(epoch, loss, col = method)) +
  geom_line() +
  scale_y_log10() +
  labs(y = "Suboptimality", color = NULL)

iteration_plot <- ggplot(conv_data, aes(iteration, loss, col = method)) +
  geom_line() +
  scale_y_log10() +
  labs(y = "Suboptimality", color = NULL)

ggpl <- epochs_plot +
  iteration_plot +
  plot_layout(axes = "collect_y", guides = "collect")

ggpl
```

##

```{r}
#| echo: false

set.seed(1)

f <- function(x) x^4 - 2 * x^2 + 1 * x
df <- function(x) 4 * x^3 - 4 * x + 1

# Parameters
eta <- 0.07
steps <- 25
x_gd <- numeric(steps)
x_sgd <- numeric(steps)
x_gd[1] <- x_sgd[1] <- 1.4

for (k in 1:(steps - 1)) {
  # Exact gradient descent
  x_gd[k + 1] <- x_gd[k] - eta * df(x_gd[k])

  # Noisy gradient descent (SGD-like)
  noise <- rnorm(1, sd = 4) # stochasticity
  x_sgd[k + 1] <- x_sgd[k] - eta * (df(x_sgd[k]) + noise)
}

plot_curves <- function(main) {
  curve(
    f,
    from = -1.6,
    to = 1.7,
    n = 200,
    lwd = 2,
    col = "gray60",
    ylab = "",
    xlab = "",
    axes = TRUE,
    yaxt = if (main == "GD") "s" else "n",
    main = main
  )
}

plot_iteration <- function(i, x_gd, x_sgd) {
  x_gd_i <- x_gd[1:i]
  x_sgd_i <- x_sgd[1:i]

  opar <- par(no.readonly = TRUE)

  par(mfrow = c(1, 2), oma = c(4.1, 4.1, 0.1, 0.3), mai = c(0.1, 0.2, 0.5, 0.0))

  plot_curves("GD")
  lines(x_gd_i, f(x_gd_i), col = "royalblue", lwd = 1)
  points(x_gd_i, f(x_gd_i), col = "royalblue", pch = 19, cex = 0.8)

  plot_curves("SGD")
  lines(x_sgd_i, f(x_sgd_i), col = "darkorange", lwd = 1)
  points(x_sgd_i, f(x_sgd_i), col = "darkorange", pch = 19, cex = 0.8)

  mtext(expression(theta), side = 1, outer = TRUE, line = 2)
  mtext(expression(H(theta)), side = 2, outer = TRUE, line = 2)

  par(opar)
}
```

```{r}
#| echo: false
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: "The stochasticity in SGD sometimes helps it escape local minima."

plot_iteration(1, x_gd, x_sgd)
```

##

```{r}
#| echo: false
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: "The stochasticity in SGD sometimes helps it escape local minima."

plot_iteration(2, x_gd, x_sgd)
```

##

```{r}
#| echo: false
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: "The stochasticity in SGD sometimes helps it escape local minima."

plot_iteration(3, x_gd, x_sgd)
```

##

```{r}
#| echo: false
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: "The stochasticity in SGD sometimes helps it escape local minima."

plot_iteration(5, x_gd, x_sgd)
```

##

```{r}
#| echo: false
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: "The stochasticity in SGD sometimes helps it escape local minima."

plot_iteration(10, x_gd, x_sgd)
```

##

```{r}
#| echo: false
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: "The stochasticity in SGD sometimes helps it escape local minima."

plot_iteration(15, x_gd, x_sgd)
```

##

```{r}
#| echo: false
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: "The stochasticity in SGD sometimes helps it escape local minima."

plot_iteration(20, x_gd, x_sgd)
```

##

```{r}
#| echo: false
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: "The stochasticity in SGD sometimes helps it escape local minima."

plot_iteration(steps, x_gd, x_sgd)
```

## Learning Rate

The learning rate $\gamma_n$ is crucial for SGD, but hard to set.

\medskip\pause

Convergence theorem only gives a few hints.

. . .

### A Class of Decay Schedules

$$
  \gamma_n = \frac{\gamma_0 K}{K + n^a} = \frac{\gamma_0}{1 + K^{-1} n^a}
$$
with initial learning rate $\gamma_0 > 0$ and constants $K, a > 0$.

\medskip\pause

Convergence is ensured by Robbins--Monro if $a \in (0.5, 1]$.

\medskip\pause

Fixing the exponent $a$ and picking a target rate, $\gamma_1$, to be reached
after $n_1$ steps, we can solve for $K$ and get
$$
  K = \frac{n_1^a \gamma_1}{\gamma_0 - \gamma_1}.
$$

##

```{r}
#| label: decay
#| cache: true

decay_scheduler <- function(
  gamma0 = 1,
  a = 1,
  K = 1,
  gamma1 = NULL,
  n1 = NULL,
  t0 = gamma0
) {
  force(a)
  # Keep the existing logreg_sgd() interface while using gamma in the slides.
  gamma0 <- t0

  if (!is.null(gamma1) && !is.null(n1)) {
    K <- n1^a * gamma1 / (gamma0 - gamma1)
  }

  b <- gamma0 * K

  function(n) b / (K + n^a)
}
```

##

```{r, dependson="decay"}
#| label: decay-fig
#| echo: false
#| fig-height: 3
#| fig-width: 4
#| fig-cap: "Examples of learning rate schedules,
#|   using $\\gamma_n = \\frac{\\gamma_0}{1 + K^{-1} n^{a}}$."

grid_par <- expand.grid(
  n = seq(0, 1000, 2),
  K = c(25, 50, 100),
  a = c(0.6, 1),
  gamma0 = 1
)

grid_par <- dplyr::mutate(grid_par, rate = gamma0 * K / (K + n^a))

ggplot(grid_par, aes(n, rate, color = factor(K), linetype = factor(a))) +
  geom_line() +
  scale_color_discrete("K") +
  scale_linetype_discrete("a") +
  guides(color = guide_legend(order = 1), linetype = guide_legend(order = 2))
```

## So, Is The Problem of Setting Learning Rate Solved?

No, not really, because it's still hard to pick $\gamma_0$, $K$, and $a$.

```{r}
#| echo: false

set.seed(123)

n <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(n, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(n, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())

loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

beta1 <- seq(-1, 3, length = 100)
beta2 <- seq(-1, 3, length = 100)

loss_2d_vectorized <- Vectorize(function(b1, b2) loss_2d(b1, b2, X, y))

z <- outer(beta1, beta2, loss_2d_vectorized)

step_size_illustration <- function(a) {
  res_sgd <- logreg_sgd(X, y, max_epochs = 100, batch_size = 1, a = a, t0 = 0.5)

  pal <- palette.colors(palette = "Okabe-Ito")

  contour(beta1, beta2, z, col = "dark grey", drawlabels = FALSE, asp = 1)

  lines(res_sgd$beta_history[1, ], res_sgd$beta_history[2, ], col = pal[3])
  points(
    res_sgd$beta_history[1, ],
    res_sgd$beta_history[2, ],
    col = pal[3],
    pch = 19,
    cex = 0.5
  )

  points(
    res_glm$coefficients[1],
    res_glm$coefficients[2],
    col = pal[2],
    pch = 19,
    cex = 1
  )
}
```

::: {.columns}
::::: {.column width="47%"}
```{r}
#| echo: false
#| fig-height: 3
#| fig-cap: SGD on the logistic regression example with $a = 0.6$ (slow decay).

step_size_illustration(0.6)
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| echo: false
#| fig-height: 3
#| fig-cap: SGD on the logistic regression example with $a = 0.9$ (fast decay).

step_size_illustration(0.9)
```
:::::
:::

##

\begin{algorithm}[H]
  \caption{Stochastic gradient descent (with mini-batches)}
  \KwData{Initial parameter $\theta_0$ and learning-rate schedule $(\gamma_n)$}
  \For{$n = 0,1,\dots$}{
    $I_n \gets$ mini-batch of $m$ indices\;

    $\rho_n \gets \frac{1}{m}\sum_{i\in I_n}\nabla_\theta L(y_i,x_i,\theta_n)$\;

    $\theta_{n+1} \gets \theta_n - \gamma_n\rho_n$\;
    \If{converged}{
      \Return{$\theta_{n+1}$}
    }
  }
\end{algorithm}

## Implementation Details

### Nested Loops

Implementing an SGD algorithm typically means implementing a nested loop, and we
know this isn't R's strongest suite.

\medskip\pause

Next lecture, when we turn to Rcpp, you will have the tools to solve this
problem.

. . .

### Data Storage

SGD algorithms typically subset rows, or blocks of rows.

\medskip\pause

In a column-major order matrix, this is not optimal. Transposing the matrix
first can help.

## Exercise: Single-Variable Ridge Regression

Minimize
$$
  H_N(\beta) = \frac{1}{N} \sum_{i = 1}^N L(y_i,x_i,\beta), \quad\text{with}\quad L(y_i,x_i,\beta)
             = \frac{1}{2} (y_i - x_i \beta)^2 + \frac{\lambda}{2} \beta^2.
$$

. . .

```{r}
#| echo: false
#| include: false

f <- function(beta, xx, yy, lambda) {
  mean(0.5 * (xx * beta - yy)^2 + 0.5 * lambda * beta^2)
}

f_grad <- function(beta, i, xx, yy, lambda) {
  (xx[i] * beta - yy[i]) * xx[i] + lambda * beta
}

beta <- 3

x <- rnorm(100)
y <- x * beta + rnorm(100)

beta_grid <- seq(-2, 5, length.out = 200)

f_vec <- Vectorize(f, vectorize.args = "beta")

plot(beta_grid, f_vec(beta_grid, x, y, 0.6), type = "l")
```

### Step 1

Generate some data, for instance $x$ from `rnorm` and
`y <- x * beta + rnorm(N)`. You can pick $\beta$ and $N$ however you want.

. . .

### Step 2

Solve this problem using stochastic gradient descent. Plot loss in terms of
epochs and time. Test your method against `optimize()` or an analytical
solution.

. . .

### Step 3

Profile your algorithm.

## Summary

Stochastic gradient descent is a popular method for large-scale optimization.

\medskip\pause

Convergence results are relatively weak, but for many applications this does not
matter.

\medskip\pause

In practice, most implementations use mini-batches.

## Next Two Lectures

### Rcpp

We will look at how to use Rcpp to speed up R code.

\medskip\pause

Because SGD typically involves nested loops, these algorithms benefit greatly
from relying on C++.

. . .

### Variations on SGD

We will look at some variations on SGD that improve convergence in practice.

\medskip\pause

These include momentum and Nesterov acceleration.
