---
title: "Monte Carlo Integration and Importance Sampling"
---

{{< include _common.qmd >}}

```{r}
#| label: setup-common2
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(patchwork)
theme_set(theme_minimal(base_size = 10))
options(width = 60)
```

```{r}
#| label: setup
#| include: false

old_options <- options(digits = 4)
```

## Today

### Monte Carlo Methods

General class of methods for estimating integrals and means.

\medskip\pause

Rejection sampling is a type of Monte Carlo method.

. . .

### Importance Sampling

Focus on integrating functions and estimating means.

\medskip\pause

Useful when we cannot analytically compute the integral.

\medskip\pause

It is also a variance reduction technique.

## Monte Carlo Methods

### Basic, Simple Idea

Use random sampling to obtain numerical results.

\medskip\pause

Useful when analytical solutions are infeasible.

. . .

### Monte Carlo Integration

With $X_1, \ldots, X_N$ i.i.d. with density $f$ and $\E|h(X_1)| < \infty$,
$$
  \hat{\mu}_{\mathrm{MC}} = \frac{1}{N} \sum_{i=1}^N h(X_i) \rightarrow \mu
                          = \E h(X_1) = \int h(x) f(x) \ \mathrm{d}x
$$
for $N \to \infty$ by the law of large numbers (LLN).

## Monte Carlo Integration

If $\sigma^2_{\mathrm{MC}} < \infty$, the central limit theorem (CLT) gives
$$
  \frac{1}{N} \sum_{i=1}^N h(X_i) \overset{\mathrm{approx}}
  \sim \mathcal{N}(\mu, \sigma^2_{\mathrm{MC}} / N)
$$
where
$$
  \sigma^2_{\mathrm{MC}} = \V h(X_1) = \int (h(x) - \mu)^2 f(x) \ \mathrm{d}x.
$$

. . .

We can estimate $\sigma^2_{\mathrm{MC}}$ using the empirical variance
$$
  \hat{\sigma}^2_{\mathrm{MC}}
  = \frac{1}{N - 1} \sum_{i=1}^N (h(X_i) - \hat{\mu}_{\mathrm{MC}})^2,
$$
then the variance of $\hat{\mu}_{\mathrm{MC}}$ is estimated as
$\hat{\sigma}^2_{\mathrm{MC}} / N$ and an approximate 95% confidence interval
for $\mu$ is
$$
  \hat{\mu}_{\mathrm{MC}} \pm 1.96 \frac{\hat{\sigma}_{\mathrm{MC}}}{\sqrt{N}}.
$$

## Example: Gamma Distribution

::: {.columns}
::::: {.column width="47%"}
We want to estimate the mean of a $\mathrm{Gamma}(8,1)$ distribution.

. . .

```{r, echo=-1}
#| label: gamma-sim

set.seed(1234)
B <- 1000
x <- rgamma(B, 8)

mu_hat <- cumsum(x) / 1:B
sigma_hat <- sd(x)
mu_hat[B] # Theoretical: 8
sigma_hat # Theoretical: sqrt(8)
```
:::::

::::: {.column width="47%"}
```{r}
#| label: gamma-fig
#| echo: false
#| fig-cap: "Density of a Gamma(8,1) distribution with mean indicated."

x <- seq(0, 20, length.out = 100)
y <- dgamma(x, 8)

tibble(x, y) |>
  ggplot(aes(x, y)) +
  geom_line() +
  geom_vline(
    xintercept = 8,
    linetype = "dashed",
    color = "steelblue"
  ) +
  annotate("text", x = 7, y = 0.02, label = "E(X)", hjust = -0.5)
```
:::::
:::

##

```{r}
#| label: gamma-fig1
#| fig-height: 2.8
#| fig-width: 5
#| echo: false
#| fig-cap: >-
#|   Convergence of the Monte Carlo estimate of the mean of a Gamma(8,1)
#|   distribution.

p <- ggplot(tibble(x = 1:B, y = mu_hat), aes(x, y)) +
  geom_hline(
    yintercept = 8,
    linetype = "dashed",
    col = "steelblue4"
  ) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(6, 10))

p +
  geom_ribbon(
    aes(
      ymin = mu_hat - 1.96 * sigma_hat / sqrt(1:B),
      ymax = mu_hat + 1.96 * sigma_hat / sqrt(1:B)
    ),
    fill = "dark orange",
    alpha = 0.4
  ) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(6, 10))
```

## Choosing the Sample Size

Monte Carlo standard errors decrease as $1 / \sqrt{N}$: halving the error
requires four times as many samples.

\medskip\pause

Use a pilot run to estimate the variance and plan a \alert{fresh run} with a
fixed sample size:
$$
  N
  \approx \left( \frac{1.96\,\hat{\sigma}_{\mathrm{pilot}}}{\varepsilon} \right)^2.
$$

. . .

This targets a 95% confidence-interval half-width of $\varepsilon$.

\pdfpcnote{
Round the planned sample size up to an integer. Use independent samples for the
fresh run, fixing its size before generating them. This is approximate
planning: the pilot variance estimate and the normal approximation can be
inaccurate, so the requested precision is not guaranteed. Repeatedly checking
an ordinary confidence interval and stopping when it is narrow does not
guarantee coverage at the resulting data-dependent stopping time.
}

## Importance Sampling

When we are only interested in Monte Carlo integration, we do not need to sample
from the target distribution.

. . .

### Basic Idea

Some regions of the target distribution are often more important than others.

\medskip\pause

So let's \alert{oversample} these regions, but correct for this by
\alert{weighting} the samples.

\medskip\pause

We want to find $\mu = \E h(X)$. Observe that
$$
  \mu = \int h(x) f(x) \ \mathrm{d}x
      = \int h(x) \frac{f(x)}{g(x)} g(x) \ \mathrm{d}x
      = \int h(x) w^*(x) g(x) \ \mathrm{d}x
$$
whenever $g$ is a density satisfying
$$
  g(x) = 0 \Rightarrow f(x) = 0.
$$

## Weights

With $X_1, \ldots, X_N$ i.i.d. with density $g$, define the *weights*
$$
  w^*(X_i) = \frac{f(X_i)}{g(X_i)}.
$$

. . .

$g$ is the *importance* density and $f/g$ is called the *likelihood ratio*.

. . .

### Estimator

$$
  \hat{\mu}_{\mathrm{IS}}^* := \frac{1}{N} \sum_{i=1}^N h(X_i)w^*(X_i).
$$

It has mean $\mu$.

##

```{r}
#| label: importance-sampling-plot
#| echo: false
#| fig-cap: "Target and importance densities with corresponding importance weights."
#| fig-width: 5
#| fig-height: 3

x <- seq(-5, 5, length.out = 200)
target <- dnorm(x, mean = 0, sd = 1)
importance <- dnorm(x, mean = 1, sd = 1)

p1 <- tibble(x, target, importance) |>
  pivot_longer(
    cols = c(target, importance),
    names_to = "Density",
    values_to = "value"
  ) |>
  ggplot(aes(x, value, color = Density, linetype = Density)) +
  geom_line() +
  scale_color_manual(
    values = c(target = "black", importance = "steelblue")
  ) +
  scale_linetype_manual(
    values = c(target = "solid", importance = "dashed")
  ) +
  theme(legend.position.inside = c(0.87, 0.82)) +
  guides(
    color = guide_legend(position = "inside"),
    linetype = guide_legend(position = "inside")
  ) +
  labs(
    y = "Density",
    x = "x",
    color = "Distribution",
    linetype = "Distribution"
  )

w <- target / importance

p2 <- tibble(x, w) |>
  ggplot(aes(x, w)) +
  geom_line() +
  labs(x = "x", y = "Weight")

p1 / p2 + plot_layout(heights = c(2, 1), axes = "collect")
```

## Gamma Importance Sampling

::: {.columns}
::::: {.column width="47%"}
Use $X = 10 + 3T$, where $T \sim t_3$. This proposal has heavier tails than the
Gamma target.

```{r, echo=-1}
#| label: gamma-sim-IS

set.seed(1234)
B <- 1000
t <- rt(B, df = 3)
x <- 10 + 3 * t
g_x <- dt(t, df = 3) / 3
w_star <- dgamma(x, 8) / g_x
cum_mean <- cumsum(x * w_star)
mu_hat_IS <- cum_mean / (1:B)
mu_hat_IS[B]
```

Close to the theoretical value (8).
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: gammasim-fig
#| echo: false
#| fig-height: 3.2
#| fig-width: 2.7

z <- seq(-12, 12, length.out = 100) + 10
f_x <- dgamma(z, 8)
g_x <- dt((z - 10) / 3, df = 3) / 3
w <- f_x / g_x

p1 <- tibble(x = z, target = f_x, importance = g_x) |>
  pivot_longer(c(target, importance)) |>
  ggplot(aes(x, value, color = name, linetype = name)) +
  geom_line() +
  guides(color = guide_legend(position = "inside")) +
  scale_color_manual(
    values = c(target = "black", importance = "steelblue")
  ) +
  scale_linetype_manual(
    values = c(target = "solid", importance = "dashed")
  ) +
  theme(legend.position.inside = c(0.8, 0.8)) +
  labs(y = "Density", color = "Density", linetype = "Density")

p2 <- tibble(x = z, w) |>
  ggplot(aes(x, w)) +
  geom_line()

p1 / p2 + plot_layout(axes = "collect")
```
:::::
:::

##

```{r}
#| label: gamma-IS-fig1
#| echo: false
#| fig-cap: >-
#|   Convergence of the importance sampling estimate of the mean of a Gamma(8,1)
#|   distribution using a Student-t proposal with 3 degrees of freedom, location
#|   10, and scale 3.
#| fig-width: 5
#| fig-height: 2.5

mu_hat_IS <- cumsum(x * w_star) / (1:B)
sigma_hat_IS <- sd(x * w_star)

p <- ggplot(tibble(x = 1:B, y = mu_hat_IS), aes(x, y)) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(6, 10)) +
  geom_hline(
    yintercept = 8,
    linetype = "dashed",
    col = "steelblue4"
  )
p
```

## Importance Sampling Variance

Again by the LLN, assuming $\E_f|h(X)| < \infty$,
$$
  \hat{\mu}_{\mathrm{IS}}^* \rightarrow \E h(X_1) w^*(X_1)
                            = \mu \quad \text{as } N \to \infty.\pause
$$
If $\sigma^{*2}_{\mathrm{IS}} < \infty$, the CLT gives
$$
  \hat{\mu}_{\mathrm{IS}}^* \overset{\mathrm{approx}}
  \sim \mathcal{N}(\mu, \sigma^{*2}_{\mathrm{IS}} / N)
$$
where
$$
  \sigma^{*2}_{\mathrm{IS}} = \V \big( h(X_1)w^*(X_1)\big)
                            = \int (h(x) w^*(x) - \mu)^2 g(x) \ \mathrm{d}x.
$$

## Importance Sampling Variance

The IS variance can be estimated analogously to the MC variance,
$$
  \hat{\sigma}^{*2}_{\mathrm{IS}}
  = \frac{1}{N - 1} \sum_{i=1}^N (h(X_i)w^*(X_i) - \hat{\mu}_{\mathrm{IS}}^*)^2.
$$

. . .

An approximate 95% confidence interval is
$$
  \hat{\mu}^*_{\mathrm{IS}}
  \pm 1.96 \frac{\hat{\sigma}^*_{\mathrm{IS}}}{\sqrt{N}}.
$$

\medskip\pause

We may have $\sigma^{*2}_{\mathrm{IS}} > \sigma^2_{\mathrm{MC}}$ or
$\sigma^{*2}_{\mathrm{IS}} < \sigma^2_{\mathrm{MC}}$ depending on $h$ and $g$.

. . .

### Goal

Choose $g$ so that $|h(x)| w^*(x)$ becomes as constant as possible.

## Gamma Importance Sampling

```{r}
#| label: gamma-IS-sd
#| echo: true

sigma_hat_IS <- sd(x * w_star)
sigma_hat_IS
```

. . .

```{r}
#| label: gamma-IS-fig2
#| echo: false
#| fig-width: 4.5
#| fig-height: 2
#| fig-cap: "Gamma mean estimate with approximate pointwise 95% intervals."

p +
  geom_ribbon(
    aes(
      ymin = mu_hat_IS - 1.96 * sigma_hat_IS / sqrt(1:B),
      ymax = mu_hat_IS + 1.96 * sigma_hat_IS / sqrt(1:B)
    ),
    fill = "darkorange",
    alpha = 0.4
  ) +
  geom_line() +
  geom_point(size = 0.5)
```

```{r}
#| label: test
#| echo: false
#| eval: false

mu_hat_IS <- replicate(10000, {
  x <- 10 + 3 * rt(B, df = 3)
  w_star <- dgamma(x, 8) / (dt((x - 10) / 3, df = 3) / 3)
  mean(x * w_star)
})
sd(mu_hat_IS) * sqrt(B)
```

## Accounting for the Integrand

Suppose we want to estimate
$$
  \E(X^4) = \int x^4 f(x) \ \mathrm{d}x = 3
$$
where $f(x)$ is the standard normal density.

\medskip\pause

Direct Monte Carlo samples from $f$, placing most draws near zero.

\medskip\pause

The \alert{integrand} $x^4f(x)$ peaks at $x = \pm2$ and decays to zero in the
far tails. We want more draws around those peaks.

\medskip\pause

Try a wider Gaussian: $g = \mathcal{N}(0,4)$, with standard deviation $2$. This
puts more mass near $\pm2$ and reduces the variance of $X^4f(X)/g(X)$.

\pdfpcnote{
The proposal should account for the shape of the integrand. Nonlinearity alone
does not tell us which proposal to choose. Differentiating x^4 f(x) gives
x^3 (4 - x^2) f(x), so the two maxima are at -2 and 2. The wider Gaussian is a
simple approximation to this shape; it is not the optimal proposal. The next
two slides illustrate the improvement and give the ideal density.
}

##

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: is-normal-plot
#| echo: false
#| fig-height: 2.65
#| fig-cap: >-
#|   Direct MC: $g = f = \mathcal{N}(0,1)$.

set.seed(1)
N <- 1000
x_norm <- rnorm(N)
target <- dnorm(x_norm, 0, 1)
importance_norm <- dnorm(x_norm, 0, 1)
weights_norm <- target / importance_norm
w_h <- x_norm^4 * weights_norm

x_grid <- seq(-6, 6, length.out = 200)
dens_df <- tibble(
  x = x_grid,
  target = dnorm(x_grid, 0, 1),
  integrand = x_grid^4 * dnorm(x_grid, 0, 1)
) |>
  pivot_longer(
    cols = c(target, integrand),
    names_to = "Type",
    values_to = "value"
  )

p1 <- ggplot(
  dens_df,
  aes(x, value, color = Type, linetype = Type)
) +
  geom_line() +
  annotate(
    "point",
    x = c(-2, 2),
    y = 2^4 * dnorm(2),
    color = "darkorange"
  ) +
  scale_x_continuous(breaks = c(-4, -2, 0, 2, 4)) +
  labs(y = "Density / Integrand") +
  scale_color_manual(
    values = c(target = "black", integrand = "darkorange"),
    labels = c(
      target = expression(f(x) == g(x)),
      integrand = expression(h(x) * f(x))
    )
  ) +
  scale_linetype_manual(
    values = c(target = "solid", integrand = "dashed"),
    labels = c(
      target = expression(f(x) == g(x)),
      integrand = expression(h(x) * f(x))
    )
  ) +
  guides(
    color = guide_legend(position = "inside"),
    linetype = guide_legend(position = "inside")
  ) +
  labs(color = NULL, linetype = NULL) +
  theme(legend.position.inside = c(0.85, 0.85))

p1b <- ggplot(tibble(x = w_h), aes(x)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  labs(x = expression(h(x) * w^"*"))

p1 / p1b + plot_layout(heights = c(2, 1))
```
:::::

::::: {.column width="47%"}
```{r}
#| label: is-wide-normal-plot
#| echo: false
#| fig-height: 2.65
#| fig-cap: >-
#|   Importance sampling: $g = \mathcal{N}(0,4)$.

set.seed(1)
x_wide <- rnorm(N, sd = 2)
target_wide <- dnorm(x_wide)
importance_wide <- dnorm(x_wide, sd = 2)
weights_wide <- target_wide / importance_wide
w_h_wide <- x_wide^4 * weights_wide

dens_df_wide <- tibble(
  x = x_grid,
  target = dnorm(x_grid, 0, 1),
  importance = dnorm(x_grid, sd = 2),
  integrand = x_grid^4 * dnorm(x_grid, 0, 1)
) |>
  pivot_longer(
    cols = c(target, importance, integrand),
    names_to = "Type",
    values_to = "value"
  )

p2 <- ggplot(
  dens_df_wide,
  aes(x, value, color = Type, linetype = Type)
) +
  geom_line() +
  annotate(
    "point",
    x = c(-2, 2),
    y = 2^4 * dnorm(2),
    color = "darkorange"
  ) +
  scale_x_continuous(breaks = c(-4, -2, 0, 2, 4)) +
  labs(y = "Density / Integrand") +
  scale_color_manual(
    values = c(
      target = "black",
      importance = "royalblue",
      integrand = "darkorange"
    ),
    labels = c(
      target = expression(f(x)),
      importance = expression(g(x)),
      integrand = expression(h(x) * f(x))
    )
  ) +
  guides(
    color = guide_legend(position = "inside"),
    linetype = guide_legend(position = "inside")
  ) +
  scale_linetype_manual(
    values = c(
      target = "solid",
      importance = "dashed",
      integrand = "dotdash"
    ),
    labels = c(
      target = expression(f(x)),
      importance = expression(g(x)),
      integrand = expression(h(x) * f(x))
    )
  ) +
  labs(color = NULL, linetype = NULL) +
  theme(legend.position.inside = c(0.85, 0.85))

p2b <- ggplot(tibble(x = w_h_wide), aes(x)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  labs(x = expression(h(x) * w^"*"))

p2 / p2b + plot_layout(heights = c(2, 1))
```
:::::
:::

. . .

For the same $N$, the estimator variances are
$$
  \V(\hat\mu_{\mathrm{MC}}) = \frac{96}{N},
  \qquad
  \V(\hat\mu_{\mathrm{IS}}^*) \approx \frac{7.93}{N}.
$$

\pdfpcnote{
Each histogram shows 1000 weighted contributions, with bin width 2. The
horizontal scales differ. Direct MC contributions are X^4, while the wider
Gaussian gives 2 X^4 exp(-3 X^2 / 8), which is bounded above by about 7.70.
The variances shown are theoretical, not estimates from these histograms:
Var under f of X^4 is E under f of X^8 minus 9 = 105 - 9 = 96.
For g = N(0,4), the second moment is
210 / (7/4)^{9/2}, giving variance approximately 7.925756.
The wider Gaussian reduces variance by a factor of about 12.
}

## The Ideal Importance Density

For $0 < \E_f|h(X)| < \infty$, the variance-minimizing density for ordinary IS
is
$$
  g_{\mathrm{opt}}(x) = \frac{|h(x)|f(x)}{\int |h(u)|f(u)\,\mathrm{d}u}.
$$

. . .

If $h \geq 0$, then $g_{\mathrm{opt}} = hf / \mu$. Every draw gives
$$
  \frac{h(X)f(X)}{g_{\mathrm{opt}}(X)} = \mu, \qquad X \sim g_{\mathrm{opt}},
$$
so the estimator has \alert{zero variance}.

\medskip\pause

But evaluating these weights requires the normalizing constant $\mu$, the
integral we want. Use the shape of $|h|f$ to guide a practical proposal.

\pdfpcnote{
Ordinary IS only requires proposal support wherever the integrand is nonzero.
For sign-changing h, the optimal proposal still minimizes variance, but the
weighted contributions generally have different signs and the variance is not
zero. This result is for ordinary IS, not self-normalized IS.
}

## Normalized Weights

If $f = c^{-1} q$ with $c$ unknown^[For instance common in Bayesian statistics.]
then
$$
  c = \int q(x) \ \mathrm{d}x = \int \frac{q(x)}{g(x)} g(x) \ \mathrm{d}x.
$$

. . .

And
$$
  \mu
  = \frac{\int h(x) w^*(x) g(x) \ \mathrm{d}x}{\int w^*(x) g(x) \ \mathrm{d}x},
$$
where
$$
  w^*(x) = \frac{q(x)}{g(x)}.
$$

\pdfpcnote{
Note that g is still the normalized density.
}

## IS with Normalized Weights

An importance sampling estimate of $\mu$ is thus
$$
  \hat{\mu}_{\mathrm{IS}}
  = \frac{\sum_{i=1}^N h(X_i) w^*(X_i)}{\sum_{i=1}^N w^*(X_i)} \pause
  = \sum_{i=1}^N h(X_i) w(X_i),
$$
where $w^*(X_i) = q(X_i) / g(X_i)$ and
$$
  w(X_i) = \frac{w^*(X_i)}{\sum_{i=1}^N w^*(X_i)}
$$
are the *normalized weights*.

\medskip\pause

This works regardless of the normalizing constant $c$, and also if an
unnormalized proposal density is used in the weights.

\medskip\pause

The estimator is generally biased for finite $N$, but is consistent under the
support and integrability assumptions.

## Gamma Sampling with Normalized Weights

Ignore the normalizing constants of the Gamma target and proposal $10 + 3T$,
$T \sim t_3$. Set $w^*(x) = 0$ for $x \leq 0$, and otherwise use
$$
  w^*(x) = x^7 e^{-x} \left( 1 + \frac{(x - 10)^2}{27} \right)^2.
$$

. . .

```{r, echo=-1}
#| label: gamma-sim-ISw

set.seed(1234)
B <- 1000
x <- 10 + 3 * rt(B, df = 3)
w_star <- numeric(B)
x_pos <- x[x > 0]
w_star[x > 0] <- exp(
  2 * log1p((x_pos - 10)^2 / 27) - x_pos + 7 * log(x_pos)
)

mu_hat_IS <- cumsum(x * w_star) / cumsum(w_star)
mu_hat_IS[B] # Theoretical value 8
```

##

```{r}
#| label: gamma-ISw-fig1
#| echo: false
#| fig-width: 4.5
#| fig-height: 3
#| fig-cap: >-
#|   Convergence of the importance sampling estimate of the mean of a Gamma(8,1)
#|   distribution using normalized weights.

p <- tibble(x = 1:B, y = mu_hat_IS) |>
  ggplot(aes(x, y)) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(6, 10)) +
  labs(y = expression(hat(mu)[IS]), x = "Number of samples")
p
```

## Variance for IS with Normalized Weights

Let $w_i = w(X_i)$ be the normalized weights. If $w^*(X)$ and $h(X)w^*(X)$ have
finite variances, the $\Delta$-method gives the standard error estimate
$$
  \widehat{\mathrm{se}}(\hat{\mu}_{\mathrm{IS}})
  = \sqrt{\frac{N}{N - 1} \sum_{i=1}^N w_i^2 \big(h(X_i) - \hat{\mu}_{\mathrm{IS}}\big)^2}.
$$

This accounts for uncertainty in both the weighted sum and the sum of weights.

\medskip\pause

An approximate 95% confidence interval is
$$
  \hat{\mu}_{\mathrm{IS}}
  \pm 1.96\,\widehat{\mathrm{se}}(\hat{\mu}_{\mathrm{IS}}).
$$

\medskip\pause

Normalizing constants cancel.

\pdfpcnote{
The factor N/(N-1) matches the sample-variance convention used earlier.
Equivalently, compute the sample standard deviation of
w-star times (h minus mu-hat), divide by the mean unnormalized weight,
and then divide by the square root of N.
}

## Gamma Normalized Weights: Standard Error

```{r}
#| label: sigma_hat_is_w

w <- w_star / sum(w_star)
weighted_residuals <- w * (x - mu_hat_IS[B])
se_hat_IS <- sqrt(B / (B - 1) * sum(weighted_residuals^2))
se_hat_IS
```

. . .

```{r}
#| label: gamma-ISw-fig2
#| echo: false
#| fig-width: 4.5
#| fig-height: 1.75
#| fig-cap: "Gamma mean estimate with approximate pointwise 95% intervals."

sigma_hat_IS_w <- se_hat_IS * sqrt(B)

p +
  geom_ribbon(
    aes(
      ymin = mu_hat_IS - 1.96 * sigma_hat_IS_w / sqrt(1:B),
      ymax = mu_hat_IS + 1.96 * sigma_hat_IS_w / sqrt(1:B)
    ),
    fill = "darkorange",
    alpha = 0.4
  ) +
  geom_line() +
  geom_point(size = 0.5)
```

## When to Use Importance Sampling?

### Rare Events

Estimating probabilities or expectations involving rare events.

. . .

### High-Variance Monte Carlo

Importance sampling can reduce variance by focusing samples where the integrand
is large.

. . .

### Difficult-to-Sample Targets

When direct sampling from the target distribution is hard, but sampling from a
related importance distribution is easy.

## Poor Importance Distribution

If the importance distribution assigns little mass to important regions,
weighted contributions can be highly variable and estimates unreliable.

\medskip\pause

### Remedies

More samples may help with poor overlap. They cannot repair missing support: if
$g = 0$ where $hf \neq 0$, ordinary IS misses part of the integral.

\medskip\pause

Redesign the importance density to cover the required support and put more mass
where the integrand is large. Self-normalized IS must cover the target support,
including where $h = 0$.

## Diagnostics for Importance Sampling

### Weight Distribution

Plot the weights and weighted integrand values to identify extreme
contributions.

. . .

### Effective Sample Size (ESS)

ESS summarizes weight concentration. A high ESS cannot rule out missed regions.

. . .

### Coefficient of Variation (CV)

CV quantifies the relative variability of the weights.

. . .

### Comparison with Standard Monte Carlo

Comparing with standard Monte Carlo shows whether importance sampling actually
improves efficiency.

## Weight Degeneracies

::: {.columns}
::::: {.column width="47%"}
Suppose that
$$
  X \sim \mathcal{N}(0, I_{2})
$$
and that we want to estimate $\E(h(X))$ where $h(X)$ is large near the origin.

\medskip\pause

If we use an importance density $g(x)$ from
$$
  \mathcal{N}(\mu, I_{2})
$$
with large $\|\mu\|$, most samples are far from the origin, so $h(X)$ is nearly
zero and weights are highly variable.
:::::

::::: {.column width="47%"}
```{r}
#| label: poor-importance-demo
#| echo: false
#| fig-height: 2.3
#| fig-cap: >-
#|   Target (blue) and importance (orange) densities with samples from the
#|   importance distribution.

library(mvtnorm)

set.seed(1)
d <- 2
N <- 1000
mu <- rep(2, d)
x <- matrix(rnorm(N * d, mean = mu), ncol = d)

colnames(x) <- c("x1", "x2")

grid <- expand.grid(
  x1 = seq(-5, 8, length.out = 100),
  x2 = seq(-5, 8, length.out = 100)
)
grid_mat <- as.matrix(grid)

grid$target <- dmvnorm(grid_mat, mean = rep(0, d))
grid$importance <- dmvnorm(grid_mat, mean = mu)

tibble(
  x1 = grid$x1,
  x2 = grid$x2,
  target = grid$target,
  importance = grid$importance
) |>
  ggplot(aes(x1, x2)) +
  geom_contour(
    aes(z = target),
    color = "steelblue",
    linetype = "dashed"
  ) +
  geom_contour(aes(z = importance), color = "darkorange") +
  geom_point(
    aes(x1, x2),
    data = as_tibble(x),
    alpha = 0.3,
    size = 0.5
  ) +
  labs(x = expression(X[1]), y = expression(X[2])) +
  coord_fixed()
```
:::::
:::

##

```{r}
#| label: weight-degeneracy-plot
#| fig-width: 5.3
#| fig-height: 3.2
#| echo: false
#| fig-cap: "Histogram and boxplot of importance weights, showing weight degeneracy."

target_density <- function(x) dmvnorm(x, mean = rep(0, d))
importance_density <- function(x) dmvnorm(x, mean = mu)
h <- function(x) exp(-sum(x^2) / 2)

w_star <- apply(
  x,
  1,
  function(row) target_density(row) / importance_density(row)
)
values <- apply(x, 1, h)
estimate <- sum(values * w_star) / sum(w_star)

df <- tibble(weights = w_star)

p1 <- ggplot(df, aes(weights)) +
  geom_histogram() +
  labs(x = "Importance Weights", y = "Frequency")

p2 <- ggplot(df, aes(weights)) +
  geom_boxplot(alpha = 0.5) +
  scale_y_discrete() +
  labs(x = "Importance Weights")

p1 + p2 + plot_layout(heights = c(2, 1), axes = "collect")
```

\pdfpcnote{
This indicates weight degeneracy and poor sampling efficiency.
}

## Effective Sample Size (ESS)

The weight-based effective sample size (ESS) summarizes how evenly the total
weight is distributed across samples:
$$
  \mathrm{ESS}
  = \frac{1}{\sum_{i=1}^N w_i^2}
  = \frac{\left( \sum_{i=1}^N w^*_i \right)^2}{\sum_{i=1}^N ({w^*_i})^2}
$$
where $w_i$ are the \alert{normalized} importance weights.

\medskip\pause

If all weights are equal, $\mathrm{ESS} = N$. If one weight dominates,
$\mathrm{ESS} \approx 1$, which warns of possible instability.

. . .

### Bivariate Normal Example

```{r}
#| label: ess-demo

ess <- (sum(w_star))^2 / sum(w_star^2)
ess # Out of N = 1000 samples.
```

## Coefficient of Variation (CV)

The coefficient of variation (CV) of the weights measures their relative
variability.

\medskip\pause

It is defined as
$$
  \mathrm{CV}
  = \frac{\sqrt{\frac{1}{N} \sum_{i=1}^N (w_i - \bar{w})^2}}{\bar{w}}
$$
where $\bar{w} = \frac{1}{N} \sum_{i=1}^N w_i$ is the mean weight.

. . .

### Interpretation

A high CV indicates uneven weights, which can make estimates unstable.

\medskip\pause

CV near 0 means nearly uniform weights. Even CV below 1 ($\mathrm{ESS} > N/2$)
does not guarantee accuracy: the integrand matters, and samples can miss
important regions.

. . .

### Connection to ESS

There is a direct relationship between ESS and CV:
$$
  \mathrm{ESS} = \frac{N}{1 + \mathrm{CV}^2}.
$$

## Choosing the Importance Distribution

We want low-variance estimates of
$$
  I = \int h(x) f(x) \ \mathrm{d}x
$$

. . .

The importance distribution determines the weights
$$
  w^*(x) = \frac{f(x)}{g(x)}
$$
but ordinary IS variance depends on $h(X)w^*(X)$, not just the weights. For
self-normalized IS, the asymptotic variance depends on $(h(X) - \mu)w^*(X)$.

## Principles for Choosing the Importance Distribution

### Cover the Support

Require $g(x) > 0$ wherever $|h(x)|f(x) > 0$ for ordinary IS, and wherever
$f(x) > 0$ for self-normalized IS, up to sets of measure zero.

. . .

### Match the Bulk

For ordinary IS, put more mass where $|h(x)|f(x)$ is large. For self-normalized
IS, the corresponding quantity is $|h(x) - \mu|f(x)$.

. . .

### Give it Heavy-Enough Tails

Use tails heavy enough that the weighted contributions have finite variance.
Heavier tails than the target are often a useful safeguard.

## Common Strategies

### Exponential Tilting

Multiply the target density by an exponential factor to favor a rare-event
region.

. . .

### Laplace Approximation

Use a Gaussian approximation centered at the mode of the target distribution.

. . .

### Adaptive Importance Sampling

Iteratively refine the importance distribution based on pilot runs.

. . .

### Mixture and Defensive Proposals

Combine components to capture multiple modes. A defensive component covers the
support and controls the tails of the weighted contributions.

## A Rare Event: No Hits

For $Z \sim \mathcal{N}(0,1)$, estimate
$$
  p = P(Z > 4) \approx 3.17 \times 10^{-5}.
$$

```{r, echo=-1}
#| label: rare-event-mc

set.seed(1)
n_rare <- 10000
rare_hits <- rnorm(n_rare) > 4
mean(rare_hits)
```

. . .

This is typical: with $N = 10{,}000$,
$$
  P(\text{no exceedances}) = (1 - p)^N \approx 0.729.
$$

The empirical standard error is also zero. We have missed the event, not
established that its probability is zero.

## A Rare Event: Exponential Tilting

For the standard normal density $f = \phi$, an \alert{exponential tilt} shifts
the mean:
$$
  g_\theta(x) = e^{\theta x-\theta^2 / 2} f(x) = \phi(x - \theta).
$$

Choose $\theta = 4$: $X_i \sim \mathcal{N}(4,1)$ and
$w^*(x) = f(x) / g_4(x) = e^{8-4x}$.

```{r}
#| label: rare-event-proposal
#| echo: false
#| fig-width: 5
#| fig-height: 1.6
#| fig-cap: "Target and proposal densities for the rare-event example."

rare_grid <- tibble(
  x = seq(-4, 8, length.out = 601),
  target = dnorm(x),
  proposal = dnorm(x, mean = 4)
)

rare_grid |>
  pivot_longer(c(target, proposal), names_to = "density") |>
  ggplot(aes(x, value, color = density, linetype = density)) +
  geom_area(
    data = filter(rare_grid, x > 4),
    aes(x = x, y = proposal),
    inherit.aes = FALSE,
    fill = "darkorange",
    alpha = 0.2
  ) +
  geom_line() +
  geom_vline(xintercept = 4, linetype = "dotted") +
  scale_color_manual(
    values = c(target = "royalblue", proposal = "darkorange")
  ) +
  scale_linetype_manual(
    values = c(target = "solid", proposal = "dashed")
  ) +
  labs(
    x = expression(x),
    y = "Density",
    color = NULL,
    linetype = NULL
  )
```

. . .

$$
  \hat{p}_{\mathrm{IS}}
  = \frac{1}{N} \sum_{i=1}^N \mathbf{1}\{X_i > 4\}\,e^{8-4X_i}.
$$

\pdfpcnote{
Exponential tilting multiplies the density by exp(theta times x) and divides
by the moment-generating function. For a standard normal, this normalizing
constant is exp(theta squared divided by 2). Completing the square gives a
normal distribution with mean theta and variance 1. At theta = 4, half the
proposal draws exceed 4. Shifting a general density need not be an exponential
tilt.

The ideal proposal is the standard normal conditional on exceeding 4. Its
normalizing constant is p. The shifted normal is a simple approximation that
makes exceedances common and has an evaluable, normalized density.
}

## A Rare Event: Repeated Runs

```{r}
#| label: rare-event-replicates
#| include: false

set.seed(2026)
p_rare <- pnorm(4, lower.tail = FALSE)
rare_estimates <- replicate(500, {
  z <- rnorm(n_rare)
  x <- rnorm(n_rare, mean = 4)
  c(MC = mean(z > 4), IS = mean((x > 4) * exp(8 - 4 * x)))
})

rare_runs <- tibble(
  run = seq_len(ncol(rare_estimates)),
  MC = rare_estimates["MC", ],
  IS = rare_estimates["IS", ]
) |>
  pivot_longer(
    c(MC, IS),
    names_to = "method",
    values_to = "estimate"
  ) |>
  mutate(
    method = factor(
      method,
      levels = c("MC", "IS"),
      labels = c("Ordinary MC", "Importance sampling")
    )
  )

rare_zero_percent <- 100 * mean(rare_estimates["MC", ] == 0)
rare_sd_ratio <- sd(rare_estimates["MC", ]) /
  sd(rare_estimates["IS", ])
```

500 independent runs per method, with $N = 10{,}000$ draws per run.

```{r}
#| label: rare-event-comparison
#| echo: false
#| fig-width: 5.2
#| fig-height: 2.3
#| fig-cap: >-
#|   Estimates of the rare-event probability $p$ from 500 independent runs of
#|   ordinary Monte Carlo and importance sampling.

ggplot(rare_runs, aes(run, estimate / p_rare, color = method)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_point(size = 0.6, alpha = 0.6) +
  facet_wrap(vars(method), nrow = 1) +
  scale_color_manual(
    values = c(
      "Ordinary MC" = "royalblue",
      "Importance sampling" = "darkorange"
    )
  ) +
  guides(color = "none") +
  labs(x = "Run", y = expression(hat(p) / p))
```

The dashed line is the truth. MC returned zero in
`r sprintf("%.1f", rare_zero_percent)`% of runs.

\medskip\pause

IS had about `r round(rare_sd_ratio)` times smaller empirical standard deviation
at the same sample size.

## Laplace Approximation

### Basic Idea

Take a Gaussian approximation of the target density at its mode $x^*$ as the
importance distribution.

. . .

### Motivation

The connection comes from the second-order Taylor expansion of $\log f(x)$ at an
interior mode $x^*$,
$$
  \log f(x) \approx \log f(x^*) - \frac{1}{2}(x - x^*)^T \Lambda (x - x^*)
$$
where $\Lambda$ is the Hessian of $-\log f(x)$ at $x^*$, assumed positive
definite.

\medskip\pause

Exponentiating both sides gives:
$$
  f(x)
  \approx f(x^*) \exp\left( -\frac{1}{2}(x - x^*)^T \Lambda (x - x^*) \right)
$$

## Laplace Approximation

### Steps

1. Find the mode $x^* = \argmax_x f(x)$.\pause
2. Compute the Hessian $\Lambda$ of $-\log f(x)$ at $x^*$.\pause
3. If $\Lambda$ is positive definite, use $\mathcal{N}(x^*, \Lambda^{-1})$ as
   the importance distribution.

. . .

### When to Use

Useful for smooth targets with an approximately Gaussian shape near the mode.
Check the tails before using it for importance sampling.

## Example: Gamma Distribution

::: {.columns}
::::: {.column width="47%"}
The density of the $\mathrm{Gamma}(3,1)$ distribution is
$$
  f(x) = \frac{1}{2} x^2 e^{-x}, \quad x > 0.
$$

. . .

The mode is at $x^* = 2$ and the Hessian of $-\log f(x)$ at $x^*$ is
$$
  \Lambda = \frac{\alpha - 1}{(x^*)^2} = \frac 1 2.
$$

\medskip\pause

So the Laplace approximation is $\mathcal{N}(2, 2)$, with standard deviation
$\sqrt{2}$.
:::::

::::: {.column width="47%"}
```{r}
#| label: laplace-approx-plot
#| echo: false
#| fig-cap: >-
#|   Target (solid black) and Laplace approximation (dashed blue) densities for
#|   a Gamma(3,1) distribution.

f_density <- function(x) dgamma(x, shape = 3, rate = 1) # Gamma(3,1)
log_f <- function(x) log(f_density(x))

x_star <- (3 - 1) / 1
lambda <- 2 / (x_star^2)
sd_laplace <- 1 / sqrt(lambda)

x_grid <- seq(-2, 10, length.out = 400)

df <- tibble(
  x = x_grid,
  f = f_density(x_grid),
  laplace = dnorm(x_grid, mean = x_star, sd = sd_laplace),
) |>
  pivot_longer(
    cols = c(f, laplace),
    names_to = "Density",
    values_to = "value"
  ) |>
  filter(value > 0)

# Plot
ggplot(
  df,
  aes(x = x, y = value, color = Density, linetype = Density)
) +
  geom_line() +
  scale_color_manual(
    values = c(f = "black", laplace = "steelblue")
  ) +
  scale_linetype_manual(
    values = c(f = "solid", laplace = "dashed")
  ) +
  theme(legend.position.inside = c(0.8, 0.8)) +
  guides(
    color = guide_legend(position = "inside"),
    linetype = guide_legend(position = "inside")
  ) +
  labs(y = "Density")
```
:::::
:::

## Adaptive Importance Sampling

Adaptive importance sampling iteratively refines the importance distribution
based on the samples obtained.

. . .

### Basic Idea

Start with an initial importance distribution and refine it based on samples and
weights.

. . .

### Steps

1. Choose an initial importance distribution $g_0(x)$ and set $n = 0$.\pause
2. Draw samples from the current proposal $g_n$ and compute weights.\pause
3. Fit $g_{n+1}$ using the weighted samples, then increment $n$. \pause
4. Repeat steps 2-3 until the proposal stabilizes.

## Moment-Matching Importance Sampling

One simple adaptive strategy is to use moment-matching to update a Gaussian
importance distribution.

\medskip\pause

At iteration $n$, draw samples $x_1, \ldots, x_N \sim g_n(x)$ and compute
weights
$$
  w_i^* = \frac{f(x_i)}{g_n(x_i)}.
$$

. . .

Then update the importance distribution parameters as $$
\begin{aligned}
  \mu_{n + 1}      &= \frac{\sum_{i=1}^N w_i^* x_i}{\sum_{i=1}^N w_i^*},\\
  \sigma^2_{n + 1} &= \frac{\sum_{i=1}^N w_i^* (x_i - \mu_{n + 1})^2}{\sum_{i=1}^N w_i^*}.
\end{aligned}
$$

\pdfpcnote{
This can be extended to multivariate targets using weighted covariance.
}

##

```{r}
#| echo: false
#| label: adaptive-importance-demo

library(ggplot2)
library(tibble)

# Target density: bimodal mixture
f_density <- function(x) {
  0.4 * dnorm(x, -2, 1) + 0.6 * dnorm(x, 3, 0.5)
}

# Adaptive importance sampling function
adaptive_is <- function(
  N = 1000,
  n_iter = 5,
  q_init_mean = 0,
  q_init_sd = 1
) {
  # Store proposal params and samples
  proposals <- list()
  samples <- list()
  weights <- list()

  mean_q <- q_init_mean
  sd_q <- q_init_sd

  for (i in 1:n_iter) {
    # Draw from current proposal
    x <- rnorm(N, mean = mean_q, sd = sd_q)
    w <- f_density(x) / dnorm(x, mean = mean_q, sd = sd_q)
    w_norm <- w / sum(w)

    # Store current step
    proposals[[i]] <- c(mean = mean_q, sd = sd_q)
    samples[[i]] <- x
    weights[[i]] <- w_norm

    # Update proposal using weighted moments
    mean_q <- sum(w_norm * x)
    sd_q <- sqrt(sum(w_norm * (x - mean_q)^2))
  }

  return(list(
    proposals = proposals,
    samples = samples,
    weights = weights
  ))
}

# Run adaptive importance sampling
set.seed(123)
res <- adaptive_is(N = 1000, n_iter = 5)

# Plotting function for a given iteration
plot_step <- function(iter) {
  proposal <- res$proposals[[iter]]
  x_grid <- seq(-6, 6, length.out = 400)
  df <- tibble(
    x = x_grid,
    target = f_density(x_grid),
    importance = dnorm(
      x_grid,
      mean = proposal["mean"],
      sd = proposal["sd"]
    )
  ) |>
    pivot_longer(
      cols = c(target, importance),
      names_to = "Density",
      values_to = "value"
    )

  # Calculate ESS and CV for current iteration
  w <- res$weights[[iter]]
  N <- length(w)
  ess <- (sum(w))^2 / sum(w^2)
  cv <- sqrt(mean((w - mean(w))^2)) / mean(w)

  ggplot(
    df,
    aes(x = x, y = value, color = Density, linetype = Density)
  ) +
    geom_line() +
    scale_color_manual(
      values = c(target = "black", importance = "steelblue")
    ) +
    scale_linetype_manual(
      values = c(target = "solid", importance = "dashed")
    ) +
    lims(x = c(-6, 6), y = c(0, 0.6)) +
    labs(
      y = "Density",
      subtitle = sprintf(
        "Iteration %d: ESS = %.0f, CV = %.1f",
        iter,
        ess,
        cv
      )
    )
}
```

```{r}
#| label: adaptive-importance-fig
#| echo: false
#| fig-show: animate
#| fig-width: 5.4
#| fig-height: 2.5
#| fig-cap: "Adaptive importance sampling with a Gaussian proposal."

for (iter in seq_along(res$proposals)) {
  print(plot_step(iter))
}
```

## Defensive Importance Sampling

Return to the $\mathrm{Gamma}(3,1)$ target and estimating $\E_f[X] = 3$. The
Laplace proposal $g_L = \mathcal{N}(2,2)$ has a right tail that is too light:
ordinary IS has \alert{infinite variance}.

\medskip\pause

Add a heavy-tailed component $g_t$, the density of $2 + \sqrt{2}T$ with
$T \sim t_3$:
$$
  g_{\mathrm{mix}}(x) = 0.9 g_L(x) + 0.1 g_t(x).
$$

. . .

Since $g_{\mathrm{mix}} \geq 0.1g_t$, the mixture inherits protection from the
Student-$t$ tails. For this Gamma mean, it restores \alert{finite variance}.

\pdfpcnote{
For h(x) = x, the second moment of an ordinary IS contribution is the integral
of x squared times f squared divided by g. For the Gaussian proposal, this
integrand is proportional to x to the sixth power times
exp(x squared divided by 4 minus 3x), whose integral diverges.

The Student-t density decays as x to the minus fourth power, so its second-moment
integrand decays as x to the tenth power times exp(-2x), which is integrable.
Since the mixture is at least 0.1 times the Student-t density, its second moment
is at most ten times the Student-t second moment. This restores finite variance,
although it need not give a smaller variance than direct Monte Carlo.
}

##

```{r}
#| label: gamma-defensive-fig
#| echo: false
#| fig-width: 5.4
#| fig-height: 3
#| fig-cap: >-
#|   Gamma(3,1) target and proposals (left); importance weights on a logarithmic
#|   scale (right).

gamma_defensive <- tibble(
  x = seq(-3, 20, length.out = 601),
  target = dgamma(x, shape = 3, rate = 1),
  laplace = dnorm(x, mean = 2, sd = sqrt(2)),
  student = dt((x - 2) / sqrt(2), df = 3) / sqrt(2),
  mixture = 0.9 * laplace + 0.1 * student
)

defensive_colors <- c(
  target = "black",
  laplace = "royalblue",
  mixture = "darkorange"
)
defensive_lines <- c(
  target = "solid",
  laplace = "dashed",
  mixture = "dotdash"
)
defensive_labels <- c(
  target = "Target",
  laplace = "Laplace",
  mixture = "Mixture"
)

p_defensive_densities <- gamma_defensive |>
  filter(x <= 12) |>
  pivot_longer(
    c(target, laplace, mixture),
    names_to = "density"
  ) |>
  ggplot(aes(x, value, color = density, linetype = density)) +
  geom_line() +
  scale_color_manual(
    values = defensive_colors,
    labels = defensive_labels
  ) +
  scale_linetype_manual(
    values = defensive_lines,
    labels = defensive_labels
  ) +
  guides(
    color = guide_legend(position = "bottom"),
    linetype = guide_legend(position = "bottom")
  ) +
  labs(
    x = expression(x),
    y = "Density",
    color = NULL,
    linetype = NULL
  )

p_defensive_weights <- gamma_defensive |>
  filter(x > 0) |>
  transmute(
    x,
    laplace = target / laplace,
    mixture = target / mixture
  ) |>
  pivot_longer(
    c(laplace, mixture),
    names_to = "proposal",
    values_to = "weight"
  ) |>
  ggplot(aes(x, weight, color = proposal, linetype = proposal)) +
  geom_line() +
  scale_y_log10(
    breaks = 10^c(0, 10, 20, 30),
    labels = function(breaks) {
      parse(text = paste0("10^", log10(breaks)))
    }
  ) +
  scale_color_manual(
    values = defensive_colors,
    labels = defensive_labels
  ) +
  scale_linetype_manual(
    values = defensive_lines,
    labels = defensive_labels
  ) +
  guides(
    color = guide_legend(position = "bottom"),
    linetype = guide_legend(position = "bottom")
  ) +
  labs(
    x = expression(x),
    y = expression(f(x) / g(x)),
    color = NULL,
    linetype = NULL
  )

p_defensive_densities + p_defensive_weights
```

## Exercise: Von Mises Importance Sampling

### Step 1

Estimate the arithmetic mean of the von Mises distribution on $[-\pi,\pi]$ with
$\mu = 0$ and $\kappa = 2$. Plot convergence toward $\E X = 0$.

```{r}
#| label: vonmises-density

dvmises <- function(x, mu = 0, kappa = 2) {
  ifelse(
    abs(x) <= pi,
    exp(kappa * cos(x - mu)) / (2 * pi * besselI(kappa, 0)),
    0
  )
}
```

. . .

### Step 2

Compute the estimate of the mean and its standard error.

. . .

### Step 3

Compute the effective sample size and coefficient of variation of the importance
weights. Experiment with different importance densities to see if you find a
better one.

```{r}
#| label: vmises-importance
#| include: false

vonmises_importance <- function(N) {
  x <- rnorm(N, 0, pi / 2)
  w <- dvmises(x) / dnorm(x, 0, pi / 2)
  cumsum(x * w) / 1:N
}

x <- seq(-pi, pi, length.out = 100)
plot(x, dvmises(x))
lines(x, dnorm(x, 0, pi / 2))
```

\pdfpcnote{
For nonzero mu, the arithmetic mean on this fixed interval generally differs
from mu, the circular mean direction. Keep mu = 0 for this exercise.
}

## Summary

### Monte Carlo Methods

Estimate integrals and means using random sampling when analytic solutions are
not possible.

. . .

### Importance Sampling

Focus samples where they matter most, but choose your proposal carefully to
avoid large or variable weights.

. . .

### Diagnostics

Examine weights, weighted integrand values, and ESS to detect instability. These
checks cannot guarantee accuracy or reveal every missed region.

. . .

### Picking Importance Distributions

It's hard! But use adaptive and defensive methods to help fix poor coverage and
improve robustness.

## Next Time

### Optimization

We introduce optimization methods for minimizing and maximizing functions.

\medskip\pause

We'll focus on **convex optimization**, a special class of optimization problems
that are easier to solve and have nice properties.

\lectureappendix

# Appendix

## Variance for IS with Normalized Weights

Computing the variance of IS with normalized weights is more complicated because
the estimator is a \alert{ratio of random variables}.

\medskip\pause

Write $c = \E_g[w^*(X)]$. If both $h(X)w^*(X)$ and $w^*(X)$ have finite
variances, the multivariate CLT gives
$$
  \frac{1}{N} \sum_{i=1}^N \begin{bmatrix} h(X_i) w^*(X_i) \\ w^*(X_i) \end{bmatrix}
  \overset{\mathrm{approx}}{\sim}
  \mathcal{N}\left(
    c \begin{bmatrix} \mu \\ 1\end{bmatrix},
    \frac{1}{N} \begin{bmatrix} \sigma^{2}_{\mathrm{IS}} & \gamma \\ \gamma & \sigma^2_{w^*} \end{bmatrix}
  \right),
$$
where $$
\begin{aligned}
  \sigma^2_{\mathrm{IS}} & = \V \big(h(X_1) w^*(X_1)\big) \\
  \gamma & = \cov (h(X_1) w^*(X_1), w^*(X_1)) \\
  \sigma^2_{w^*} & = \V (w^*(X_1))
\end{aligned}
$$

## Importance Sampling Variance

We can then apply the $\Delta$-method with $t(x, y) = x / y$.

\medskip\pause

Observe that $\nabla t(x, y) = \begin{bmatrix} 1 / y \\ -x / y^2 \end{bmatrix}$
, whence

$$
  \nabla t(c\mu, c)^T \left(\begin{array}{cc} \sigma^{2}_{\mathrm{IS}} & \gamma \\ \gamma & \sigma^2_{w^*}
  \end{array} \right) \nabla t(c\mu, c) = \frac{\sigma^{2}_{\mathrm{IS}} + \mu^2 \sigma_{w^*}^2 - 2 \mu \gamma}{c^{2}}.
$$

. . .

By the $\Delta$-method,
$$
  \hat{\mu}_{\mathrm{IS}} \overset{\mathrm{approx}}{\sim} \mathcal{N}\left( \mu, \frac{\sigma^{2}_{\mathrm{IS}} + \mu^2 \sigma_{w^*}^2 - 2 \mu \gamma}{c^2 N} \right).
$$

. . .

Plug in empirical estimates, estimating $c$ when it is unknown by
$$
  \hat{c} = \frac{1}{N} \sum_{i=1}^N w^*(X_i).
$$

If both densities in the weights are unnormalized, $c$ is the ratio of their
normalizing constants.

\pdfpcnote{
hat(c) is just the MC estimate of c.
}

##

```{r}
#| label: delta-method-plot3D
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| fig-cap: Tangent plane approximation of the function $t(x,y) = x/y$ at the
#|   point $(2,3)$.

library(plot3D)

# Define function and means
t_fun <- function(x, y) x / y
mu_x <- 2
mu_y <- 3

# Grid for surface
x_seq <- seq(1, 3, length.out = 26)
y_seq <- seq(2, 4, length.out = 26)
z <- outer(x_seq, y_seq, t_fun)

# Tangent plane at (mu_x, mu_y)
grad_x <- 1 / mu_y
grad_y <- -mu_x / mu_y^2
tangent_plane <- function(x, y) {
  t_fun(mu_x, mu_y) + grad_x * (x - mu_x) + grad_y * (y - mu_y)
}
z_tangent <- outer(x_seq, y_seq, tangent_plane)

# Plot both surfaces together
persp3D(
  x_seq,
  y_seq,
  z,
  col = "steelblue",
  xlab = "x",
  ylab = "y",
  zlab = "t(x, y)",
  theta = 30,
  phi = 10,
  border = "black"
)

# Add tangent plane
persp3D(
  x_seq,
  y_seq,
  z_tangent,
  add = TRUE,
  col = "grey",
  alpha = 0.6
)

# Add mean point
points3D(
  mu_x,
  mu_y,
  t_fun(mu_x, mu_y),
  col = "darkorange",
  pch = 19,
  add = TRUE,
  cex = 1
)
```
