---
title: "Random Number Generation and Rejection Sampling"
---

{{< include _common.qmd >}}

```{r}
#| label: setup
#| include: false

library(profvis)

knitr::read_chunk(here::here("R", "vMSim.R"))

source(here::here("R", "vMSim.R"), keep.source = TRUE)
```

## Today

### Pseudo-Random Numbers

How do you simulate random numbers in R?

. . .

### Rejection Sampling

General and useful method for sampling from a target distribution

## Pseudo-Random Numbers

R uses deterministic algorithms to generate **pseudo-random** numbers.

\medskip\pause

The sequences are reproducible from a seed but designed to appear random.

\medskip\pause

A research field in itself. See `?RNG` in R for available algorithms.

. . .

### Mersenne Twister

The default in R, which generates integers in the range
$$
  \{0, 1, \ldots, 2^{32} - 1\}.
$$

\medskip\pause

Long period: $2^{19937} - 1$. Blocks of up to 623 consecutive integers occur
nearly equally often over a full period.

. . .

### Uniform Random Variables in R

`runif()` scales these pseudo-random integers to the unit interval, with
adjustments to exclude 0 and 1.

\pdfpcnote{
  Random numbers are used in many applications.
}

## Transformation Methods

What if we want to sample from a different distribution?

\medskip\pause

We could use **the transformation method**, which relies on the following fact:

\medskip

If $T : \mathcal{Z} \to \mathbb{R}$ is a map and $Z \in \mathcal{Z}$ is a random
variable we can sample, then we can sample $X = T(Z)$.

\begin{figure}
\centering
\begin{tikzpicture}[node distance=1cm, auto, >=stealth, thick]
  \node[draw, rectangle, rounded corners, fill=blue!10] (A) {Sample $Z$};
  \node[draw, rectangle, rounded corners, right = of A, fill=green!10] (B) {Apply $T$};
  \node[draw, rectangle, rounded corners, right = of B, fill=orange!10] (C) {Get $X = T(Z)$};
  \draw[->] (A) -- (B);
  \draw[->] (B) -- (C);
\end{tikzpicture}
\end{figure}

\pause

Transformation methods let us generate samples from complex distributions by
applying a function to samples from a simpler distribution.

\medskip

### Example {.example}

To sample $X \sim \mathrm{Exp}(\lambda)$ with rate $\lambda > 0$, use
$X = -\log(U)/\lambda$ with $U \sim \mathcal{U}(0, 1)$.

## Inversion Sampling

A type of **transformation** sampling.

\medskip\pause

If $F^{-1} : (0,1) \to \mathbb{R}$ is the generalized inverse of a distribution
function $F$ and $U$ is uniformly distributed on $(0, 1)$, then
$$
  F^{-1}(U)
$$
has distribution function $F$.

\medskip\pause

Computing $F^{-1}$ can be simple (e.g., for the exponential distribution) or
require a search or numerical approximation.

## Following the Inverse CDF

For $X \sim \mathrm{Exp}(1)$, $F(x) = 1 - e^{-x}$ for $x \geq 0$.

```{r}
#| label: inversion-cdf-arrows
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 2.2

inverse_example <- tibble(u = c(0.2, 0.6, 0.9)) |>
  mutate(x = -log1p(-u))

inverse_curve <- tibble(x = seq(0, 3, length.out = 300)) |>
  mutate(cdf = pexp(x))

for (stage in 1:3) {
  p <- ggplot(inverse_curve, aes(x, cdf)) +
    geom_line() +
    geom_point(
      data = inverse_example,
      aes(x = 0, y = u),
      color = "royalblue"
    ) +
    scale_y_continuous(breaks = c(0, 0.2, 0.6, 0.9, 1)) +
    coord_cartesian(xlim = c(0, 3), ylim = c(0, 1)) +
    labs(x = expression(x), y = expression(F(x)))
  if (stage >= 2) {
    p <- p +
      geom_segment(
        data = inverse_example,
        aes(x = 0, xend = x, y = u, yend = u),
        arrow = grid::arrow(length = grid::unit(0.07, "inches")),
        color = "royalblue"
      )
  }
  if (stage >= 3) {
    p <- p +
      geom_segment(
        data = inverse_example,
        aes(x = x, xend = x, y = u, yend = 0),
        arrow = grid::arrow(length = grid::unit(0.07, "inches")),
        color = "darkorange"
      )
  }
  print(p)
}
```

Move right from $U$ to the CDF, then down to $X = F^{-1}(U) = -\log(1 - U)$.

## Inversion for a Discrete Distribution

The **generalized inverse** is $F^{-1}(u) = \inf\{x : F(x) \geq u\}$.

```{r}
#| label: inversion-discrete-cdf
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 2.2

discrete_cdf <- tibble(
  left = c(-0.4, 0, 1, 2),
  right = c(0, 1, 2, 2.4),
  cdf = c(0, 0.2, 0.7, 1)
)

for (stage in 1:2) {
  p <- ggplot(discrete_cdf) +
    geom_segment(aes(
      x = left,
      xend = right,
      y = cdf,
      yend = cdf
    )) +
    geom_point(
      data = discrete_cdf[-1, ],
      aes(x = left, y = cdf)
    ) +
    geom_point(
      data = discrete_cdf[-4, ],
      aes(x = right, y = cdf),
      shape = 21,
      fill = "white"
    ) +
    scale_x_continuous(breaks = 0:2) +
    scale_y_continuous(breaks = c(0, 0.2, 0.7, 1)) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = expression(x), y = expression(F(x)))
  if (stage >= 2) {
    p <- p +
      annotate(
        "segment",
        x = -0.2,
        xend = -0.2,
        y = 0.2,
        yend = 0.7,
        color = "darkorange"
      ) +
      annotate(
        "segment",
        x = -0.2,
        xend = 1,
        y = 0.55,
        yend = 0.55,
        arrow = grid::arrow(length = grid::unit(0.07, "inches")),
        color = "darkorange"
      ) +
      annotate(
        "segment",
        x = 1,
        xend = 1,
        y = 0.55,
        yend = 0,
        arrow = grid::arrow(length = grid::unit(0.07, "inches")),
        color = "darkorange"
      ) +
      annotate("text", x = 0.4, y = 0.6, label = "u = 0.55")
  }
  print(p)
}
```

Every $u \in (0.2, 0.7]$ maps to 1, so $\Pr(X = 1) = 0.7 - 0.2 = 0.5$.

## Gaussian Random Variables

### Box--Muller Method

A transformation of two independent uniforms into two independent Gaussian
random variables (polar coordinates).

. . .

### Inverse Transform

$X = \Phi^{-1}(U)$, where $U \sim \mathcal{U}(0, 1)$ and $\Phi$ is the standard
Gaussian distribution function.

. . .

### Rejection Sampling

See [Exercise 6.1 in
CSwR](https://cswr.nrhstat.org/reject-samp#exr:reject-gaussian) or the [Ziggurat
algorithm](https://en.wikipedia.org/wiki/Ziggurat_algorithm).

## Computing $\Phi^{-1}$

Recall that
$$
  \Phi(x) = \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^x e^{-z^2/2} \ \mathrm{d}z.
$$

. . .

But $\Phi^{-1}$ (the quantile function) has no elementary closed-form
expression.

\medskip\pause

We need to approximate $\Phi^{-1}$ numerically.

\medskip\pause

[R's default normal
sampler](https://github.com/wch/r-source/blob/af7f52f70101960861e5d995d3a4bec010bc89e6/src/nmath/snorm.c#L265)
uses a [numerical approximation of
$\Phi^{-1}$](https://github.com/wch/r-source/blob/af7f52f70101960861e5d995d3a4bec010bc89e6/src/nmath/qnorm.c#L52)
to generate samples from $\mathcal{N}(0,1)$.

## Sampling in R

R has three basic options for customizing the behavior, set through `RNGkind()`.

```{r}
#| label: rng-kind

RNGkind()
```

. . .

### Options

- `kind`: type of pseudo-random number generator
- `normal.kind`: type of sampler for normal random variables
- `sample.kind`: type of sampler for `sample()`

##

\begin{figure}
  \begin{tikzpicture}[node distance=1cm, auto, >=stealth, thick]
    % Nodes
    \node[draw, rectangle, rounded corners, fill=gray!20] (A) {Integers from Pseudo-RNG};
    \node[draw, rectangle, rounded corners, below=of A, fill=green!10] (B) {$\mathcal{U}(0, 1)$};
    \node[draw, rectangle, rounded corners, below right=of B, fill=orange!10] (D1) {Transformation methods};
    \node[draw, rectangle, rounded corners, below left=of B, fill=red!10] (D2) {Rejection sampling};

    % Arrows
    \draw[->] (A) -- (B);
    \draw[->] (B) -- (D1);
    \draw[->] (B) -- (D2);
  \end{tikzpicture}
  \caption{Main categories of sampling methods (in this course).}
\end{figure}

## Example: Sampling from a $t$-Distribution

If we let
$$
  T(z,w) = \frac{z}{\sqrt{w/k}},
$$
and take independent $Z \sim \mathcal{N}(0,1)$ and $W \sim \chi^2_k$ with
$k > 0$, then
$$
  X = T(Z, W) = \frac{Z}{\sqrt{W/k}} \sim t_k.
$$

. . .

This transformation is used in R and other software to simulate $t$-distributed
random variables.

\medskip\pause

\begin{algorithm}[H]
  \caption{Transformation sampling for the $t$-distribution}
  $z \gets $ sample from $\mathcal{N}(0, 1)$\;
  $w \gets $ sample from $\chi^2_k$, independently of $z$\;
  \KwRet{$z / \sqrt{w / k}$}\;
\end{algorithm}

## Von Mises Distribution

The density on $(-\pi, \pi]$ is
$$
  f(x) = \frac{1}{2 \pi I_0(\kappa)} \ e^{\kappa \cos(x-\mu)}
$$
where $\kappa \geq 0$ is the concentration, $\mu \in (-\pi, \pi]$ is the
location, and $I_0$ is the modified Bessel function of the first kind of order
0.

\medskip

```{r}
#| label: vMdensity
#| fig-cap: The von Mises density for $\mu = 2$ and $\kappa = 9$.
#| echo: false
#| fig-width: 5
#| fig-height: 2

library(ggplot2)

dvm <- function(x, k) {
  exp(k * cos(x - 2)) / (2 * pi * besselI(k, 0))
}

xy <- tibble(x = seq(0, 2 * pi, length.out = 400), y = dvm(x, 9))

p1 <- ggplot(xy, aes(x, y)) +
  geom_line() +
  coord_fixed(ratio = 2 * pi / 1.2) +
  labs(y = "Density") +
  theme_minimal()

p2 <- ggplot(xy, aes(x = x, y = y)) +
  geom_line() +
  coord_polar(theta = "x", start = 0, direction = 1) +
  labs(x = "Angle (radians)", y = NULL) +
  theme_minimal()

p1 + p2
```

## Sampling from the von Mises Distribution

### Using movMF

```{r}
#| label: vMsim

library(movMF)
xy <- rmovMF(500, 0.5 * c(cos(-1.5), sin(-1.5)))

# atan2 preserves the quadrant of each point on the unit circle.
x <- atan2(xy[, 2], xy[, 1])
```

. . .

### Problem {.alert}

Can we use inversion sampling? Yes, but it requires numerical inversion of the
distribution function.

##

```{r}
#| label: vMhist
#| fig-height: 2.5
#| fig-width: 5.5
#| echo: false
#| fig-cap: Histogram of samples from `rmovMF()`.

ggplot(tibble(x = x), aes(x)) +
  geom_rug(alpha = 0.1) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 15,
    fill = "grey70",
    color = "white"
  ) +
  stat_function(
    aes(color = "Theoretical Density"),
    fun = function(x) {
      exp(0.5 * cos(x + 1.5)) / (2 * pi * besselI(0.5, 0))
    }
  ) +
  geom_density(
    aes(color = "SJ density estimate"),
    bw = "sj",
    key_glyph = "path"
  ) +
  labs(y = "Density", x = "x", color = NULL)
```

## Rejection Sampling

We want to sample from a target density $f$ using a proposal density $g$ that we
can sample from easily.

\medskip\pause

Let $Y_1, Y_2, \ldots$ be i.i.d. with density $g$ on $\mathbb{R}$ and
$U_1, U_2, \ldots$ be i.i.d. $\mathcal{U}(0, 1)$, with the two sequences
independent.

\medskip\pause

Define
$$
  \sigma
  = \inf\left\{ n \geq 1 \mid U_n \leq \alpha \frac{f(Y_n)}{g(Y_n)} \right\},
$$
for $\alpha \in (0, 1]$, where $f$ is a density.

. . .

### Theorem

If $f(y) \leq g(y)/\alpha$ for all $y \in \mathbb{R}$, then the distribution of
$Y_\sigma$ has density $f$.

\medskip\pause

$\alpha$ is the **acceptance probability** and $g(y)/\alpha$ the **envelope** of
$f$.

##

```{r}
#| label: rejection-sampling-ggplot
#| echo: false
#| fig-height: 2.5
#| fig-width: 5
#| fig-cap: Rejection sampling for a normal target truncated to $(0, 9)$, using
#|   a uniform proposal. Points have height $U g(Y)/\alpha$.

set.seed(1)
N <- 1000
y <- runif(N, 0, 9)
u <- runif(N, 0, 0.43)
target <- function(x) {
  ifelse(
    x > 0 & x < 9,
    dnorm(x, 5, 1) / (pnorm(9, 5, 1) - pnorm(0, 5, 1)),
    0
  )
}
envelope <- function(x) rep(0.43, length(x))

accepted <- u <= target(y)
df <- data.frame(y = y, u = u, accepted = accepted)

curve_df <- data.frame(
  x = seq(0, 9, length.out = 200),
  target = target(seq(0, 9, length.out = 200)),
  envelope = envelope(seq(0, 9, length.out = 200))
)

ggplot() +
  geom_line(data = curve_df, aes(x = x, y = envelope)) +
  geom_area(
    data = curve_df,
    aes(x = x, y = target),
    col = "black",
    fill = "grey90"
  ) +
  geom_point(
    data = df,
    aes(x = y, y = u, color = accepted),
    size = 0.5
  ) +
  scale_color_manual(
    values = c("red", "steelblue2"),
    labels = c("Rejected", "Accepted")
  ) +
  labs(x = "y", y = "Density", color = NULL) +
  theme_minimal() +
  theme(legend.position = "right")
```

## Why Rejection Sampling Works

For one proposal $Y \sim g$, let $A = \{U \leq \alpha f(Y)/g(Y)\}$ be the
acceptance event.

\medskip\pause

Writing $F$ for the target CDF,
$$
  \Pr(Y \leq x, A) = \int_{-\infty}^x g(y)\frac{\alpha f(y)}{g(y)}\,\mathrm{d}y
                   = \alpha F(x).
$$

. . .

Taking $x \to \infty$ gives $\Pr(A) = \alpha$. Therefore,
$$
  \Pr(Y \leq x \mid A) = \frac{\alpha F(x)}{\alpha} = F(x).
$$

. . .

Each accepted proposal has the target distribution. Fresh independent attempts
give the first accepted proposal, $Y_\sigma$, this same distribution.

## The Number of Proposals

Each attempt succeeds with probability $\alpha$, independently:
$$
  \Pr(\sigma = n) = (1 - \alpha)^{n-1}\alpha, \qquad n = 1, 2, \ldots.
$$

. . .

```{r}
#| label: rejection-geometric-cost
#| echo: false
#| fig-width: 5
#| fig-height: 1.7

proposal_counts <- tidyr::expand_grid(
  n = 1:25,
  alpha = c(0.2, 0.8)
) |>
  mutate(
    probability = dgeom(n - 1, alpha),
    label = paste("alpha ==", alpha)
  )

ggplot(proposal_counts, aes(n, probability)) +
  geom_segment(aes(xend = n, yend = 0), color = "royalblue") +
  facet_wrap(~label, labeller = "label_parsed") +
  labs(x = expression(n), y = expression(Pr(sigma == n)))
```

. . .

$\sigma$ counts attempts **including the accepted one**:
$\mathbb{E}[\sigma] = 1/\alpha$.

\medskip

Stopping at the $N$th acceptance takes $N/\alpha$ proposals on average.

## Normalizing Constants

Rejection sampling works with unnormalized densities!

\medskip\pause

Suppose:

- Target: $f(y) = c q(y)$ (unknown $c$)
- Proposal: $g(y) = d p(y)$ (can sample from $g$, unknown $d$)

\medskip\pause

Acceptance test:
$$
  U \leq \frac{\alpha f(y)}{g(y)} = \frac{\alpha c q(y)}{d p(y)}
$$

\medskip\pause

If $\alpha = \frac{\alpha' d}{c}$, then
$$
  \frac{\alpha f(y)}{g(y)} = \frac{\alpha' q(y)}{p(y)}
$$

\pause

### Takeaway

You do not need to know $c$ or $d$!

## How to Find $\alpha'$?

Given functions $q$ and $p$, we need to find $\alpha'$ such that
$$
  \alpha' q(y) \leq p(y) \quad \text{for all } y.
$$

\medskip\pause

### How?

We can try to solve the problem
$$
  \alpha'_{\max} = \inf_{y: q(y)>0} \frac{p(y)}{q(y)}.
$$

We need $0 < \alpha' \leq \alpha'_{\max}$.

\medskip\pause

An analytic bound guarantees validity.

\medskip\pause

Numerical optimization (e.g., `optimize()` in R) can suggest a bound, but the
envelope inequality must hold everywhere.

## When No Envelope Exists

We need $g(y) > 0$ wherever $f(y) > 0$ and $\sup_y f(y)/g(y) < \infty$.

\bigskip

```{r}
#| label: rejection-invalid-proposals
#| echo: false
#| fig-width: 5.5
#| fig-height: 2.2

invalid_proposals <- tibble(x = seq(-4, 4, length.out = 500)) |>
  mutate(
    target = dnorm(x),
    proposal = dunif(x, -2, 2),
    log_ratio = dnorm(x, log = TRUE) -
      dnorm(x, sd = 0.5, log = TRUE)
  )

support_plot <- ggplot(invalid_proposals, aes(x)) +
  geom_line(aes(y = target), color = "royalblue") +
  geom_line(aes(y = proposal), color = "darkorange") +
  labs(
    title = "Missing support",
    subtitle = "Blue: target; orange: proposal",
    x = expression(y),
    y = "Density"
  )

tail_plot <- ggplot(invalid_proposals, aes(x, log_ratio)) +
  geom_line(color = "darkorange") +
  labs(
    title = "Tails too light",
    subtitle = "Normal proposal with SD 0.5",
    x = expression(y),
    y = expression(log(f(y) / g(y)))
  )

support_plot + tail_plot
```

. . .

For a standard normal target, neither a $\mathcal{U}(-2,2)$ proposal nor a
$\mathcal{N}(0,1/4)$ proposal admits a finite envelope.

## Von Mises Rejection Sampling

For $\mu = 0$, use the uniform proposal on $(-\pi, \pi)$: $g(y) = 1/(2\pi)$.

\medskip\pause

Since
$$
  e^{\kappa(\cos(y)-1)} = \alpha' e^{\kappa \cos(y)} \leq 1,
$$
where $\alpha' = \exp(-\kappa)$, we reject if
$$
  U > e^{\kappa(\cos(Y)-1)}.
$$

. . .

\begin{algorithm}[H]
  \caption{Rejection sampling for the von Mises distribution}
  \Repeat{sample is accepted}{
    Generate $Y \sim \mathcal{U}(-\pi, \pi)$\;
    Generate $U \sim \mathcal{U}(0, 1)$\;
    \If{$U \leq e^{\kappa(\cos(Y) - 1)}$}{
      Accept $Y$\;
    }
  }
\end{algorithm}

## Implementation

```{r}
#| label: vMsim-slow

get_one_sample_vmises <- function(kappa) {
  repeat {
    Y <- runif(1, -pi, pi)
    U <- runif(1)
    if (U <= exp(kappa * (cos(Y) - 1))) {
      break
    }
  }
  Y
}
```

. . .

```{r}
#| label: vMsim-slow2

sample_vonmises_slow <- function(N, kappa) {
  replicate(N, get_one_sample_vmises(kappa))
}
```

##

```{r}
#| label: vMsim2
#| fig-width: 5.4
#| fig-height: 3
#| echo: false
#| fig-cap: Histogram of samples from `sample_vonmises_slow()`, with theoretical
#|   density overlaid.

f <- function(y, kappa) {
  exp(kappa * cos(y)) / (2 * pi * besselI(kappa, 0))
}

# Simulate data
x1 <- sample_vonmises_slow(100000, 0.5)
x2 <- sample_vonmises_slow(100000, 2)

df <- tibble(
  x = c(x1, x2),
  kappa = factor(rep(c(0.5, 2), each = 100000))
)

ggplot(df, aes(x)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "grey60",
    color = "white"
  ) +
  stat_function(
    fun = f,
    args = list(kappa = 0.5),
    color = "steelblue3",
    data = data.frame(
      x = seq(-pi, pi, length.out = 200),
      kappa = 0.5
    )
  ) +
  stat_function(
    fun = f,
    args = list(kappa = 2),
    color = "steelblue3",
    data = data.frame(
      x = seq(-pi, pi, length.out = 200),
      kappa = 2
    )
  ) +
  labs(y = "Density", x = "x") +
  facet_wrap(~kappa, ncol = 2, labeller = "label_both")
```

##

```{r}
#| label: profile-vsim
#| echo: false
#| eval: false

source(here::here("R", "vmises-sim-slow.R"))
profvis(sample_vonmises_slow(100000, 5))
```

![A line profile of the naive von Mises sampling
implementation.](../images/vmises-slow-profiling.png){width=82%}

## Findings from Profiling

The bottleneck is the repeated calls to `runif(1)`.

\medskip\pause

Both loop overhead and calling the RNG for one number at a time add cost.

\medskip\pause

It is faster to generate random numbers in batches and store them in vectors.

\medskip\pause

But how can we do that for rejection sampling with an \alert{unknown} number of
rejections?

## Vectorized (but Random Length) Rejection Sampler

We can speed up the rejection sampler by generating $m$ proposal samples at
once.

```{r}
#| label: vM-rejection-vec

fixed_vonmises_sampler <- function(m, kappa) {
  Y <- runif(m, -pi, pi)
  u <- runif(m)
  accept <- u <= exp(kappa * (cos(Y) - 1))
  Y[accept]
}
```

. . .

Simple and fast, but returns a **random** number of samples!

\medskip\pause

We need a function that returns exactly $N$ samples. We can estimate $\alpha$ to
choose batch sizes without computing it in advance.

## How Many Samples Does a Batch Return?

From $m$ independent proposals, the accepted count satisfies
$$
  B_m \sim \mathrm{Binomial}(m,\alpha), \qquad \mathbb{E}[B_m] = m\alpha.
$$

```{r}
#| label: rejection-binomial-batches
#| echo: false
#| fig-width: 5
#| fig-height: 1.9

batch_counts <- tidyr::expand_grid(
  count = 60:170,
  m = c(500, 600)
) |>
  mutate(
    probability = dbinom(count, m, 0.2),
    enough = count >= 100,
    label = sprintf(
      "m = %d: %.0f%% reach 100",
      m,
      100 * pbinom(99, m, 0.2, lower.tail = FALSE)
    )
  )

ggplot(batch_counts, aes(count, probability, color = enough)) +
  geom_segment(aes(xend = count, yend = 0)) +
  geom_vline(xintercept = 100, linetype = "dashed") +
  scale_color_manual(
    values = c("darkorange", "royalblue"),
    guide = "none"
  ) +
  facet_wrap(~label) +
  labs(x = "Accepted samples", y = "Probability")
```

. . .

Here $N = 100$ and $\alpha = 0.2$. Even with $m = N/\alpha = 500$, a batch may
fall short. Generate another batch when needed.

##

\begin{algorithm}[H]
  \caption{Sampling exactly $N$ values via rejection sampling with unknown
    acceptance probability $\alpha$.}
  $\widehat{\alpha} \gets 1$\;
  $\symbf{y} \gets$ empty vector\;
  $n_\text{accepted} \gets 0$\;
  \While{$n_\text{accepted} < N$}{
    $m \gets \lceil (N - n_\text{accepted}) / \widehat{\alpha} \rceil$\;
    Generate $m$ proposals and apply the rejection test\;
    $\symbf{y}_\text{new} \gets$ accepted proposals\;
    $n_\text{accepted} \gets n_\text{accepted} +$ length$(\symbf{y}_\text{new})$\;
    \If{$\symbf{y}$ is empty}{
      $\widehat{\alpha} \gets (n_\text{accepted} + 1) / (m + 1)$ \tcp*[f]{Keep the estimate positive.}
    }
    Append $\symbf{y}_\text{new}$ to $\symbf{y}$\;
  }
  \KwRet{$\symbf{y}_{1:N}$}
\end{algorithm}

## Function Factory for Rejection Sampling

```{r}
#| label: new_rejection_sampler

new_rejection_sampler <- function(generator) {
  function(N, ...) {
    alpha <- 1
    y <- numeric(0)
    n_accepted <- 0
    while (n_accepted < N) {
      m <- ceiling((N - n_accepted) / alpha)
      y_new <- generator(m, ...)
      n_accepted <- n_accepted + length(y_new)
      if (length(y) == 0) {
        alpha <- (n_accepted + 1) / (m + 1)
      }
      y <- c(y, y_new)
    }
    list(x = y[seq_len(N)], alpha = alpha)
  }
}
```

##

```{r}
#| label: vmises-bench
#| fig-width: 5
#| fig-height: 1.5

sample_vonmises <- new_rejection_sampler(fixed_vonmises_sampler)

bench::mark(
  sample_vonmises(1e5, 5),
  sample_vonmises_slow(1e5, 5),
  check = FALSE
) |>
  plot()
```

## Piecewise Exponential Envelopes

::: {.columns}
::::: {.column width="47%"}
When $f$ is positive and *log-concave* on an interval $I$, we can construct an
integrable envelope of the form
$$
  f(x) \leq e^{V(x)}
$$
where
$$
  V(x) = \sum_{i=1}^m (a_i x + b_i) \symbf{1}_{I_i}(x)
$$
for intervals $I_i$ forming a partition of $I$.
:::::

::::: {.column width="47%"}
```{r}
#| label: log-concave-envelope-general
#| echo: false

library(ggplot2)

curve_fun <- function(x) dnorm(x)
log_curve <- function(x) dnorm(x, log = TRUE)

# Tangent points
x1 <- -1.5
x2 <- 1.5

# Compute tangent lines in log-space
slope1 <- -x1
intercept1 <- log_curve(x1) - slope1 * x1
slope2 <- -x2
intercept2 <- log_curve(x2) - slope2 * x2

xseq <- seq(-3, 3, length.out = 400)
envelope_log <- ifelse(
  xseq < 0,
  slope1 * xseq + intercept1,
  slope2 * xseq + intercept2
)
envelope <- exp(envelope_log)

df <- data.frame(
  x = xseq,
  curve = curve_fun(xseq),
  envelope = envelope
)

ggplot(df, aes(x)) +
  geom_line(aes(y = curve), color = "grey60") +
  geom_line(aes(y = envelope)) +
  geom_point(aes(x = x1, y = curve_fun(x1))) +
  geom_point(aes(x = x2, y = curve_fun(x2))) +
  scale_y_continuous(
    transform = scales::log_trans(),
    labels = scales::label_math(e^.x, format = log)
  ) +
  labs(y = "Density") +
  theme_minimal()
```
:::::
:::

\bigskip\pause

Typically, $a_i x + b_i$ is tangent to the graph of $\log(f)$ at
$x_i \in I_i = (z_{i-1}, z_i]$ for
$$
  z_0 < x_1 < z_1 < x_2 < \ldots < z_{m-1} < x_m < z_m.
$$

## Why Tangents Give Upper Bounds

If $h = \log f$ is differentiable and concave, then
$$
  h(x) \leq h(x_i) + h'(x_i)(x - x_i) = a_i x + b_i.
$$

```{r}
#| label: envelope-tangent-bounds
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 2.1

normal_log_tangent <- function(x, at) {
  dnorm(at, log = TRUE) - at * (x - at)
}

tangent_example <- tibble(x = seq(-3, 3, length.out = 401)) |>
  mutate(
    log_density = dnorm(x, log = TRUE),
    left = normal_log_tangent(x, -1.5),
    right = normal_log_tangent(x, 1.5),
    envelope = pmin(left, right)
  )

for (stage in 1:2) {
  p <- ggplot(tangent_example, aes(x)) +
    geom_line(aes(y = log_density), color = "grey50") +
    geom_line(aes(y = left), color = "royalblue") +
    annotate(
      "point",
      x = -1.5,
      y = dnorm(-1.5, log = TRUE),
      color = "royalblue"
    ) +
    coord_cartesian(ylim = c(-5.5, 2)) +
    labs(x = expression(x), y = expression(h(x)))
  if (stage >= 2) {
    p <- p +
      geom_line(aes(y = right), color = "darkorange") +
      annotate(
        "point",
        x = 1.5,
        y = dnorm(1.5, log = TRUE),
        color = "darkorange"
      )
  }
  print(p)
}
```

Choose $a_i = h'(x_i)$ and $b_i = h(x_i) - a_i x_i$. Exponentiating gives
$f(x) \leq e^{a_i x+b_i}$.

## Where Should the Pieces Meet?

Use the smallest tangent bound at each point: $V(x) = \min_i\{a_i x + b_i\}$.

```{r}
#| label: envelope-tangent-intersections
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 1.9

for (stage in 1:2) {
  p <- ggplot(tangent_example, aes(x)) +
    geom_line(aes(y = log_density), color = "grey50") +
    geom_line(
      aes(y = left),
      color = "royalblue",
      linetype = "dashed"
    ) +
    geom_line(
      aes(y = right),
      color = "darkorange",
      linetype = "dashed"
    ) +
    geom_vline(xintercept = 0, color = "grey70") +
    coord_cartesian(ylim = c(-5.5, 2)) +
    labs(x = expression(x), y = expression(h(x)))
  if (stage >= 2) {
    p <- p + geom_line(aes(y = envelope))
  }
  print(p)
}
```

Adjacent tangents meet at
$$
  z_i = \frac{b_{i+1} - b_i}{a_i - a_{i+1}} \qquad (a_i \neq a_{i+1}).
$$
Here $z_1 = 0$. Keep the blue tangent to the left and the orange tangent to the
right.

## Sampling from the Envelope

For each piece $I_i = (L_i,R_i]$, compute its finite mass
$$
  w_i = \int_{L_i}^{R_i} e^{a_i x+b_i}\,\mathrm{d}x, \qquad C = \sum_i w_i.
$$

. . .

Choose a piece $J$ with $\Pr(J = i) = w_i/C$.

\begin{center}
\begin{tikzpicture}[node distance=0.6cm, >=stealth]
  \node[draw, rounded corners, fill=blue!10] (piece) {Choose $J$};
  \node[draw, rounded corners, fill=orange!10, right=of piece] (draw) {Draw $Y$ in $I_J$};
  \node[draw, rounded corners, fill=green!10, right=of draw] (test) {Rejection test};
  \draw[->] (piece) -- (draw);
  \draw[->] (draw) -- (test);
\end{tikzpicture}
\end{center}

. . .

Given $J = i$ and $a_i \neq 0$, draw $U \sim \mathcal{U}(0,1)$ and set
$$
  Y = \frac{1}{a_i}\log\!\left[ (1 - U)e^{a_i L_i} + Ue^{a_i R_i} \right].
$$
Use endpoint limits for infinite bounds. If $a_i = 0$, sample uniformly on the
finite interval.

\medskip\pause

This gives $g(y) = e^{V(y)}/C$. Accept with probability $f(y)/e^{V(y)}$.

## Beta Distribution

For shape parameters $\alpha, \beta > 1$, the beta density is log-concave.

```{r}
#| echo: false
#| fig-width: 5.5
#| fig-height: 2.5

fixed_beta_sampler <- function(m, x1, x2, alpha, beta) {
  lf <- function(x) {
    (alpha - 1) * log(x) + (beta - 1) * log(1 - x)
  }
  lf_deriv <- function(x) (alpha - 1) / x - (beta - 1) / (1 - x)
  a1 <- lf_deriv(x1)
  a2 <- lf_deriv(x2)
  if (a1 == 0 || a2 == 0 || a1 - a2 == 0) {
    stop(
      "\nThe implementation requires a_1 and a_2 different and both different from zero. Choose different values of x_1 and x_2.",
      call. = FALSE
    )
  }
  b1 <- lf(x1) - a1 * x1
  b2 <- lf(x2) - a2 * x2
  z1 <- (b2 - b1) / (a1 - a2)
  Q1 <- exp(b1) * (exp(a1 * z1) - 1) / a1
  c <- Q1 + exp(b2) * (exp(a2 * 1) - exp(a2 * z1)) / a2

  y <- ratio <- numeric(m)
  uy <- c * runif(m)
  u <- runif(m)
  i <- uy < Q1
  y[i] <- z <- log(a1 * exp(-b1) * uy[i] + 1) / a1
  y[!i] <- zz <- log(
    a2 * exp(-b2) * (uy[!i] - Q1) + exp(a2 * z1)
  ) /
    a2
  ratio[i] <- exp(lf(z) - a1 * z - b1)
  ratio[!i] <- exp(lf(zz) - a2 * zz - b2)
  accept <- u <= ratio
  y[accept]
}

sample_beta <- new_rejection_sampler(fixed_beta_sampler)
```

```{r}
#| label: Beta-fig
#| fig-cap: Samples from a piecewise exponential envelope sampler, with
#|   densities (blue) and envelopes (orange) overlaid.
#| fig-width: 5.5
#| fig-height: 2.65
#| echo: false

params <- data.frame(
  x1 = c(0.3, 0.3, 0.2, 0.2),
  x2 = c(0.7, 0.7, 0.5, 0.5),
  alpha = c(4, 1.8, 4, 1.8),
  beta = c(2, 2.4, 2, 2.4),
  label = c(
    "x1=0.3, x2=0.7, alpha=4, beta=2",
    "x1=0.3, x2=0.7, alpha=1.8, beta=2.4",
    "x1=0.2, x2=0.5, alpha=4, beta=2",
    "x1=0.2, x2=0.5, alpha=1.8, beta=2.4"
  )
)

make_beta_df <- function(N, x1, x2, alpha, beta, label) {
  x <- sample_beta(
    N,
    x1 = x1,
    x2 = x2,
    alpha = alpha,
    beta = beta
  )$x
  data.frame(x = x, label = label)
}

sample_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_beta_df(100000, x1, x2, alpha, beta, label)
      )
    }
  )
)

xseq <- seq(0, 1, length.out = 400)

make_density_df <- function(xseq, alpha, beta, label) {
  data.frame(
    x = xseq,
    density = xseq^(alpha - 1) *
      (1 - xseq)^(beta - 1) /
      beta(alpha, beta),
    label = label
  )
}

make_envelope_df <- function(xseq, x1, x2, alpha, beta, label) {
  envelope <- function(x, x1, x2, alpha, beta) {
    lf <- function(x) {
      (alpha - 1) * log(x) + (beta - 1) * log(1 - x)
    }
    lf_deriv <- function(x) {
      (alpha - 1) / x - (beta - 1) / (1 - x)
    }
    a1 <- lf_deriv(x1)
    a2 <- lf_deriv(x2)
    b1 <- lf(x1) - a1 * x1
    b2 <- lf(x2) - a2 * x2
    z1 <- (b2 - b1) / (a1 - a2)
    ifelse(x < z1, exp(a1 * x + b1), exp(a2 * x + b2))
  }
  data.frame(
    x = xseq,
    envelope = envelope(xseq, x1, x2, alpha, beta) /
      beta(alpha, beta),
    label = label
  )
}

dens_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_density_df(xseq, alpha, beta, label)
      )
    }
  )
)
env_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_envelope_df(xseq, x1, x2, alpha, beta, label)
      )
    }
  )
)

ggplot(sample_df, aes(x)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 50,
    fill = "grey80"
  ) +
  geom_line(
    data = dens_df,
    aes(x = x, y = density),
    color = "steelblue3"
  ) +
  geom_line(
    data = env_df,
    aes(x = x, y = envelope),
    color = "dark orange"
  ) +
  coord_cartesian(ylim = c(0, 3)) +
  facet_wrap(~label, ncol = 2)
```

##

```{r}
#| label: beta-envelopes
#| echo: false
#| warning: false
#| fig-width: 5.5
#| fig-height: 3
#| fig-cap: Theoretical densities (blue) and envelopes (orange) for different
#|   parameter values and choices of $x_1$ and $x_2$.

params <- data.frame(
  alpha = c(4, 1.8),
  beta = c(2, 2.4),
  x1 = c(0.3, 0.2),
  x2 = c(0.7, 0.5),
  label = c(
    "alpha=4, beta=2, x1=0.3, x2=0.7",
    "alpha=1.8, beta=2.4, x1=0.2, x2=0.5"
  )
)

xseq <- seq(0.05, 0.95, length.out = 400)

make_df <- function(xseq, alpha, beta, x1, x2, label) {
  envelope_fun <- function(x, x1, x2, alpha, beta) {
    lf <- function(x) {
      (alpha - 1) * log(x) + (beta - 1) * log(1 - x)
    }
    lf_deriv <- function(x) {
      (alpha - 1) / x - (beta - 1) / (1 - x)
    }
    a1 <- lf_deriv(x1)
    a2 <- lf_deriv(x2)
    b1 <- lf(x1) - a1 * x1
    b2 <- lf(x2) - a2 * x2
    z1 <- (b2 - b1) / (a1 - a2)
    ifelse(x < z1, exp(a1 * x + b1), exp(a2 * x + b2))
  }
  data.frame(
    x = xseq,
    density = xseq^(alpha - 1) *
      (1 - xseq)^(beta - 1) /
      beta(alpha, beta),
    envelope = envelope_fun(xseq, x1, x2, alpha, beta) /
      beta(alpha, beta),
    label = label
  )
}

plot_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_df(xseq, alpha, beta, x1, x2, label)
      )
    }
  )
)

ggplot(plot_df, aes(x = x)) +
  geom_line(aes(y = density), color = "steelblue3") +
  geom_line(aes(y = envelope), color = "dark orange") +
  scale_y_log10() +
  facet_wrap(~label, ncol = 2)
```

## Von Mises Envelopes

::: {.columns}
::::: {.column width="47%"}
For $\mu = 0$ and $\kappa > 0$, the von Mises density is not globally
log-concave.

\medskip\pause

Since $(\log f)''(x) = -\kappa\cos(x)$, it is

- log-concave on $(-\pi/2, \pi/2)$ and
- log-convex on $(-\pi,-\pi/2)$ and $(\pi/2, \pi)$.

Use tangents in the center and chords in the tails to bound $\log f$.
:::::

::::: {.column width="47%"}
```{r}
#| echo: false
#| fig-cap: The von Mises log-density with $\mu = 0$ and $\kappa = 0.5$.

dvm <- function(x, k) exp(k * cos(x)) / (2 * pi * besselI(k, 0))

xy <- tibble(
  x = seq(-pi, pi, length.out = 400),
  y = log(dvm(x, 0.5))
)

ggplot(xy, aes(x = x, y = y)) +
  geom_vline(
    xintercept = c(-pi, -pi / 2, pi / 2, pi),
    color = "grey60"
  ) +
  geom_line() +
  labs(y = "log-density") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
:::::
:::

##

```{r}
#| label: vonmises-adaptive
#| echo: false

dvm <- function(x, k) exp(k * cos(x)) / (2 * pi * besselI(k, 0))

von_mises_adapt_random <- function(N, x1, x2, kappa) {
  lf <- function(x) kappa * cos(x)
  lf_deriv <- function(x) -kappa * sin(x)
  a1 <- 2 * kappa / pi
  a2 <- lf_deriv(x1)
  a3 <- lf_deriv(x2)
  a4 <- -a1

  b1 <- kappa
  b2 <- lf(x1) - a2 * x1
  b3 <- lf(x2) - a3 * x2
  b4 <- kappa

  z0 <- -pi
  z1 <- -pi / 2
  z2 <- (b3 - b2) / (a2 - a3)
  z3 <- pi / 2
  z4 <- pi

  Q1 <- exp(b1) * (exp(a1 * z1) - exp(a1 * z0)) / a1
  Q2 <- Q1 + exp(b2) * (exp(a2 * z2) - exp(a2 * z1)) / a2
  Q3 <- Q2 + exp(b3) * (exp(a3 * z3) - exp(a3 * z2)) / a3
  c <- Q3 + exp(b4) * (exp(a4 * z4) - exp(a4 * z3)) / a4

  u0 <- c * runif(N)
  u <- runif(N)

  I1 <- (u0 < Q1)
  I2 <- (u0 >= Q1) & (u0 < Q2)
  I3 <- (u0 >= Q2) & (u0 < Q3)
  I4 <- (u0 >= Q3)

  x <- numeric(N)
  accept <- logical(N)
  x[I1] <- log(a1 * exp(-b1) * u0[I1] + exp(a1 * z0)) / a1
  accept[I1] <- u[I1] <= exp(lf(x[I1]) - a1 * x[I1] - b1)
  x[I2] <- log(a2 * exp(-b2) * (u0[I2] - Q1) + exp(a2 * z1)) / a2
  accept[I2] <- u[I2] <= exp(lf(x[I2]) - a2 * x[I2] - b2)
  x[I3] <- log(a3 * exp(-b3) * (u0[I3] - Q2) + exp(a3 * z2)) / a3
  accept[I3] <- u[I3] <= exp(lf(x[I3]) - a3 * x[I3] - b3)
  x[I4] <- log(a4 * exp(-b4) * (u0[I4] - Q3) + exp(a4 * z3)) / a4
  accept[I4] <- u[I4] <= exp(lf(x[I4]) - a4 * x[I4] - b4)

  x[accept]
}

von_mises_adapt <- new_rejection_sampler(von_mises_adapt_random)

envelope <- function(x, x1, x2, kappa) {
  lf <- function(x) kappa * cos(x)
  lf_deriv <- function(x) -kappa * sin(x)
  a1 <- 2 * kappa / pi
  a2 <- lf_deriv(x1)
  a3 <- lf_deriv(x2)
  a4 <- -a1

  b1 <- kappa
  b2 <- lf(x1) - a2 * x1
  b3 <- lf(x2) - a3 * x2
  b4 <- kappa

  z0 <- -pi
  z1 <- -pi / 2
  z2 <- (b3 - b2) / (a2 - a3)
  z3 <- pi / 2
  z4 <- pi

  env <- numeric(length(x))
  i1 <- x < z1
  i2 <- x >= z1 & x < z2
  i3 <- x >= z2 & x < z3
  i4 <- x >= z3

  env[i1] <- exp(a1 * x[i1] + b1)
  env[i2] <- exp(a2 * x[i2] + b2)
  env[i3] <- exp(a3 * x[i3] + b3)
  env[i4] <- exp(a4 * x[i4] + b4)

  env
}
```

```{r}
#| label: vonmises-adaptive-plot
#| fig-cap: Von Mises samples from a piecewise exponential envelope sampler,
#|   with densities (blue) and envelopes (orange) overlaid.
#| fig-width: 5.5
#| fig-height: 3
#| echo: false

params <- data.frame(
  x1 = c(-0.4, -0.1, -0.4, -1),
  x2 = c(0.4, 0.1, 0.4, 1),
  kappa = c(5, 5, 2, 2),
  label = c(
    "kappa=5, x1=-0.4, x2=0.4",
    "kappa=5, x1=-0.1, x2=0.1",
    "kappa=2, x1=-0.4, x2=0.4",
    "kappa=2, x1=-1, x2=1"
  )
)

make_vm_df <- function(N, x1, x2, kappa, label) {
  x <- von_mises_adapt(N, x1, x2, kappa)$x
  data.frame(x = x, label = label)
}

sample_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(params[i, ], make_vm_df(100000, x1, x2, kappa, label))
    }
  )
)

xseq <- seq(-pi, pi, length.out = 400)

make_density_df <- function(xseq, kappa, label) {
  data.frame(x = xseq, density = dvm(xseq, kappa), label = label)
}

make_envelope_df <- function(xseq, x1, x2, kappa, label) {
  env <- envelope(xseq, x1, x2, kappa) /
    (2 * pi * besselI(kappa, 0))
  data.frame(x = xseq, envelope = env, label = label)
}

dens_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(params[i, ], make_density_df(xseq, kappa, label))
    }
  )
)
env_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_envelope_df(xseq, x1, x2, kappa, label)
      )
    }
  )
)

ggplot(sample_df, aes(x)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "grey80"
  ) +
  geom_line(
    data = dens_df,
    aes(x = x, y = density),
    color = "steelblue3"
  ) +
  geom_line(
    data = env_df,
    aes(x = x, y = envelope),
    color = "dark orange"
  ) +
  coord_cartesian(ylim = c(0, 1.5)) +
  facet_wrap(~label, ncol = 2)
```

## Adapting the Envelope

The examples above use fixed envelopes. Adaptive rejection sampling updates the
envelope using new points.

\medskip\pause

Efficiency depends on both the number of rejections and the cost of generating
and evaluating proposals.

\medskip\pause

A tighter envelope reduces rejections, but updating it adds computational cost.

## One Adaptive Update

For a log-concave target, use a rejected proposal $Y$ as a new tangent point.

```{r}
#| label: envelope-adaptive-update
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 2.1

adaptive_example <- tangent_example |>
  mutate(
    new_tangent = normal_log_tangent(x, 0),
    updated = pmin(envelope, new_tangent)
  )

for (stage in 1:3) {
  p <- ggplot(adaptive_example, aes(x)) +
    geom_line(aes(y = exp(log_density))) +
    geom_line(aes(y = exp(envelope)), color = "royalblue") +
    coord_cartesian(ylim = c(0, 1.3)) +
    labs(x = expression(x), y = "Density")
  if (stage >= 2) {
    p <- p +
      annotate("point", x = 0, y = 0.9, color = "darkorange") +
      annotate(
        "text",
        x = 1.1,
        y = 0.95,
        label = "Rejected at Y = 0"
      )
  }
  if (stage >= 3) {
    p <- p +
      geom_line(aes(y = exp(updated)), color = "darkorange")
  }
  print(p)
}
```

$$
  V_{\mathrm{new}}(x) = \min\{V_{\mathrm{old}}(x),\ h(Y) + h'(Y)(x - Y)\}.
$$
\uncover<3->{The orange envelope is tighter. Update the piece masses before the
next draw.}

## Transformation Methods vs. Rejection Sampling

::: {.columns}
::::: {.column width="47%"}
### Transformation Methods

Transform samples from a simple distribution to the target distribution.

\medskip\pause

Efficient if the transformation is available and cheap to compute.

\medskip\pause

Finding or evaluating a transformation can be difficult for complex or
unnormalized distributions.
:::::

. . .

::::: {.column width="47%"}
### Rejection Sampling

Uses a proposal distribution and accepts/rejects samples.

\medskip\pause

Works even if the quantile function is unavailable or expensive.

\medskip\pause

Efficiency depends on the proposal and acceptance probability.

\medskip\pause

Works with unnormalized densities, but acceptance can be very low in high
dimensions.
:::::
:::

. . .

### Takeaway {.example}

Use a simple transformation when available. Otherwise, consider rejection
sampling with a suitable envelope.

## Check the Distribution Before Timing

For independent $Z_1,\ldots,Z_5 \sim \mathcal{N}(0,1)$,
$\sum_{j=1}^5 Z_j^2 \sim \chi^2_5$. What if we reuse one normal draw five times?

```{r}
#| label: sampler-validation-example
#| echo: false
#| fig-width: 5
#| fig-height: 2
#| fig-cap: Empirical CDFs of independent and dependent $\chi^2_5$ samples, with
#|   the theoretical CDF overlaid in black.

set.seed(510)
validation_n <- 20000
independent_chisq <- rowSums(matrix(
  rnorm(5 * validation_n)^2,
  ncol = 5
))
# Reusing one draw preserves the mean but breaks independence.
dependent_chisq <- 5 * rnorm(validation_n)^2

validation_samples <- tibble(
  x = c(independent_chisq, dependent_chisq),
  method = rep(
    c("Independent normals", "Same normal reused"),
    each = validation_n
  )
)

ggplot(validation_samples, aes(x)) +
  stat_ecdf(aes(color = method)) +
  stat_function(fun = pchisq, args = list(df = 5)) +
  scale_color_manual(
    values = c("royalblue", "darkorange"),
    guide = "none"
  ) +
  coord_cartesian(xlim = c(0, 20)) +
  facet_wrap(~method) +
  labs(x = expression(x), y = "CDF")
```

. . .

Both constructions have mean 5, but their variances are 10 and 50. The black
curve is the target CDF. Check support, distribution, and moments before
benchmarking.

## Exercises

### Exercise 1

For a positive integer $k$, use the fact that
$$
  W = X_1^2 + \ldots + X_k^2 \sim \chi^2_k \quad \text{for} \quad X_1,
  \ldots, X_k \sim_{\text{i.i.d.}} \mathcal{N}(0, 1)
$$
to implement a function, `my_rchisq()`, such that `my_rchisq(N, k)` returns a
sample of $N$ i.i.d. observations from $\chi^2_k$.

```{r}
#| label: rchisq2-implementation
#| include: false

my_rchisq <- function(N, k) {
  y <- numeric(N)
  for (i in seq_len(N)) {
    x <- rnorm(k)
    y[i] <- sum(x^2)
  }
  y
}
```

```{r}
#| label: rchisq2-implementation-faster
#| include: false

my_fast_rchisq <- function(N, k) {
  x <- matrix(rnorm(k * N), k)
  colSums(x^2)
}
```

. . .

### Exercise 2

Make another function, `my_other_rchisq()`, that uses inversion sampling.

```{r}
#| label: rchisq2
#| include: false

my_other_rchisq <- function(N, k) {
  x <- runif(N)
  qchisq(x, k)
}
```

. . .

### Exercise 3

Benchmark your implementations against one another and `rchisq()`.

```{r}
#| label: rchisq-bench
#| include: false

N <- 1e4
k <- 5
bench::mark(
  rchisq(N, k),
  my_rchisq(N, k),
  my_fast_rchisq(N, k),
  my_other_rchisq(N, k),
  check = FALSE
) |>
  plot()
```

## Summary

### Random Number Generation

### Rejection Sampling

## Next Time

### Monte Carlo Methods

### Importance Sampling
