---
title: "A Deeper Look at the EM Algorithm"
execute:
  cache: false
---

{{< include _common.qmd >}}

```{r}
#| label: setup
#| include: false

set.seed(1618)
library(patchwork)
```

```{r}
#| label: moth_likelihood
#| echo: false

M <- function(y, group) as.vector(tapply(y, group, sum))

mult_likelihood <- function(
  x,
  group,
  pi_map,
  constraint = function(par) TRUE
) {
  function(par) {
    pr <- pi_map(par)
    if (!constraint(par) || any(pr > 1) || any(pr < 0)) {
      return(Inf)
    }
    -sum(x * log(M(pr, group)))
  }
}

pi_map <- function(p) {
  p[3] <- 1 - p[1] - p[2]
  c(
    p[1]^2,
    2 * p[1] * p[2],
    2 * p[1] * p[3],
    p[2]^2,
    2 * p[2] * p[3],
    p[3]^2
  )
}

e_step_mult <- function(theta, x, group) {
  p <- pi_map(theta)
  x[group] * p / M(p, group)[group]
}

m_step_mult <- function(y) {
  N <- sum(y)
  pC <- (y[1] + (y[2] + y[3]) / 2) / N
  pI <- (y[4] + (y[2] + y[5]) / 2) / N
  c(pC, pI)
}

constraint <- function(par) {
  par[1] <= 1 &&
    par[1] >= 0 &&
    par[2] <= 1 &&
    par[2] >= 0 &&
    1 - par[1] - par[2] >= 0
}

neg_loglik_obs <- mult_likelihood(
  x = c(85, 196, 341),
  group = c(1, 1, 1, 2, 2, 3),
  pi_map = pi_map,
  constraint = constraint
)

moth_optim <- optim(c(0.3, 0.3), neg_loglik_obs)
```

## Last Time

### Likelihood Optimization

We introduced the peppered moths problem and maximized the observed-data
likelihood using general-purpose and constrained optimization.

\medskip\pause

We talked about barrier methods for constrained optimization.

. . .

### The EM Algorithm

We introduced the EM algorithm for likelihood optimization in models with
missing data or latent variables.

## Recap: The EM Algorithm

The EM algorithm iteratively optimizes
$$
  \theta_{n+1} = \argmax_{\theta \in \Theta} Q(\theta \mid \theta_n),
$$
with
$$
  Q(\theta \mid \theta_n)
  = \E_{Z \mid X, \theta_n}\big(\log p(X, Z \mid \theta)\big).
$$

\medskip\pause

Each iteration is guaranteed not to decrease the observed-data log-likelihood.

## Today

### More on the EM Algorithm

We tackle the peppered moths example with the EM algorithm.

. . .

### Fisher Information

After we have optimized the likelihood, we often want to estimate the variance
of our estimators. Fisher information helps us do this.

. . .

### Convergence and Generalized EM

We connect the EM convergence rate to missing information and introduce the
generalized EM algorithm.

## Recap: The Peppered Moths

::: {.columns}
::::: {.column width="47%"}
### Alleles

C, I, and T with frequencies $p_C$, $p_I$, and $p_T$, where
$$
  p_C + p_I + p_T = 1.
$$

### Genotypes

CC, CI, CT, II, IT, TT

### Phenotypes

Black, Mottled, Light-Colored
:::::

::::: {.column width="47%"}
```{r}
#| label: moth-genotype-phenotype
#| echo: false
#| fig-width: 2.8
#| fig-height: 1.6

alleles <- c("C", "I", "T")
moth_genotypes <- tidyr::expand_grid(
  row = alleles,
  column = alleles
) |>
  dplyr::mutate(
    genotype = paste0(pmin(row, column), pmax(row, column)),
    phenotype = case_when(
      row == "C" | column == "C" ~ "Black",
      row == "I" | column == "I" ~ "Mottled",
      .default = "Light-colored"
    )
  )

ggplot(moth_genotypes, aes(column, row, fill = phenotype)) +
  geom_tile(color = "white") +
  geom_text(
    aes(
      label = genotype,
      color = if_else(phenotype == "Black", "white", "black")
    ),
    size = 3
  ) +
  scale_x_discrete(
    limits = alleles,
    position = "top",
    expand = c(0, 0)
  ) +
  scale_y_discrete(limits = rev(alleles), expand = c(0, 0)) +
  scale_fill_manual(
    name = "Phenotype",
    values = c(
      "Black" = "grey20",
      "Mottled" = "darkorange",
      "Light-colored" = "grey90"
    ),
    breaks = c("Black", "Mottled", "Light-colored")
  ) +
  scale_color_identity() +
  coord_fixed() +
  labs(x = NULL, y = NULL)
```
:::::
:::

## Latent Variables

### Genotype Labels

Let each moth $i = 1,\ldots,N$ have an (unobserved) genotype label
$$
  Z_i \in \{1,2,3,4,5,6\} \quad\text{(CC, CI, CT, II, IT, TT)}.
$$

. . .

Via the Hardy--Weinberg principle, the **genotype** probabilities are
$$
  p = (p_C^2,\ 2p_Cp_I,\ 2p_Cp_T,\ p_I^2,\ 2p_Ip_T,\ p_T^2).
$$

. . .

### Genotype Counts

The genotype counts $Y_k$ are sufficient statistics for the complete data. They
are also latent, with
$$
  Y_k = \sum_{i=1}^N \mathbf{1}(Z_i = k),
  \quad Y = (Y_1,\ldots,Y_6), \quad \sum_k Y_k = N.
$$

## Observed Data

What we actually observe are the **phenotype** counts
$$
  X_j = \sum_{k \in A_j} Y_k,\quad X = M(Y),
$$
where the genotype categories are partitioned into
$$
  A_1 = \{1,2,3\}\ (\text{Black}),\quad A_2 = \{4,5\}\ (\text{Mottled}),
  \quad A_3 = \{6\}\ (\text{Light}).
$$

. . .

Here, $M : \mathbb{R}^6 \to \mathbb{R}^3$ is the cell-collapsing map
$$
  M(y) = (y_1 + y_2 + y_3,\ y_4 + y_5,\ y_6).
$$

. . .

### Multinomial Distribution

If $Y \sim \textrm{Mult}(N, p)$ with $p = (p_1, \ldots, p_6)$, then
$$
  X = M(Y) \sim \textrm{Mult}(N, M(p)).
$$

## Complete-Data Log-Likelihood

The complete-data log-likelihood, with $c(Y)$ independent of $p$, is
$$
  \ell(p \mid Y) = \sum_{k=1}^6 Y_k \log p_k + c(Y).
$$

. . .

### Optimization Variable

For our EM algorithm, we will optimize over the **allele frequencies**
$$
  \theta = (p_C, p_I), \qquad p_T = 1 - p_C - p_I.
$$

\medskip\pause

Write $\pi(\theta)$ for the genotype probabilities given above, with
$p_T = 1 - p_C - p_I$.

## Multinomial Conditional Distributions

Given phenotype counts $X$ and current allele frequencies $\theta'$, the
genotype counts $Y_{A_j} = (Y_k)_{k \in A_j}$ within phenotype $j$ satisfy
$$
  Y_{A_j} \mid X, \theta'
  \sim \textrm{Mult}\left( X_j, \left( \frac{\pi_k(\theta')}{M(\pi(\theta'))_j} \right)_{k \in A_j} \right).
$$

\pause

### Expected Genotype Counts

Hence for $k \in A_j$,
$$
  \E_{Y \mid X, \theta'}(Y_k) = \frac{X_j \pi_k(\theta')}{M(\pi(\theta'))_j}.
$$

\medskip

The expected counts divide each phenotype's observed count among its compatible
genotypes in proportion to their current probabilities---the oberved count
$\times$ the **responsibility**.

## The Q Function

EM $Q$-function:
$$
  \begin{aligned}
    Q(\theta \mid \theta')
     & = \E_{Y \mid X, \theta'} \log p (Y \mid \theta)                                                        \\
    \pause
     & = \sum_{k=1}^6 \E_{Y \mid X, \theta'}(Y_k) \log \pi_k(\theta) + c(\theta')                             \\
    \pause
     & = \sum_{k=1}^6 \frac{X_{j(k)} \pi_k(\theta')}{M(\pi(\theta'))_{j(k)}} \log \pi_k(\theta) + c(\theta').
  \end{aligned}
$$

. . .

Here, $k \in A_{j(k)}$. The expected counts are computed at $\theta'$ and held
fixed while we maximize over $\theta$.

\medskip

The term $c(\theta') = \E_{Y \mid X,\theta'}c(Y)$ does not affect this
maximization.

\pdfpcnote{
  - The E-step computes expected genotype counts using the current allele frequencies.\\
}

## M-Step for the Multinomial Model

We need to maximize $Q(\theta \mid \theta')$ w.r.t. $\theta$.

\medskip\pause

With $y = (n_{CC}, n_{CI}, n_{CT}, n_{II}, n_{IT}, n_{TT})^T$ a complete
observation, it can be shown that the MLE is
$$
  \begin{aligned}
    \hat{p}_C & = \frac{n_{CC} + (n_{CI} + n_{CT}) / 2}{N} \\
    \hat{p}_I & = \frac{n_{II} + (n_{CI} + n_{IT}) / 2}{N}
  \end{aligned}
$$
where $N = \symbf{1}^T y$ is the total number of observations.

\medskip\pause

This is basically just **counting alleles**: the number of C alleles divided by
the total number of alleles.

\medskip

In the M-step, replace $y$ by its conditional expectation from the E-step.

## EM Algorithm Implementation

```{r}
#| label: em-mult
#| code-style:
#|   line-width: 60

em_mult <- function(
  par,
  x,
  group,
  eps = 1e-12,
  maxit = 20,
  cb = NULL
) {
  for (i in seq_len(maxit)) {
    par0 <- par
    y <- e_step_mult(par, x, group)
    par <- m_step_mult(y)
    if (!is.null(cb)) {
      cb() # Record the current state for later inspection.
    }
    if (sum((par - par0)^2) <= eps * (sum(par^2) + eps)) {
      break
    }
  }
  par
}
```

## Peppered Moths EM Algorithm

The observed counts are 85 black, 196 mottled, and 341 light-colored moths.

```{r}
#| label: moth-em-data
#| results: "hold"

moth_x <- c(85, 196, 341)
moth_group <- c(1, 1, 1, 2, 2, 3)
```

. . .

::: {.columns}
::::: {.column width="47%"}
Then we can run the EM algorithm.

```{r}
#| label: fit-moth-em
#| code-style:
#|   line-width: 30

em_mult(
  c(0.3, 0.3),
  moth_x,
  moth_group
)
```
:::::

. . .

::::: {.column width="47%"}
This corresponds to the MLE found via `optim()`.

```{r}
#| label: moth-optim-estimate

moth_optim$par
```
:::::
:::

## Inside the EM Algorithm

The `tracer()` function in the `CSwR` package can be used to trace the
iterations of an algorithm.^[Use `Delta` to control verbosity. `0`: no output.]

. . .

```{r}
#| label: initialize-moth-tracer

library(CSwR)
em_tracer <- tracer("par", Delta = 0)
```

. . .

Each call to `tracer()` returns an S3 object of class `tracer`. The object has a
field `tracer`, which is a function that will record the current value of given
parameters each time it is called.

. . .

```{r}
#| label: run-moth-tracer

em_mult(c(0.3, 0.3), moth_x, moth_group, cb = em_tracer$tracer)
```

##

You can then summarize the collected values using `summary()`.

```{r}
#| label: summarize-moth-tracer

summary(em_tracer)
```

## Tracing the EM Algorithm

```{r}
#| label: em-tracer

em_tracer <- tracer(c("par0", "par"), Delta = 0)
p_hat <- em_mult(
  c(0.3, 0.3),
  moth_x,
  moth_group,
  eps = 0,
  cb = em_tracer$tracer
)
p_hat

em_trace <- summary(em_tracer)
```

##

```{r}
#| label: plot-curves
#| echo: false
#| warning: false
#| fig-width: 4.7
#| fig-height: 2.8
#| fig-show: hide

nll_pep <- Vectorize(function(p1, p2) neg_loglik_obs(c(p1, p2)))

em_trace <- transform(
  em_trace,
  n = seq_len(nrow(em_trace)),
  par_norm_diff = sqrt((par0.1 - par.1)^2 + (par0.2 - par.2)^2),
  nll = nll_pep(par.1, par.2)
)

grid <- expand.grid(
  pC = seq(0.02, 0.34, length.out = 100),
  pI = seq(0.15, 0.34, length.out = 80)
)
grid$loglik_gap <- vapply(
  seq_len(nrow(grid)),
  function(j) {
    neg_loglik_obs(unlist(grid[j, c("pC", "pI")])) -
      moth_optim$value
  },
  numeric(1)
)
mle_point <- data.frame(
  pC = moth_optim$par[1],
  pI = moth_optim$par[2]
)
em_path <- data.frame(pC = rep(NA_real_, 5), pI = rep(NA_real_, 5))
em_path[1, ] <- c(0.3, 0.3)

for (i in seq_len(4)) {
  current <- em_path[i, , drop = FALSE]
  y <- e_step_mult(unlist(current), moth_x, moth_group)
  next_par <- m_step_mult(y)
  next_point <- data.frame(pC = next_par[1], pI = next_par[2])
  em_path[i + 1, ] <- next_par

  neg_q <- function(par) -sum(y * log(pi_map(par)))
  grid$gap <- vapply(
    seq_len(nrow(grid)),
    function(j) {
      neg_q(unlist(grid[j, c("pC", "pI")])) - neg_q(next_par)
    },
    numeric(1)
  )

  plot <- ggplot(grid, aes(pC, pI)) +
    geom_contour(
      aes(z = gap),
      breaks = c(1, 5, 10, 20, 50, 100, 200),
      color = "grey65"
    ) +
    geom_contour(
      aes(z = loglik_gap),
      breaks = c(5, 50, 150),
      color = "darkorange",
      linetype = "dashed",
      alpha = 0.7
    ) +
    geom_point(
      data = mle_point,
      color = "darkorange",
      size = 3,
      shape = 4
    ) +
    geom_point(data = current, color = "black", size = 2.5) +
    coord_equal(xlim = c(0.02, 0.34), ylim = c(0.15, 0.34)) +
    labs(x = expression(p[C]), y = expression(p[I]))

  if (i > 1) {
    plot <- plot +
      geom_path(data = em_path[seq_len(i), ], color = "royalblue") +
      geom_point(
        data = em_path[seq_len(i - 1), ],
        color = "royalblue"
      )
  }

  print(plot + labs(title = paste("Iteration", i, ": E-step")))

  print(
    plot +
      geom_segment(
        data = data.frame(current, next_point),
        aes(x = pC, y = pI, xend = pC.1, yend = pI.1),
        arrow = grid::arrow(length = grid::unit(0.08, "inches")),
        color = "royalblue"
      ) +
      geom_point(
        data = next_point,
        color = "royalblue",
        size = 2.5
      ) +
      labs(title = paste("Iteration", i, ": M-step"))
  )
}
```

```{=latex}
\begin{figure}
\centering
\multiinclude[<+>][format=pdf,start=1,end=8]{lecture10_files/figure-beamer/plot-curves}
\caption{Gray contours show $-Q$; dashed orange contours show $-\ell$. The cross marks the MLE.}
\end{figure}
```

##

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: tracer-plot1
#| echo: false
#| fig-cap: "Parameter convergence of the EM algorithm applied to the peppered
#|   moths problem."

ggplot(em_trace, aes(n, par_norm_diff)) +
  geom_point() +
  geom_line() +
  labs(
    y = expression(group("||", theta[n] - theta[n - 1], "||"))
  ) +
  scale_y_log10()
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: loglikelihood-convergence
#| warning: false
#| echo: false
#| fig-cap: "Negative log-likelihood gap relative to the best recorded iterate."

mutate(em_trace, diff_nll = 1e-14 + nll - min(nll)) |>
  ggplot(aes(n, diff_nll)) +
  labs(
    y = expression(
      max[k] * italic(l)(theta[k]) - italic(l)(theta[n])
    )
  ) +
  geom_point() +
  geom_line() +
  scale_y_log10()
```
:::::
:::

## Benchmarking

How does EM compare with the two methods we used for the peppered moths in the
previous lecture?

```{r}
#| label: benchmark-moth-setup
#| include: false

moth_start <- c(1 / 3, 1 / 3)
moth_A <- rbind(c(1, 0), c(-1, 0), c(0, 1), c(0, -1), c(-1, -1))
moth_b <- c(0, -1, 0, -1, -1)

moth_grad <- function(par) {
  pC <- par[1]
  pI <- par[2]
  pT <- 1 - pC - pI
  qC <- pC * (2 - pC)
  qI <- pI * (2 - 2 * pC - pI)

  c(
    -2 * moth_x[1] * (1 - pC) / qC +
      2 * moth_x[2] * pI / qI +
      2 * moth_x[3] / pT,
    -2 * moth_x[2] * pT / qI + 2 * moth_x[3] / pT
  )
}
```

\medskip\pause

```{r}
#| label: benchmark-moth-methods

moth_bm <- bench::mark(
  em = em_mult(moth_start, moth_x, moth_group, maxit = 100),
  nelder_mead = optim(moth_start, neg_loglik_obs),
  bfgs_barrier = constrOptim(
    moth_start,
    neg_loglik_obs,
    moth_grad,
    ui = moth_A,
    ci = moth_b,
    method = "BFGS"
  ),
  check = FALSE
)
```

##

```{r}
#| label: plot-moth-methods-benchmark
#| fig-cap: "Benchmarking EM, Nelder--Mead, and the BFGS barrier method for the
#|   peppered moths."
#| fig-width: 4.5

plot(moth_bm)
```

. . .

But there is something fundamentally wrong about this comparison. What?

## Convergence of the Three Methods

```{r}
#| label: trace-moth-methods-benchmark
#| include: false

start_nll <- neg_loglik_obs(moth_start)
moth_minimum <- -sum(moth_x * log(moth_x / sum(moth_x)))

em_benchmark_tracer <- tracer(
  "nll",
  Delta = 0,
  expr = quote(nll <- neg_loglik_obs(par))
)
em_mult(
  moth_start,
  moth_x,
  moth_group,
  maxit = 100,
  cb = em_benchmark_tracer$tracer
)

optim_benchmark_tracer <- tracer("nll", Delta = 0)
traced_optim_nll <- function(par) {
  nll <- neg_loglik_obs(par)
  optim_benchmark_tracer$tracer()
  nll
}
optim(moth_start, traced_optim_nll, method = "Nelder-Mead")

barrier_benchmark_tracer <- tracer("nll", Delta = 0)
traced_barrier_nll <- function(par) {
  nll <- neg_loglik_obs(par)
  barrier_benchmark_tracer$tracer()
  nll
}
constrOptim(
  moth_start,
  traced_barrier_nll,
  moth_grad,
  ui = moth_A,
  ci = moth_b,
  method = "BFGS"
)

trace_path <- function(tracer, method) {
  trace <- tracer$get(simplify = TRUE)
  data.frame(
    elapsed_ms = c(0, cumsum(trace[, ".time"])) * 1000,
    nll = cummin(c(start_nll, trace[, "nll"])),
    method = method
  )
}

convergence <- rbind(
  trace_path(em_benchmark_tracer, "EM"),
  trace_path(optim_benchmark_tracer, "Nelder-Mead"),
  trace_path(barrier_benchmark_tracer, "BFGS barrier")
)
```

```{r}
#| label: plot-moth-methods-convergence
#| echo: false
#| fig-width: 4
#| fig-height: 2.8
#| fig-cap: "Best negative log-likelihood gap against elapsed time. The EM trace
#|   records iterations; the other traces record objective evaluations."

convergence$gap <- pmax(convergence$nll - moth_minimum, 1e-10)

ggplot(convergence, aes(elapsed_ms, gap, color = method)) +
  geom_line() +
  scale_color_manual(
    values = c(
      EM = "darkorange",
      "Nelder-Mead" = "royalblue",
      "BFGS barrier" = "forestgreen"
    )
  ) +
  scale_y_log10() +
  labs(
    x = "Elapsed time (ms)",
    y = "Best negative log-likelihood gap",
    color = NULL
  )
```

# Fisher Information

## The Fisher Information

When doing statistical inference, we need to measure the uncertainty of
estimates.

\medskip\pause

The Fisher information is a key quantity for this. It is defined as
$$
  \mathcal{I}(\theta) = -\E_\theta\left( D_\theta^2 \ell(\theta) \right).
$$

\medskip\pause

This is the expected negative Hessian of the log-likelihood.

### Complete, Observed, and Missing Information

In the presence of latent variables, we can define three types of Fisher
information:

- **Complete-data information** $\mathcal{I}_Y(\theta)$
- **Observed-data information** $\mathcal{I}_X(\theta)$
- **Missing information**
  $\mathcal{I}_{Y \mid X}(\theta) = \mathcal{I}_Y(\theta) - \mathcal{I}_X(\theta)$

The unobserved genotypes reduce the information available about the allele
frequencies. We can see the effect on uncertainty in the moth example.

## Observed Fisher Information

The observed information is the negative Hessian of the observed-data
log-likelihood:
$$
  \hat{i}_X(\theta) = -D_\theta^2 \ell(\theta).
$$

At the MLE, it estimates the expected Fisher information
$\mathcal{I}_X(\theta)$.

\medskip\pause

For the moth data, we already have the observed likelihood, so we can
differentiate it directly.

## Directly from the Observed Likelihood

Since `neg_loglik_obs` is the negative observed-data log-likelihood, its Hessian
at the MLE is the observed information.

```{r}
#| label: observed-information-direct

theta_hat <- em_mult(
  c(0.3, 0.3),
  moth_x,
  moth_group,
  eps = 1e-20
)
iX_direct <- numDeriv::hessian(neg_loglik_obs, theta_hat)
iX_direct
```

\medskip\pause

The inverse of this matrix estimates the covariance matrix of $\hat{\theta}$.

##

```{r}
#| label: fisher-ellipses
#| warning: false
#| message: false
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "Approximate 95% confidence ellipses at the observed-data MLE:
#|   observed phenotypes (solid blue) versus hypothetical complete genotypes
#|   (dashed orange). Their near overlap shows that little information is lost
#|   by observing only phenotype. The cross marks the MLE."

alpha <- 0.05

ellipse_points <- function(I, center, level = 0.95, n = 200) {
  Sigma <- solve(I)
  eig <- eigen(Sigma, symmetric = TRUE)
  A <- eig$vectors %*% diag(sqrt(pmax(eig$values, 0)))
  r <- sqrt(qchisq(level, df = 2))
  t <- seq(0, 2 * pi, length.out = n)
  pts <- t(sapply(
    t,
    function(phi) center + r * A %*% c(cos(phi), sin(phi))
  ))
  colnames(pts) <- c("pC", "pI")
  as.data.frame(pts)
}

clip_simplex <- function(df) {
  subset(df, pC >= 0 & pI >= 0 & pC + pI <= 1)
}

df_obs <- clip_simplex(ellipse_points(
  iX_direct,
  theta_hat,
  1 - alpha
))

# Complete genotypes reveal two independent allele draws per moth.
pC <- theta_hat[1]
pI <- theta_hat[2]
pT <- 1 - pC - pI
i_complete <- 2 *
  sum(moth_x) *
  matrix(
    c(1 / pC + 1 / pT, 1 / pT, 1 / pT, 1 / pI + 1 / pT),
    nrow = 2
  )
df_complete <- clip_simplex(ellipse_points(
  i_complete,
  theta_hat,
  1 - alpha
))

theta_df <- data.frame(pC = theta_hat[1], pI = theta_hat[2])

ggplot() +
  geom_path(data = df_obs, aes(pC, pI), color = "royalblue") +
  geom_path(
    data = df_complete,
    aes(pC, pI),
    color = "darkorange",
    linetype = 2
  ) +
  geom_point(
    data = theta_df,
    aes(pC, pI),
    color = "darkorange",
    pch = 4,
    size = 3
  ) +
  coord_equal() +
  lims(x = c(0, 0.15), y = c(0.1, 0.25)) +
  labs(x = expression(p[C]), y = expression(p[I]))
```

## Convergence of the EM Algorithm

We have already shown that the EM algorithm monotonically increases the observed
data log-likelihood:
$$
  \ell(\theta_{n+1}) \ge \ell(\theta_n).
$$

. . .

Parameter convergence requires additional assumptions. If the iterates converge
and the EM mapping $\Phi$ is continuous at their limit, then
$$
  \theta_{n+1} = \Phi(\theta_n), \quad \hat{\theta} = \Phi(\hat{\theta}).
$$

. . .

By the Taylor expansion of $\Phi$ around $\hat{\theta}$, we have
$$
  \theta_{n+1}
  - \hat{\theta} \approx D\Phi(\hat{\theta}) (\theta_n - \hat{\theta}).
$$

. . .

So the error at iteration $n + 1$ is approximately a linear transformation of
the error at iteration $n$.

\medskip\pause

And the rate of convergence is determined by the matrix $D\Phi(\hat{\theta})$.

## Convergence: Connection to Fisher Information

Let $\rho(A)$ be the spectral radius of a matrix $A$, i.e., the largest absolute
value of its eigenvalues.

\medskip\pause

At a regular optimum, a standard local result gives the convergence rate in
terms of complete and missing information. We state it without proof:
$$
  \rho(D\Phi(\hat{\theta}))
  = \rho\left( \hat{i}_{Y \mid X} \left( \hat{i}_Y \right)^{-1} \right).
$$

Here $\hat{i}_Y$ and $\hat{i}_{Y \mid X}$ denote complete and missing
information at the fitted parameter value.

\medskip\pause

If $\rho(D\Phi(\hat{\theta}))$ is close to 0, the EM algorithm converges
quickly, and if it is close to 1, the EM algorithm converges slowly.

. . .

### Relation to Missing Information

If the missing information is small **relative to the complete-data
information**, then $\rho(D\Phi(\hat{\theta}))$ is small and EM converges
quickly.

## Convergence Estimation

Near a regular optimum, EM typically converges linearly. If the dominant error
component is present, then asymptotically
$$
  \|\theta_{n+1} - \hat{\theta}\| \approx \rho \|\theta_n - \hat{\theta}\|,
  \qquad
  0 < \rho < 1.
$$

\medskip\pause

The rate can be estimated from successive parameter differences, which typically
decay at the same asymptotic rate:
$$
  \log \|\theta_n - \theta_{n-1}\| \approx n \log \rho + \text{constant}.
$$

. . .

```{r}
#| label: linear-convergence

log_lm <- lm(log(par_norm_diff) ~ n, data = em_trace)
exp(coefficients(log_lm)["n"])
```

\medskip\pause

It is small in this case, implying fast convergence. This agrees with the near
overlap of the observed- and complete-information ellipses.

## A Mixture of Two Gaussians

::: {.columns align="center"}
::::: {.column width="47%"}
### Goal

First fit a two-component Gaussian mixture model by ordinary EM.

. . .

### Model

We assume that the data $X_i \sim \operatorname{\mathcal{N}}(\mu_k, \sigma_k^2)$
with probability $\pi_k$, $k = 1, 2$.
:::::

::::: {.column width="47%"}
```{r}
#| label: gaussian-mixture-density
#| echo: false
#| fig-cap: "A mixture of two Gaussian distributions."

gausdens <- function(x, mu1, mu2, sigma1, sigma2, p) {
  (p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
    (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2) /
    sqrt(2 * pi)
}

xx <- seq(-3, 11, 0.1)

sigma1 <- 1
sigma2 <- 2
mu1 <- -0.5
mu2 <- 4
p <- 0.5

il_df <- data.frame(
  x = xx,
  y1 = dnorm(xx, mu1, sigma1) * p,
  y2 = dnorm(xx, mu2, sigma2) * (1 - p),
  ymix = gausdens(xx, mu1, mu2, sigma1, sigma2, p)
)

pl <- ggplot(il_df, aes(x, ymix)) +
  geom_line() +
  geom_line(
    aes(y = y1),
    linetype = "dashed",
    color = "steelblue"
  ) +
  geom_line(
    aes(y = y2),
    linetype = "dashed",
    color = "darkorange"
  ) +
  labs(y = "Density", x = "x")

pl
```
:::::
:::

## Finite Mixtures

In a finite mixture, the marginal density of $X$ is
$$
  p(x \mid \psi) = \sum_{k=1}^K \pi_k f_k(x \mid \psi_k)
$$
where $\pi_k \geq 0$, $\sum_k \pi_k = 1$, and $f_k$ are densities.

\medskip\pause

### Latent Variable

Assume each $X_i$ comes from one of the $K$ components, but we don't know which
one.

\medskip\pause

We introduce a latent variable $Z_i \in \{1, \ldots, K\}$ such that
$$
  \Pr(Z_i = k) = \pi_k.
$$

. . .

### Complete-Data Model

Then the complete-data model is
$$
  p(x_i, z_i \mid \psi) = \pi_{z_i} f_{z_i}(x_i \mid \psi_{z_i}).
$$

## Gaussian Mixture

For Gaussian mixtures, the components are Gaussian:
$$
  f_k(x \mid \mu_k, \sigma_k^2)
  = \frac{1}{\sqrt{2 \pi \sigma_k^2}} e^{-\frac{(x-\mu_k)^2}{2 \sigma_k^2}}.
$$
with $\psi_k = (\mu_k, \sigma_k^2)$.

\medskip\pause

### Complete-Data Log-Likelihood

Letting $\theta = (\pi_1, \ldots, \pi_K, \psi_1, \ldots, \psi_K)$, the
complete-data log-likelihood is
$$
  \begin{aligned}
    \ell_c (\theta) & = \sum_{i=1}^N \log p(x_i, z_i \mid \theta)                                                                     \\
    \pause
                    & = \sum_{i=1}^N \log \left( f(x_i \mid z_i, \theta) f(z_i \mid \theta) \right)                                   \\
    \pause
                    & = \sum_{i=1}^N \sum_{k=1}^K \symbf{1}(Z_i = k) \left( \log \pi_k + \log f_k(x_i \mid \mu_k, \sigma_k^2) \right)
  \end{aligned}
$$

## Simple Case: Fixed Component Variances, $K = 2$

::: {.columns align="center"}
::::: {.column width="47%"}
Let's simplify and assume $K = 2$ and that $\sigma_1$ and $\sigma_2$ are known.

\medskip\pause

Then the model has three unknown parameters:
$$
  \theta = (\pi, \mu_1, \mu_2).
$$
:::::

::::: {.column width="47%"}
\begin{tikzpicture}[
    node distance=1.6cm,
    latent/.style={circle,draw,minimum size=10mm,inner sep=0pt},
    observed/.style={circle,draw,fill=gray!20,minimum size=10mm,inner sep=0pt},
    param/.style={rectangle,draw,minimum size=7mm,inner sep=2pt},
  ]

  % Parameters for components
  \node[param] (pi) {$\pi$};
  \node[param, right=2.5cm of pi, yshift=1cm] (mu1) {$\mu_1$};
  \node[param, right=2.5cm of pi, yshift=-1cm] (mu2) {$\mu_2$};

  % Latent and observed variables
  \node[latent, below=1.8cm of pi] (z) {$Z_i$};
  \node[observed, below=1.8cm of z] (x) {$X_i$};

  % Arrows
  \draw[->] (pi) -- (z);
  \draw[->] (z) -- (x);
  \draw[->] (mu1) -- (x);
  \draw[->] (mu2) -- (x);

  % Plate for data points
  \node[plate={below:{$i = 1,\dots,N$}}, fit=(z)(x)] (plateN) {};
\end{tikzpicture}
:::::
:::

## The E-Step

The complete-data log-likelihood is now
$$
  \begin{aligned}
    \ell_c(\theta) & = \sum_{i=1}^N \symbf{1}(Z_i = 1) \left( \log(\pi) - \frac{(x_i - \mu_1)^2}{2 \sigma_1^2} \right)                       \\
                   & \phantom{ = {}} + \symbf{1}(Z_i = 2)\left( \log(1 - \pi) - \frac{(x_i - \mu_2)^2}{2 \sigma_2^2} \right) + \text{const}.
  \end{aligned}
$$

. . .

Taking expectations gives
$$
  Q(\theta \mid \theta')
  = \sum_{i=1}^N \tau_{i} \left( \log(\pi) - \frac{(x_i - \mu_1)^2}{2 \sigma_1^2} \right)
    + (1 - \tau_{i})\left( \log(1 - \pi) - \frac{(x_i - \mu_2)^2}{2 \sigma_2^2} \right)
$$
where $\tau_i = \Pr(Z_i = 1 \mid X_i = x_i, \theta')$ is the **responsibility**
of component 1 for observation $i$.

\medskip\pause

The responsibility tells us how responsible component 1 is for generating $x_i$.

## The Responsibilities

The conditional probability in a mixture model is generally
$$
  \Pr(Z = z \mid X = x)
  = \frac{\pi_z f_z(x \mid \psi_z)}{\sum_{k=1}^K \pi_k f_k(x \mid \psi_k)}.
$$

\medskip\pause

For the $K = 2$ Gaussian case, this gives us
$$
  \tau_i
  = \Pr(Z_i = 1 \mid X = x_i, \theta')
  = \frac{\pi'_1 e^{-\frac{(x_i-\mu_1')^2}{2 \sigma_1^2}}}{\left( \pi'_1 e^{-\frac{(x_i-\mu_1')^2}{2 \sigma_1^2}} + \frac{\sigma_1}{\sigma_2} (1 - \pi'_1) e^{-\frac{(x_i-\mu_2')^2}{2 \sigma_2^2}} \right)}.
$$

## The Q-Step

The MLEs of $\theta$ given $Q(\theta \mid \theta')$ are
$$
  \hat{\theta}
  = \left( \frac{1}{N} \sum_{i=1}^N \tau_i, \quad \frac{1}{\sum_{i=1}^N \tau_i} \sum_{i=1}^N \tau_i x_i, \quad \frac{1}{\sum_{i=1}^N (1 - \tau_i)} \sum_{i=1}^N (1 - \tau_i) x_i \right).
$$

. . .

::: {.columns}
::::: {.column width="47%"}
### Updating $\pi$

We update $\pi$ as
$$
  \hat{\pi} = \frac{1}{N} \sum_{i=1}^N \tau_i,
$$
which is the fraction of all data points "assigned" to component 1.
:::::

. . .

::::: {.column width="47%"}
### Updating $\mu_1$ and $\mu_2$

We update $\mu_1$ as
$$
  \hat{\mu}_1 = \frac{\sum_{i=1}^N \tau_i x_i}{\sum_{i=1}^N \tau_i},
$$
and similarly for $\mu_2$.

\medskip\pause

This is a weighted mean, weighted by how much each component is responsible for
each observation.
:::::
:::

## Example: Simulation

We start by setting the parameters. Then we simulate data.

::: {.columns}
::::: {.column width="42%"}
```{r}
#| label: mixture-gaus-sim

sigma1 <- 1
sigma2 <- 2
mu1 <- -0.5
mu2 <- 4
p <- 0.5
```
:::::

. . .

::::: {.column width="51%"}
```{r}
#| label: simulate-gaussian-mixture

N <- 5000
z <- sample(
  c(TRUE, FALSE),
  N,
  replace = TRUE,
  prob = c(p, 1 - p)
)
x <- numeric(N)
N1 <- sum(z)
x[z] <- rnorm(N1, mu1, sigma1)
x[!z] <- rnorm(N - N1, mu2, sigma2)
```
:::::
:::

##

```{r}
#| label: gaussian-mixture-histogram
#| echo: false
#| fig-cap: "Histogram of simulated data from a two-component Gaussian mixture model."
#| fig-width: 4
#| fig-height: 2.7

pl <- ggplot(data.frame(x), aes(x)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 40,
    fill = "lightgray"
  ) +
  geom_line(
    aes(x, gausdens(x, mu1, mu2, sigma1, sigma2, p)),
    color = "black",
    linewidth = 1
  ) +
  labs(y = "Density", x = "x")

pl
```

```{r}
#| label: gaus-mix-loglik
#| echo: false

neg_loglik <- function(par, x) {
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]

  if (p < 0 || p > 1) {
    return(Inf)
  }

  -sum(log(
    p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
      (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2
  ))
}
```

## General-Purpose Optimization

```{r}
#| label: gaus-mix-example

optim_par <- optim(c(0.5, -0.5, 4), neg_loglik, x = x)
optim_par[1:2]
```

. . .

But the starting value matters!

```{r}
#| label: gaus-mix-example-bad

optim(c(0.9, 3, 1), neg_loglik, x = x)[1:2]
```

## The EM Algorithm

```{r}
#| label: em-alg
#| code-style:
#|   line-width: 72

em <- function(par, em_step, eps = 1e-12, maxit = 20, cb = NULL) {
  for (i in seq_len(maxit)) {
    par0 <- par
    par <- em_step(par) # not shown here

    if (!is.null(cb)) {
      cb() # Callback function (tracer)
    }

    if (sum((par - par0)^2) <= eps * (sum(par^2) + eps)) {
      break
    }
  }
  par
}
```

```{r}
#| label: source-em-mix
#| echo: false

library(numDeriv)

e_step_gauss <- function(par, x, sigma1 = 1, sigma2 = sigma1) {
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  a <- p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1
  b <- (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2
  a / (a + b)
}

m_step_gauss <- function(p_hat, x) {
  N <- length(x)
  N1 <- sum(p_hat)
  N2 <- N - N1
  c(N1 / N, sum(p_hat * x) / N1, sum((1 - p_hat) * x) / N2)
}

create_em_gauss_step <- function(
  x,
  sigma1 = 1,
  sigma2 = sigma1
) {
  force(x)
  force(sigma1)
  force(sigma2)

  function(par) {
    m_step_gauss(e_step_gauss(par, x, sigma1, sigma2), x)
  }
}

Q <- function(par, par_prime) {
  p_hat <- e_step_gauss(par_prime, x, sigma1, sigma2)
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  sum(
    p_hat * (log(p) - (x - mu1)^2 / (2 * sigma1^2)) +
      (1 - p_hat) * (log(1 - p) - (x - mu2)^2 / (2 * sigma2^2))
  )
}

grad1 <- function(par) {
  grad(function(par) neg_loglik(par, x), par)
}
grad2 <- function(par) -grad(Q, par, par_prime = par)

em_gauss_step <- create_em_gauss_step(x, sigma1, sigma2)
```

## Testing

```{r}
#| label: test-em-gauss-mix

em_par <- em(c(0.5, -0.5, 4), em_gauss_step)
```

. . .

```{r}
#| label: gaus-mix-example2

testthat::test_that("EM and optim give similar results", {
  testthat::expect_equal(em_par, optim_par$par, tolerance = 1e-3)
})
```

. . .

```{r}
#| label: run-em-gauss-mix-again

em(c(0.9, 3, 1), em_gauss_step) # Starting value still matters
```

## Tracing Convergence

The `tracer` object from the **CSwR** package lets us evaluate arbitrary
expressions at each iteration of the EM algorithm via `expr`.

```{r}
#| label: tracing-convergence

library(CSwR)
em_tracer <- tracer(
  c("par0", "par", "nll"),
  Delta = 0, # No printing
  expr = quote({
    nll <- neg_loglik(par, x)
  })
)
```

. . .

```{r}
#| label: fit-gaussian-mixture-trace

par_hat <- em(c(0.2, 2, 2), em_gauss_step, cb = em_tracer$tracer)

em_trace <- summary(em_tracer)
```

##

::: {.columns}
::::: {.column width="47%"}
**CSwR** contains an `autoplot` S3 method for `trace` objects.

. . .

```{r}
#| label: gaussmix-convergence-plot
#| eval: false
#| code-style:
#|   line-width: 35

autoplot(
  em_trace,
  y = nll - min(nll)
)
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: convergence-plots-2
#| warning: false
#| echo: false
#| ref-label: gaussmix-convergence-plot
#| fig-cap: "Convergence of the EM algorithm for Gaussian mixture model."
```
:::::
:::

##

```{r}
#| label: mixture-convergence
#| echo: false

gaussdens2 <- function(x, p, mu1, mu2) {
  (p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
    (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2) /
    sqrt(2 * pi)
}

em_gauss_illustration <- function(i, em_trace) {
  p_i <- em_trace$par0.1[i]
  mu1_i <- em_trace$par0.2[i]
  mu2_i <- em_trace$par0.3[i]

  pl +
    geom_line(
      aes(y = gaussdens2(x, p_i, mu1_i, mu2_i)),
      color = "darkorange",
      size = 1
    ) +
    lims(y = c(0, 0.25)) +
    labs(title = paste0("k = ", i))
}
```

```{r}
#| label: em-gauss-iterations
#| fig-width: 4.2
#| fig-height: 2.8
#| echo: false
#| fig-show: animate
#| fig-cap: "Iterations of the EM algorithm"

for (i in c(1, 2, 3, 5, 10, 15, 18, 19, 20)) {
  print(em_gauss_illustration(i, em_trace))
}
```

## Label Switching

If we set $\sigma_1 = \sigma_2$, then the likelihood is invariant under
switching the labels of the components, i.e., $(p, \mu_1, \mu_2)$ and
$(1 - p, \mu_2, \mu_1)$ give the same likelihood.

\medskip\pause

So the EM algorithm may converge to either of the two symmetric modes.

```{r}
#| label: mixture-gaus-sim2
#| echo: false

sigma1 <- 1
sigma2 <- 1
mu1 <- -0.5
mu2 <- 4
p <- 0.8
N <- 5000
z <- sample(
  c(TRUE, FALSE),
  N,
  replace = TRUE,
  prob = c(p, 1 - p)
)
x <- numeric(N)
N1 <- sum(z)
x[z] <- rnorm(N1, mu1, sigma1)
x[!z] <- rnorm(N - N1, mu2, sigma2)

em_gauss_step_alt <- create_em_gauss_step(x, sigma1)
```

. . .

```{r}
#| label: fit-label-switched-mixture-first

em(c(0.2, 3, -0.5), em_gauss_step_alt)
```

. . .

```{r}
#| label: fit-label-switched-mixture-second

em(c(0.8, -3, -0.5), em_gauss_step_alt)
```

. . .

In this example, we generated data with $p = 0.8$ and $\sigma_1 = \sigma_2$.

## A Mixture with Covariate-Dependent Weights

We now observe a covariate $w_i$ alongside each $x_i$. Let the probability of
component 1 depend on it:
$$
  \pi_i = \Pr(Z_i = 1 \mid w_i) = \frac{1}{1 + e^{-(a+b w_i)}}.
$$

\medskip\pause

The component distributions are still Gaussian. To focus on the mixing weights,
we treat $\mu_1 = -0.5$, $\mu_2 = 4$, and $\sigma_1 = \sigma_2 = 1$ as known.

\medskip\pause

With a constant mixing weight, we had a closed-form M-step. Here, estimating
$(a,b)$ requires a logistic regression fit.

## Example: A Mixture with Varying Weights

::: {.columns}
::::: {.column width="47%"}
We simulate a covariate and use it to choose the component for each observation.

```{r}
#| label: simulate-varying-mixture
#| code-style:
#|   line-width: 35

w_gem <- runif(500, -2, 2)
z_gem <- rbinom(
  500,
  1,
  plogis(-0.5 + 1.5 * w_gem)
)
x_gem <- rnorm(
  500,
  ifelse(z_gem == 1, -0.5, 4),
  1
)
```
:::::

::::: {.column width="47%"}
```{r}
#| label: plot-varying-mixture
#| echo: false
#| fig-height: 2.6
#| fig-cap: "The fraction of observations near each Gaussian component changes
#|   with the covariate."

ggplot(data.frame(w = w_gem, x = x_gem), aes(w, x)) +
  geom_point(alpha = 0.45, color = "royalblue") +
  labs(x = expression(w), y = expression(x))
```
:::::
:::

## E-Step with Varying Mixture Weights

Let $\phi_k(x)$ denote the density of the known Gaussian component $k$. At the
current parameters $(a_n,b_n)$, compute
$$
  \pi_i^{(n)} = \frac{1}{1 + e^{-(a_n+b_n w_i)}}
$$
and the responsibility
$$
  \tau_i
  = \Pr(Z_i = 1 \mid x_i,w_i,a_n,b_n)
  = \frac{\pi_i^{(n)} \phi_1(x_i)} {\pi_i^{(n)} \phi_1(x_i) + (1 - \pi_i^{(n)})\phi_2(x_i)}.
$$

\medskip\pause

As before, the E-step assigns each observation a fractional component label.

## The M-Step with Varying Mixture Weights

Hold the responsibilities $\tau_i$ fixed. With $\eta_i = a + b w_i$, the part of
$Q$ that depends on $(a,b)$ is
$$
  Q(a,b \mid a_n,b_n)
  = \sum_{i=1}^N \left[ \tau_i \eta_i - \log(1 + e^{\eta_i}) \right]
    + \text{const}.
$$

\medskip\pause

There is no closed-form maximizer. We can fit the weighted logistic regression
to convergence while holding $\tau_i$ fixed, then recompute the
responsibilities.

```{r}
#| label: varying-mixture-responsibilities
#| echo: false

gate_responsibility <- function(coef) {
  p <- plogis(coef[1] + coef[2] * w_gem)
  f1 <- dnorm(x_gem, -0.5, 1)
  f2 <- dnorm(x_gem, 4, 1)
  p * f1 / (p * f1 + (1 - p) * f2)
}

gate_q <- function(coef, tau) {
  eta <- coef[1] + coef[2] * w_gem
  sum(tau * eta - pmax(eta, 0) - log1p(exp(-abs(eta))))
}
```

## Fit the M-Step to Convergence

For each E-step, maximize $Q$ with the responsibilities held fixed. The outer EM
algorithm then recomputes them and repeats.

```{r}
#| label: varying-mixture-full-step

gate_m_step <- function(coef, reltol = 1e-12) {
  tau <- gate_responsibility(coef)
  optim(
    coef,
    function(candidate) -gate_q(candidate, tau),
    method = "BFGS",
    control = list(maxit = 100, reltol = reltol)
  )$par
}
```

. . .

We are not showing `gate_responsibility()` and `gate_q()` here.

```{r}
#| label: varying-mixture-fit-helper
#| include: false

gate_loglik <- function(coef) {
  p <- plogis(coef[1] + coef[2] * w_gem)
  sum(log(p * dnorm(x_gem, -0.5, 1) + (1 - p) * dnorm(x_gem, 4, 1)))
}

fit_gate <- function(
  step,
  start = c(0, 0),
  tol = 1e-8,
  maxit = 1000
) {
  coef <- start
  start_time <- Sys.time()
  loglik <- gate_loglik(coef)
  path <- loglik
  elapsed_ms <- 0
  for (iteration in seq_len(maxit)) {
    coef <- step(coef)
    next_loglik <- gate_loglik(coef)
    path <- c(path, next_loglik)
    elapsed_ms <- c(
      elapsed_ms,
      as.numeric(difftime(Sys.time(), start_time, units = "secs")) *
        1000
    )
    if (next_loglik < loglik - 1e-8) {
      stop("Observed-data log-likelihood decreased")
    }
    if (next_loglik - loglik < tol) {
      break
    }
    loglik <- next_loglik
  }
  list(
    coef = coef,
    loglik = next_loglik,
    iterations = iteration,
    path = path,
    elapsed_ms = elapsed_ms
  )
}
```

## Run Ordinary EM

The same outer loop repeatedly applies the E-step and the fully optimized M-step
until the observed-data log-likelihood gain falls below $10^{-8}$.

```{r}
#| label: varying-mixture-em-fit

full_fit <- fit_gate(gate_m_step)
unlist(full_fit[c("coef", "loglik", "iterations")])
```

## What If We Stop the M-Step Early?

The weighted logistic regression is solved afresh at **every** EM iteration.
What if we limit the inner `optim()` call before recomputing the
responsibilities?

\medskip\pause

The **generalized EM algorithm** (GEM) only requires the M-step to increase $Q$:
$$
  Q(\theta_{n+1} \mid \theta_n) \ge Q(\theta_n \mid \theta_n).
$$

\medskip\pause

The E-step is unchanged. If the truncated M-step increases $Q$, the
observed-data log-likelihood remains nondecreasing. Cheaper iterations may
require more outer iterations.

## Generalized EM Iterations

```{r}
#| label: gem-lower-bound
#| echo: false
#| fig-show: animate
#| fig-width: 5.5
#| fig-height: 3
#| fig-cap: Four generalized EM iterations.

# Use the same Gaussian latent-variable model as the EM illustration in
# lecture 9, so the difference lies entirely in the M-step.
gem_loglik <- function(theta) -(theta - 2)^2 / 4
gem_bound <- function(theta, current) {
  gem_loglik(theta) - (theta - current)^2 / 4
}

gem_theta <- numeric(5)
gem_theta[1] <- -1
for (i in seq_len(4)) {
  bound_maximizer <- (gem_theta[i] + 2) / 2
  gem_theta[i + 1] <- (gem_theta[i] + bound_maximizer) / 2
}

gem_curve <- data.frame(theta = seq(-1.4, 4.5, length.out = 401))
gem_curve$loglik <- gem_loglik(gem_curve$theta)

for (i in seq_len(4)) {
  current <- gem_theta[i]
  next_theta <- gem_theta[i + 1]
  bound_maximizer <- (current + 2) / 2
  gem_curve$bound <- gem_bound(gem_curve$theta, current)
  history <- data.frame(
    theta = gem_theta[seq_len(i)],
    value = gem_loglik(gem_theta[seq_len(i)])
  )

  p <- ggplot(gem_curve, aes(x = theta)) +
    geom_line(aes(y = loglik, color = "likelihood")) +
    geom_line(aes(y = bound, color = "bound")) +
    geom_point(data = history, aes(y = value), color = "grey60") +
    geom_vline(
      xintercept = current,
      color = "grey60",
      linetype = "dashed"
    ) +
    scale_color_manual(
      name = NULL,
      values = c(likelihood = "royalblue", bound = "darkorange"),
      breaks = c("likelihood", "bound"),
      labels = expression(
        italic(l) * (theta),
        italic("F") * (q[n] * "," ~ theta)
      )
    ) +
    coord_cartesian(xlim = c(-1.4, 4.5), ylim = c(-3, 0.25)) +
    labs(x = expression(theta), y = "Value")

  print(
    p +
      annotate(
        "point",
        x = current,
        y = gem_loglik(current),
        color = "royalblue",
        size = 2.5
      ) +
      labs(title = sprintf("Iteration %d: E-step", i))
  )

  print(
    p +
      geom_vline(
        xintercept = next_theta,
        color = "grey60",
        linetype = "dashed"
      ) +
      annotate(
        "point",
        x = bound_maximizer,
        y = gem_bound(bound_maximizer, current),
        color = "grey60",
        shape = 4,
        size = 2.5
      ) +
      annotate(
        "segment",
        x = next_theta,
        xend = next_theta,
        y = gem_bound(next_theta, current),
        yend = gem_loglik(next_theta),
        color = "royalblue"
      ) +
      annotate(
        "point",
        x = next_theta,
        y = gem_bound(next_theta, current),
        color = "darkorange",
        fill = "white",
        shape = 21,
        size = 2.5
      ) +
      annotate(
        "point",
        x = next_theta,
        y = gem_loglik(next_theta),
        color = "royalblue",
        size = 2.5
      ) +
      labs(title = sprintf("Iteration %d: GEM M-step", i))
  )
}
```

## Tune the Inner Tolerance

The M-step passes `reltol` to `optim()`. We can change its relative convergence
tolerance while keeping the E-step and outer stopping rule fixed.

\medskip\pause

Larger values stop the inner BFGS fit sooner. Smaller values make each M-step
more accurate, but may spend time solving a problem whose responsibilities will
change at the next E-step.

. . .

### Three M-Step Tolerances

```{r}
#| label: varying-mixture-gem-step

loose_gate_step <- function(coef) gate_m_step(coef, reltol = 1e-2)
moderate_gate_step <- function(coef) {
  gate_m_step(coef, reltol = 1e-9)
}
tight_gate_step <- function(coef) gate_m_step(coef, reltol = 1e-14)
```

Each outer iteration still recomputes the responsibilities before calling
`optim()` again.

## Compare the Fits

All fits start at $(a_0,b_0) = (0,0)$ and use the same stopping rule for the
observed-data log-likelihood gain.

```{r}
#| label: varying-mixture-compare-fits
#| echo: false

loose_fit <- fit_gate(loose_gate_step)
moderate_fit <- fit_gate(moderate_gate_step)
tight_fit <- fit_gate(tight_gate_step)
gate_fits <- list(
  "1e-2" = loose_fit,
  "1e-9" = moderate_fit,
  "1e-14" = tight_fit
)
data.frame(
  reltol = names(gate_fits),
  iterations = vapply(gate_fits, `[[`, integer(1), "iterations"),
  loglik = vapply(gate_fits, `[[`, numeric(1), "loglik"),
  row.names = NULL
)
```

## Convergence over Time

```{r}
#| label: plot-varying-mixture-convergence
#| echo: false
#| fig-width: 4.5
#| fig-height: 2.5
#| fig-cap: "Observed-data log-likelihood gap against median elapsed time over
#|   30 complete fits."

best_loglik <- max(vapply(gate_fits, `[[`, numeric(1), "loglik"))
gate_steps <- list(
  loose_gate_step,
  moderate_gate_step,
  tight_gate_step
)
gate_trace <- do.call(
  rbind,
  Map(
    function(fit, step, reltol) {
      elapsed_runs <- replicate(30, fit_gate(step)$elapsed_ms)
      data.frame(
        elapsed_ms = apply(elapsed_runs, 1, median),
        gap = best_loglik - fit$path,
        reltol = reltol
      )
    },
    gate_fits,
    gate_steps,
    names(gate_fits)
  )
)
gate_trace$reltol <- factor(
  gate_trace$reltol,
  levels = names(gate_fits)
)

ggplot(
  gate_trace,
  aes(elapsed_ms, pmax(gap, 1e-10), color = reltol)
) +
  geom_line() +
  scale_color_manual(
    values = c(
      "1e-2" = "darkorange",
      "1e-9" = "forestgreen",
      "1e-14" = "royalblue"
    )
  ) +
  scale_y_log10() +
  labs(
    x = "Elapsed time (ms)",
    y = "Log-likelihood gap",
    color = "reltol"
  )
```

## Exercise

Implement the EM algorithm for a missing data problem:

```{r}
#| label: bernoulli-missing-data

x <- c(1, 0, 1, NA, 0, NA, 1, 1, 0)
```

Assume that $X_i \sim \text{Bernoulli}(\theta)$ i.i.d. and that missingness is
ignorable (missing at random, with separate parameters for the missingness
mechanism). The complete-data log-likelihood is
$$
  \ell_C(\theta)
  = \sum_{i=1}^N \left( X_i \log(\theta) + (1 - X_i) \log(1 - \theta) \right).
$$

. . .

### Step 1 (Derive)

- E-step: Compute
  $Q(\theta \mid \theta_n) = \E_{\theta_n}\left( \log p(X_\text{obs}, X_\text{miss} \mid \theta) \mid X_\text{obs} \right)$.
- M-step: Maximize $Q(\theta \mid \theta_n)$ over $\theta$.

. . .

### Step 2 (Implement)

Implement the EM algorithm and run it from two different starting points. Record
the path of $\theta_n$ and the log-likelihood values $\ell(\theta_n)$.

```{r}
#| label: em-bernoulli-missing-data
#| include: false

x <- c(1, 0, 1, NA, 0, NA, 1, 1, 0)

# EM for Bernoulli with missing entries (treated as latent)
em_bern_missing <- function(x, p0 = 0.5, maxit = 6) {
  obs <- !is.na(x)
  ll <- function(p) {
    sum(x[obs] * log(p) + (1 - x[obs]) * log(1 - p))
  }
  p <- p0
  ps <- numeric(maxit + 1)
  lls <- numeric(maxit + 1)

  for (t in 0:maxit) {
    ps[t + 1] <- p
    lls[t + 1] <- ll(p)
    if (t < maxit) {
      # E-step: expected value of each missing Xi is p
      k <- sum(!obs)
      p <- (sum(x[obs]) + k * p) / length(x) # M-step: mean of completed data
    }
  }
  list(
    p_final = p,
    path = ps,
    loglik = lls,
    loglik_diff = diff(lls)
  )
}

# Run from two starts
res1 <- em_bern_missing(x, p0 = 0.5)
res2 <- em_bern_missing(x, p0 = 0.1)

res1$path
res1$loglik_diff # ≥ 0 (monotone)
res2$path

# Closed-form fixed point: solves p = (sum(obs 1s) + 2 p)/9 -> p = 4/7
# Plug-in expected Fisher information for each sample size.
p_hat <- res1$p_final
n_obs <- sum(!is.na(x))
i_obs <- n_obs / (p_hat * (1 - p_hat)) # observed-data info
i_complete <- length(x) / (p_hat * (1 - p_hat)) # complete-data info
c(p_hat = p_hat, i_obs = i_obs, i_complete = i_complete)
```

## Summary

We applied the EM algorithm to the peppered moths example.

\medskip\pause

We estimated the covariance of the allele frequencies from the observed Fisher
information.

\medskip\pause

We connected the convergence rate to missing information and introduced the
generalized EM algorithm.

## Next Time

### Stochastic Gradient Descent (SGD)

A version of gradient descent that uses a single observation (or a small batch)
to compute the gradient at each iteration. This is useful for large datasets.
