---
title: "Inversion and Rejection Sampling"
---

{{< include _common.qmd >}}

```{r}
#| label: setup
#| include: false

library(profvis)

knitr::read_chunk(here::here("R", "vMSim.R"))

source(here::here("R", "vMSim.R"), keep.source = TRUE)
```

## Today

### Inversion Sampling

Sample from a target distribution using its inverse CDF.

. . .

### Rejection Sampling

Sample using a density that may be known only up to normalization.

## Why Do We Need a Sampler?

Simulation lets us explore the consequences of a statistical model when analytic
calculations are difficult.

\medskip\pause

How biased is an estimator in a small sample? Does a nominal 95% confidence
interval actually achieve 95% coverage?

\medskip\pause

\begin{center}
\begin{tikzpicture}[node distance=0.4cm, >=stealth, font=\small]
  \node[draw, rounded corners, fill=blue!10] (data) {Simulate data};
  \node[draw, rounded corners, fill=orange!10, right=of data] (fit) {Compute estimates};
  \node[draw, rounded corners, fill=green!10, right=of fit] (repeat) {Assess bias, coverage, etc.};
  \draw[->] (data) -- (fit);
  \draw[->] (fit) -- (repeat);
\end{tikzpicture}
\end{center}

\medskip\pause

Draws also let us estimate probabilities and expectations, including tail
probabilities and posterior summaries.

\medskip\pause

All of these applications need **reliable and efficient sampling from a
specified distribution.**

## Pseudo-Random Numbers

The basis for all sampling methods is a source of random numbers and R uses
deterministic algorithms to generate **pseudo-random** numbers.

\medskip\pause

The sequences are reproducible from a seed but designed to appear random.

\medskip\pause

A research field in itself. See `?RNG` in R for available algorithms.

. . .

### Mersenne Twister

The default in R, which generates integers in the range
$$
  \{0, 1, \ldots, 2^{32} - 1\}.
$$

\medskip\pause

Long period: $2^{19937} - 1$. Blocks of up to 623 consecutive integers occur
nearly equally often over a full period.

. . .

### Uniform Random Variables in R

`runif()` scales these pseudo-random integers to the unit interval, with
adjustments to exclude 0 and 1.

\pdfpcnote{
  Random numbers are used in many applications.
  Use \texttt{RNGkind()} to inspect the selected pseudo-random number generator,
  normal sampler, and discrete sampling method used by \texttt{sample()}.
}

## From Uniform Random Numbers to Target Distributions

\begin{figure}
  \begin{tikzpicture}[node distance=1cm, auto, >=stealth, thick]
    \node[draw, rectangle, rounded corners, fill=gray!20] (A) {Integers from Pseudo-RNG};
    \node[draw, rectangle, rounded corners, below=of A, fill=green!10] (B) {$\mathcal{U}(0, 1)$};
    \node[draw, rectangle, rounded corners, below right=of B, fill=orange!10] (D1) {Transformation methods};
    \node[draw, rectangle, rounded corners, below left=of B, fill=red!10] (D2) {Rejection sampling};

    \draw[->] (A) -- (B);
    \draw[->] (B) -- (D1);
    \draw[->] (B) -- (D2);
  \end{tikzpicture}
  \caption{Main categories of sampling methods (in this course).}
\end{figure}

## Transformation Methods

What if we want to sample from a different distribution?

\medskip\pause

We could use **the transformation method**, which relies on the following fact:

\medskip

If $T : \mathcal{Z} \to \mathbb{R}$ is a map and $Z \in \mathcal{Z}$ is a random
variable we can sample, then we can sample $X = T(Z)$.

\begin{figure}
\centering
\begin{tikzpicture}[node distance=1cm, auto, >=stealth, thick]
  \node[draw, rectangle, rounded corners, fill=blue!10] (A) {Sample $Z$};
  \node[draw, rectangle, rounded corners, right = of A, fill=green!10] (B) {Apply $T$};
  \node[draw, rectangle, rounded corners, right = of B, fill=orange!10] (C) {Get $X = T(Z)$};
  \draw[->] (A) -- (B);
  \draw[->] (B) -- (C);
\end{tikzpicture}
\end{figure}

\pause

Transformation methods let us generate samples from complex distributions by
applying a function to samples from a simpler distribution.

\medskip\pause

### Example {.example}

To sample $X \sim \mathrm{Exp}(\lambda)$ with rate $\lambda > 0$, use
$X = -\log(U)/\lambda$ with $U \sim \mathcal{U}(0, 1)$.

## Inversion Sampling

A type of **transformation** sampling.

\medskip\pause

If $F^{-1} : (0,1) \to \mathbb{R}$ is the generalized inverse of a distribution
function $F$ and $U$ is uniformly distributed on $(0, 1)$, then
$$
  F^{-1}(U)
$$
has distribution function $F$.

\medskip\pause

Computing $F^{-1}$ can be simple (e.g., for the exponential distribution) or
require a search or numerical approximation.

## Following the Inverse CDF

For $X \sim \mathrm{Exp}(1)$, $F(x) = 1 - e^{-x}$ for $x \geq 0$.

```{r}
#| label: inversion-cdf-arrows
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 2.2

inverse_example <- tibble(u = c(0.2, 0.6, 0.9)) |>
  mutate(x = -log1p(-u))

inverse_curve <- tibble(x = seq(0, 3, length.out = 300)) |>
  mutate(cdf = pexp(x))

for (stage in 1:3) {
  p <- ggplot(inverse_curve, aes(x, cdf)) +
    geom_line() +
    geom_point(
      data = inverse_example,
      aes(x = 0, y = u),
      color = "royalblue"
    ) +
    scale_y_continuous(breaks = c(0, 0.2, 0.6, 0.9, 1)) +
    coord_cartesian(xlim = c(0, 3), ylim = c(0, 1)) +
    labs(x = expression(x), y = expression(F(x)))
  if (stage >= 2) {
    p <- p +
      geom_segment(
        data = inverse_example,
        aes(x = 0, xend = x, y = u, yend = u),
        arrow = grid::arrow(length = grid::unit(0.07, "inches")),
        color = "royalblue"
      )
  }
  if (stage >= 3) {
    p <- p +
      geom_segment(
        data = inverse_example,
        aes(x = x, xend = x, y = u, yend = 0),
        arrow = grid::arrow(length = grid::unit(0.07, "inches")),
        color = "darkorange"
      )
  }
  print(p)
}
```

Move right from $U$ to the CDF, then down to $X = F^{-1}(U) = -\log(1 - U)$.

## Inversion for a Discrete Distribution

The **generalized inverse** is $F^{-1}(u) = \inf\{x : F(x) \geq u\}$.

```{r}
#| label: inversion-discrete-cdf
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 2.2

discrete_cdf <- tibble(
  left = c(-0.4, 0, 1, 2),
  right = c(0, 1, 2, 2.4),
  cdf = c(0, 0.2, 0.7, 1)
)

for (stage in 1:2) {
  p <- ggplot(discrete_cdf) +
    geom_segment(aes(
      x = left,
      xend = right,
      y = cdf,
      yend = cdf
    )) +
    geom_point(
      data = discrete_cdf[-1, ],
      aes(x = left, y = cdf)
    ) +
    geom_point(
      data = discrete_cdf[-4, ],
      aes(x = right, y = cdf),
      shape = 21,
      fill = "white"
    ) +
    scale_x_continuous(breaks = 0:2) +
    scale_y_continuous(breaks = c(0, 0.2, 0.7, 1)) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = expression(x), y = expression(F(x)))
  if (stage >= 2) {
    p <- p +
      annotate(
        "segment",
        x = -0.2,
        xend = -0.2,
        y = 0.2,
        yend = 0.7,
        color = "darkorange"
      ) +
      annotate(
        "segment",
        x = -0.2,
        xend = 1,
        y = 0.55,
        yend = 0.55,
        arrow = grid::arrow(length = grid::unit(0.07, "inches")),
        color = "darkorange"
      ) +
      annotate(
        "segment",
        x = 1,
        xend = 1,
        y = 0.55,
        yend = 0,
        arrow = grid::arrow(length = grid::unit(0.07, "inches")),
        color = "darkorange"
      ) +
      annotate("text", x = 0.4, y = 0.63, label = "u = 0.55")
  }
  print(p)
}
```

Every $u \in (0.2, 0.7]$ maps to 1, so $\Pr(X = 1) = 0.7 - 0.2 = 0.5$.

## Inversion Without a Closed Form

The standard Gaussian distribution function is
$$
  \Phi(x) = \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^x e^{-z^2/2} \ \mathrm{d}z.
$$

. . .

Inversion gives $X = \Phi^{-1}(U)$ with $U \sim \mathcal{U}(0, 1)$, but
$\Phi^{-1}$ has no elementary closed-form expression.

\medskip\pause

We need to approximate $\Phi^{-1}$ numerically.

\medskip\pause

[R's default normal
sampler](https://github.com/wch/r-source/blob/af7f52f70101960861e5d995d3a4bec010bc89e6/src/nmath/snorm.c#L265)
uses a [numerical approximation of
$\Phi^{-1}$](https://github.com/wch/r-source/blob/af7f52f70101960861e5d995d3a4bec010bc89e6/src/nmath/qnorm.c#L52)
to generate samples from $\mathcal{N}(0,1)$.

## Numerical Inversion Is a Root Search

For $u = 0.9$, solve $\Phi(x) - u = 0$.

```{r}
#| label: inversion-numerical-bracket
#| echo: false
#| fig-width: 5
#| fig-height: 1.25

ggplot(tibble(x = seq(0, 2, length.out = 200)), aes(x)) +
  geom_line(aes(y = pnorm(x) - 0.9), color = "royalblue") +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey50"
  ) +
  geom_point(
    data = tibble(x = c(0, 2)),
    aes(y = pnorm(x) - 0.9),
    color = "darkorange"
  ) +
  labs(x = expression(x), y = expression(Phi(x) - u))
```

```{r}
#| label: inversion-numerical-root
#| results: hide

u <- 0.9
uniroot(function(x) pnorm(x) - u, interval = c(0, 2), tol = 1e-8)
```

\pause

The bracket must contain the root for the chosen $u$; `tol` controls accuracy in
$x$. General root finding repeatedly evaluates the CDF; `qnorm()` uses a
specialized approximation.

## Example: Sampling from a $t$-Distribution

If we let
$$
  T(z,w) = \frac{z}{\sqrt{w/k}},
$$
and take independent $Z \sim \mathcal{N}(0,1)$ and $W \sim \chi^2_k$ with
$k > 0$, then
$$
  X = T(Z, W) = \frac{Z}{\sqrt{W/k}} \sim t_k.
$$

. . .

This transformation is used in R and other software to simulate $t$-distributed
random variables.

\medskip\pause

\begin{algorithm}[H]
  \caption{Transformation sampling for the $t$-distribution}
  $z \gets $ sample from $\mathcal{N}(0, 1)$\;
  $w \gets $ sample from $\chi^2_k$, independently of $z$\;
  \KwRet{$z / \sqrt{w / k}$}\;
\end{algorithm}

## Von Mises Distribution

The density on $(-\pi, \pi]$ is
$$
  f(x) = \frac{1}{2 \pi I_0(\kappa)} \ e^{\kappa \cos(x-\mu)}
$$
where $\kappa \geq 0$ is the concentration, $\mu \in (-\pi, \pi]$ is the
location, and $I_0$ is the modified Bessel function of the first kind of order
0.

\medskip

```{r}
#| label: vMdensity
#| fig-cap: The von Mises density for $\mu = 2$ and $\kappa = 9$.
#| echo: false
#| fig-width: 5
#| fig-height: 2

library(ggplot2)

dvm <- function(x, k) {
  exp(k * cos(x - 2)) / (2 * pi * besselI(k, 0))
}

xy <- tibble(x = seq(0, 2 * pi, length.out = 400), y = dvm(x, 9))

p1 <- ggplot(xy, aes(x, y)) +
  geom_line() +
  coord_fixed(ratio = 2 * pi / 1.2) +
  labs(y = "Density") +
  theme_minimal()

p2 <- ggplot(xy, aes(x = x, y = y)) +
  geom_line() +
  coord_polar(theta = "x", start = 0, direction = 1) +
  labs(x = "Angle (radians)", y = NULL) +
  theme_minimal()

p1 + p2
```

## Sampling from the von Mises Distribution

We can evaluate the unnormalized density
$$
  q(x) = e^{\kappa \cos(x-\mu)}.
$$

. . .

For $\kappa > 0$, the CDF has no elementary closed-form inverse. Inversion
sampling is possible, but requires numerical approximation.

. . .

### Sampling Problem {.alert}

Can we sample using $q$ directly, without computing or inverting the CDF?

\medskip\pause

Yes! We can use **rejection sampling**.

## Rejection Sampling

We want to sample from a target density $f$ using a proposal density $g$ that we
can sample from easily.

\medskip\pause

Let $Y_1, Y_2, \ldots$ be i.i.d. with density $g$ on $\mathbb{R}$ and
$U_1, U_2, \ldots$ be i.i.d. $\mathcal{U}(0, 1)$, with the two sequences
independent.

\medskip\pause

Define
$$
  \sigma
  = \inf\left\{ n \geq 1 \mid U_n \leq \alpha \frac{f(Y_n)}{g(Y_n)} \right\},
$$
for $\alpha \in (0, 1]$, where $f$ is a density.

. . .

### Theorem

If $f(y) \leq g(y)/\alpha$ for all $y \in \mathbb{R}$, then the distribution of
$Y_\sigma$ has density $f$.

\medskip\pause

$\alpha$ is the **acceptance probability** and $g(y)/\alpha$ the **envelope** of
$f$.

##

```{r}
#| label: rejection-sampling-ggplot
#| echo: false
#| fig-height: 2.5
#| fig-width: 5
#| fig-cap: Rejection sampling for a normal target truncated to $(0, 9)$, using
#|   a uniform proposal. Points have height $U g(Y)/\alpha$.

set.seed(1)
N <- 1000
y <- runif(N, 0, 9)
u <- runif(N, 0, 0.43)
target <- function(x) {
  ifelse(
    x > 0 & x < 9,
    dnorm(x, 5, 1) / (pnorm(9, 5, 1) - pnorm(0, 5, 1)),
    0
  )
}
envelope <- function(x) rep(0.43, length(x))

accepted <- u <= target(y)
df <- data.frame(y = y, u = u, accepted = accepted)

curve_df <- data.frame(
  x = seq(0, 9, length.out = 200),
  target = target(seq(0, 9, length.out = 200)),
  envelope = envelope(seq(0, 9, length.out = 200))
)

ggplot() +
  geom_line(data = curve_df, aes(x = x, y = envelope)) +
  geom_area(
    data = curve_df,
    aes(x = x, y = target),
    col = "black",
    fill = "grey90"
  ) +
  geom_point(
    data = df,
    aes(x = y, y = u, color = accepted),
    size = 0.5
  ) +
  scale_color_manual(
    values = c("red", "steelblue2"),
    labels = c("Rejected", "Accepted")
  ) +
  labs(x = "y", y = "Density", color = NULL) +
  theme_minimal() +
  theme(legend.position = "right")
```

## Why Rejection Sampling Works

For one proposal $Y \sim g$, let $A = \{U \leq \alpha f(Y)/g(Y)\}$ be the
acceptance event.

\medskip\pause

Writing $F$ for the target CDF,
$$
  \Pr(Y \leq x, A) = \int_{-\infty}^x g(y)\frac{\alpha f(y)}{g(y)}\,\mathrm{d}y
                   = \alpha F(x).
$$

. . .

Taking $x \to \infty$ gives $\Pr(A) = \alpha$. Therefore,
$$
  \Pr(Y \leq x \mid A) = \frac{\alpha F(x)}{\alpha} = F(x).
$$

. . .

Each accepted proposal has the target distribution. Fresh independent attempts
give the first accepted proposal, $Y_\sigma$, this same distribution.

## The Number of Proposals

Each attempt succeeds with probability $\alpha$, independently:
$$
  \Pr(\sigma = n) = (1 - \alpha)^{n-1}\alpha, \qquad n = 1, 2, \ldots.
$$

. . .

```{r}
#| label: rejection-geometric-cost
#| echo: false
#| fig-width: 5
#| fig-height: 1.7

proposal_counts <- tidyr::expand_grid(
  n = 1:25,
  alpha = c(0.2, 0.8)
) |>
  mutate(
    probability = dgeom(n - 1, alpha),
    label = paste("alpha ==", alpha)
  )

ggplot(proposal_counts, aes(n, probability)) +
  geom_segment(aes(xend = n, yend = 0), color = "royalblue") +
  facet_wrap(~label, labeller = "label_parsed") +
  labs(x = expression(n), y = expression(Pr(sigma == n)))
```

. . .

$\sigma$ counts attempts **including the accepted one**:
$\mathbb{E}[\sigma] = 1/\alpha$.

\medskip

Stopping at the $N$th acceptance takes $N/\alpha$ proposals on average.

## Normalizing Constants

Rejection sampling works with unnormalized densities!

\medskip\pause

Suppose:

- Target: $f(y) = c q(y)$ (unknown $c$)
- Proposal: $g(y) = d p(y)$ (can sample from $g$, unknown $d$)

\medskip\pause

Acceptance test:
$$
  U \leq \frac{\alpha f(y)}{g(y)} = \frac{\alpha c q(y)}{d p(y)}
$$

\medskip\pause

If $\alpha = \frac{\alpha' d}{c}$, then
$$
  \frac{\alpha f(y)}{g(y)} = \frac{\alpha' q(y)}{p(y)}
$$

\pause

### Takeaway

You do not need to know $c$ or $d$!

## How to Find $\alpha'$?

Given functions $q$ and $p$, we need to find $\alpha'$ such that
$$
  \alpha' q(y) \leq p(y) \quad \text{for all } y.
$$

\medskip\pause

### How?

We can try to solve the problem
$$
  \alpha'_{\max} = \inf_{y: q(y)>0} \frac{p(y)}{q(y)}.
$$

We need $0 < \alpha' \leq \alpha'_{\max}$.

\medskip\pause

An analytic bound guarantees validity.

\medskip\pause

Numerical optimization (e.g., `optimize()` in R) can suggest a bound, but the
envelope inequality must hold everywhere.

## Finding the von Mises Bound

For $\mu = 0$, take $q(y) = e^{\kappa\cos y}$ and $p(y) = 1$ on $(-\pi,\pi)$.

```{r}
#| label: vonmises-bound
#| echo: false
#| fig-show: animate
#| fig-width: 4
#| fig-height: 2.4

bound_curve <- tibble(y = seq(-pi, pi, length.out = 401)) |>
  mutate(tight_bound = exp(2 * (cos(y) - 1)))

bound_scales <- c(0.5, 1, 1.25)
bound_labels <- c(
  "Valid, with room to spare",
  "Largest valid value",
  "Invalid"
)

for (stage in 1:3) {
  bound_color <- if (stage < 3) "royalblue" else "darkorange"
  p <- ggplot(
    bound_curve,
    aes(y, bound_scales[stage] * tight_bound)
  ) +
    geom_hline(
      yintercept = 1,
      color = "grey40",
      linetype = "dashed"
    ) +
    geom_line(color = bound_color) +
    annotate(
      "point",
      x = 0,
      y = bound_scales[stage],
      color = bound_color
    ) +
    annotate(
      "text",
      x = -2.9,
      y = 1.2,
      label = "p(y) == 1",
      parse = TRUE,
      hjust = 0
    ) +
    scale_x_continuous(
      breaks = c(-pi, 0, pi),
      labels = expression(-pi, 0, pi)
    ) +
    coord_cartesian(ylim = c(0, 1.4)) +
    labs(
      title = bound_labels[stage],
      subtitle = bquote(
        kappa == 2 ~
          "," ~
          alpha^minute ==
            .(bound_scales[stage]) *
              e^{
                -kappa
              }
      ),
      x = expression(y),
      y = expression(alpha^minute * q(y))
    )
  print(p)
}
```

\uncover<2->{Since $\cos y \leq 1$,
$\alpha'_{\max} = \inf_y e^{-\kappa\cos y} = e^{-\kappa}$, attained at $y = 0$.}

## Von Mises Rejection Sampling

For $\mu = 0$, use the uniform proposal on $(-\pi, \pi)$: $g(y) = 1/(2\pi)$.

\medskip\pause

Since
$$
  e^{\kappa(\cos(y)-1)} = \alpha' e^{\kappa \cos(y)} \leq 1,
$$
where $\alpha' = \exp(-\kappa)$, we reject if
$$
  U > e^{\kappa(\cos(Y)-1)}.
$$

. . .

\begin{algorithm}[H]
  \caption{Rejection sampling for the von Mises distribution}
  \Repeat{sample is accepted}{
    Generate $Y \sim \mathcal{U}(-\pi, \pi)$\;
    Generate $U \sim \mathcal{U}(0, 1)$\;
    \If{$U \leq e^{\kappa(\cos(Y) - 1)}$}{
      Accept $Y$\;
    }
  }
\end{algorithm}

## Compute the Acceptance Test on the Log Scale

The same rejection test can be written as
$$
  \log U \leq \log\alpha' + \log q(Y) - \log p(Y).
$$

\medskip\pause

For the posterior in Assignment 2A, compute the log-density as a sum:

```{r}
#| label: posterior-log-density

log_q <- function(theta, x, z) {
  if (theta < 0) {
    return(-Inf)
  }
  eta <- theta * x
  sum(z * eta - exp(eta))
}
```

\pause

Multiplying many small likelihood contributions can underflow to zero. Work with
log-densities and subtract them before exponentiating, if needed.

\medskip\pause

For von Mises sampling, the log acceptance ratio is simply $\kappa(\cos Y - 1)$.

## When No Envelope Exists

We need $g(y) > 0$ wherever $f(y) > 0$ and $\sup_y f(y)/g(y) < \infty$.

\bigskip

```{r}
#| label: rejection-invalid-proposals
#| echo: false
#| fig-width: 5.5
#| fig-height: 2.2

invalid_proposals <- tibble(x = seq(-4, 4, length.out = 500)) |>
  mutate(
    target = dnorm(x),
    proposal = dunif(x, -2, 2),
    log_ratio = dnorm(x, log = TRUE) -
      dnorm(x, sd = 0.5, log = TRUE)
  )

support_plot <- ggplot(invalid_proposals, aes(x)) +
  geom_line(aes(y = target), color = "royalblue") +
  geom_line(aes(y = proposal), color = "darkorange") +
  labs(
    title = "Missing support",
    subtitle = "Blue: target; orange: proposal",
    x = expression(y),
    y = "Density"
  )

tail_plot <- ggplot(invalid_proposals, aes(x, log_ratio)) +
  geom_line(color = "darkorange") +
  labs(
    title = "Tails too light",
    subtitle = "Normal proposal with SD 0.5",
    x = expression(y),
    y = expression(log(f(y) / g(y)))
  )

support_plot + tail_plot
```

. . .

For a standard normal target, neither a $\mathcal{U}(-2,2)$ proposal nor a
$\mathcal{N}(0,1/4)$ proposal admits a finite envelope.

## Implementation

```{r}
#| label: vMsim-slow

get_one_sample_vmises <- function(kappa) {
  repeat {
    Y <- runif(1, -pi, pi)
    U <- runif(1)
    if (log(U) <= kappa * (cos(Y) - 1)) {
      break
    }
  }
  Y
}
```

. . .

```{r}
#| label: vMsim-slow2

sample_vonmises_slow <- function(N, kappa) {
  replicate(N, get_one_sample_vmises(kappa))
}
```

##

```{r}
#| label: vMsim2
#| fig-width: 5.4
#| fig-height: 3
#| echo: false
#| fig-cap: Histogram of samples from `sample_vonmises_slow()`, with theoretical
#|   density overlaid.

f <- function(y, kappa) {
  exp(kappa * cos(y)) / (2 * pi * besselI(kappa, 0))
}

# Simulate data
x1 <- sample_vonmises_slow(100000, 0.5)
x2 <- sample_vonmises_slow(100000, 2)

df <- tibble(
  x = c(x1, x2),
  kappa = factor(rep(c(0.5, 2), each = 100000))
)

ggplot(df, aes(x)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "grey60",
    color = "white"
  ) +
  stat_function(
    fun = f,
    args = list(kappa = 0.5),
    color = "steelblue3",
    data = data.frame(
      x = seq(-pi, pi, length.out = 200),
      kappa = 0.5
    )
  ) +
  stat_function(
    fun = f,
    args = list(kappa = 2),
    color = "steelblue3",
    data = data.frame(
      x = seq(-pi, pi, length.out = 200),
      kappa = 2
    )
  ) +
  labs(y = "Density", x = "x") +
  facet_wrap(~kappa, ncol = 2, labeller = "label_both")
```

##

```{r}
#| label: profile-vsim
#| echo: false
#| eval: false

source(here::here("R", "vmises-sim-slow.R"))
profvis(sample_vonmises_slow(100000, 5))
```

![A line profile of the naive von Mises sampling
implementation.](../images/vmises-slow-profiling.png){width=82%}

## Findings from Profiling

The bottleneck is the repeated calls to `runif(1)`.

\medskip\pause

Both loop overhead and calling the RNG for one number at a time add cost.

\medskip\pause

It is faster to generate random numbers in batches and store them in vectors.

\medskip\pause

But how can we do that for rejection sampling with an \alert{unknown} number of
rejections?

## Vectorized (but Random Length) Rejection Sampler

We can speed up the rejection sampler by generating $m$ proposal samples at
once.

```{r}
#| label: vM-rejection-vec

fixed_vonmises_sampler <- function(m, kappa) {
  Y <- runif(m, -pi, pi)
  u <- runif(m)
  accept <- log(u) <= kappa * (cos(Y) - 1)
  Y[accept]
}
```

. . .

Simple and fast, but returns a **random** number of samples!

\medskip\pause

We need a function that returns exactly $N$ samples. We can estimate $\alpha$ to
choose batch sizes without computing it in advance.

## How Many Samples Does a Batch Return?

We want **100 accepted samples**. How many proposals should we generate?

. . .

Suppose $\alpha = 0.2$. From $m$ independent proposals, the accepted count is
$$
  B_m \sim \mathrm{Binomial}(m,\alpha), \qquad \mathbb{E}[B_m] = m\alpha.
$$

```{r}
#| label: rejection-binomial-batches
#| echo: false
#| fig-width: 5
#| fig-height: 1.9

batch_counts <- tidyr::expand_grid(
  count = 60:170,
  m = c(500, 600)
) |>
  mutate(
    probability = dbinom(count, m, 0.2),
    enough = count >= 100,
    label = sprintf(
      "m = %d: %.0f%% reach 100",
      m,
      100 * pbinom(99, m, 0.2, lower.tail = FALSE)
    )
  )

ggplot(batch_counts, aes(count, probability, color = enough)) +
  geom_segment(aes(xend = count, yend = 0)) +
  geom_vline(xintercept = 100, linetype = "dashed") +
  scale_color_manual(
    values = c("darkorange", "royalblue"),
    guide = "none"
  ) +
  facet_wrap(~label) +
  labs(x = "Accepted samples", y = "Probability")
```

. . .

$m = 100/0.2 = 500$ gives 100 accepted samples **on average**. If a batch falls
short, generate another batch.

##

\begin{algorithm}[H]
  \caption{Sampling exactly $N$ values via rejection sampling with unknown
    acceptance probability $\alpha$.}
  $\widehat{\alpha} \gets 1$\;
  $\symbf{y} \gets$ empty vector\;
  $n_\text{accepted} \gets 0$\;
  \While{$n_\text{accepted} < N$}{
    $m \gets \lceil (N - n_\text{accepted}) / \widehat{\alpha} \rceil$\;
    Generate $m$ proposals and apply the rejection test\;
    $\symbf{y}_\text{new} \gets$ accepted proposals\;
    $n_\text{accepted} \gets n_\text{accepted} +$ length$(\symbf{y}_\text{new})$\;
    \If{$\symbf{y}$ is empty}{
      $\widehat{\alpha} \gets (n_\text{accepted} + 1) / (m + 1)$ \tcp*[f]{Keep the estimate positive.}
    }
    Append $\symbf{y}_\text{new}$ to $\symbf{y}$\;
  }
  \KwRet{$\symbf{y}_{1:N}$}
\end{algorithm}

## Function Factory for Rejection Sampling

\footnotesize

```{r}
#| label: new_rejection_sampler

new_rejection_sampler <- function(generator) {
  function(N, ...) {
    alpha <- 1
    y <- numeric(0)
    n_accepted <- n_proposed <- 0
    while (n_accepted < N) {
      m <- ceiling((N - n_accepted) / alpha)
      y_new <- generator(m, ...)
      n_proposed <- n_proposed + m
      n_accepted <- n_accepted + length(y_new)
      if (length(y) == 0) {
        alpha <- (n_accepted + 1) / (m + 1)
      }
      y <- c(y, y_new)
    }
    list(
      x = y[seq_len(N)],
      proposed = n_proposed,
      accepted = n_accepted
    )
  }
}
```

## Sampling with movMF

We can compare our samplers with the implementation in `movMF`.

\medskip\pause

For $\mu = 0$, use the parameter vector $(\kappa, 0)$. The package returns
points on the unit circle, which we convert to angles.

```{r}
#| label: vMsim

library(movMF)

sample_vonmises_movMF <- function(N, kappa) {
  xy <- rmovMF(N, c(kappa, 0))
  # atan2 preserves the quadrant of each point on the unit circle.
  atan2(xy[, 2], xy[, 1])
}

x <- sample_vonmises_movMF(500, 5)
```

##

```{r}
#| label: vMhist
#| fig-height: 2.5
#| fig-width: 5.5
#| echo: false
#| fig-cap: Histogram of samples from `rmovMF()` for $\mu = 0$ and $\kappa = 5$,
#|   with the theoretical density in blue.

ggplot(tibble(x = x), aes(x)) +
  geom_rug(alpha = 0.1) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 15,
    fill = "grey70",
    color = "white"
  ) +
  stat_function(
    fun = f,
    args = list(kappa = 5),
    color = "royalblue"
  ) +
  labs(y = "Density", x = "x")
```

## Check the Distribution Before Timing

For independent $Z_1,\ldots,Z_5 \sim \mathcal{N}(0,1)$,
$\sum_{j=1}^5 Z_j^2 \sim \chi^2_5$. What if we reuse one normal draw five times?

```{r}
#| label: sampler-validation-example
#| echo: false
#| fig-width: 5
#| fig-height: 2
#| fig-cap: Empirical CDFs of $\chi^2_5$ and $5\chi^2_1$ samples, with the
#|   target $\chi^2_5$ CDF overlaid in black.

set.seed(510)
validation_n <- 20000
independent_chisq <- rowSums(matrix(
  rnorm(5 * validation_n)^2,
  ncol = 5
))
# Dependence among terms within each sum changes the output distribution.
reused_normal_chisq <- 5 * rnorm(validation_n)^2

validation_samples <- tibble(
  x = c(independent_chisq, reused_normal_chisq),
  method = rep(
    c("Independent normals", "Same normal reused"),
    each = validation_n
  )
)

ggplot(validation_samples, aes(x)) +
  stat_ecdf(aes(color = method)) +
  stat_function(fun = pchisq, args = list(df = 5)) +
  scale_color_manual(
    values = c("royalblue", "darkorange"),
    guide = "none"
  ) +
  coord_cartesian(xlim = c(0, 20)) +
  facet_wrap(~method) +
  labs(x = expression(x), y = "CDF")
```

. . .

Both constructions give independent outputs with mean 5, but their variances are
10 and 50. Reusing the normal changes the marginal distribution. Check support,
distribution, and moments before benchmarking.

## Benchmarking the von Mises Samplers

Each method returns $N = 10^5$ angles with $\mu = 0$ and $\kappa = 5$.

```{r}
#| label: vmises-bench-setup

sample_vonmises <- new_rejection_sampler(fixed_vonmises_sampler)
```

```{r}
#| label: vmises-bench
#| fig-width: 5
#| fig-height: 1.5
#| echo: false

bench::mark(
  vectorized = sample_vonmises(1e5, 5)$x,
  scalar = sample_vonmises_slow(1e5, 5),
  movMF = sample_vonmises_movMF(1e5, 5),
  check = FALSE
) |>
  plot()
```

## Piecewise Exponential Envelopes

::: {.columns}
::::: {.column width="47%"}
When $f$ is positive and *log-concave* on an interval $I$, we can construct an
integrable envelope of the form
$$
  f(x) \leq e^{V(x)}
$$
where
$$
  V(x) = \sum_{i=1}^m (a_i x + b_i) \symbf{1}_{I_i}(x)
$$
for intervals $I_i$ forming a partition of $I$.
:::::

::::: {.column width="47%"}
```{r}
#| label: log-concave-envelope-general
#| echo: false

library(ggplot2)

curve_fun <- function(x) dnorm(x)
log_curve <- function(x) dnorm(x, log = TRUE)

# Tangent points
x1 <- -1.5
x2 <- 1.5

# Compute tangent lines in log-space
slope1 <- -x1
intercept1 <- log_curve(x1) - slope1 * x1
slope2 <- -x2
intercept2 <- log_curve(x2) - slope2 * x2

xseq <- seq(-3, 3, length.out = 400)
envelope_log <- ifelse(
  xseq < 0,
  slope1 * xseq + intercept1,
  slope2 * xseq + intercept2
)
envelope <- exp(envelope_log)

df <- data.frame(
  x = xseq,
  curve = curve_fun(xseq),
  envelope = envelope
)

ggplot(df, aes(x)) +
  geom_line(aes(y = curve), color = "grey60") +
  geom_line(aes(y = envelope)) +
  geom_point(aes(x = x1, y = curve_fun(x1))) +
  geom_point(aes(x = x2, y = curve_fun(x2))) +
  scale_y_continuous(
    transform = scales::log_trans(),
    labels = scales::label_math(e^.x, format = log)
  ) +
  labs(y = "Density") +
  theme_minimal()
```
:::::
:::

\bigskip\pause

Typically, $a_i x + b_i$ is tangent to the graph of $\log(f)$ at
$x_i \in I_i = (z_{i-1}, z_i]$ for
$$
  z_0 < x_1 < z_1 < x_2 < \ldots < z_{m-1} < x_m < z_m.
$$

## Why Tangents Give Upper Bounds

If $h = \log f$ is differentiable and concave, then
$$
  h(x) \leq h(x_i) + h'(x_i)(x - x_i) = a_i x + b_i.
$$

```{r}
#| label: envelope-tangent-bounds
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 2.1

normal_log_tangent <- function(x, at) {
  dnorm(at, log = TRUE) - at * (x - at)
}

tangent_example <- tibble(x = seq(-3, 3, length.out = 401)) |>
  mutate(
    log_density = dnorm(x, log = TRUE),
    left = normal_log_tangent(x, -1.5),
    right = normal_log_tangent(x, 1.5),
    envelope = pmin(left, right)
  )

for (stage in 1:2) {
  p <- ggplot(tangent_example, aes(x)) +
    geom_line(aes(y = log_density), color = "grey50") +
    geom_line(aes(y = left), color = "royalblue") +
    annotate(
      "point",
      x = -1.5,
      y = dnorm(-1.5, log = TRUE),
      color = "royalblue"
    ) +
    coord_cartesian(ylim = c(-5.5, 2)) +
    labs(x = expression(x), y = expression(h(x)))
  if (stage >= 2) {
    p <- p +
      geom_line(aes(y = right), color = "darkorange") +
      annotate(
        "point",
        x = 1.5,
        y = dnorm(1.5, log = TRUE),
        color = "darkorange"
      )
  }
  print(p)
}
```

Choose $a_i = h'(x_i)$ and $b_i = h(x_i) - a_i x_i$. Exponentiating gives
$f(x) \leq e^{a_i x+b_i}$.

## Where Should the Pieces Meet?

Use the smallest tangent bound at each point: $V(x) = \min_i\{a_i x + b_i\}$.

```{r}
#| label: envelope-tangent-intersections
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 1.9

for (stage in 1:2) {
  p <- ggplot(tangent_example, aes(x)) +
    geom_line(aes(y = log_density), color = "grey50") +
    geom_line(
      aes(y = left),
      color = "royalblue",
      linetype = "dashed"
    ) +
    geom_line(
      aes(y = right),
      color = "darkorange",
      linetype = "dashed"
    ) +
    geom_vline(xintercept = 0, color = "grey70") +
    coord_cartesian(ylim = c(-5.5, 2)) +
    labs(x = expression(x), y = expression(h(x)))
  if (stage >= 2) {
    p <- p + geom_line(aes(y = envelope))
  }
  print(p)
}
```

Adjacent tangents meet at
$$
  z_i = \frac{b_{i+1} - b_i}{a_i - a_{i+1}} \qquad (a_i \neq a_{i+1}).
$$
Here $z_1 = 0$. Keep the blue tangent to the left and the orange tangent to the
right.

## Sampling from the Envelope

For each piece $I_i = (L_i,R_i]$, compute its finite mass
$$
  w_i = \int_{L_i}^{R_i} e^{a_i x+b_i}\,\mathrm{d}x, \qquad C = \sum_i w_i.
$$

. . .

Choose a piece $J$ with $\Pr(J = i) = w_i/C$.

\begin{center}
\begin{tikzpicture}[node distance=0.6cm, >=stealth]
  \node[draw, rounded corners, fill=blue!10] (piece) {Choose $J$};
  \node[draw, rounded corners, fill=orange!10, right=of piece] (draw) {Draw $Y$ in $I_J$};
  \node[draw, rounded corners, fill=green!10, right=of draw] (test) {Rejection test};
  \draw[->] (piece) -- (draw);
  \draw[->] (draw) -- (test);
\end{tikzpicture}
\end{center}

. . .

Given $J = i$ and $a_i \neq 0$, draw $U \sim \mathcal{U}(0,1)$ and set
$$
  Y = \frac{1}{a_i}\log\!\left[ (1 - U)e^{a_i L_i} + Ue^{a_i R_i} \right].
$$
Use endpoint limits for infinite bounds. If $a_i = 0$, sample uniformly on the
finite interval.

\medskip\pause

This gives $g(y) = e^{V(y)}/C$. Accept with probability $f(y)/e^{V(y)}$.

## Beta Distribution

For shape parameters $\alpha, \beta > 1$, the beta density is log-concave.

```{r}
#| echo: false
#| fig-width: 5.5
#| fig-height: 2.5

fixed_beta_sampler <- function(m, x1, x2, alpha, beta) {
  lf <- function(x) (alpha - 1) * log(x) + (beta - 1) * log1p(-x)
  lf_deriv <- function(x) (alpha - 1) / x - (beta - 1) / (1 - x)
  a1 <- lf_deriv(x1)
  a2 <- lf_deriv(x2)
  if (a1 == 0 || a2 == 0 || a1 - a2 == 0) {
    stop(
      "\nThe implementation requires a_1 and a_2 different and both different from zero. Choose different values of x_1 and x_2.",
      call. = FALSE
    )
  }
  b1 <- lf(x1) - a1 * x1
  b2 <- lf(x2) - a2 * x2
  z1 <- (b2 - b1) / (a1 - a2)
  Q1 <- exp(b1) * (exp(a1 * z1) - 1) / a1
  c <- Q1 + exp(b2) * (exp(a2 * 1) - exp(a2 * z1)) / a2

  y <- log_ratio <- numeric(m)
  uy <- c * runif(m)
  u <- runif(m)
  i <- uy < Q1
  y[i] <- z <- log(a1 * exp(-b1) * uy[i] + 1) / a1
  y[!i] <- zz <- log(
    a2 * exp(-b2) * (uy[!i] - Q1) + exp(a2 * z1)
  ) /
    a2
  log_ratio[i] <- lf(z) - a1 * z - b1
  log_ratio[!i] <- lf(zz) - a2 * zz - b2
  accept <- log(u) <= log_ratio
  y[accept]
}

sample_beta <- new_rejection_sampler(fixed_beta_sampler)
```

```{r}
#| label: Beta-fig
#| fig-cap: Samples from a piecewise exponential envelope sampler, with
#|   densities (blue) and envelopes (orange) overlaid.
#| fig-width: 5.5
#| fig-height: 2.65
#| echo: false

params <- data.frame(
  x1 = c(0.3, 0.3, 0.2, 0.2),
  x2 = c(0.7, 0.7, 0.5, 0.5),
  alpha = c(4, 1.8, 4, 1.8),
  beta = c(2, 2.4, 2, 2.4),
  label = c(
    "x1=0.3, x2=0.7, alpha=4, beta=2",
    "x1=0.3, x2=0.7, alpha=1.8, beta=2.4",
    "x1=0.2, x2=0.5, alpha=4, beta=2",
    "x1=0.2, x2=0.5, alpha=1.8, beta=2.4"
  )
)

make_beta_df <- function(N, x1, x2, alpha, beta, label) {
  x <- sample_beta(
    N,
    x1 = x1,
    x2 = x2,
    alpha = alpha,
    beta = beta
  )$x
  data.frame(x = x, label = label)
}

sample_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_beta_df(100000, x1, x2, alpha, beta, label)
      )
    }
  )
)

xseq <- seq(0, 1, length.out = 400)

make_density_df <- function(xseq, alpha, beta, label) {
  data.frame(
    x = xseq,
    density = xseq^(alpha - 1) *
      (1 - xseq)^(beta - 1) /
      beta(alpha, beta),
    label = label
  )
}

make_envelope_df <- function(xseq, x1, x2, alpha, beta, label) {
  envelope <- function(x, x1, x2, alpha, beta) {
    lf <- function(x) {
      (alpha - 1) * log(x) + (beta - 1) * log1p(-x)
    }
    lf_deriv <- function(x) {
      (alpha - 1) / x - (beta - 1) / (1 - x)
    }
    a1 <- lf_deriv(x1)
    a2 <- lf_deriv(x2)
    b1 <- lf(x1) - a1 * x1
    b2 <- lf(x2) - a2 * x2
    z1 <- (b2 - b1) / (a1 - a2)
    ifelse(x < z1, exp(a1 * x + b1), exp(a2 * x + b2))
  }
  data.frame(
    x = xseq,
    envelope = envelope(xseq, x1, x2, alpha, beta) /
      beta(alpha, beta),
    label = label
  )
}

dens_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_density_df(xseq, alpha, beta, label)
      )
    }
  )
)
env_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_envelope_df(xseq, x1, x2, alpha, beta, label)
      )
    }
  )
)

ggplot(sample_df, aes(x)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 50,
    fill = "grey80"
  ) +
  geom_line(
    data = dens_df,
    aes(x = x, y = density),
    color = "steelblue3"
  ) +
  geom_line(
    data = env_df,
    aes(x = x, y = envelope),
    color = "dark orange"
  ) +
  coord_cartesian(ylim = c(0, 3)) +
  facet_wrap(~label, ncol = 2)
```

##

```{r}
#| label: beta-envelopes
#| echo: false
#| warning: false
#| fig-width: 5.5
#| fig-height: 3
#| fig-cap: Theoretical densities (blue) and envelopes (orange) for different
#|   parameter values and choices of $x_1$ and $x_2$.

params <- data.frame(
  alpha = c(4, 1.8),
  beta = c(2, 2.4),
  x1 = c(0.3, 0.2),
  x2 = c(0.7, 0.5),
  label = c(
    "alpha=4, beta=2, x1=0.3, x2=0.7",
    "alpha=1.8, beta=2.4, x1=0.2, x2=0.5"
  )
)

xseq <- seq(0.05, 0.95, length.out = 400)

make_df <- function(xseq, alpha, beta, x1, x2, label) {
  envelope_fun <- function(x, x1, x2, alpha, beta) {
    lf <- function(x) {
      (alpha - 1) * log(x) + (beta - 1) * log1p(-x)
    }
    lf_deriv <- function(x) {
      (alpha - 1) / x - (beta - 1) / (1 - x)
    }
    a1 <- lf_deriv(x1)
    a2 <- lf_deriv(x2)
    b1 <- lf(x1) - a1 * x1
    b2 <- lf(x2) - a2 * x2
    z1 <- (b2 - b1) / (a1 - a2)
    ifelse(x < z1, exp(a1 * x + b1), exp(a2 * x + b2))
  }
  data.frame(
    x = xseq,
    density = xseq^(alpha - 1) *
      (1 - xseq)^(beta - 1) /
      beta(alpha, beta),
    envelope = envelope_fun(xseq, x1, x2, alpha, beta) /
      beta(alpha, beta),
    label = label
  )
}

plot_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_df(xseq, alpha, beta, x1, x2, label)
      )
    }
  )
)

ggplot(plot_df, aes(x = x)) +
  geom_line(aes(y = density), color = "steelblue3") +
  geom_line(aes(y = envelope), color = "dark orange") +
  scale_y_log10() +
  facet_wrap(~label, ncol = 2)
```

## Adapting the Envelope

The examples above use fixed envelopes. Adaptive rejection sampling updates the
envelope using new points.

\medskip\pause

For a fixed envelope, a simple cost model is
$$
  \mathbb{E}[T] \approx T_{\mathrm{setup}}
                        + \frac{N}{\alpha}\,c_{\mathrm{attempt}},
$$
where $c_{\mathrm{attempt}}$ includes generating and testing a proposal.

\medskip\pause

Batching reduces overhead per proposal. A tighter envelope reduces rejections,
but constructing, sampling, and updating it also cost time.

## One Adaptive Update

For a log-concave target, use a rejected proposal $Y$ as a new tangent point.

```{r}
#| label: envelope-adaptive-update
#| echo: false
#| fig-show: animate
#| fig-width: 5
#| fig-height: 2.1

adaptive_example <- tangent_example |>
  mutate(
    new_tangent = normal_log_tangent(x, 0),
    updated = pmin(envelope, new_tangent)
  )

for (stage in 1:3) {
  p <- ggplot(adaptive_example, aes(x)) +
    geom_line(aes(y = exp(log_density))) +
    geom_line(aes(y = exp(envelope)), color = "royalblue") +
    coord_cartesian(ylim = c(0, 1.3)) +
    labs(x = expression(x), y = "Density")
  if (stage >= 2) {
    p <- p +
      annotate("point", x = 0, y = 0.9, color = "darkorange") +
      annotate(
        "text",
        x = 1.1,
        y = 0.95,
        label = "Rejected at Y = 0"
      )
  }
  if (stage >= 3) {
    p <- p +
      geom_line(aes(y = exp(updated)), color = "darkorange")
  }
  print(p)
}
```

$$
  V_{\mathrm{new}}(x) = \min\{V_{\mathrm{old}}(x),\ h(Y) + h'(Y)(x - Y)\}.
$$
\uncover<3->{The orange envelope is tighter. Update the piece masses before the
next draw.}

## Does a Tighter Envelope Run Faster?

Compare uniform and two-piece exponential proposals for $\mathrm{Beta}(a,a)$.
Larger $a$ concentrates the target near $1/2$.

```{r}
#| label: envelope-benchmark-helpers
#| include: false

compare_rejection_pair <- function(N, uniform, piecewise) {
  draws <- list(uniform = uniform(N), piecewise = piecewise(N))
  timings <- bench::mark(
    uniform = uniform(N)$x,
    piecewise = piecewise(N)$x,
    check = FALSE,
    min_iterations = 10,
    min_time = 0.1
  )
  methods <- as.character(timings$expression)
  tibble(
    method = methods,
    acceptance = vapply(
      draws[methods],
      function(draw) draw$accepted / draw$proposed,
      numeric(1)
    ),
    microseconds = 1e6 * as.numeric(timings$median) / N
  )
}

plot_envelope_comparison <- function(results, x_label) {
  results |>
    pivot_longer(
      c(acceptance, microseconds),
      names_to = "metric",
      values_to = "value"
    ) |>
    mutate(
      metric = factor(
        metric,
        levels = c("acceptance", "microseconds"),
        labels = c("Acceptance", "Time")
      )
    ) |>
    ggplot(aes(parameter, value, color = method)) +
    geom_line() +
    geom_point() +
    facet_grid(
      metric ~ N,
      scales = "free_y",
      labeller = labeller(N = function(x) paste("N =", x))
    ) +
    scale_x_continuous(
      breaks = sort(unique(results$parameter))
    ) +
    scale_color_manual(
      values = c(
        piecewise = "darkorange",
        uniform = "royalblue"
      ),
      labels = c(piecewise = "Piecewise", uniform = "Uniform")
    ) +
    labs(x = x_label, y = NULL, color = NULL)
}

fixed_beta_uniform_sampler <- function(m, shape) {
  log_target <- function(x) (shape - 1) * (log(x) + log1p(-x))
  y <- runif(m)
  log_ratio <- log_target(y) - log_target(0.5)
  y[log(runif(m)) <= log_ratio]
}

sample_beta_uniform <- new_rejection_sampler(
  fixed_beta_uniform_sampler
)

set.seed(511)
beta_comparison <- pmap_dfr(
  expand_grid(N = c(100, 10000), shape = c(2, 5, 20)),
  function(N, shape) {
    # Points one standard deviation from the mode follow its concentration.
    delta <- sqrt(1 / (4 * (2 * shape + 1)))
    compare_rejection_pair(
      N,
      uniform = function(n) sample_beta_uniform(n, shape),
      piecewise = function(n) {
        sample_beta(n, 0.5 - delta, 0.5 + delta, shape, shape)
      }
    ) |>
      mutate(N = N, parameter = shape)
  }
)
```

```{r}
#| label: beta-envelope-benchmark
#| echo: false
#| fig-width: 5.5
#| fig-height: 2.25

plot_envelope_comparison(beta_comparison, expression(a))
```

\pause

Time is median microseconds per returned sample, including setup and all
batches. Acceptance is accepted/proposed, including surplus draws.

\pdfpcnote{
  The uniform envelope has height equal to the beta density at its mode, 1/2.
  For the piecewise envelope, place tangent points one standard deviation
  either side of 1/2. The acceptance proportions come from separate sampler
  runs; the timings are medians over repeated runs. A higher acceptance rate
  need not give a lower runtime.
}

## Transformation Methods vs. Rejection Sampling

::: {.columns}
::::: {.column width="47%"}
### Transformation Methods

Transform samples from a simple distribution to the target distribution.

\medskip\pause

Efficient if the transformation is available and cheap to compute.

\medskip\pause

Finding or evaluating a transformation can be difficult for complex or
unnormalized distributions.
:::::

. . .

::::: {.column width="47%"}
### Rejection Sampling

Uses a proposal distribution and accepts/rejects samples.

\medskip\pause

Works even if the quantile function is unavailable or expensive.

\medskip\pause

Efficiency depends on the proposal and acceptance probability.

\medskip\pause

Works with unnormalized densities, but acceptance can be very low in high
dimensions.
:::::
:::

. . .

### Takeaway {.example}

Use a simple transformation when available. Otherwise, consider rejection
sampling with a suitable envelope.

## Exercises

### Exercise 1

For a positive integer $k$, use the fact that
$$
  W = X_1^2 + \ldots + X_k^2 \sim \chi^2_k \quad \text{for} \quad X_1,
  \ldots, X_k \sim_{\text{i.i.d.}} \mathcal{N}(0, 1)
$$
to implement a function, `my_rchisq()`, such that `my_rchisq(N, k)` returns a
sample of $N$ i.i.d. observations from $\chi^2_k$.

```{r}
#| label: rchisq2-implementation
#| include: false

my_rchisq <- function(N, k) {
  y <- numeric(N)
  for (i in seq_len(N)) {
    x <- rnorm(k)
    y[i] <- sum(x^2)
  }
  y
}
```

```{r}
#| label: rchisq2-implementation-faster
#| include: false

my_fast_rchisq <- function(N, k) {
  x <- matrix(rnorm(k * N), k)
  colSums(x^2)
}
```

. . .

### Exercise 2

Make another function, `my_other_rchisq()`, that uses inversion sampling.

```{r}
#| label: rchisq2
#| include: false

my_other_rchisq <- function(N, k) {
  x <- runif(N)
  qchisq(x, k)
}
```

. . .

### Exercise 3

Benchmark your implementations against one another and `rchisq()`.

```{r}
#| label: rchisq-bench
#| include: false

N <- 1e4
k <- 5
bench::mark(
  rchisq(N, k),
  my_rchisq(N, k),
  my_fast_rchisq(N, k),
  my_other_rchisq(N, k),
  check = FALSE
) |>
  plot()
```

## Exercise: Rejection Sampling

The $\mathrm{Beta}(2,2)$ density is
$$
  f(x) = 6x(1 - x), \qquad 0 < x < 1.
$$

Use a $\mathcal{U}(0,1)$ proposal. Find a valid envelope and predict the
acceptance probability.

\medskip\pause

1. Implement a sampler that returns exactly $N$ values, using
   `new_rejection_sampler()` to handle batches.
2. Check support, compare the empirical CDF with `pbeta(x, 2, 2)`, and check the
   mean $1/2$ and variance $1/20$.
3. Compare accepted/proposed with your prediction, then benchmark against
   `rbeta()` for small and large $N$.

```{r}
#| label: beta-rejection-exercise-solution
#| include: false

fixed_beta22_sampler <- function(m) {
  y <- runif(m)
  y[log(runif(m)) <= log(4) + log(y) + log1p(-y)]
}
sample_beta22 <- new_rejection_sampler(fixed_beta22_sampler)
```

## Summary

### Inversion Sampling

- Use the quantile function to transform uniform samples to the target
  distribution.\pause
- Efficient if the quantile function is available and cheap to compute.\pause

### Rejection Sampling

- Use a proposal distribution and accept/reject samples.\pause
- Works even if the quantile function is unavailable or expensive.

## Next Time

### Monte Carlo Integration

Sample from a distribution to estimate expectations of functions of the random
variable.

### Importance Sampling

Weight samples from a proposal distribution to estimate expectations with
respect to a target distribution.

\appendix

# Appendix

## Von Mises Envelopes

::: {.columns}
::::: {.column width="47%"}
For $\mu = 0$ and $\kappa > 0$, the von Mises density is not globally
log-concave.

\medskip\pause

Since $(\log f)''(x) = -\kappa\cos(x)$, it is

- log-concave on $(-\pi/2, \pi/2)$ and
- log-convex on $(-\pi,-\pi/2)$ and $(\pi/2, \pi)$.

\medskip\pause

Use tangents in the center and chords in the tails to bound $\log f$.
:::::

::::: {.column width="47%"}
```{r}
#| echo: false
#| fig-cap: The von Mises log-density with $\mu = 0$ and $\kappa = 0.5$.

dvm <- function(x, k) exp(k * cos(x)) / (2 * pi * besselI(k, 0))

xy <- tibble(
  x = seq(-pi, pi, length.out = 400),
  y = log(dvm(x, 0.5))
)

ggplot(xy, aes(x = x, y = y)) +
  geom_vline(
    xintercept = c(-pi, -pi / 2, pi / 2, pi),
    color = "grey60"
  ) +
  geom_line() +
  labs(y = "log-density") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
:::::
:::

##

```{r}
#| label: vonmises-adaptive
#| echo: false

dvm <- function(x, k) exp(k * cos(x)) / (2 * pi * besselI(k, 0))

von_mises_adapt_random <- function(N, x1, x2, kappa) {
  lf <- function(x) kappa * cos(x)
  lf_deriv <- function(x) -kappa * sin(x)
  a1 <- 2 * kappa / pi
  a2 <- lf_deriv(x1)
  a3 <- lf_deriv(x2)
  a4 <- -a1

  b1 <- kappa
  b2 <- lf(x1) - a2 * x1
  b3 <- lf(x2) - a3 * x2
  b4 <- kappa

  z0 <- -pi
  z1 <- -pi / 2
  z2 <- (b3 - b2) / (a2 - a3)
  z3 <- pi / 2
  z4 <- pi

  Q1 <- exp(b1) * (exp(a1 * z1) - exp(a1 * z0)) / a1
  Q2 <- Q1 + exp(b2) * (exp(a2 * z2) - exp(a2 * z1)) / a2
  Q3 <- Q2 + exp(b3) * (exp(a3 * z3) - exp(a3 * z2)) / a3
  c <- Q3 + exp(b4) * (exp(a4 * z4) - exp(a4 * z3)) / a4

  u0 <- c * runif(N)
  u <- runif(N)

  I1 <- (u0 < Q1)
  I2 <- (u0 >= Q1) & (u0 < Q2)
  I3 <- (u0 >= Q2) & (u0 < Q3)
  I4 <- (u0 >= Q3)

  x <- numeric(N)
  accept <- logical(N)
  x[I1] <- log(a1 * exp(-b1) * u0[I1] + exp(a1 * z0)) / a1
  accept[I1] <- log(u[I1]) <= lf(x[I1]) - a1 * x[I1] - b1
  x[I2] <- log(a2 * exp(-b2) * (u0[I2] - Q1) + exp(a2 * z1)) / a2
  accept[I2] <- log(u[I2]) <= lf(x[I2]) - a2 * x[I2] - b2
  x[I3] <- log(a3 * exp(-b3) * (u0[I3] - Q2) + exp(a3 * z2)) / a3
  accept[I3] <- log(u[I3]) <= lf(x[I3]) - a3 * x[I3] - b3
  x[I4] <- log(a4 * exp(-b4) * (u0[I4] - Q3) + exp(a4 * z3)) / a4
  accept[I4] <- log(u[I4]) <= lf(x[I4]) - a4 * x[I4] - b4

  x[accept]
}

von_mises_adapt <- new_rejection_sampler(von_mises_adapt_random)

envelope <- function(x, x1, x2, kappa) {
  lf <- function(x) kappa * cos(x)
  lf_deriv <- function(x) -kappa * sin(x)
  a1 <- 2 * kappa / pi
  a2 <- lf_deriv(x1)
  a3 <- lf_deriv(x2)
  a4 <- -a1

  b1 <- kappa
  b2 <- lf(x1) - a2 * x1
  b3 <- lf(x2) - a3 * x2
  b4 <- kappa

  z0 <- -pi
  z1 <- -pi / 2
  z2 <- (b3 - b2) / (a2 - a3)
  z3 <- pi / 2
  z4 <- pi

  env <- numeric(length(x))
  i1 <- x < z1
  i2 <- x >= z1 & x < z2
  i3 <- x >= z2 & x < z3
  i4 <- x >= z3

  env[i1] <- exp(a1 * x[i1] + b1)
  env[i2] <- exp(a2 * x[i2] + b2)
  env[i3] <- exp(a3 * x[i3] + b3)
  env[i4] <- exp(a4 * x[i4] + b4)

  env
}
```

```{r}
#| label: vonmises-adaptive-plot
#| fig-cap: Von Mises samples from a piecewise exponential envelope sampler,
#|   with densities (blue) and envelopes (orange) overlaid.
#| fig-width: 5.5
#| fig-height: 3
#| echo: false

params <- data.frame(
  x1 = c(-0.4, -0.1, -0.4, -1),
  x2 = c(0.4, 0.1, 0.4, 1),
  kappa = c(5, 5, 2, 2),
  label = c(
    "kappa=5, x1=-0.4, x2=0.4",
    "kappa=5, x1=-0.1, x2=0.1",
    "kappa=2, x1=-0.4, x2=0.4",
    "kappa=2, x1=-1, x2=1"
  )
)

make_vm_df <- function(N, x1, x2, kappa, label) {
  x <- von_mises_adapt(N, x1, x2, kappa)$x
  data.frame(x = x, label = label)
}

sample_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(params[i, ], make_vm_df(100000, x1, x2, kappa, label))
    }
  )
)

xseq <- seq(-pi, pi, length.out = 400)

make_density_df <- function(xseq, kappa, label) {
  data.frame(x = xseq, density = dvm(xseq, kappa), label = label)
}

make_envelope_df <- function(xseq, x1, x2, kappa, label) {
  env <- envelope(xseq, x1, x2, kappa) /
    (2 * pi * besselI(kappa, 0))
  data.frame(x = xseq, envelope = env, label = label)
}

dens_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(params[i, ], make_density_df(xseq, kappa, label))
    }
  )
)
env_df <- do.call(
  rbind,
  lapply(
    seq_len(nrow(params)),
    function(i) {
      with(
        params[i, ],
        make_envelope_df(xseq, x1, x2, kappa, label)
      )
    }
  )
)

ggplot(sample_df, aes(x)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "grey80"
  ) +
  geom_line(
    data = dens_df,
    aes(x = x, y = density),
    color = "steelblue3"
  ) +
  geom_line(
    data = env_df,
    aes(x = x, y = envelope),
    color = "dark orange"
  ) +
  coord_cartesian(ylim = c(0, 1.5)) +
  facet_wrap(~label, ncol = 2)
```

## Comparing the von Mises Envelopes

Compare the uniform proposal with the four-piece envelope. Use central tangent
points $\pm\min(1,\kappa^{-1/2})$ to follow the target's concentration.

```{r}
#| label: vonmises-envelope-benchmark
#| echo: false
#| fig-width: 5.5
#| fig-height: 2.25

set.seed(512)
vonmises_comparison <- pmap_dfr(
  expand_grid(N = c(100, 10000), kappa = c(0.5, 5, 20)),
  function(N, kappa) {
    delta <- min(1, 1 / sqrt(kappa))
    compare_rejection_pair(
      N,
      uniform = function(n) sample_vonmises(n, kappa),
      piecewise = function(n) {
        von_mises_adapt(n, -delta, delta, kappa)
      }
    ) |>
      mutate(N = N, parameter = kappa)
  }
)
plot_envelope_comparison(vonmises_comparison, expression(kappa))
```

\pause

These envelopes stay fixed during sampling. Time is median microseconds per
returned sample, including setup and all batches. Compare both rows as $N$ and
$\kappa$ change.
