---
title: "Parallelization and Scatterplot Smoothing"
---

{{< include _common.qmd >}}

## Today

### Parallelization

We use multiple cores to speed up computations.

. . .

### Scatterplot Smoothing

We will cover **nearest-neighbor smoothing** and **smoothing splines**.

# Parallelization

## Parallelization

Most processors have multiple cores.

\medskip\pause

Unless instructed otherwise, only a single core is used.

\medskip\pause

The computer doesn't automatically know that your computations are safe to do in
parallel.

## Embarrassingly Parallel Tasks

These tasks separate easily into independent subtasks.

\pause\medskip

### Examples {.example}

- Summing a vector (or matrix): `sum()`\pause
- Linear algebra operations: `%*%` (`crossprod()`)\pause
- Running iterations of a simulation\pause
- Cross-validation

## The foreach Package

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: foreach-load
#| warning: false
#| message: false

library(foreach)
```

```{r}
#| label: foreach-sqrt
#| eval: false

foreach(i = seq_len(3)) %dopar%
  {
    sqrt(i)
  }
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| echo: false
#| ref-label: "foreach-sqrt"
```
:::::
:::

\pause\medskip

### Notes {.alert}

- Returns a list, so it is not really a `for` loop.\pause
- Nothing is actually parallel yet!\pause
- First, we need to register a **backend**.

## Backends

Several backends are available as separate packages (**doParallel**, **doMC**,
**doFuture**). Load and **register** one before using `%dopar%`.

. . .

::: {.columns}
::::: {.column width="43%"}
```{r}
#| label: doparallel

library(doParallel)
cl <- makeCluster(2)
registerDoParallel(cl)
```
:::::

. . .

::::: {.column width="51%"}
```{r}
#| label: doparallel-execute
#| message: false

foreach(i = seq_len(3)) %dopar%
  {
    paste("Hi from", Sys.getpid())
  }
```
:::::
:::

## Combining Results

`foreach()` returns a list by default.

\medskip\pause

If you want to reduce your result, use either the `.combine` argument or
manually reduce the resulting list.

. . .

```{r}
x <- c(4, 1, 1e2, 2)

foreach(i = seq_len(4), .combine = c) %dopar%
  {
    log(x[[i]])
  }
```

## Caveats

### Realistic Expectations

In practice, twice as many cores $\neq$ twice the speed.

. . .

### Overhead

Parallelization comes with overhead from starting workers and communicating
between them.

. . .

### Thread Safety

Some operations are not thread-safe---for example, writing to the same file from
multiple workers.

. . .

### Redundancies

Parallelizing a task that is already parallelized may not help and may even
\alert{hurt} performance.

## Parallelization on One Computer

\begin{center}
\begin{tikzpicture}[
    font=\small,
    box/.style={draw, rounded corners=2pt, minimum height=0.45cm,
      minimum width=1.7cm, inner sep=3pt},
    flow/.style={-{Stealth}, draw=RoyalBlue3}
  ]
  \draw[rounded corners=4pt, draw=black!50, fill=black!2]
    (-2, -1.35) rectangle (2, 3.25);
  \node at (0, 2.95) {One computer};
  \draw[rounded corners=2pt, draw=RoyalBlue3, fill=RoyalBlue3!5]
    (-1.35, -0.2) rectangle (1.35, 2.6);
  \node at (0, 2.3) {Processor};
  \node at (-4, 2.95) {Tasks};
  \node at (4, 2.95) {Results};

  \foreach \i in {1,...,4} {
    \node[box, draw=RoyalBlue3, fill=RoyalBlue3!10]
      (task-\i) at (-4, {2.35 - 0.55*\i}) {Task \i};
    \node[box, draw=RoyalBlue3, fill=RoyalBlue3!15]
      (core-\i) at (0, {2.35 - 0.55*\i}) {Core \i};
    \node[box, draw=DarkOrange3, fill=DarkOrange1!15]
      (result-\i) at (4, {2.35 - 0.55*\i}) {Result \i};
    \draw[flow] (task-\i) -- (core-\i);
    \draw[flow, draw=DarkOrange3] (core-\i) -- (result-\i);
  }

  \node[box, draw=black!50, fill=black!5, minimum width=2.7cm]
    (memory) at (0, -0.85) {Shared memory};
  \draw[{Stealth}-{Stealth}, draw=black!50] (0, -0.2) -- (memory.north);
\end{tikzpicture}
\end{center}

\pause

Independent tasks can run simultaneously on different cores of one processor.

\pdfpcnote{
  The cores share the memory of one computer.

  Most R code uses only one core unless we explicitly parallelize it.
}

## Parallelization on an HPC Cluster

\begin{center}
\begin{tikzpicture}[
    font=\small,
    box/.style={draw, rounded corners=2pt, align=center, inner sep=4pt},
    flow/.style={-{Stealth}, draw=RoyalBlue3}
  ]
  \node[box, draw=RoyalBlue3, fill=RoyalBlue3!10, minimum width=1.5cm]
    (jobs) at (-5.45, 0) {Job\\queue};
  \node[box, draw=RoyalBlue3, fill=RoyalBlue3!10, minimum height=0.9cm]
    (scheduler) at (-3, 0) {Scheduler};
  \node[box, draw=DarkOrange3, fill=DarkOrange1!15, minimum height=0.9cm]
    (results) at (5.25, 0) {Results};
  \draw[flow] (jobs) -- (scheduler);

  \foreach \i/\y in {1/1.7, 2/0, 3/-1.7} {
    \node[box, draw=black!50, fill=black!2,
      minimum width=4.3cm, minimum height=1.5cm]
      (node-\i) at (1.3, \y) {};
    \node at (1.3, {\y + 0.52}) {Node \i};
    \node[box, font=\scriptsize, draw=black!50, fill=black!5,
      minimum width=3.8cm, minimum height=0.34cm, inner sep=1pt]
      (memory-\i) at (1.3, {\y - 0.49}) {Shared memory};

    \foreach \j in {1,...,4} {
      \node[box, font=\scriptsize, draw=RoyalBlue3, fill=RoyalBlue3!15,
        minimum width=0.88cm, minimum height=0.42cm, inner sep=1pt]
        (core-\i-\j) at ({1.3 + (\j - 2.5)}, {\y + 0.02}) {Core \j};
      \draw[draw=black!50] (core-\i-\j.south)
        -- (core-\i-\j.south |- memory-\i.north);
    }

    \draw[flow] (scheduler.east) to[out=0, in=180] (node-\i.west);
    \draw[flow, draw=DarkOrange3]
      (node-\i.east) to[out=0, in=180] (results.west);
  }
\end{tikzpicture}
\end{center}

\pause

A scheduler assigns independent jobs to cores across several **nodes**.

\pdfpcnote{
  A node is one physical computer containing several cores and shared memory.

  Communication between nodes is usually more expensive than communication
  within a node.

  Job arrays are ideal for independent simulation scenarios or bootstrap
  batches.

  The program must not start more workers than the number of allocated cores.

  Have each job write to a separate output file.
}

# Scatterplot Smoothing

## Nuuk Temperature Data

```{r}
#| label: load-nuuk-data
#| echo: false
#| message: false

nuuk <- read_table(
  here::here("data", "nuuk.txt"),
  col_names = c("Year", 1:12),
  na = "-999",
  skip = 1
) |>
  gather(
    key = "Month",
    value = "Temperature",
    -Year,
    convert = TRUE
  ) |>
  mutate(Temperature = Temperature / 10) |>
  filter(Year > 1866)

nuuk_year <- group_by(nuuk, Year) |>
  summarise(
    Temperature = mean(Temperature),
    Median = median(Temperature),
    High = max(Temperature),
    Low = min(Temperature)
  )
N <- nrow(nuuk_year)
```

```{r}
#| fig-cap: Annual average temperature in Nuuk, Greenland
#| fig-width: 4

nuuk_plot <- ggplot(nuuk_year, aes(Year, Temperature)) +
  geom_point()
nuuk_plot
```

Data is available [here](../data/nuuk.txt).

## Nearest-Neighbor Estimation

Given data points $(x_1, y_1), \ldots, (x_N, y_N)$, the $k$-nearest-neighbor
estimator at $x_i$ is
$$
  \hat{f}_i = \frac{1}{k} \sum_{j \in \mathcal{N}_i} y_j
$$
where $\mathcal{N}_i$ is the set of indices for the $k$ nearest neighbors of
$x_i$.

\medskip\pause

This is an estimator of
$$
  f(x) = \E(Y \mid X = x).
$$

\pdfpcnote{
 N_i is the neighborhood of x_i.
}

\pdfpcnote{
  Show example with k = 3.
}

## Matrix Formulation

We can express all fitted values as a matrix-vector product:
$$
  \hat{\symbf{f}} = \symbf{S} \symbf{y}
$$
where $\symbf{y} = (y_1, \ldots, y_N)^T$ and $\symbf{S}$ is the **smoother
matrix**.

\pause

### Smoother Matrix

Each entry of $\symbf{S}$ is
$$
  S_{ij} = \frac{1}{k} \cdot \symbf{1}_{\{j \in \mathcal{N}_i\}}
$$
where $\symbf{1}_{\{j \in \mathcal{N}_i\}}$ is 1 if $j$ is a neighbor of $i$ and
0 otherwise.

\pdfpcnote{
  - Each fitted value is a weighted average of the y-values.\\
   -S is a relatively sparse N x N matrix.
}

## Smoother Matrix: An Example

Take $x_i = i$ for $i = 1, \ldots, 5$ and $k = 3$. Include each point itself
among its neighbors.

$$
  \symbf{S} = \frac{1}{3} \begin{pmatrix}
                            1                 & 1                 & 1                 & 0                 & 0                 \\
                            1                 & 1                 & 1                 & 0                 & 0                 \\
                            \alert < 2- > {0} & \alert < 2- > {1} & \alert < 2- > {1} & \alert < 2- > {1} & \alert < 2- > {0} \\
                            0                 & 0                 & 1                 & 1                 & 1                 \\
                            0                 & 0                 & 1                 & 1                 & 1
                          \end{pmatrix}.
$$

\pause

The third row selects and averages the neighbors of $x_3$:
$$
  \hat{f}_3 = \frac{y_2 + y_3 + y_4}{3}.
$$

\pdfpcnote{
  Each row has three nonzero entries, each with weight 1/3.

  At the endpoints, we still select the three closest points. This explains
  why the first two rows are identical, as are the last two rows.
}

## Sparse Storage

When $k \ll N$, most entries of $\symbf{S}$ are zero: it is a **sparse
matrix**.

The third row of our example is
$$
  \begin{pmatrix} 0 & 1/3 & 1/3 & 1/3 & 0 \end{pmatrix}.
$$

\pause

Store only the nonzero values and their positions:
$$
  \begin{array}{c|ccc}
    \text{Column index} & 2 & 3 & 4 \\
    \hline
    \text{Value} & 1/3 & 1/3 & 1/3
  \end{array}
$$

\pause

For $N$ rows with $k$ nonzeros each, storage grows as $O(Nk)$.

\pdfpcnote{
  A dense matrix stores all N squared entries, including the zeros.

  Sparse storage also needs indices to locate the nonzero entries. The table
  illustrates one row; there are several formats for organizing a whole matrix.

  Our five-point example is small. The savings become substantial when k is
  small relative to N; sparse storage can cost more for tiny matrices.
}

## Sparse Matrices in R

Construct the smoother matrix directly from its nonzero entries.

::: {.columns}
::::: {.column width="56%"}
```{r}
#| label: sparse-smoother-example

S_example <- Matrix::sparseMatrix(
  i = rep(1:5, each = 3),
  j = c(1:3, 1:3, 2:4, 3:5, 3:5),
  x = 1 / 3,
  dims = c(5, 5)
)
```
:::::

. . .

::::: {.column width="40%"}
```{r}
#| label: sparse-smoother-display
#| echo: false

Matrix::printSpMatrix(3 * S_example)
```
:::::
:::

The display is scaled by 3. Dots represent zeros that are not stored.

\pause\medskip

The usual `%*%` operator also works with sparse matrices, using only the stored
entries when multiplying by a vector.

\pdfpcnote{
  The i and j arguments give row and column indices. The x argument gives the
  corresponding values; a single value is repeated for every listed position.

  A base R matrix containing many zeros still uses dense storage. Constructing
  the sparse matrix directly avoids allocating that dense matrix first.
}

## Linear Smoothers

Smoothers of the form $\hat{\symbf{f}} = \symbf{S} \symbf{y}$ are called *linear
smoothers*.

Here, linear means linear in $\symbf{y}$, with the $x$-values and $k$ fixed.

\pause\medskip
The $k$-nearest-neighbor smoother is a simple example of a linear smoother that
works for $x$-values in any metric space.

\pause\medskip

The representation of a linear smoother as a matrix-vector product,
$$
  \symbf{S} \symbf{y}
$$
is theoretically useful, but often not the best way to actually compute
$\hat{\symbf{f}}$.

## Running Mean

When $x_i \in \mathbb{R}$, we can sort the data according to the $x$-values and
then use a *symmetric* neighbor definition:
$$
  \mathcal{N}_i
  = \left\{ i - \frac{k - 1}{2}, i - \frac{k - 1}{2} + 1, \ldots, i - 1 , i, i + 1, \ldots, i + \frac{k - 1}{2} \right\}
$$
for odd $k$.

\pause\medskip

This is a version of nearest-neighbor smoothing that assumes ordered and
symmetric neighborhoods.

\pause

### Simplifies Computations

We don't need to keep track of metric comparisons. Only the order matters.

\pdfpcnote{
- Symmetric neighborhoods do not take distance into account.

- We only need to sort.

- Only the order matters.
}

## Running Mean: An Illustration

```{r}
#| label: running-mean-illustration
#| echo: false
#| warning: false
#| fig-show: animate
#| fig-width: 5.5
#| fig-height: 2.8
#| fig-cap: Illustration of the running mean with $k = 3$. The orange points are
#|   the three nearest neighbors of the blue point.

running_mean_data <- tibble(
  x = seq_len(6),
  y = c(1, 7, 4, 10, 4, 7)
) |>
  mutate(pair = sprintf("(%d, %d)", x, y))

running_mean_centers <- 2:5
running_mean_fits <- tibble(
  x = running_mean_centers,
  y = map_dbl(
    running_mean_centers,
    \(i) mean(running_mean_data$y[(i - 1):(i + 1)])
  )
)

running_mean_frame <- function(step) {
  center <- running_mean_centers[step]
  neighbors <- (center - 1):(center + 1)
  current_y <- running_mean_fits$y[step]
  window_values <- running_mean_data$y[neighbors]
  displayed_fits <- running_mean_fits[seq_len(step), ]
  equation <- sprintf(
    paste0(
      "paste(hat(f)[%d] == frac(y[%d] + y[%d] + y[%d], 3), ",
      "' = ', frac(%d + %d + %d, 3), ' = ', %d)"
    ),
    center,
    neighbors[1],
    neighbors[2],
    neighbors[3],
    window_values[1],
    window_values[2],
    window_values[3],
    current_y
  )

  running_mean_data |>
    mutate(in_window = x %in% neighbors) |>
    ggplot(aes(x, y)) +
    geom_line(color = "grey80") +
    geom_point(aes(color = in_window), size = 3) +
    geom_text(
      aes(label = pair, color = in_window),
      nudge_y = 0.8,
      show.legend = FALSE
    ) +
    geom_line(
      aes(x, y),
      data = displayed_fits,
      color = "royalblue"
    ) +
    geom_point(
      aes(x, y),
      data = displayed_fits,
      color = "royalblue",
      size = 2.5
    ) +
    annotate(
      "point",
      x = center,
      y = current_y,
      color = "royalblue",
      size = 4
    ) +
    annotate(
      "segment",
      x = neighbors[1],
      xend = neighbors[3],
      y = 0,
      yend = 0,
      color = "darkorange"
    ) +
    annotate(
      "segment",
      x = c(neighbors[1], neighbors[3]),
      xend = c(neighbors[1], neighbors[3]),
      y = 0,
      yend = 0.35,
      color = "darkorange"
    ) +
    annotate(
      "text",
      x = center,
      y = -0.55,
      label = sprintf("N[%d]", center),
      parse = TRUE,
      color = "darkorange"
    ) +
    annotate(
      "text",
      x = 2,
      y = 10,
      label = equation,
      parse = TRUE,
      color = "royalblue"
    ) +
    scale_color_manual(
      values = c(`FALSE` = "grey60", `TRUE` = "darkorange"),
      guide = "none"
    ) +
    scale_x_continuous(breaks = running_mean_data$x) +
    coord_cartesian(
      xlim = c(0.65, 6.35),
      ylim = c(-1.3, 12.7),
      expand = FALSE
    ) +
    labs(x = expression(x[i]), y = expression(y[i]))
}

for (step in seq_along(running_mean_centers)) {
  print(running_mean_frame(step))
}
```

\pdfpcnote{
  Each click shifts the three-point neighborhood one position to the right.

  Only the first and last observations in the neighborhood change.

  This sets up the efficient update on the following slides.
}

## Running Mean (Naive Implementation)

Assume $y$ is sorted and $k$ is odd.

```{r}
#| label: runmean-simple

run_mean_naive <- function(y, k) {
  N <- length(y)
  m <- (k - 1) / 2
  f_hat <- rep(NA, N)

  for (i in (m + 1):(N - m)) {
    f_hat[i] <- mean(y[(i - m):(i + m)])
  }

  f_hat
}
```

\pdfpcnote{
- We can speed this up by replacing the mean computation.
}

## Efficient Running Mean

We can achieve a more efficient implementation by noting that
$$
  \hat{f}_{i+1} = \hat{f}_{i} - \frac{y_{i-(k-1)/2}}{k}
                  + \frac{y_{i+(k+1)/2}}{k}.
$$

. . .

```{r}
#| label: run_mean_def

run_mean <- function(y, k) {
  N <- length(y)
  m <- floor((k - 1) / 2)
  k <- 2 * m + 1 # Ensures k is odd and m = (k - 1) / 2
  y <- y / k
  f_hat <- rep(NA, N)
  f_hat[m + 1] <- sum(y[1:k])

  for (i in (m + 1):(N - m - 1)) {
    f_hat[i + 1] <- f_hat[i] - y[i - m] + y[i + 1 + m]
  }
  f_hat
}
```

\pdfpcnote{
  - A sliding window of size k is used to compute the running mean.\\
  - This ensures that k is odd.\\
  - The complexity decreases from O(Nk) to O(N).
}

## Visualization

```{r}
#| label: nuuk-NN-plot2
#| warning: false
#| fig-width: 5
#| fig-height: 2.2
#| fig-cap: Running mean smoother with $k = 11$.

f_hat <- run_mean(nuuk_year$Temperature, 11)
nuuk_plot + geom_line(aes(y = f_hat), color = "blue")
```

\pdfpcnote{
  This takes five observations to the right and five to the left.
}

## `stats::filter()`

`stats::filter()` applies linear filtering (moving averages or autoregression).

. . .

```{r, echo=2}
#| label: run-mean-nas

op <- options(digits = 2)
f_hat_filter <- stats::filter(
  nuuk_year$Temperature,
  rep(1 / 11, 11)
)
f_hat_filter[c(1:10, 137:147)]
options(digits = op$digits)
```

. . .

```{r}
#| label: runmean-test

f_hat <- run_mean(nuuk_year$Temperature, 11)

all.equal(f_hat, as.numeric(f_hat_filter))
```

\pdfpcnote{
  Warn about the name clash with dplyr::filter().

  `rep(1 / 11, 11)` contains the moving-average coefficients.
}

##

```{r}
#| label: runMeanBench
#| echo: false
#| warning: false
#| message: false
#| fig-width: 5.5
#| fig-height: 3

nn_bench <- bench::press(N = c(512, 1024, 2048, 4096), {
  k <- 11
  m <- (k - 1) / 2
  y <- rnorm(N)
  S <- matrix(0, N - 2 * m, N)
  for (i in seq_len(nrow(S))) {
    S[i, i:(i + k - 1)] <- 1 / k
  }
  S_sparse <- Matrix::sparseMatrix(
    i = rep(seq_len(nrow(S)), each = k),
    j = rep(seq_len(nrow(S)), each = k) +
      rep(0:(k - 1), times = nrow(S)),
    x = 1 / k,
    dims = dim(S)
  )
  bench::mark(
    dense = S %*% y,
    sparse = S_sparse %*% y,
    running = run_mean(y, k = k),
    filter = stats::filter(y, rep(1 / k, k)),
    check = FALSE
  )
})

plot(nn_bench)
```

Dense matrix-vector multiplication is $O(N^2)$.

\medskip\pause

For fixed $k$, sparse multiplication and the other two algorithms are $O(N)$.

\pdfpcnote{
  Both matrix representations are constructed before timing. The benchmark
  measures applying an already constructed smoother to a vector.

  The matrix products return only the interior fitted values; the other methods
  also return missing values at the boundaries.
}

## Three Ways to Compute a Running Mean

| Method | Storage for $\symbf{S}$ | Work for all fitted values |
|:--|:--|:--|
| Dense matrix | $O(N^2)$ | $O(N^2)$ |
| Sparse matrix | $O(Nk)$ | $O(Nk)$ |
| Sliding-window update | No matrix | $O(N)$ |

\pause\medskip

Sparse multiplication skips the zero weights.

\pause\medskip

The sliding-window update also reuses the overlap between neighborhoods.

\pdfpcnote{
  These costs assume the observations are already ordered and k is the window
  size. Matrix construction and finding neighborhoods are not included.

  All methods still need storage for the data and fitted values. The table
  compares the additional storage used to represent the smoother matrix.

  With k fixed, both sparse multiplication and the sliding-window update are
  linear in N. Varying k reveals their different dependence on window size.
}

## Bias--Variance Trade-Off

There is a bias-variance trade-off in nearest-neighbor smoothing with respect to
$k$.

\medskip\pause

If the data are i.i.d. with $\V(Y \mid X) = \sigma^2$ and
$f(x) = \E(Y \mid X = x)$,

\pause

$$
  \begin{aligned}
    \mathrm{MSE}(x) & = \E(f(x) - \hat{f}(x))^2                                                                                                                              \\
    \pause
                    & = \V\big(f(x) - \hat{f}(x)\big) + \big(\E(f(x) - \hat{f}(x))\big)^2                                                                                    \\
    \pause
                    & = \underbrace{\frac{\sigma^2}{k}}_{\text{variance}} + \bigg( \underbrace{f(x) - \frac{1}{k} \sum_{l \in \mathcal{N}(x)} f(x_l)}_{\text{bias}}\bigg)^2.
  \end{aligned}
$$

. . .

### Bias-Variance Trade-Off

- Small $k$ gives large variance and small bias (if $f$ is smooth).\pause
- Large $k$ gives small variance and potentially large bias (if $f$ is not
  constant).

## How to Choose $k$?

We know there is a bias-variance trade-off in $k$---so how do we choose $k$?

\pause

### Leave-One-Out Cross-Validation (LOOCV)

Leave out one observation at a time, predict using the remaining data, and
measure the prediction error.

\pause\medskip

::: {.columns}
::::: {.column width="47%"}
The general LOOCV formula is
$$
  \mathrm{LOOCV} = \frac{1}{N} \sum_{i=1}^N (y_i - \hat{f}^{-i}_i)^2.
$$

\pause

Keeping the neighborhoods fixed, we can express the leave-one-out fitted values
as
$$
  \hat{f}^{-i}_i = \sum_{j \neq i} \frac{S_{ij}y_j}{1 - S_{ii}},
$$
:::::

. . .

::::: {.column width="47%"}
This leads to the simplified formula
$$
  \mathrm{LOOCV}
  = \frac{1}{N} \sum_{i=1}^N \left( \frac{y_i - \hat{f}_i}{1 - S_{ii}} \right)^2.
$$
:::::
:::

\pdfpcnote{
For the running mean, we can, in principle, compute this without leaving out any
observations.
}

## LOOCV for Nearest Neighbors

For nearest neighbors, $S_{ii} = \frac{1}{k}$ for all $i$, so
$$
  \mathrm{LOOCV}
  = \frac{1}{N} \sum_{i=1}^N \left( \frac{y_i - \hat{f}_i}{1 - 1/k} \right)^2.
$$

. . .

### Implementation

```{r}
#| label: LOOCV-runMean

loocv <- function(k, y) {
  f_hat <- run_mean(y, k)
  mean(((y - f_hat) / (1 - 1 / k))^2, na.rm = TRUE)
}
```

. . .

### Note {.alert}

The implementation ignores the missing fitted values introduced by our boundary
handling. It uses `mean()` rather than `sum()` so that the criterion is averaged
over the remaining observations.

\pdfpcnote{
- For nearest neighbors, $S_{ii} = \frac{1}{k}$ for all $i$.

- **Derive the formula on the blackboard!**
}

## LOOCV for Nearest Neighbors

Define
$$
  \hat{f}_i^{-i}
  = \sum_{j \neq i} \frac{S_{ij}y_j}{1 - S_{ii}}, \qquad \hat{f}_i
                 = \frac{1}{k} \sum_{j \in \mathcal{N}_i} y_j.
$$

Then
$$
  \begin{aligned}
    \mathrm{LOOCV}
     & = \frac{1}{N} \sum_{i=1}^N (y_i - \hat{f}^{-i}_i)^2                            \\
     & = \frac{1}{N} \sum_{i=1}^N \left( \frac{y_i - \hat{f}_i}{1 - S_{ii}} \right)^2 \\
     & = \frac{1}{N} \sum_{i=1}^N \left( \frac{y_i - \hat{f}_i}{1 - 1/k} \right)^2.
  \end{aligned}
$$

##

```{r}
#| label: nuuk-running-loocv

k <- seq(3, 40, 2)
cv_error <- sapply(
  k,
  function(kk) loocv(kk, nuuk_year$Temperature)
)
k_opt <- k[which.min(cv_error)]
```

\medskip\pause

```{r}
#| label: nuuk-running-loocv-plot
#| fig-width: 5
#| fig-height: 2
#| echo: false
#| fig-cap: LOOCV error for nearest-neighbor smoothing as a function of $k$.

ggplot(tibble(k, cv_error), aes(k, cv_error)) +
  geom_vline(xintercept = k_opt, color = "dark orange") +
  geom_line() +
  geom_point() +
  labs(y = "LOOCV Error", x = expression(k))
```

##

```{r}
#| label: nuuk-NN-plot3
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 2.5
#| fig-cap: Nearest-neighbor smoother with LOOCV-optimal $k=15$.

nuuk_plot +
  geom_line(
    aes(y = run_mean(nuuk_year$Temperature, k_opt)),
    color = "blue"
  )
```

## Wiggliness

The running mean is wiggly!

\pause\medskip

We can use kernel smoothing (as we did last week) to fix this.

\pause\medskip

But another idea is to use \alert{smoothing splines} instead.

# Smoothing Splines

## Smoothing Splines

We want a smooth function $f$ that balances goodness of fit and wiggliness.

\medskip\pause

To do this, we minimize an objective function that penalizes wiggliness:
$$
  L(f) = \sum_{i=1}^N (y_i - f(x_i))^2
         + \lambda \underbrace{\int f''(z)^2 \ \mathrm{d}z}_{\|f''\|_2^2}
$$
where $\lambda \geq 0$ is a tuning parameter that controls the trade-off.

\pause\medskip

The minimizer of this objective is a **natural cubic spline** with **knots** at
the data points $x_i$.

\medskip\pause

That is, it has the representation
$$
  f = \sum_j \beta_j \varphi_j,
$$
where $\{\varphi_j\}$ is a basis for the $N$-dimensional space of such splines.

\medskip\pause

The problem reduces to finding the coefficients $\symbf{\beta}$.

\pdfpcnote{
Without the penalty term, any interpolating function would work well.

The penalty term penalizes solutions whose second derivatives have large norms.
}

## Loss Function

In vector notation,
$$
  \symbf{f} = \symbf{\Phi}\symbf{\beta}
$$
where $\symbf{\Phi}_{ij} = \varphi_j(x_i)$. Then \pause
$$
  \begin{aligned}
    L(\symbf{f}) & = (\symbf{y} - \symbf{f})^T (\symbf{y} - \symbf{f}) + \lambda \|f''\|_2^2                                                                  \\
                 & = (\symbf{y} - \symbf{\Phi}\symbf{\beta})^T (\symbf{y} - \symbf{\Phi}\symbf{\beta}) + \lambda \symbf{\beta}^T \symbf{\Omega} \symbf{\beta}
  \end{aligned}
$$
with
$$
  \symbf{\Omega}_{jk} = \int \varphi_j''(z) \varphi_k''(z) \ \mathrm{d}z.
$$

\pause

### Solution

The solution is a *linear smoother* with smoother matrix $\symbf{S}_{\lambda}$
and penalty matrix $\symbf{\Omega}$.

\pdfpcnote{
  This is similar to ridge (Tikhonov) regression: OLS plus a squared $\ell_2$
  penalty.

  We can write the penalty term in the quadratic form
  $\lambda \symbf{\beta}^T \symbf{\Omega} \symbf{\beta}$.

  Derive the minimizer on the blackboard.
}

## Splines in R

```{r}
#| label: splines
#| message: false
#| warning: false

knots <- c(0, 0, 0, seq(0, 1, 0.2), 1, 1, 1)
xx <- seq(0, 1, 0.005)
b_splines <- splines::splineDesign(knots, xx)
```

. . .

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 2
#| fig-cap: Cubic B-spline basis functions with knots at 0, 0.2, 0.4, 0.6, 0.8,
#|   and 1.

df <- data.frame(
  x = rep(xx, times = ncol(b_splines)),
  y = as.vector(b_splines),
  group = as.character(rep(
    seq_len(ncol(b_splines)),
    each = length(xx)
  ))
)

knot_points <- c(0, 0.2, 0.4, 0.6, 0.8, 1)

ggplot(df, aes(x = x, y = y, group = group, color = group)) +
  scale_x_continuous(breaks = knot_points, minor_breaks = NULL) +
  geom_line() +
  labs(x = "x", y = "Basis Value") +
  guides(color = "none")
```

## Penalty Matrix $\symbf{\Omega}$

```{r}
#| label: Omega-Simpson
#| echo: false
#| fig-cap: The penalty matrix $\symbf{\Omega}$ for a cubic spline with knots at
#|   0, 0.2, ..., 0.8, 1.
#| fig-width: 5
#| fig-height: 3

library(Matrix)
pen_mat <- function(inner_knots) {
  knots <- sort(c(rep(range(inner_knots), 3), inner_knots))
  d <- diff(inner_knots) # the vector of knot differences; b - a
  g_ab <- splines::splineDesign(knots, inner_knots, derivs = 2)
  knots_mid <- inner_knots[-length(inner_knots)] + d / 2
  g_ab_mid <- splines::splineDesign(knots, knots_mid, derivs = 2)
  g_a <- g_ab[-nrow(g_ab), ]
  g_b <- g_ab[-1, ]
  (crossprod(d * g_a, g_a) +
    4 * crossprod(d * g_ab_mid, g_ab_mid) +
    crossprod(d * g_b, g_b)) /
    6
}

omega <- pen_mat(seq(0, 1, 0.1))

omega_mat <- as.matrix(omega)

n_row <- nrow(omega_mat)
n_col <- ncol(omega_mat)

omega_df <- expand.grid(
  Row = seq_len(n_row),
  Col = seq_len(n_col)
)
omega_df$Value <- as.vector(omega_mat)

ggplot(omega_df, aes(x = Col, y = Row, fill = Value)) +
  geom_tile() +
  scale_y_reverse() +
  scale_fill_gradient2() +
  coord_fixed()
```

## Sparsity in Spline Computations

B-spline basis functions have **local support**: each is nonzero only on a few
neighboring knot intervals. This makes $\symbf{\Phi}$ and $\symbf{\Omega}$ sparse.

\pause\medskip

The coefficients solve
$$
  (\symbf{\Phi}^T \symbf{\Phi} + \lambda \symbf{\Omega})
  \hat{\symbf{\beta}} = \symbf{\Phi}^T \symbf{y}.
$$
With the basis ordered by knot location, this system is **banded**: its nonzero
entries lie near the diagonal.

\pause\medskip

Solve for $\hat{\symbf{\beta}}$, then compute
$\hat{\symbf{f}} = \symbf{\Phi}\hat{\symbf{\beta}}$.
The full smoother matrix $\symbf{S}_{\lambda}$ is generally dense.

\pause\medskip

In R, `splineDesign(..., sparse = TRUE)` creates a sparse design matrix.

\pdfpcnote{
  At any x, at most four cubic B-spline basis functions are nonzero.

  Basis functions with disjoint supports also have zero entries in the penalty
  matrix. Its band structure is visible in the preceding figure.

  Sparse structure must be reflected in the storage to obtain the computational
  benefits. The following implementation uses dense storage to show the algebra.

  The inverse of a sparse matrix need not be sparse. Solving the system avoids
  explicitly constructing the inverse and the full smoother matrix.
}

## Fitting a Smoothing Spline

We implement the matrix algebra directly to compute
$\symbf{S}_{\lambda} \symbf{y}$.

```{r}
#| label: nuuk-smooth-spline

inner_knots <- nuuk_year$Year

# Note that the order does not matter.
knots <- c(rep(range(inner_knots), 3), inner_knots)

Phi <- splines::splineDesign(knots, inner_knots)
```

. . .

```{r}
#| label: nuuk-smooth-spline2

# The penalty matrix can be reused because it depends only on the knots.
omega <- pen_mat(inner_knots)
smoother <- function(lambda) {
  Phi %*%
    solve(
      crossprod(Phi) + lambda * omega, # Phi^T Phi + lambda Omega
      crossprod(Phi, nuuk_year$Temperature) # Phi^T y
    )
}
```

##

```{r}
#| label: nuuk-smoother-plot
#| fig-width: 5
#| fig-height: 2.5

nuuk_plot +
  geom_line(aes(y = smoother(10)), color = "blue") +
  geom_line(aes(y = smoother(1000)), color = "red") +
  geom_line(aes(y = smoother(100000)), color = "orange")
```

## Generalized Cross-Validation (GCV)

For smoothing splines, $S_{ii}$ varies with $i$, which makes LOOCV expensive to
compute.

\medskip\pause

### Idea

Replace $S_{ii}$ in LOOCV with $\mathrm{df} / N$, where
$\mathrm{df} = \operatorname{trace}(\symbf{S}) = \sum_{i=1}^N S_{ii}$. This
gives the **generalized** cross-validation criterion
$$
  \mathrm{GCV}
  = \frac{1}{N} \sum_{i=1}^N \left( \frac{y_i - \hat{f}_i}{1 - \mathrm{df} / N} \right)^2.
$$

. . .

### Implementation

```{r}
#| label: gcv-smooth-spline

gcv <- function(lambda, y) {
  S <- Phi %*% solve(crossprod(Phi) + lambda * omega, t(Phi))
  df <- sum(diag(S))
  mean(((y - S %*% y) / (1 - df / length(y)))^2, na.rm = TRUE)
}
```

\pdfpcnote{
  - The df approximation relies on S_ii being nearly constant.\\
  - When S_ii is constant, GCV coincides with the LOOCV formula.
}

## GCV-Optimal $\lambda$

Apply `gcv()` across a grid of $\lambda$-values and choose the $\lambda$ that
minimizes GCV.

```{r}
#| label: nuuk-spline-gcv

lambda <- seq(50, 250, 2)
gcv <- sapply(lambda, gcv, y = nuuk_year$Temperature)
lambda_opt <- lambda[which.min(gcv)]
```

```{r}
#| label: gc-plot
#| echo: false
#| fig-width: 5
#| fig-height: 2

tibble(lambda, gcv) |>
  ggplot(aes(lambda, gcv)) +
  geom_vline(xintercept = lambda_opt, color = "dark orange") +
  geom_line() +
  labs(y = "GCV Error", x = expression(lambda))
```

## GCV-Optimal Smoothing Spline

```{r}
#| label: nuuk-spline-opt
#| fig-width: 5
#| fig-height: 2.5

smooth_opt <- smoother(lambda_opt)
nuuk_plot + geom_line(aes(y = smooth_opt), color = "blue")
```

## Using `smooth.spline()`

```{r}
#| label: nuuk-spline-opt2
#| fig-width: 5
#| fig-height: 2.5

smooth <- smooth.spline(nuuk_year$Year, nuuk_year$Temperature)
nuuk_plot + geom_line(aes(y = smooth$y), color = "blue")
```

\pdfpcnote{
- Disable the fast heuristic to obtain results similar to ours.
}

## Efficient Computations

When $N$ is large, computing $\symbf{S}_{\lambda} \symbf{y}$ directly is
expensive.

\medskip\pause

In practice, we therefore often use fewer basis functions than data points,
$p < N$.

\pause

### Singular Value Decomposition

Using the singular value decomposition
$$
  \Phi = \symbf{U} D \symbf{V}^T
$$
\pause we can rewrite the smoother matrix as
$$
  \symbf{S}_{\lambda}
  = \widetilde{\symbf{U}} (I + \lambda \Gamma)^{-1} \widetilde{\symbf{U}}^T
$$
where $\widetilde{\symbf{U}} = \symbf{U} \symbf{W}$ and \pause
$$
  D^{-1} \symbf{V}^T \symbf{\Omega} \symbf{V} D^{-1}
  = \symbf{W} \Gamma \symbf{W}^T.
$$

\medskip\pause

The columns of $\widetilde{\symbf{U}}$ form the **Demmler-Reinsch basis**.

\pdfpcnote{
  This can be done whether we use all or a subset of the basis functions.

  The smoother matrix S_\lambda is diagonalized.

  Derive $\symbf{S}_\lambda$ on blackboard.
}

## Interpretation

Project the data $\symbf{y}$ onto the Demmler-Reinsch basis (the columns of
$\widetilde{\symbf{U}}$) to obtain the coefficients
$\hat{\symbf{\beta}} = \widetilde{\symbf{U}}^T \symbf{y}$.

\pause\medskip

Each coefficient is shrunk according to the corresponding eigenvalue $\gamma_i$
and smoothing parameter $\lambda$:
$$
  \hat{\symbf{\beta}}_i(\lambda) = \frac{\hat{\beta}_i}{1 + \lambda \gamma_i}.
$$

\pause\medskip

The final smoothed values are reconstructed by combining the shrunk coefficients
with the basis functions:
$$
  \hat{\symbf{f}} = \widetilde{\symbf{U}} \hat{\symbf{\beta}}(\lambda).
$$

\pdfpcnote{
  The $\hat\beta_i$ are the original coefficients when $\lambda = 0$.
}

## The Demmler-Reinsch Basis (Columns of $\widetilde{\symbf{U}}$)

```{r}
#| label: spline-diagonalization
#| echo: false
#| fig-width: 5.5
#| fig-height: 3

inner_knots <- seq(1867, 2013, length.out = 18)
Phi <- splines::splineDesign(
  c(rep(range(inner_knots), 3), inner_knots),
  nuuk_year$Year
)
omega <- pen_mat(inner_knots)
Phi_svd <- svd(Phi)
omega_tilde <- t(
  t(crossprod(Phi_svd$v, omega %*% Phi_svd$v)) / Phi_svd$d
) /
  Phi_svd$d

Omega_tilde_svd <- svd(omega_tilde)
U_tilde <- Phi_svd$u %*% Omega_tilde_svd$u

colnames(U_tilde) <- paste0(rep("u", 20), 1:20)
bind_cols(select(nuuk_year, Year), as_tibble(U_tilde)) |>
  gather(
    key = "term",
    value = "value",
    -Year,
    factor_key = TRUE
  ) |>
  ggplot(aes(Year, value, color = term)) +
  geom_line() +
  facet_wrap(~term) +
  theme(legend.position = "none") +
  ylab("")
```

##

```{r}
#| label: spline-gamma
#| echo: false
#| fig-width: 5.5
#| fig-height: 3
#| fig-cap: The eigenvalues $\gamma_i$ of the penalty matrix in the
#|   Demmler-Reinsch basis.

library(patchwork)

d <- Omega_tilde_svd$d

eigen_data <- tibble(i = 1:20, Eigenvalues = d)

p1 <- ggplot(eigen_data, aes(i, Eigenvalues)) +
  geom_point() +
  labs(y = expression(gamma[i]))

p2 <- p1 + scale_y_log10()

p1 / p2 + plot_layout(axes = "collect")
```

# Exercises

## Exercises

### Exercise 1: Parallel Square

Use the `foreach` and `doParallel` packages to compute the squares of a vector
in parallel. Benchmark the parallel version against a regular `for` loop using
`bench::mark()`.

```{r}
#| echo: false
#| eval: false

library(foreach)
library(doParallel)
library(bench)

x <- rnorm(10000)

loop_square <- function(x) {
  n <- length(x)
  out <- numeric(n)

  for (i in seq_len(n)) {
    out[i] <- x[i]^2
  }

  out
}

parallel_square <- function(x) {
  out <- foreach(i = seq_along(x), .combine = c) %dopar%
    {
      x[i]^2
    }
  out
}

cl <- makeCluster(detectCores() / 2)
registerDoParallel(cl)

bench::mark(
  loop = loop_square(x),
  parallel = parallel_square(x),
  check = FALSE
)
```

. . .

### Exercise 2: Future

Create a future that waits for five seconds using `Sys.sleep(5)`. When it is
done, it should print "The future is here!" and return the value 42. Try
accessing the value of the future before and after the computation is complete.

```{r}
#| echo: false
#| eval: false

library(future)
plan(multicore) # Not on Windows or RStudio!
v %<-%
  {
    Sys.sleep(5)
    cat("The future is here!\n")
    42
  }

v # This line blocks until v is available
```

. . .

### Exercise 3: Smoothing Air Quality Data

Apply `stats::filter()` and `smooth.spline()` to the `airquality$Temp` data. Try
different window sizes for the moving average and different `spar` values for
the spline. Plot and compare the results.

## Summary

### Parallelization

- Parallelization delegates work across your CPU cores to speed up
  computations.\pause
- Use `foreach` and `%dopar%` to parallelize computations.

. . .

### Smoothing Splines
