---
title: "Monte Carlo Integration and Importance Sampling"
---

{{< include _common.qmd >}}

```{r}
#| label: setup-common2
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(patchwork)
theme_set(theme_minimal(base_size = 10))
options(width = 60)
```

```{r}
#| label: setup
#| include: false

old_options <- options(digits = 4)
```

## Today

### Monte Carlo Methods

General class of methods for estimating integrals and means.

\medskip\pause

Rejection sampling is a type of Monte Carlo method.

. . .

### Importance Sampling

Focus on integrating functions and estimating means.

\medskip\pause

Useful when we cannot analytically compute the integral.

\medskip\pause

It is also a variance reduction technique.

## Monte Carlo Methods

### Basic, Simple Idea

Use random sampling to obtain numerical results.

\medskip\pause

Useful when analytical solutions are infeasible.

. . .

### Monte Carlo Integration

With $X_1, \ldots, X_N$ i.i.d. with density $f$ and $\E|h(X_1)| < \infty$,
$$
  \hat{\mu}_{\mathrm{MC}} = \frac{1}{N} \sum_{i=1}^N h(X_i) \rightarrow \mu
                          = \E h(X_1) = \int h(x) f(x) \ \mathrm{d}x
$$
for $N \to \infty$ by the law of large numbers (LLN).

## Monte Carlo Integration

If $\sigma^2_{\mathrm{MC}} < \infty$, the central limit theorem (CLT) gives
$$
  \frac{1}{N} \sum_{i=1}^N h(X_i) \overset{\mathrm{approx}}
  \sim \mathcal{N}(\mu, \sigma^2_{\mathrm{MC}} / N)
$$
where
$$
  \sigma^2_{\mathrm{MC}} = \V h(X_1) = \int (h(x) - \mu)^2 f(x) \ \mathrm{d}x.
$$

## Monte Carlo Integration

We can estimate $\sigma^2_{\mathrm{MC}}$ using the empirical variance
$$
  \hat{\sigma}^2_{\mathrm{MC}}
  = \frac{1}{N - 1} \sum_{i=1}^N (h(X_i) - \hat{\mu}_{\mathrm{MC}})^2,
$$
then the variance of $\hat{\mu}_{\mathrm{MC}}$ is estimated as
$\hat{\sigma}^2_{\mathrm{MC}} / N$ and an approximate 95% confidence interval
for $\mu$ is
$$
  \hat{\mu}_{\mathrm{MC}} \pm 1.96 \frac{\hat{\sigma}_{\mathrm{MC}}}{\sqrt{N}}.
$$

## Example: Gamma Distribution

::: {.columns}
::::: {.column width="47%"}
We want to estimate the mean of a $\mathrm{Gamma}(8,1)$ distribution.

. . .

```{r, echo=-1}
#| label: gamma-sim

set.seed(1234)
B <- 1000
x <- rgamma(B, 8)

mu_hat <- cumsum(x) / 1:B
sigma_hat <- sd(x)
mu_hat[B] # Theoretical: 8
sigma_hat # Theoretical: sqrt(8)
```
:::::

::::: {.column width="47%"}
```{r}
#| label: gamma-fig
#| echo: false
#| fig-cap: "Density of a Gamma(8,1) distribution with mean indicated."

x <- seq(0, 20, length.out = 100)
y <- dgamma(x, 8)

tibble(x, y) |>
  ggplot(aes(x, y)) +
  geom_line() +
  geom_vline(
    xintercept = 8,
    linetype = "dashed",
    color = "steelblue"
  ) +
  annotate("text", x = 7, y = 0.02, label = "E(X)", hjust = -0.5)
```
:::::
:::

##

```{r}
#| label: gamma-fig1
#| fig-height: 2.8
#| fig-width: 5
#| echo: false
#| fig-cap: >-
#|   Convergence of the Monte Carlo estimate of the mean of a Gamma(8,1)
#|   distribution.

p <- ggplot(tibble(x = 1:B, y = mu_hat), aes(x, y)) +
  geom_hline(
    yintercept = 8,
    linetype = "dashed",
    col = "steelblue4"
  ) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(6, 10))

p +
  geom_ribbon(
    aes(
      ymin = mu_hat - 1.96 * sigma_hat / sqrt(1:B),
      ymax = mu_hat + 1.96 * sigma_hat / sqrt(1:B)
    ),
    fill = "dark orange",
    alpha = 0.4
  ) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(6, 10))
```

## When to Stop?

Concentration inequalities can give a sample size for a desired error bound.

\medskip\pause

Chebyshev requires a known variance or upper bound; Chernoff requires control of
moment-generating functions. Plugging in a variance estimate loses the
guarantee.

\medskip\pause

Stopping when an empirical confidence interval is narrow is a practical
heuristic. Ordinary fixed-sample intervals do not guarantee coverage at a
data-dependent stopping time.

. . .

### Trade-Offs

Overshooting wastes computing time; stopping too early risks an imprecise
estimate. Sequential guarantees require a suitable stopping procedure.

## Importance Sampling

When we are only interested in Monte Carlo integration, we do not need to sample
from the target distribution.

. . .

### Basic Idea

Some regions of the target distribution are often more important than others.

\medskip\pause

So let's \alert{oversample} these regions, but correct for this by
\alert{weighting} the samples.

\medskip\pause

We want to find $\mu = \E h(X)$. Observe that
$$
  \mu = \int h(x) f(x) \ \mathrm{d}x
      = \int h(x) \frac{f(x)}{g(x)} g(x) \ \mathrm{d}x
      = \int h(x) w^*(x) g(x) \ \mathrm{d}x
$$
whenever $g$ is a density satisfying
$$
  g(x) = 0 \Rightarrow f(x) = 0.
$$

## Weights

With $X_1, \ldots, X_N$ i.i.d. with density $g$, define the *weights*
$$
  w^*(X_i) = \frac{f(X_i)}{g(X_i)}.
$$

. . .

$g$ is the *importance* density and $f/g$ is called the *likelihood ratio*.

. . .

### Estimator

$$
  \hat{\mu}_{\mathrm{IS}}^* := \frac{1}{N} \sum_{i=1}^N h(X_i)w^*(X_i).
$$

It has mean $\mu$.

##

```{r}
#| label: importance-sampling-plot
#| echo: false
#| fig-cap: "Target and importance densities with corresponding importance weights."
#| fig-width: 5
#| fig-height: 3

x <- seq(-5, 5, length.out = 200)
target <- dnorm(x, mean = 0, sd = 1)
importance <- dnorm(x, mean = 1, sd = 1)

p1 <- tibble(x, target, importance) |>
  pivot_longer(
    cols = c(target, importance),
    names_to = "Density",
    values_to = "value"
  ) |>
  ggplot(aes(x, value, color = Density, linetype = Density)) +
  geom_line() +
  scale_color_manual(
    values = c(target = "black", importance = "steelblue")
  ) +
  scale_linetype_manual(
    values = c(target = "solid", importance = "dashed")
  ) +
  theme(legend.position.inside = c(0.87, 0.82)) +
  guides(
    color = guide_legend(position = "inside"),
    linetype = guide_legend(position = "inside")
  ) +
  labs(
    y = "Density",
    x = "x",
    color = "Distribution",
    linetype = "Distribution"
  )

w <- target / importance

p2 <- tibble(x, w) |>
  ggplot(aes(x, w)) +
  geom_line() +
  labs(x = "x", y = "Weight")

p1 / p2 + plot_layout(heights = c(2, 1), axes = "collect")
```

## Gamma Importance Sampling

::: {.columns}
::::: {.column width="47%"}
Use $X = 10 + 3T$, where $T \sim t_3$. This proposal has heavier tails than the
Gamma target.

```{r, echo=-1}
#| label: gamma-sim-IS

set.seed(1234)
B <- 1000
t <- rt(B, df = 3)
x <- 10 + 3 * t
g_x <- dt(t, df = 3) / 3
w_star <- dgamma(x, 8) / g_x
mu_hat_IS <- cumsum(x * w_star) / (1:B)
mu_hat_IS[B]
```

Close to the theoretical value (8).
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: gammasim-fig
#| echo: false
#| fig-height: 3.2
#| fig-width: 2.7

z <- seq(-12, 12, length.out = 100) + 10
f_x <- dgamma(z, 8)
g_x <- dt((z - 10) / 3, df = 3) / 3
w <- f_x / g_x

p1 <- tibble(x = z, target = f_x, importance = g_x) |>
  pivot_longer(c(target, importance)) |>
  ggplot(aes(x, value, color = name, linetype = name)) +
  geom_line() +
  guides(color = guide_legend(position = "inside")) +
  scale_color_manual(
    values = c(target = "black", importance = "steelblue")
  ) +
  scale_linetype_manual(
    values = c(target = "solid", importance = "dashed")
  ) +
  theme(legend.position.inside = c(0.8, 0.8)) +
  labs(y = "Density", color = "Density", linetype = "Density")

p2 <- tibble(x = z, w) |>
  ggplot(aes(x, w)) +
  geom_line()

p1 / p2 + plot_layout(axes = "collect")
```
:::::
:::

##

```{r}
#| label: gamma-IS-fig1
#| echo: false
#| fig-cap: >-
#|   Convergence of the importance sampling estimate of the mean of a Gamma(8,1)
#|   distribution using a Student-t proposal with 3 degrees of freedom, location
#|   10, and scale 3.
#| fig-width: 5
#| fig-height: 2.5

mu_hat_IS <- cumsum(x * w_star) / (1:B)
sigma_hat_IS <- sd(x * w_star)

p <- ggplot(tibble(x = 1:B, y = mu_hat_IS), aes(x, y)) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(6, 10)) +
  geom_hline(
    yintercept = 8,
    linetype = "dashed",
    col = "steelblue4"
  )
p
```

## Importance Sampling Variance

Again by the LLN, assuming $\E_f|h(X)| < \infty$,
$$
  \hat{\mu}_{\mathrm{IS}}^* \rightarrow \E h(X_1) w^*(X_1)
                            = \mu \quad \text{as } N \to \infty.\pause
$$
If $\sigma^{*2}_{\mathrm{IS}} < \infty$, the CLT gives
$$
  \hat{\mu}_{\mathrm{IS}}^* \overset{\mathrm{approx}}
  \sim \mathcal{N}(\mu, \sigma^{*2}_{\mathrm{IS}} / N)
$$
where
$$
  \sigma^{*2}_{\mathrm{IS}} = \V \big( h(X_1)w^*(X_1)\big)
                            = \int (h(x) w^*(x) - \mu)^2 g(x) \ \mathrm{d}x.
$$

## Importance Sampling Variance

The IS variance can be estimated analogously to the MC variance,
$$
  \hat{\sigma}^{*2}_{\mathrm{IS}}
  = \frac{1}{N - 1} \sum_{i=1}^N (h(X_i)w^*(X_i) - \hat{\mu}_{\mathrm{IS}}^*)^2.
$$

. . .

An approximate 95% confidence interval is
$$
  \hat{\mu}^*_{\mathrm{IS}}
  \pm 1.96 \frac{\hat{\sigma}^*_{\mathrm{IS}}}{\sqrt{N}}.
$$

\medskip\pause

We may have $\sigma^{*2}_{\mathrm{IS}} > \sigma^2_{\mathrm{MC}}$ or
$\sigma^{*2}_{\mathrm{IS}} < \sigma^2_{\mathrm{MC}}$ depending on $h$ and $g$.

. . .

### Goal

Choose $g$ so that $|h(x)| w^*(x)$ becomes as constant as possible.

## Gamma Importance Sampling

```{r}
#| label: gamma-IS-sd
#| echo: true

sigma_hat_IS <- sd(x * w_star)
sigma_hat_IS
```

. . .

```{r}
#| label: gamma-IS-fig2
#| echo: false
#| fig-width: 4.5
#| fig-height: 2
#| fig-cap: "Gamma mean estimate with approximate pointwise 95% intervals."

p +
  geom_ribbon(
    aes(
      ymin = mu_hat_IS - 1.96 * sigma_hat_IS / sqrt(1:B),
      ymax = mu_hat_IS + 1.96 * sigma_hat_IS / sqrt(1:B)
    ),
    fill = "darkorange",
    alpha = 0.4
  ) +
  geom_line() +
  geom_point(size = 0.5)
```

```{r}
#| label: test
#| echo: false
#| eval: false

mu_hat_IS <- replicate(10000, {
  x <- 10 + 3 * rt(B, df = 3)
  w_star <- dgamma(x, 8) / (dt((x - 10) / 3, df = 3) / 3)
  mean(x * w_star)
})
sd(mu_hat_IS) * sqrt(B)
```

## Nonlinear Functions

It is not always the case that the best importance distribution is close to the
target distribution.

. . .

### Example

Suppose we want to estimate
$$
  \E(X^4) = \int x^4 f(x) \ \mathrm{d}x = 3
$$
where $f(x)$ is the standard normal density.

\medskip\pause

If we use $g(x) = f(x) = \mathcal{N}(0,1)$, the weights are $w^*(x) = 1$, so the
estimator is just the sample mean of $x^4$.

\medskip\pause

But $h(x) = x^4$ is large in the tails, so sampling from a normal may not be
efficient, as most samples are near zero.

\medskip\pause

Instead, we can try a heavy-tailed importance distribution, such as Student-$t$.

##

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: is-normal-plot
#| echo: false
#| fig-height: 3
#| fig-cap: >-
#|   Densities, integrand, and weighted integrand values for the example with a
#|   normal importance distribution.

set.seed(1)
N <- 1000
x_norm <- rnorm(N)
target <- dnorm(x_norm, 0, 1)
importance_norm <- dnorm(x_norm, 0, 1)
weights_norm <- target / importance_norm # All 1
w_h <- x_norm^4 * weights_norm

x_grid <- seq(-6, 6, length.out = 200)
dens_df <- tibble(
  x = x_grid,
  target = dnorm(x_grid, 0, 1),
  integrand = x_grid^4 * dnorm(x_grid, 0, 1)
) |>
  pivot_longer(
    cols = c(target, integrand),
    names_to = "Type",
    values_to = "value"
  )

p1 <- ggplot(
  dens_df,
  aes(x, value, color = Type, linetype = Type)
) +
  geom_line() +
  labs(y = "Density / Integrand") +
  scale_color_manual(
    values = c(target = "black", integrand = "darkorange"),
    labels = c(
      target = expression(f(x) == g(x)),
      integrand = expression(h(x) * f(x))
    )
  ) +
  scale_linetype_manual(
    values = c(target = "solid", integrand = "dashed"),
    labels = c(
      target = expression(f(x) == g(x)),
      integrand = expression(h(x) * f(x))
    )
  ) +
  guides(
    color = guide_legend(position = "inside"),
    linetype = guide_legend(position = "inside")
  ) +
  labs(color = NULL, linetype = NULL) +
  theme(legend.position.inside = c(0.85, 0.85))

p1b <- ggplot(tibble(x = w_h), aes(x)) +
  geom_histogram() +
  labs(x = expression(abs(h(x)) * w^"*"))

p1 / p1b + plot_layout(heights = c(2, 1), axes = "collect")
```
:::::

::::: {.column width="47%"}
```{r}
#| label: is-t-plot
#| echo: false
#| fig-height: 3
#| fig-cap: >-
#|   Densities, integrand, and weighted integrand values for the example with
#|   Student-$t$ as the importance distribution.

dens_df_t <- tibble(
  x = x_grid,
  target = dnorm(x_grid, 0, 1),
  importance = dt(x_grid, df = 3)
) |>
  pivot_longer(
    cols = c(target, importance),
    names_to = "Density",
    values_to = "value"
  )

set.seed(1)
x_t <- rt(N, df = 3)
target_t <- dnorm(x_t, 0, 1)
importance_t <- dt(x_t, df = 3)
weights_t <- target_t / importance_t
w_h_t <- x_t^4 * weights_t

dens_df_t <- tibble(
  x = x_grid,
  target = dnorm(x_grid, 0, 1),
  importance = dt(x_grid, df = 3),
  integrand = x_grid^4 * dnorm(x_grid, 0, 1)
) |>
  pivot_longer(
    cols = c(target, importance, integrand),
    names_to = "Type",
    values_to = "value"
  )

p2 <- ggplot(
  dens_df_t,
  aes(x, value, color = Type, linetype = Type)
) +
  geom_line() +
  labs(y = "Density / Integrand") +
  scale_color_manual(
    values = c(
      target = "black",
      importance = "steelblue",
      integrand = "darkorange"
    ),
    labels = c(
      target = expression(f(x)),
      importance = expression(g(x)),
      integrand = expression(h(x) * f(x))
    )
  ) +
  guides(
    color = guide_legend(position = "inside"),
    linetype = guide_legend(position = "inside")
  ) +
  scale_linetype_manual(
    values = c(
      target = "solid",
      importance = "dashed",
      integrand = "dotdash"
    ),
    labels = c(
      target = expression(f(x)),
      importance = expression(g(x)),
      integrand = expression(h(x) * f(x))
    )
  ) +
  labs(color = NULL, linetype = NULL) +
  theme(legend.position.inside = c(0.85, 0.85))

p2b <- ggplot(tibble(x = abs(w_h_t)), aes(x)) +
  geom_histogram() +
  labs(x = expression(abs(h(x)) * w^"*"))

p2 / p2b + plot_layout(heights = c(2, 1), axes = "collect")
```
:::::
:::

## The Ideal Importance Density

For $0 < \E_f|h(X)| < \infty$, the variance-minimizing density for ordinary IS is
$$
  g_{\mathrm{opt}}(x)
  = \frac{|h(x)|f(x)}{\int |h(u)|f(u)\,\mathrm{d}u}.
$$

. . .

If $h \geq 0$, then $g_{\mathrm{opt}} = hf / \mu$. Every draw gives
$$
  \frac{h(X)f(X)}{g_{\mathrm{opt}}(X)} = \mu,
  \qquad X \sim g_{\mathrm{opt}},
$$
so the estimator has \alert{zero variance}.

. . .

But evaluating these weights requires the normalizing constant $\mu$, the
integral we want. Use the shape of $|h|f$ to guide a practical proposal.

\pdfpcnote{
Ordinary IS only requires proposal support wherever the integrand is nonzero.
For sign-changing h, the optimal proposal still minimizes variance, but the
weighted contributions generally have different signs and the variance is not
zero. This result is for ordinary IS, not self-normalized IS.
}

## A Rare Event: No Hits

For $Z \sim \mathcal{N}(0,1)$, estimate
$$
  p = P(Z > 4) \approx 3.17 \times 10^{-5}.
$$

```{r, echo=-1}
#| label: rare-event-mc

set.seed(1)
n_rare <- 10000
rare_hits <- rnorm(n_rare) > 4
mean(rare_hits)
```

. . .

This is typical: with $N = 10{,}000$,
$$
  P(\text{no exceedances}) = (1 - p)^N \approx 0.729.
$$

The empirical standard error is also zero. We have missed the event, not
established that its probability is zero.

## A Rare Event: Shift the Proposal

Draw $X_i \sim g = \mathcal{N}(4,1)$, so half the draws exceed 4. The weights are
$$
  w^*(x) = \frac{f(x)}{g(x)} = e^{8 - 4x}.
$$

```{r}
#| label: rare-event-proposal
#| echo: false
#| fig-width: 5
#| fig-height: 1.8

rare_grid <- tibble(
  x = seq(-4, 8, length.out = 601),
  target = dnorm(x),
  proposal = dnorm(x, mean = 4)
)

rare_grid |>
  pivot_longer(c(target, proposal), names_to = "density") |>
  ggplot(aes(x, value, color = density, linetype = density)) +
  geom_area(
    data = filter(rare_grid, x > 4),
    aes(x = x, y = proposal),
    inherit.aes = FALSE,
    fill = "darkorange",
    alpha = 0.2
  ) +
  geom_line() +
  geom_vline(xintercept = 4, linetype = "dotted") +
  scale_color_manual(
    values = c(target = "royalblue", proposal = "darkorange")
  ) +
  scale_linetype_manual(values = c(target = "solid", proposal = "dashed")) +
  labs(x = expression(x), y = "Density", color = NULL, linetype = NULL)
```

. . .

$$
  \hat{p}_{\mathrm{IS}}
  = \frac{1}{N} \sum_{i=1}^N \mathbf{1}\{X_i > 4\}\,e^{8 - 4X_i}.
$$

\pdfpcnote{
The ideal proposal is the standard normal conditional on exceeding 4. Its
normalizing constant is p. The shifted normal is a simple approximation that
makes exceedances common and has an evaluable, normalized density.
}

## A Rare Event: Repeated Runs

```{r}
#| label: rare-event-replicates
#| include: false

set.seed(2026)
p_rare <- pnorm(4, lower.tail = FALSE)
rare_estimates <- replicate(500, {
  z <- rnorm(n_rare)
  x <- rnorm(n_rare, mean = 4)
  c(
    MC = mean(z > 4),
    IS = mean((x > 4) * exp(8 - 4 * x))
  )
})

rare_runs <- tibble(
  run = seq_len(ncol(rare_estimates)),
  MC = rare_estimates["MC", ],
  IS = rare_estimates["IS", ]
) |>
  pivot_longer(c(MC, IS), names_to = "method", values_to = "estimate") |>
  mutate(method = factor(
    method,
    levels = c("MC", "IS"),
    labels = c("Ordinary MC", "Importance sampling")
  ))

rare_zero_percent <- 100 * mean(rare_estimates["MC", ] == 0)
rare_sd_ratio <- sd(rare_estimates["MC", ]) / sd(rare_estimates["IS", ])
```

500 independent runs per method, with $N = 10{,}000$ draws per run.

```{r}
#| label: rare-event-comparison
#| echo: false
#| fig-width: 5.2
#| fig-height: 2.3

ggplot(rare_runs, aes(run, estimate / p_rare, color = method)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_point(size = 0.6, alpha = 0.6) +
  facet_wrap(vars(method), nrow = 1) +
  scale_color_manual(values = c(
    "Ordinary MC" = "royalblue",
    "Importance sampling" = "darkorange"
  )) +
  guides(color = "none") +
  labs(x = "Run", y = expression(hat(p) / p))
```

The dashed line is the truth. MC returned zero in
`r sprintf("%.1f", rare_zero_percent)`% of runs.

. . .

IS had about `r round(rare_sd_ratio)` times smaller empirical standard deviation
at the same sample size.

## Normalized Weights

If $f = c^{-1} q$ with $c$ unknown then
$$
  c = \int q(x) \ \mathrm{d}x = \int \frac{q(x)}{g(x)} g(x) \ \mathrm{d}x.
$$

. . .

And
$$
  \mu
  = \frac{\int h(x) w^*(x) g(x) \ \mathrm{d}x}{\int w^*(x) g(x) \ \mathrm{d}x},
$$
where
$$
  w^*(x) = \frac{q(x)}{g(x)}.
$$

\pdfpcnote{
Note that g is still the normalized density.
}

## IS with Normalized Weights

An importance sampling estimate of $\mu$ is thus
$$
  \hat{\mu}_{\mathrm{IS}}
  = \frac{\sum_{i=1}^N h(X_i) w^*(X_i)}{\sum_{i=1}^N w^*(X_i)} \pause
  = \sum_{i=1}^N h(X_i) w(X_i),
$$
where $w^*(X_i) = q(X_i) / g(X_i)$ and
$$
  w(X_i) = \frac{w^*(X_i)}{\sum_{i=1}^N w^*(X_i)}
$$
are the *normalized weights*.

\medskip\pause

This works regardless of the normalizing constant $c$, and also if an
unnormalized proposal density is used in the weights.

\medskip\pause

The estimator is generally biased for finite $N$, but is consistent under the
support and integrability assumptions.

## Gamma Sampling with Normalized Weights

Ignore the normalizing constants of the Gamma target and proposal $10 + 3T$,
$T \sim t_3$. Set $w^*(x) = 0$ for $x \leq 0$, and otherwise use
$$
  w^*(x) = x^7 e^{-x} \left( 1 + \frac{(x - 10)^2}{27} \right)^2.
$$

. . .

```{r, echo=-1}
#| label: gamma-sim-ISw

set.seed(1234)
B <- 1000
x <- 10 + 3 * rt(B, df = 3)
w_star <- numeric(B)
x_pos <- x[x > 0]
w_star[x > 0] <- exp(
  2 * log1p((x_pos - 10)^2 / 27) - x_pos + 7 * log(x_pos)
)

mu_hat_IS <- cumsum(x * w_star) / cumsum(w_star)
mu_hat_IS[B] # Theoretical value 8
```

##

```{r}
#| label: gamma-ISw-fig1
#| echo: false
#| fig-width: 4.5
#| fig-height: 3
#| fig-cap: >-
#|   Convergence of the importance sampling estimate of the mean of a Gamma(8,1)
#|   distribution using normalized weights.

p <- tibble(x = 1:B, y = mu_hat_IS) |>
  ggplot(aes(x, y)) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(6, 10)) +
  labs(y = expression(hat(mu)[IS]), x = "Number of samples")
p
```

## Variance for IS with Normalized Weights

Let $w_i = w(X_i)$ be the normalized weights. If $w^*(X)$ and $h(X)w^*(X)$ have
finite variances, the $\Delta$-method gives the standard error estimate
$$
  \widehat{\mathrm{se}}(\hat{\mu}_{\mathrm{IS}})
  = \sqrt{\frac{N}{N - 1} \sum_{i=1}^N w_i^2 \big(h(X_i) - \hat{\mu}_{\mathrm{IS}}\big)^2}.
$$

This accounts for uncertainty in both the weighted sum and the sum of weights.

. . .

An approximate 95% confidence interval is
$$
  \hat{\mu}_{\mathrm{IS}}
  \pm 1.96\,\widehat{\mathrm{se}}(\hat{\mu}_{\mathrm{IS}}).
$$

\medskip\pause

Normalizing constants cancel. The derivation is in the appendix.

\pdfpcnote{
The factor N/(N-1) matches the sample-variance convention used earlier.
Equivalently, compute the sample standard deviation of
w-star times (h minus mu-hat), divide by the mean unnormalized weight,
and then divide by the square root of N.
}

## Gamma Normalized Weights: Standard Error

```{r}
#| label: sigma_hat_is_w

w <- w_star / sum(w_star)
weighted_residuals <- w * (x - mu_hat_IS[B])
se_hat_IS <- sqrt(B / (B - 1) * sum(weighted_residuals^2))
se_hat_IS
```

. . .

```{r}
#| label: gamma-ISw-fig2
#| echo: false
#| fig-width: 4.5
#| fig-height: 1.75
#| fig-cap: "Gamma mean estimate with approximate pointwise 95% intervals."

sigma_hat_IS_w <- se_hat_IS * sqrt(B)

p +
  geom_ribbon(
    aes(
      ymin = mu_hat_IS - 1.96 * sigma_hat_IS_w / sqrt(1:B),
      ymax = mu_hat_IS + 1.96 * sigma_hat_IS_w / sqrt(1:B)
    ),
    fill = "darkorange",
    alpha = 0.4
  ) +
  geom_line() +
  geom_point(size = 0.5)
```

## When to Use Importance Sampling?

### Rare Events

Estimating probabilities or expectations involving rare events.

. . .

### High-Variance Monte Carlo

Importance sampling can reduce variance by focusing samples where the integrand
is large.

. . .

### Difficult-to-Sample Targets

When direct sampling from the target distribution is hard, but sampling from a
related importance distribution is easy.

## Poor Importance Distribution

If the importance distribution assigns little mass to important regions,
weighted contributions can be highly variable and estimates unreliable.

\medskip\pause

### Remedies

More samples may help with poor overlap. They cannot repair missing support: if
$g = 0$ where $hf \neq 0$, ordinary IS misses part of the integral.

\medskip\pause

Redesign the importance density to cover the required support and put more mass
where the integrand is large. Self-normalized IS must cover the target support,
including where $h = 0$.

## Diagnostics for Importance Sampling

### Weight Distribution

Plot the weights and weighted integrand values to identify extreme
contributions.

. . .

### Effective Sample Size (ESS)

ESS summarizes weight concentration. A high ESS cannot rule out missed regions.

. . .

### Coefficient of Variation (CV)

CV quantifies the relative variability of the weights.

. . .

### Comparison with Standard Monte Carlo

Comparing with standard Monte Carlo shows whether importance sampling actually
improves efficiency.

## Weight Degeneracies

::: {.columns}
::::: {.column width="47%"}
Suppose that
$$
  X \sim \mathcal{N}(0, I_{2})
$$
and that we want to estimate $\E(h(X))$ where $h(X)$ is large near the origin.

\medskip\pause

If we use an importance density $g(x)$ from
$$
  \mathcal{N}(\mu, I_{2})
$$
with large $\|\mu\|$, most samples are far from the origin, so $h(X)$ is nearly
zero and weights are highly variable.
:::::

::::: {.column width="47%"}
```{r}
#| label: poor-importance-demo
#| echo: false
#| fig-height: 2.3
#| fig-cap: >-
#|   Target (blue) and importance (orange) densities with samples from the
#|   importance distribution.

library(mvtnorm)

set.seed(1)
d <- 2
N <- 1000
mu <- rep(2, d)
x <- matrix(rnorm(N * d, mean = mu), ncol = d)

colnames(x) <- c("x1", "x2")

grid <- expand.grid(
  x1 = seq(-5, 8, length.out = 100),
  x2 = seq(-5, 8, length.out = 100)
)
grid_mat <- as.matrix(grid)

grid$target <- dmvnorm(grid_mat, mean = rep(0, d))
grid$importance <- dmvnorm(grid_mat, mean = mu)

tibble(
  x1 = grid$x1,
  x2 = grid$x2,
  target = grid$target,
  importance = grid$importance
) |>
  ggplot(aes(x1, x2)) +
  geom_contour(
    aes(z = target),
    color = "steelblue",
    linetype = "dashed"
  ) +
  geom_contour(aes(z = importance), color = "darkorange") +
  geom_point(
    aes(x1, x2),
    data = as_tibble(x),
    alpha = 0.3,
    size = 0.5
  ) +
  labs(x = expression(X[1]), y = expression(X[2])) +
  coord_fixed()
```
:::::
:::

##

```{r}
#| label: weight-degeneracy-plot
#| fig-width: 5.3
#| fig-height: 3.2
#| echo: false
#| fig-cap: "Histogram and boxplot of importance weights, showing weight degeneracy."

target_density <- function(x) dmvnorm(x, mean = rep(0, d))
importance_density <- function(x) dmvnorm(x, mean = mu)
h <- function(x) exp(-sum(x^2) / 2)

w_star <- apply(
  x,
  1,
  function(row) target_density(row) / importance_density(row)
)
values <- apply(x, 1, h)
estimate <- sum(values * w_star) / sum(w_star)

df <- tibble(weights = w_star)

p1 <- ggplot(df, aes(weights)) +
  geom_histogram() +
  labs(x = "Importance Weights", y = "Frequency")

p2 <- ggplot(df, aes(weights)) +
  geom_boxplot(alpha = 0.5) +
  scale_y_discrete() +
  labs(x = "Importance Weights")

p1 + p2 + plot_layout(heights = c(2, 1), axes = "collect")
```

\pdfpcnote{
This indicates weight degeneracy and poor sampling efficiency.
}

## Effective Sample Size (ESS)

The weight-based effective sample size (ESS) summarizes how evenly the total
weight is distributed across samples:
$$
  \mathrm{ESS}
  = \frac{1}{\sum_{i=1}^N w_i^2}
  = \frac{\left( \sum_{i=1}^N w^*_i \right)^2}{\sum_{i=1}^N ({w^*_i})^2}
$$
where $w_i$ are the \alert{normalized} importance weights.

\medskip\pause

If all weights are equal, $\mathrm{ESS} = N$. If one weight dominates,
$\mathrm{ESS} \approx 1$, which warns of possible instability.

. . .

### Bivariate Normal Example

```{r}
#| label: ess-demo

ess <- (sum(w_star))^2 / sum(w_star^2)
ess # Out of N = 1000 samples.
```

## Coefficient of Variation (CV)

The coefficient of variation (CV) of the weights measures their relative
variability.

\medskip\pause

It is defined as
$$
  \mathrm{CV}
  = \frac{\sqrt{\frac{1}{N} \sum_{i=1}^N (w_i - \bar{w})^2}}{\bar{w}}
$$
where $\bar{w} = \frac{1}{N} \sum_{i=1}^N w_i$ is the mean weight.

. . .

### Interpretation

A high CV indicates uneven weights, which can make estimates unstable.

\medskip\pause

CV near 0 means nearly uniform weights. Even CV below 1 ($\mathrm{ESS} > N/2$)
does not guarantee accuracy: the integrand matters, and samples can miss
important regions.

. . .

### Connection to ESS

There is a direct relationship between ESS and CV:
$$
  \mathrm{ESS} = \frac{N}{1 + \mathrm{CV}^2}.
$$

## Choosing the Importance Distribution

We want low-variance estimates of
$$
  I = \int h(x) f(x) \ \mathrm{d}x
$$

. . .

The importance distribution determines the weights
$$
  w^*(x) = \frac{f(x)}{g(x)}
$$
but ordinary IS variance depends on $h(X)w^*(X)$, not just the weights. For
self-normalized IS, the asymptotic variance depends on $(h(X) - \mu)w^*(X)$.

## Principles for Choosing the Importance Distribution

### Cover the Support

Require $g(x) > 0$ wherever $|h(x)|f(x) > 0$ for ordinary IS, and wherever
$f(x) > 0$ for self-normalized IS, up to sets of measure zero.

. . .

### Match the Bulk

For ordinary IS, put more mass where $|h(x)|f(x)$ is large. For self-normalized
IS, the corresponding quantity is $|h(x) - \mu|f(x)$.

. . .

### Give it Heavy-Enough Tails

Use tails heavy enough that the weighted contributions have finite variance.
Heavier tails than the target are often a useful safeguard.

## Common Strategies

### Laplace Approximation

Use a Gaussian approximation centered at the mode of the target distribution.

. . .

### Adaptive Importance Sampling

Iteratively refine the importance distribution based on pilot runs.

. . .

### Defensive Mixtures

Add a component that covers the target support and controls the tails of the
weighted contributions.

. . .

### Mixture Distributions

Use a mixture of distributions to capture multiple modes or complex shapes.

## Laplace Approximation

### Basic Idea

Take a Gaussian approximation of the target density at its mode $x^*$ as the
importance distribution.

. . .

### Motivation

The connection comes from the second-order Taylor expansion of $\log f(x)$ at an
interior mode $x^*$,
$$
  \log f(x) \approx \log f(x^*) - \frac{1}{2}(x - x^*)^T \Lambda (x - x^*)
$$
where $\Lambda$ is the Hessian of $-\log f(x)$ at $x^*$, assumed positive
definite.

\medskip\pause

Exponentiating both sides gives:
$$
  f(x)
  \approx f(x^*) \exp\left( -\frac{1}{2}(x - x^*)^T \Lambda (x - x^*) \right)
$$

## Laplace Approximation

### Steps

1. Find the mode $x^* = \argmax_x f(x)$.\pause
2. Compute the Hessian $\Lambda$ of $-\log f(x)$ at $x^*$.\pause
3. If $\Lambda$ is positive definite, use $\mathcal{N}(x^*, \Lambda^{-1})$ as
   the importance distribution.

. . .

### When to Use

Useful for smooth targets with an approximately Gaussian shape near the mode.
Check the tails before using it for importance sampling.

## Example: Gamma Distribution

::: {.columns}
::::: {.column width="47%"}
The density of the $\mathrm{Gamma}(3,1)$ distribution is
$$
  f(x) = \frac{1}{2} x^2 e^{-x}, \quad x > 0.
$$

. . .

The mode is at $x^* = 2$ and the Hessian of $-\log f(x)$ at $x^*$ is
$$
  \Lambda = \frac{\alpha - 1}{(x^*)^2} = \frac 1 2.
$$

\medskip\pause

So the Laplace approximation is $\mathcal{N}(2, 2)$, with standard deviation
$\sqrt{2}$.
:::::

::::: {.column width="47%"}
```{r}
#| label: laplace-approx-plot
#| echo: false
#| fig-cap: >-
#|   Target (solid black) and Laplace approximation (dashed blue) densities for
#|   a Gamma(3,1) distribution.

f_density <- function(x) dgamma(x, shape = 3, rate = 1) # Gamma(3,1)
log_f <- function(x) log(f_density(x))

x_star <- (3 - 1) / 1
lambda <- 2 / (x_star^2)
sd_laplace <- 1 / sqrt(lambda)

x_grid <- seq(-2, 10, length.out = 400)

df <- tibble(
  x = x_grid,
  f = f_density(x_grid),
  laplace = dnorm(x_grid, mean = x_star, sd = sd_laplace),
) |>
  pivot_longer(
    cols = c(f, laplace),
    names_to = "Density",
    values_to = "value"
  ) |>
  filter(value > 0)

# Plot
ggplot(
  df,
  aes(x = x, y = value, color = Density, linetype = Density)
) +
  geom_line() +
  scale_color_manual(
    values = c(f = "black", laplace = "steelblue")
  ) +
  scale_linetype_manual(
    values = c(f = "solid", laplace = "dashed")
  ) +
  theme(legend.position.inside = c(0.8, 0.8)) +
  guides(
    color = guide_legend(position = "inside"),
    linetype = guide_legend(position = "inside")
  ) +
  labs(y = "Density")
```
:::::
:::

## Adaptive Importance Sampling

Adaptive importance sampling iteratively refines the importance distribution
based on the samples obtained.

. . .

### Basic Idea

Start with an initial importance distribution and refine it based on samples and
weights.

. . .

### Steps

1. Choose an initial importance distribution $g_0(x)$ and set $n = 0$.\pause
2. Draw samples from the current proposal $g_n$ and compute weights.\pause
3. Fit $g_{n+1}$ using the weighted samples, then increment $n$. \pause
4. Repeat steps 2-3 until the proposal stabilizes.

## Moment-Matching Importance Sampling

One simple adaptive strategy is to use moment-matching to update a Gaussian
importance distribution.

\medskip\pause

At iteration $n$, draw samples $x_1, \ldots, x_N \sim g_n(x)$ and compute
weights
$$
  w_i^* = \frac{f(x_i)}{g_n(x_i)}.
$$

. . .

Then update the importance distribution parameters as $$
\begin{aligned}
  \mu_{n + 1}      &= \frac{\sum_{i=1}^N w_i^* x_i}{\sum_{i=1}^N w_i^*},\\
  \sigma^2_{n + 1} &= \frac{\sum_{i=1}^N w_i^* (x_i - \mu_{n + 1})^2}{\sum_{i=1}^N w_i^*}.
\end{aligned}
$$

\pdfpcnote{
This can be extended to multivariate targets using weighted covariance.
}

##

```{r}
#| echo: false
#| label: adaptive-importance-demo

library(ggplot2)
library(tibble)

# Target density: bimodal mixture
f_density <- function(x) {
  0.4 * dnorm(x, -2, 1) + 0.6 * dnorm(x, 3, 0.5)
}

# Adaptive importance sampling function
adaptive_is <- function(
  N = 1000,
  n_iter = 5,
  q_init_mean = 0,
  q_init_sd = 1
) {
  # Store proposal params and samples
  proposals <- list()
  samples <- list()
  weights <- list()

  mean_q <- q_init_mean
  sd_q <- q_init_sd

  for (i in 1:n_iter) {
    # Draw from current proposal
    x <- rnorm(N, mean = mean_q, sd = sd_q)
    w <- f_density(x) / dnorm(x, mean = mean_q, sd = sd_q)
    w_norm <- w / sum(w)

    # Store current step
    proposals[[i]] <- c(mean = mean_q, sd = sd_q)
    samples[[i]] <- x
    weights[[i]] <- w_norm

    # Update proposal using weighted moments
    mean_q <- sum(w_norm * x)
    sd_q <- sqrt(sum(w_norm * (x - mean_q)^2))
  }

  return(list(
    proposals = proposals,
    samples = samples,
    weights = weights
  ))
}

# Run adaptive importance sampling
set.seed(123)
res <- adaptive_is(N = 1000, n_iter = 5)

# Plotting function for a given iteration
plot_step <- function(iter) {
  proposal <- res$proposals[[iter]]
  x_grid <- seq(-6, 6, length.out = 400)
  df <- tibble(
    x = x_grid,
    target = f_density(x_grid),
    importance = dnorm(
      x_grid,
      mean = proposal["mean"],
      sd = proposal["sd"]
    )
  ) |>
    pivot_longer(
      cols = c(target, importance),
      names_to = "Density",
      values_to = "value"
    )

  # Calculate ESS and CV for current iteration
  w <- res$weights[[iter]]
  N <- length(w)
  ess <- (sum(w))^2 / sum(w^2)
  cv <- sqrt(mean((w - mean(w))^2)) / mean(w)

  ggplot(
    df,
    aes(x = x, y = value, color = Density, linetype = Density)
  ) +
    geom_line() +
    scale_color_manual(
      values = c(target = "black", importance = "steelblue")
    ) +
    scale_linetype_manual(
      values = c(target = "solid", importance = "dashed")
    ) +
    lims(x = c(-6, 6), y = c(0, 0.6)) +
    labs(
      y = "Density",
      subtitle = sprintf(
        "Iteration %d: ESS = %.0f, CV = %.1f",
        iter,
        ess,
        cv
      )
    )
}
```

```{r}
#| label: adaptive-importance-fig
#| echo: false
#| fig-show: animate
#| fig-width: 5.4
#| fig-height: 2.5
#| fig-cap: "Adaptive importance sampling with a Gaussian proposal."

for (iter in seq_along(res$proposals)) {
  print(plot_step(iter))
}
```

## Defensive Importance Sampling

Defensive importance sampling adds a component that covers the target support
and controls extreme weighted contributions.

\medskip\pause

For the Gamma(8,1) target, let $g_s$ be the density of $10 + sT$, where
$T \sim t_3$. Our earlier proposal was $g_3$.

\medskip\pause

Instead we can use a mixture
$$
  g_{\mathrm{mix}}(x) = 0.8 g_3(x) + 0.2 g_{10}(x).
$$
Both components have tails heavy enough for finite variance in this example; the
broader component increases sampling away from the bulk.

##

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: gamma-importance-fig
#| echo: false
#| fig-cap: >-
#|   Target (black) and importance (blue) densities, with importance weights
#|   below.
#| ref-label: gammasim-fig
#| fig-height: 3
#| fig-width: 2.7
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: gamma-defensive-fig
#| fig-height: 3
#| fig-width: 2.7
#| fig-cap: >-
#|   Target (black) and defensive mixture (orange) densities, with importance
#|   weights below.
#| echo: false

set.seed(1234)
B <- 1000
mix_indicator <- rbinom(B, 1, 0.8)
x <- 10 + 3 * rt(B, df = 3)
x[mix_indicator == 0] <- 10 +
  10 * rt(sum(mix_indicator == 0), df = 3)
g_mix <- 0.8 * dt((x - 10) / 3, df = 3) / 3 +
  0.2 * dt((x - 10) / 10, df = 3) / 10
w_star <- dgamma(x, 8) / g_mix

z <- seq(-12, 12, length.out = 100) + 10
f_x <- dgamma(z, 8)
g_mix <- 0.8 * dt((z - 10) / 3, df = 3) / 3 +
  0.2 * dt((z - 10) / 10, df = 3) / 10
w <- f_x / g_mix

p1 <- tibble(x = z, target = f_x, mixture = g_mix) |>
  pivot_longer(c(target, mixture)) |>
  ggplot(aes(x, value, color = name, linetype = name)) +
  geom_line() +
  guides(color = guide_legend(position = "inside")) +
  scale_color_manual(
    values = c(target = "black", mixture = "darkorange")
  ) +
  scale_linetype_manual(
    values = c(target = "solid", mixture = "dashed")
  ) +
  theme(legend.position.inside = c(0.8, 0.8)) +
  labs(y = "Density", color = "Density", linetype = "Density")

p2 <- tibble(x = z, w) |>
  ggplot(aes(x, w)) +
  geom_line()

p1 / p2 + plot_layout(axes = "collect")
```
:::::
:::

## Exercise: Von Mises Importance Sampling

### Step 1

Estimate the arithmetic mean of the von Mises distribution on $[-\pi,\pi]$ with
$\mu = 0$ and $\kappa = 2$. Plot convergence toward $\E X = 0$.

```{r}
#| label: vonmises-density

dvmises <- function(x, mu = 0, kappa = 2) {
  ifelse(
    abs(x) <= pi,
    exp(kappa * cos(x - mu)) / (2 * pi * besselI(kappa, 0)),
    0
  )
}
```

\pdfpcnote{
For nonzero mu, the arithmetic mean on this fixed interval generally differs
from mu, the circular mean direction. Keep mu = 0 for this exercise.
}

## Exercise: Von Mises Importance Sampling

### Step 2

Compute the estimate of the mean and its standard error.

. . .

### Step 3

Compute the effective sample size and coefficient of variation of the importance
weights. Experiment with different importance densities to see if you find a
better one.

```{r}
#| label: vmises-importance
#| include: false

vonmises_importance <- function(N) {
  x <- rnorm(N, 0, pi / 2)
  w <- dvmises(x) / dnorm(x, 0, pi / 2)
  cumsum(x * w) / 1:N
}

x <- seq(-pi, pi, length.out = 100)
plot(x, dvmises(x))
lines(x, dnorm(x, 0, pi / 2))
```

## Summary

### Monte Carlo Methods

Estimate integrals and means using random sampling when analytic solutions are
not possible.

. . .

### Importance Sampling

Focus samples where they matter most, but choose your proposal carefully to
avoid large or variable weights.

. . .

### Diagnostics

Examine weights, weighted integrand values, and ESS to detect instability. These
checks cannot guarantee accuracy or reveal every missed region.

. . .

### Picking Importance Distributions

It's hard! But use adaptive and defensive methods to help fix poor coverage and
improve robustness.

## Next Time

### Optimization

We introduce optimization methods for minimizing and maximizing functions.

\medskip\pause

We'll focus on **convex optimization**, a special class of optimization problems
that are easier to solve and have nice properties.

\lectureappendix

# Appendix

## Variance for IS with Normalized Weights

Computing the variance of IS with normalized weights is more complicated because
the estimator is a \alert{ratio of random variables}.

\medskip\pause

Write $c = \E_g[w^*(X)]$. If both $h(X)w^*(X)$ and $w^*(X)$ have finite
variances, the multivariate CLT gives
$$
  \frac{1}{N} \sum_{i=1}^N \begin{bmatrix} h(X_i) w^*(X_i) \\ w^*(X_i) \end{bmatrix}
  \overset{\mathrm{approx}}{\sim}
  \mathcal{N}\left(
    c \begin{bmatrix} \mu \\ 1\end{bmatrix},
    \frac{1}{N} \begin{bmatrix} \sigma^{2}_{\mathrm{IS}} & \gamma \\ \gamma & \sigma^2_{w^*} \end{bmatrix}
  \right),
$$
where $$
\begin{aligned}
  \sigma^2_{\mathrm{IS}} & = \V \big(h(X_1) w^*(X_1)\big) \\
  \gamma & = \cov (h(X_1) w^*(X_1), w^*(X_1)) \\
  \sigma^2_{w^*} & = \V (w^*(X_1))
\end{aligned}
$$

## Importance Sampling Variance

We can then apply the $\Delta$-method with $t(x, y) = x / y$.

\medskip\pause

Observe that $\nabla t(x, y) = \begin{bmatrix} 1 / y \\ -x / y^2 \end{bmatrix}$
, whence

$$
  \nabla t(c\mu, c)^T \left(\begin{array}{cc} \sigma^{2}_{\mathrm{IS}} & \gamma \\ \gamma & \sigma^2_{w^*}
  \end{array} \right) \nabla t(c\mu, c) = \frac{\sigma^{2}_{\mathrm{IS}} + \mu^2 \sigma_{w^*}^2 - 2 \mu \gamma}{c^{2}}.
$$

. . .

By the $\Delta$-method,
$$
  \hat{\mu}_{\mathrm{IS}} \overset{\mathrm{approx}}{\sim} \mathcal{N}\left( \mu, \frac{\sigma^{2}_{\mathrm{IS}} + \mu^2 \sigma_{w^*}^2 - 2 \mu \gamma}{c^2 N} \right).
$$

. . .

Plug in empirical estimates, estimating $c$ when it is unknown by
$$
  \hat{c} = \frac{1}{N} \sum_{i=1}^N w^*(X_i).
$$

If both densities in the weights are unnormalized, $c$ is the ratio of their
normalizing constants.

\pdfpcnote{
hat(c) is just the MC estimate of c.
}

##

```{r}
#| label: delta-method-plot3D
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| fig-cap: Tangent plane approximation of the function $t(x,y) = x/y$ at the
#|   point $(2,3)$.

library(plot3D)

# Define function and means
t_fun <- function(x, y) x / y
mu_x <- 2
mu_y <- 3

# Grid for surface
x_seq <- seq(1, 3, length.out = 26)
y_seq <- seq(2, 4, length.out = 26)
z <- outer(x_seq, y_seq, t_fun)

# Tangent plane at (mu_x, mu_y)
grad_x <- 1 / mu_y
grad_y <- -mu_x / mu_y^2
tangent_plane <- function(x, y) {
  t_fun(mu_x, mu_y) + grad_x * (x - mu_x) + grad_y * (y - mu_y)
}
z_tangent <- outer(x_seq, y_seq, tangent_plane)

# Plot both surfaces together
persp3D(
  x_seq,
  y_seq,
  z,
  col = "steelblue",
  xlab = "x",
  ylab = "y",
  zlab = "t(x, y)",
  theta = 30,
  phi = 10,
  border = "black"
)

# Add tangent plane
persp3D(
  x_seq,
  y_seq,
  z_tangent,
  add = TRUE,
  col = "grey",
  alpha = 0.6
)

# Add mean point
points3D(
  mu_x,
  mu_y,
  t_fun(mu_x, mu_y),
  col = "darkorange",
  pch = 19,
  add = TRUE,
  cex = 1
)
```
