---
title: "Stochastic Gradient Descent"
---

{{< include _common.qmd >}}

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(mvtnorm)

set.seed(1422)

source(here::here("R", "sgd.R"))
```

## Last Time

### A Deeper Look Into the EM Algorithm

We looked at convergence properties and the generalized EM algorithm.

. . .

### Fisher Information

We talked about computing the Fisher information and how it relates to the EM
algorithm.

## Today

### Stochastic Gradient Descent

How can we optimize a loss when each full gradient is expensive?

\medskip\pause

We will see why a noisy gradient can be useful, when its noise matters, and how
the learning rate controls it.

# Stochastic Gradient Descent

## Empirical Risk Minimization

For data $(y_i,x_i)$, $i = 1,\ldots,N$, the empirical risk is
$$
  H_N(\theta) = \frac{1}{N} \sum_{i=1}^N L(y_i,x_i,\theta).
$$

. . .

Gradient descent uses the mean of all $N$ observation-level gradients:
$$
  \nabla H_N(\theta) = \frac{1}{N}\sum_{i=1}^N \nabla_\theta L(y_i,x_i,\theta),
  \qquad
  \theta_{n+1} = \theta_n - \gamma_n\nabla H_N(\theta_n).
$$

## The Cost of a Full Gradient

For least squares with $p$ predictors, one observation contributes a gradient in
$O(p)$ operations.

\medskip\pause

A full gradient therefore costs $O(Np)$ operations. Gradient descent pays that
cost at **every** update.

\bigskip\pause

If $N$ is large, can we estimate this mean without computing every term?

## Estimate a Mean by Sampling

Write $g_i(\theta) = \nabla_\theta L(y_i,x_i,\theta)$. The full gradient is
$$
  \nabla H_N(\theta) = \frac{1}{N}\sum_{i=1}^N g_i(\theta).
$$

\medskip\pause

Draw one index $J_n$ uniformly from $\{1,\ldots,N\}$ and use $g_{J_n}(\theta_n)$
as the estimate.

\medskip\pause

What does this save, and what do we lose?

## The Stochastic Gradient Descent Update

Set $\rho_n = \nabla_\theta L(y_{J_n},x_{J_n},\theta_n)$, then update
$$
  \theta_{n+1} = \theta_n - \gamma_n\rho_n.
$$

\medskip\pause

One update costs $O(p)$ rather than $O(Np)$, but its direction depends on the
sampled observation.

## Example: Ordinary Least Squares

The empirical risk is
$$
  H_N(\beta) = \frac{1}{N} \sum_{i=1}^N L(y_i,x_i,\beta),
$$
where
$$
  L(y_i,x_i,\beta) = \frac{1}{2} (y_i - \beta^T x_i)^2.
$$

And the gradient is
$$
  \nabla_\beta L(y_i,x_i,\beta) = -x_i(y_i - \beta^T x_i).
$$

## Basic Implementation of SGD

```{r}
#| label: sgd-basic

sgd_basic <- function(
  par,
  grad,
  N,                                      # number of data points
  gamma,                                  # learning rate
  sampler = function(n) sample.int(n, 1), # sample one index
  maxit = 100,
  ...
) {
  for (k in seq_len(maxit)) {
    for (i in seq_len(N)) {
      ind <- sampler(N)
      par <- par - gamma * grad(par, ind, ...)
    }
  }
  par
}
```

## Unbiasedness

Conditioning on the current iterate $\theta_n$ gives
$$
  \begin{aligned}
    \E(\rho_n \mid \theta_n)
     & = \sum_{i=1}^N \Pr(J_n = i)\,g_i(\theta_n) \\
     & = \frac{1}{N}\sum_{i=1}^N g_i(\theta_n)    \\
     & = \nabla H_N(\theta_n).
  \end{aligned}
$$

Thus each stochastic gradient is **unbiased**, even though a single one can
point far from the full gradient.

\medskip\pause

An **epoch** measures $N$ observations processed. With independent sampling,
some observations can appear more than once during an epoch.

## Before We Look at the Path

Suppose GD and SGD both start at the same point on a logistic-regression loss
surface.

\medskip\pause

Which path will be smoother? Which method will reach a useful loss first if we
count observations processed rather than updates?

## Example: Logistic Regression

::: {.columns}
::::: {.column width="47%"}
With $y_i \in \{0, 1\}$, the logistic regression loss is
$$
  L(y_i,x_i,\beta) = -y_i \beta^T x_i + \log\left( 1 + e^{\beta^T x_i} \right).
$$
and
$$
  \nabla_\beta L(y_i,x_i,\beta)
  = x_i \left( \frac{e^{\beta^T x_i}}{1 + e^{\beta^T x_i}} - y_i \right).
$$
:::::

::::: {.column width="47%"}
```{r}
#| label: logistic-regression-paths
#| echo: false
#| fig-show: animate
#| fig-height: 3.6
#| fig-width: 3.2

set.seed(123)

N <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(N, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(N, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())

loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

beta1 <- seq(-1, 3, length = 100)
beta2 <- seq(-1, 3, length = 100)

loss_2d_vectorized <- Vectorize(function(b1, b2) {
  loss_2d(b1, b2, X, y)
})

z <- outer(beta1, beta2, loss_2d_vectorized)

res_sgd <- logreg_sgd(X, y, max_epochs = 100, batch_size = 1)
res_sgd2 <- logreg_sgd(
  X,
  y,
  max_epochs = 100,
  batch_size = 1,
  K = 5,
  a = 0.5
)
res_gd <- logreg_sgd(X, y, max_epochs = 100, batch_size = N)

loss_optim <- loss_2d(
  res_glm$coefficients[1],
  res_glm$coefficients[2],
  X,
  y
)

res_gd <- logreg_sgd(X, y, max_epochs = 100, batch_size = N)
loss1 <- loss_2d_vectorized(
  res_gd$beta_history[1, ],
  res_gd$beta_history[2, ]
)
loss2 <- loss_2d_vectorized(
  res_sgd$beta_history[1, ],
  res_sgd$beta_history[2, ]
)
loss3 <- loss_2d_vectorized(
  res_sgd2$beta_history[1, ],
  res_sgd2$beta_history[2, ]
)

n1 <- length(loss1)
n2 <- length(loss2)
n3 <- length(loss3)

pal <- palette.colors(palette = "Okabe-Ito")

for (show_sgd in c(FALSE, TRUE)) {
  contour(
    beta1,
    beta2,
    z,
    col = "dark grey",
    drawlabels = FALSE,
    asp = 1
  )

  lines(
    res_gd$beta_history[1, ],
    res_gd$beta_history[2, ],
    col = pal[1]
  )
  points(
    res_gd$beta_history[1, ],
    res_gd$beta_history[2, ],
    col = pal[1],
    pch = 19,
    cex = 0.5
  )

  if (show_sgd) {
    lines(
      res_sgd$beta_history[1, ],
      res_sgd$beta_history[2, ],
      col = pal[3]
    )
    points(
      res_sgd$beta_history[1, ],
      res_sgd$beta_history[2, ],
      col = pal[3],
      pch = 19,
      cex = 0.5
    )
  }

  points(
    res_glm$coefficients[1],
    res_glm$coefficients[2],
    col = pal[2],
    pch = 19,
    cex = 1
  )

  legend(
    "bottomright",
    legend = if (show_sgd) c("SGD", "GD") else "GD",
    col = if (show_sgd) pal[c(3, 1)] else pal[1],
    pch = 19,
    cex = 0.8
  )
}
```
:::::
:::

## The Same Runs, Counted by Iterations and Epochs

::: {.columns}
::::: {.column width="47%"}
Over the first 100 updates, GD reduces the loss much faster. But each GD update
processes all $N$ observations.

```{r}
#| label: convergence-by-updates
#| echo: false
#| fig-cap: "The first 100 updates of the same GD and SGD runs."

conv_data <- tibble(
  method = rep(c("GD", "SGD"), times = c(n1, n2)),
  iteration = c(seq_len(n1) - 1L, seq_len(n2) - 1L),
  observations = c((seq_len(n1) - 1L) * N, seq_len(n2) - 1L),
  loss = c(loss1 - loss_optim, loss2 - loss_optim)
)

ggplot(
  dplyr::filter(conv_data, iteration <= 100),
  aes(iteration, loss, color = method)
) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Updates", y = "Suboptimality", color = NULL)
```
:::::

::::: {.column width="47%"}
One GD update and $N$ single-observation SGD updates each process $N$
observations. This is one **epoch of work** for either method.

```{r}
#| label: convergence-by-epochs
#| echo: false
#| fig-cap: "The same GD and SGD runs, measured in epochs of work."

ggplot(
  dplyr::mutate(conv_data, epoch = observations / N),
  aes(epoch, loss, color = method)
) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Epochs of work", y = "Suboptimality", color = NULL)
```

SGD can reach a useful loss sooner when we compare equal amounts of data
processed. The result depends on the problem and learning rates.
:::::
:::

## If SGD Reaches the Minimizer

Suppose $\theta_n = \theta^*$, so the **full** gradient is zero.

\bigskip\pause

Will the next SGD update stay at $\theta^*$?

\medskip\pause

Individual observation gradients usually do not vanish there. Their mean is
zero, but the sampled gradient can still move the iterate.

## Unbiased Does Not Mean Noiseless

At $\theta^*$, let
$\sigma_*^2 = \E(\|\rho_n\|_2^2 \mid \theta_n = \theta^*) > 0$. For a fixed
learning rate $\gamma > 0$,
$$
  \theta_{n+1} - \theta^* = -\gamma\rho_n.
$$

\medskip\pause

Consequently,
$$
  \E(\|\theta_{n+1} - \theta^*\|_2^2 \mid \theta_n = \theta^*)
  = \gamma^2\sigma_*^2 > 0.
$$

Even a step taken **from** the minimizer moves away from it on average in
squared distance.

## Learning Rate

::: {.columns align = "center"}
::::: {.column width="47%"}
With a suitable fixed step length, GD can converge to the minimizer.

\medskip\pause

When observation gradients remain noisy at the minimizer, fixed-rate SGD keeps
fluctuating around it.
:::::

::::: {.column width="47%"}
```{r}
#| label: fixed-learning-rate
#| fig-cap: "Convergence of SGD for logistic regression with a fixed learning rate."
#| echo: false

L <- 0.25 * norm(crossprod(X), "2") / N
gamma_fixed <- 0.05 / L
fixed_updates <- 30L * N
record_every <- 100L
fixed_beta <- numeric(p)
fixed_loss <- numeric(fixed_updates / record_every + 1L)
fixed_loss[1] <- loss_2d(fixed_beta[1], fixed_beta[2], X, y)

set.seed(481)
for (k in seq_len(fixed_updates)) {
  ind <- sample.int(N, 1L)
  x_ind <- X[ind, ]
  gradient <- x_ind * (plogis(sum(x_ind * fixed_beta)) - y[ind])
  fixed_beta <- fixed_beta - gamma_fixed * gradient

  if (k %% record_every == 0L) {
    fixed_loss[k / record_every + 1L] <- loss_2d(
      fixed_beta[1],
      fixed_beta[2],
      X,
      y
    )
  }
}

fixed_data <- dplyr::bind_rows(
  tibble(
    method = "Fixed-rate SGD",
    epoch = seq(0, fixed_updates, by = record_every) / N,
    loss = pmax(fixed_loss - loss_optim, 1e-12)
  ),
  tibble(
    method = "GD",
    epoch = seq_len(n1) - 1L,
    loss = pmax(loss1 - loss_optim, 1e-12)
  )
)

ggplot(fixed_data, aes(epoch, loss, color = method)) +
  geom_line() +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 30)) +
  labs(x = "Epochs", y = "Suboptimality", color = NULL) +
  theme(legend.position = c(0.8, 0.8))
```
:::::
:::

\medskip

But first, we need to introduce the concept of **strong convexity**.

## Strong Convexity

A differentiable function $H: \mathbb{R}^p \to \mathbb{R}$ is **strongly
convex** if
$$
  H(\theta) \geq H(\theta') + \nabla H(\theta')^T (\theta - \theta')
                 + \frac{l}{2} \| \theta - \theta'\|_2^2
$$
for all $\theta,\theta' \in \mathbb{R}^p$ and some $l > 0$.

\medskip\pause

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        width=7cm, height=4.3cm,
        axis lines=none,
        xlabel={$\theta$},
        ylabel={$H(\theta)$},
        ymin=-0.5, ymax=4,
        xmin=-1, xmax=2.2,
        samples=200,
        domain=-0.8:2.2,
        clip=false
      ]
      % f(x) = x^2
      \addplot[thick] {x^2};
      % Tangent at x0=1: 2*(x-1) + 1
      \addplot[DarkOrange, dashed, thick] {2*(x-1) + 1};
      % Strong convexity lower bound: tangent + 0.5*(x-1)^2
      \addplot[RoyalBlue, dotted, thick] {2*(x-1) + 1 + 0.5*(x-1)^2};
      % Point at (1,1)
      \addplot[only marks, mark=*] coordinates {(1,1)};
      % Annotations
      \node[anchor=west] at (axis cs:2.2,5) {$H(\theta)$};
      \node[anchor=west, RoyalBlue] at (axis cs:2.2,4.3) {$H(\theta') + \nabla H(\theta')^T(\theta-\theta') + \frac{l}{2}\|\theta-\theta' \|^2$};
      \node[anchor=west, DarkOrange] at (axis cs:2.2,3.5) {$H(\theta') + \nabla H(\theta')^T(\theta-\theta')$};
    \end{axis}
  \end{tikzpicture}
\end{figure}

## Example: Strong Convexity

The function
$$
  H(\theta) = \theta^2
$$
\pause is strongly convex with $l = 2$ since $$
\begin{aligned}
H(\theta) - H(\theta') - \nabla H(\theta')^T(\theta - \theta') & = \theta^2 - (\theta')^2 - 2\theta'(\theta - \theta') \\ \pause
                                         & = (\theta - \theta')^2 \\ \pause
                                         & \geq \frac{l}{2}(\theta - \theta')^2.
\end{aligned}
$$

\medskip\pause

The inequality holds for $l \leq 2$, and in particular for $l = 2$.

## Relationship to Lipschitz Gradient

If $H$ is strongly convex with parameter $l$ and has Lipschitz gradient with
constant $L$, then
$$
  l I \preceq D^2H(\theta) \preceq L I.
$$

\medskip\pause

- Strong convexity provides a **lower** bound on the curvature of $H$\pause
- Lipschitz gradient provides an **upper** bound on the curvature of $H$

. . .

### Example

The function $H(\theta) = \theta^2$ is strongly convex with $l = 2$ and has
Lipschitz gradient with $L = 2$, since $D^2H(\theta) = 2$.

\medskip\pause

This is *identical* to the strong convexity parameter.

## Example of Strong Convexity--Lipschitz Gap

::: {.columns}
::::: {.column width="47%"}
Let's take the function
$$
  H(\theta) = \theta^2 + \sin(\theta).
$$

\medskip\pause

This is twice differentiable with
$$
  D^2H(\theta) = 2 - \sin(\theta).
$$

\medskip\pause

And since
$$
  1 \leq 2 - \sin(x) \leq 3,
$$
$H$ is strongly convex with $l = 1$ and has Lipschitz gradient with $L = 3$.

\medskip\pause
:::::

::::: {.column width="47%"}
\begin{figure}
  \begin{tikzpicture}
    \pgfmathsetmacro{\a}{1}
    \pgfmathsetmacro{\xzero}{4.712389} % 3*pi/2
    \pgfmathsetmacro{\mu}{2 - \a}      % =1
    \pgfmathsetmacro{\Lips}{2 + \a}    % =3
    \pgfmathsetmacro{\fz}{\xzero*\xzero + sin(deg(\xzero))}
    \pgfmathsetmacro{\dfz}{2*\xzero + cos(deg(\xzero))}
    \begin{axis}[
        width=5cm,
        height=6.2cm,
        domain=-0.5:12.5,
        clip=false,
        samples=200,
        axis lines=none,
        xlabel={$\theta$},
        ylabel={$H(\theta)$},
        legend style={draw=none,at={(0.02,0.98)},anchor=north west}
      ]
      % f
      \addplot[thick, line label={$H(\theta)$}] {x^2 + sin(deg(x))};
      % lower quadratic
      \addplot[
        DarkOrange,
        dashed,
        thick,
        line label={Strong convexity bound},
      ] {
        \fz + \dfz*(x-\xzero) + 0.5*\mu*(x-\xzero)^2
      };
      % upper quadratic
      \addplot[
        RoyalBlue,
        dashed,
        thick,
        line label={Lipschitz bound}
      ] {
        \fz + \dfz*(x-\xzero) + 0.5*\Lips*(x-\xzero)^2
      };
      % point
      \addplot[only marks,mark=*] coordinates {(\xzero,\fz)};
    \end{axis}
  \end{tikzpicture}
  \caption{The function $H(\theta) = \theta^2 + \sin(\theta)$ is strongly convex with $l = 1$
    and has Lipschitz gradient with $L = 3$.}
\end{figure}
:::::
:::

## Conditioning

The **condition number** of a strongly convex function with Lipschitz gradient
is
$$
  \kappa = \frac{L}{l} \geq 1.
$$

. . .

The closer $\kappa$ is to $1$, the better conditioned the problem is.

. . .

### Example

For $H(\theta) = \theta^2$, we have $\kappa = 2/2 = 1$, so the problem is
perfectly conditioned.

\medskip\pause

We reach the minimum in one step with gradient descent and step length
$t = 1/L = 1/2$.

## The Robbins--Monro Convergence Theorem

Let $H(\theta) = \E L(Y,\theta)$ be the population risk. Suppose $H$ is strongly
convex and
$$
  \E\left( \|\nabla_\theta L(Y,\theta)\|_2^2 \right) \leq A + B \| \theta \|_2^2
$$
for some constants $A$ and $B$.

\medskip\pause

If $\theta^*$ is the global minimizer of $H$, then $\theta_n$ converges almost
surely toward $\theta^*$ if
$$
  \sum_{n=1}^{\infty} \gamma_n^2
  < \infty \quad \text{and} \quad \sum_{n=1}^{\infty} \gamma_n = \infty.
$$

. . .

The theorem applies to learning rates satisfying
$$
  \gamma_n \propto \frac{1}{n}.
$$

## What Do the Two Sums Mean?

The Robbins--Monro conditions ask the learning rates to do two jobs:
$$
  \sum_n \gamma_n = \infty \quad\text{and}\quad \sum_n \gamma_n^2 < \infty.
$$

\medskip\pause

The first sum keeps the algorithm moving far enough to reach the minimum. The
second makes the accumulated gradient noise finite.

\medskip\pause

For $\gamma_n \propto n^{-a}$, both hold when $1/2 < a \leq 1$. That still
leaves a wide range of finite-run behavior.

## Learning Rate

The learning rate $\gamma_n$ is crucial for SGD, but hard to set.

\medskip\pause

Convergence theorem only gives a few hints.

. . .

### A Class of Decay Schedules

$$
  \gamma_n = \frac{\gamma_0 K}{K + n^a} = \frac{\gamma_0}{1 + K^{-1} n^a}
$$
with initial learning rate $\gamma_0 > 0$ and constants $K, a > 0$.

\medskip\pause

Convergence is ensured by Robbins--Monro if $a \in (0.5, 1]$.

\medskip\pause

Fixing the exponent $a$ and picking a target rate, $\gamma_1$, to be reached
after $n_1$ steps, we can solve for $K$ and get
$$
  K = \frac{n_1^a \gamma_1}{\gamma_0 - \gamma_1}.
$$

##

```{r}
#| label: decay
#| cache: true

decay_scheduler <- function(
  gamma0 = 1,
  a = 1,
  K = 1,
  gamma1 = NULL,
  n1 = NULL,
  t0 = gamma0
) {
  force(a)
  gamma0 <- t0

  if (!is.null(gamma1) && !is.null(n1)) {
    K <- n1^a * gamma1 / (gamma0 - gamma1)
  }

  b <- gamma0 * K

  function(n) b / (K + n^a)
}
```

##

```{r, dependson="decay"}
#| label: decay-fig
#| echo: false
#| fig-height: 3
#| fig-width: 4
#| fig-cap: "Examples of learning rate schedules,
#|   using $\\gamma_n = \\frac{\\gamma_0}{1 + K^{-1} n^{a}}$."

grid_par <- expand.grid(
  n = seq(0, 1000, 2),
  K = c(25, 50, 100),
  a = c(0.6, 1),
  gamma0 = 1
)

grid_par <- dplyr::mutate(
  grid_par,
  rate = gamma0 * K / (K + n^a)
)

ggplot(
  grid_par,
  aes(n, rate, color = factor(K), linetype = factor(a))
) +
  geom_line() +
  scale_color_discrete("K") +
  scale_linetype_discrete("a") +
  guides(
    color = guide_legend(order = 1),
    linetype = guide_legend(order = 2)
  )
```

## So, Is The Problem of Setting Learning Rate Solved?

No, not really, because it's still hard to pick $\gamma_0$, $K$, and $a$.

```{r}
#| label: step-size-illustration
#| echo: false

set.seed(123)

N <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(N, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(N, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())

loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

beta1 <- seq(-1, 3, length = 100)
beta2 <- seq(-1, 3, length = 100)

loss_2d_vectorized <- Vectorize(function(b1, b2) {
  loss_2d(b1, b2, X, y)
})

z <- outer(beta1, beta2, loss_2d_vectorized)

step_size_illustration <- function(a) {
  old_par <- par(mar = c(2.5, 2.5, 0.5, 0.5))
  on.exit(par(old_par))

  res_sgd <- logreg_sgd(
    X,
    y,
    max_epochs = 100,
    batch_size = 1,
    a = a,
    t0 = 0.5
  )

  pal <- palette.colors(palette = "Okabe-Ito")

  contour(
    beta1,
    beta2,
    z,
    col = "dark grey",
    drawlabels = FALSE,
    asp = 1
  )

  lines(
    res_sgd$beta_history[1, ],
    res_sgd$beta_history[2, ],
    col = pal[3]
  )
  points(
    res_sgd$beta_history[1, ],
    res_sgd$beta_history[2, ],
    col = pal[3],
    pch = 19,
    cex = 0.5
  )

  points(
    res_glm$coefficients[1],
    res_glm$coefficients[2],
    col = pal[2],
    pch = 19,
    cex = 1
  )
}
```

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: step-size-illustration-06
#| echo: false
#| dev: png
#| dev.args: { pointsize: 9 }
#| dpi: 300
#| crop: false
#| fig-height: 2.4
#| fig-cap: SGD on the logistic regression example with $a = 0.6$ (slow decay).

step_size_illustration(0.6)
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: step-size-illustration-09
#| echo: false
#| dev: png
#| dev.args: { pointsize: 9 }
#| dpi: 300
#| crop: false
#| fig-height: 2.4
#| fig-cap: SGD on the logistic regression example with $a = 0.9$ (fast decay).

step_size_illustration(0.9)
```
:::::
:::

## Convergence Rates

### Gradient Descent

For convex $H_N$, GD with diminishing step lengths converges at rate
$$
  H_N(\theta_n) - H_N^* = O\left( \frac{1}{\sqrt{n}} \right).
$$

. . .

When $H_N$ is differentiable with Lipschitz gradient, we get
$$
  H_N(\theta_n) - H_N^* = O\left( \frac{1}{n} \right).
$$

. . .

### Stochastic Gradient Descent

For convex $H_N$ with diminishing learning rates, SGD converges at rate
$$
  \E\big(H_N(\theta_n) - H_N^*\big) = O\left( \frac{1}{\sqrt{n}} \right).
$$

. . .

But this **does not improve** when $H_N$ is differentiable with Lipschitz
gradient.

## Rates for Strongly Convex Functions

::: {.columns}
::::: {.column width="47%"}
When $H_N$ is strongly convex and has Lipschitz gradient, GD satisfies
$$
  H_N(\theta_n) - H_N^* = O(\delta^n)
$$
with $\delta \in (0, 1)$ and $\delta = 1 - l/L$.

\medskip\pause

Under the **same conditions**, SGD gives us
$$
  \E\big(H_N(\theta_n) - H_N^*\big) = O\left( \frac{1}{n} \right).
$$

. . .

So SGD does **not** enjoy linear convergence rates under strong convexity.
:::::

::::: {.column width="47%"}
\begin{figure}
\begin{tikzpicture}
  \begin{axis}[
      width=6.5cm, height=6cm,
      xlabel={$n$},
      ylabel={Suboptimality},
      legend pos=north east,
      ymode=log,
      xmax=36,
      domain=1:20
    ]
    % O(1/sqrt(n))
    \addplot[
      thick,
      line label={$O(1/\sqrt{n})$}
    ] {1/sqrt(x)};

    % O(1/n)
    \addplot[
      thick,
      dashed,
      line label={$O(1/n)$}
    ] {1/x};

    % O(delta^n) with delta = 0.5
    \addplot[
      thick,
      dotted,
      samples=50,
      line label={$O(\delta^n)$}
    ] {0.5^x};
  \end{axis}
\end{tikzpicture}
\caption{Illustration of convergence rates. Here, $\delta = 0.5$.}
\end{figure}
:::::
:::

## Mini-Batch Gradient Descent

Take a mini-batch $I_n$ of size $m$ and compute
$$
  \rho_n = \frac{1}{m}\sum_{i \in I_n} \nabla_\theta L(y_i,x_i,\theta_n),
$$
then update $\theta_{n+1} = \theta_n - \gamma_n\rho_n$.

\medskip\pause

Estimates are still unbiased,
$$
  \E(\rho_n \mid \theta_n) = \nabla H_N(\theta_n),
$$
but variance is reduced by a factor of $1/m$.

\medskip\pause

Under Lipschitz gradient, the rate goes from $O(1/\sqrt{n})$ to
$O(1/\sqrt{mn}+1/n)$.

\medskip\pause

After $E$ epochs, $n = EN/m$, so the bound becomes $O(1/\sqrt{EN}+m/(EN))$. Its
leading term does not improve with batch size.

\medskip\pause

Mini-batches can still reduce runtime through vectorized and parallel
computation.

\pdfpcnote{
Connect to law of large numbers: averaging reduces variance.
}

## How Large Should a Mini-Batch Be?

There is no universal best $m$: useful sizes depend on gradient noise, the
learning-rate schedule, and the hardware.

\medskip\pause

Try a few feasible sizes across orders of magnitude---for example,
$m \in \{1,10,100,N\}$ here---and tune the learning rate for each.

\medskip\pause

Choose the size that reaches a useful loss (or validation error) fastest in
**wall-clock time**. Stop increasing $m$ once larger batches no longer shorten
that time.

--------------------------------------------------------------------------------

```{r}
#| label: batch-size-convergence
#| echo: false
#| fig-width: 5
#| fig-height: 2.8
#| fig-cap: "The best tested rate for each batch size minimizes mean loss at
#|   epoch 100 over three pilot runs. Curves show medians of five new runs;
#|   ribbons show interquartile ranges. Mini-batch rates decay per update, while
#|   the full-batch rate stays fixed. Gaps below $10^{-10}$ are clipped."

set.seed(123)

N <- 1000
p <- 2
Sigma <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
mu <- c(0, 0)

X <- rmvnorm(N, mean = mu, sigma = Sigma)
beta <- c(1, 2)
y <- rbinom(N, 1, plogis(X %*% beta))

res_glm <- glm.fit(X, y, family = binomial())
loss_2d <- function(beta1, beta2, X, y) {
  beta <- c(beta1, beta2)
  z <- X %*% beta
  p_hat <- 1 / (1 + exp(-z))
  -mean(y * log(p_hat) + (1 - y) * log(1 - p_hat))
}

loss_optim <- loss_2d(
  res_glm$coefficients[1],
  res_glm$coefficients[2],
  X,
  y
)

batch_sizes <- c(1L, 10L, 100L, N)
rate_grid <- c(0.3, 1, 3, 5, 10, 15)
tuning_seeds <- 101:103
plot_seeds <- 201:205

tune_rate <- function(b) {
  scores <- vapply(
    rate_grid,
    function(rate) {
      gaps <- vapply(
        tuning_seeds,
        function(seed) {
          set.seed(seed)
          result <- logreg_sgd(
            X,
            y,
            max_epochs = 25,
            batch_size = b,
            a = 1,
            t0 = rate
          )
          tail(result$loss, 1) - loss_optim
        },
        numeric(1)
      )
      mean(pmax(gaps, 0))
    },
    numeric(1)
  )

  rate_grid[which.min(scores)]
}

best_rates <- vapply(batch_sizes, tune_rate, numeric(1))

conv_data <- purrr::map2_dfr(
  batch_sizes,
  best_rates,
  function(b, rate) {
    purrr::map_dfr(plot_seeds, function(seed) {
      set.seed(seed)
      result <- logreg_sgd(
        X,
        y,
        max_epochs = 100,
        batch_size = b,
        a = 1,
        t0 = rate,
        record_time = TRUE
      )

      tibble(
        b = b,
        seed = seed,
        epoch = seq_len(100),
        seconds = result$elapsed,
        loss = pmax(result$loss - loss_optim, 1e-10)
      )
    })
  }
)

plot_data <- conv_data |>
  group_by(b, epoch) |>
  summarize(
    seconds = median(seconds),
    loss = median(loss),
    lower = quantile(loss, 0.25),
    upper = quantile(loss, 0.75),
    .groups = "drop"
  ) |>
  mutate(
    batch = factor(
      b,
      levels = batch_sizes,
      labels = paste0(batch_sizes, " (", best_rates, ")")
    )
  )

ggplot(
  plot_data,
  aes(epoch, loss, color = batch, fill = batch)
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.15,
    color = NA
  ) +
  geom_line() +
  scale_y_log10() +
  labs(
    color = expression("Batch size (" * gamma[0] * ")"),
    fill = expression("Batch size (" * gamma[0] * ")"),
    y = "Suboptimality"
  )
```

##

```{r}
#| label: batch-size-wall-time
#| echo: false
#| fig-width: 5
#| fig-height: 2.8
#| fig-cap: "The same tuned runs for $N = 1000$ and $p = 2$, measured in this R
#|   implementation. Timings include updates and full-loss evaluation each
#|   epoch, but exclude rate tuning."

ggplot(
  plot_data,
  aes(seconds, loss, color = batch, fill = batch)
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.15,
    color = NA
  ) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Elapsed time (s)",
    color = expression("Batch size (" * gamma[0] * ")"),
    fill = expression("Batch size (" * gamma[0] * ")"),
    y = "Suboptimality"
  )
```

## So, Why Use SGD?

We have seen that convergence rates of SGD are worse than those of GD.

\medskip\pause

So why even care about SGD?

. . .

### Reasons to Use SGD

- On large data, mini-batches can reach a useful loss before GD completes many
  full-gradient steps. \pause
- In many applications (e.g. deep learning), we don't *care* about optimizing to
  high accuracy. \pause
- Sometimes, the stochasticity in SGD helps it escape saddle points and local
  minima.

## When SGD Wins: A Moderate Accuracy Target

```{r}
#| label: large-data-sgd-vs-gd
#| echo: false
#| fig-width: 5
#| fig-height: 2.8
#| fig-cap: "Logistic regression with $N = 10^5$ and $p = 20$; both methods use
#|   step length $1/L$. The dashed line marks a 0.01 loss gap. Curves show
#|   medians of five runs. Timing excludes full-loss checks and reference fitting."

set.seed(1729)

N_large <- 100000L
p_large <- 20L
X_large <- matrix(rnorm(N_large * p_large), nrow = N_large)
beta_true_large <- seq(0.8, 0.1, length.out = p_large)
y_large <- rbinom(
  N_large,
  1,
  plogis(drop(X_large %*% beta_true_large))
)

full_loss_large <- function(theta) {
  z <- drop(X_large %*% theta)
  mean(log1p(exp(z)) - y_large * z)
}

beta_opt_large <- glm.fit(
  X_large,
  y_large,
  family = binomial()
)$coefficients
loss_opt_large <- full_loss_large(beta_opt_large)
L_large <- 0.25 * norm(crossprod(X_large), "2") / N_large
target_gap <- 0.01

run_large_logreg <- function(batch_size, seed) {
  set.seed(seed)
  theta <- numeric(p_large)
  elapsed <- 0
  max_updates <- if (batch_size == N_large) 15L else 150L
  history <- vector("list", max_updates)

  for (k in seq_len(max_updates)) {
    start <- as.numeric(Sys.time())

    if (batch_size == N_large) {
      X_batch <- X_large
      y_batch <- y_large
    } else {
      ind <- sample.int(N_large, batch_size, replace = TRUE)
      X_batch <- X_large[ind, , drop = FALSE]
      y_batch <- y_large[ind]
    }

    gradient <- drop(crossprod(
      X_batch,
      plogis(drop(X_batch %*% theta)) - y_batch
    )) /
      batch_size
    theta <- theta - gradient / L_large
    elapsed <- elapsed + as.numeric(Sys.time()) - start

    if (batch_size == N_large || k <= 20L || k %% 5L == 0L) {
      history[[k]] <- tibble(
        update = k,
        milliseconds = 1000 * elapsed,
        gap = pmax(
          full_loss_large(theta) - loss_opt_large,
          1e-10
        )
      )
    }
  }

  dplyr::bind_rows(history)
}

comparison_large <- purrr::map_dfr(
  c(N_large, 1000L),
  function(batch_size) {
    purrr::map_dfr(201:205, function(seed) {
      run_large_logreg(batch_size, seed) |>
        mutate(
          method = if (batch_size == N_large) {
            "GD"
          } else {
            "SGD (m = 1000)"
          }
        )
    })
  }
)

comparison_large <- comparison_large |>
  group_by(method, update) |>
  summarize(
    milliseconds = median(milliseconds),
    gap = median(gap),
    lower = quantile(gap, 0.25),
    upper = quantile(gap, 0.75),
    .groups = "drop"
  )

ggplot(
  comparison_large,
  aes(milliseconds, gap, color = method, fill = method)
) +
  geom_hline(
    yintercept = target_gap,
    linetype = "dashed",
    color = "gray50"
  ) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.15,
    color = NA
  ) +
  geom_line() +
  scale_y_log10() +
  coord_cartesian(xlim = c(0.1, 20), ylim = c(1e-3, 0.3)) +
  labs(
    x = "Optimizer time (ms)",
    y = "Suboptimality",
    color = NULL,
    fill = NULL
  )
```

##

```{r}
#| label: escape-minima-setup
#| echo: false

set.seed(1)

f <- function(x) x^4 - 2 * x^2 + 1 * x
df <- function(x) 4 * x^3 - 4 * x + 1

# Parameters
eta <- 0.07
steps <- 25
x_gd <- numeric(steps)
x_sgd <- numeric(steps)
x_gd[1] <- x_sgd[1] <- 1.4

for (k in 1:(steps - 1)) {
  # Exact gradient descent
  x_gd[k + 1] <- x_gd[k] - eta * df(x_gd[k])

  # Noisy gradient descent (SGD-like)
  noise <- rnorm(1, sd = 4) # stochasticity
  x_sgd[k + 1] <- x_sgd[k] - eta * (df(x_sgd[k]) + noise)
}

plot_curves <- function(main) {
  curve(
    f,
    from = -1.6,
    to = 1.7,
    n = 200,
    lwd = 2,
    col = "gray60",
    ylab = "",
    xlab = "",
    axes = TRUE,
    yaxt = if (main == "GD") "s" else "n",
    main = main
  )
}

plot_iteration <- function(i, x_gd, x_sgd) {
  x_gd_i <- x_gd[1:i]
  x_sgd_i <- x_sgd[1:i]

  opar <- par(no.readonly = TRUE)

  par(
    mfrow = c(1, 2),
    oma = c(4.1, 4.1, 0.1, 0.3),
    mai = c(0.1, 0.2, 0.5, 0.0)
  )

  plot_curves("GD")
  lines(x_gd_i, f(x_gd_i), col = "royalblue", lwd = 1)
  points(
    x_gd_i,
    f(x_gd_i),
    col = "royalblue",
    pch = 19,
    cex = 0.8
  )

  plot_curves("SGD")
  lines(x_sgd_i, f(x_sgd_i), col = "darkorange", lwd = 1)
  points(
    x_sgd_i,
    f(x_sgd_i),
    col = "darkorange",
    pch = 19,
    cex = 0.8
  )

  mtext(expression(theta), side = 1, outer = TRUE, line = 2)
  mtext(expression(H(theta)), side = 2, outer = TRUE, line = 2)

  par(opar)
}
```

```{r}
#| label: sgd-escape-minima
#| echo: false
#| fig-show: animate
#| fig-width: 5.2
#| fig-height: 3.2
#| fig-cap: >-
#|   The stochasticity in SGD sometimes helps it escape local minima and saddle
#|   points.

for (i in c(1, 2, 3, 5, 10, 15, 20, steps)) {
  plot_iteration(i, x_gd, x_sgd)
}
```

##

\begin{algorithm}[H]
  \caption{Stochastic gradient descent (with mini-batches)}
  \KwData{Initial parameter $\theta_0$ and learning-rate schedule $(\gamma_n)$}
  \For{$n = 0,1,\dots$}{
    $I_n \gets$ mini-batch of $m$ indices\;

    $\rho_n \gets \frac{1}{m}\sum_{i\in I_n}\nabla_\theta L(y_i,x_i,\theta_n)$\;

    $\theta_{n+1} \gets \theta_n - \gamma_n\rho_n$\;
    \If{converged}{
      \Return{$\theta_{n+1}$}
    }
  }
\end{algorithm}

## Implementation Details

### Nested Loops

Implementing an SGD algorithm typically means implementing a nested loop, and we
know this isn't R's strongest suite.

\medskip\pause

Next lecture, when we turn to Rcpp, you will have the tools to solve this
problem.

. . .

### Data Storage

SGD algorithms typically subset rows, or blocks of rows.

\medskip\pause

In a column-major order matrix, this is not optimal. Transposing the matrix
first can help.

## Exercise: Estimate a Mean with SGD

::: {.columns}
::::: {.column width="47%"}
For observations $y_1,\ldots,y_N$, minimize
$$
  H_N(\theta) = \frac{1}{N}\sum_{i=1}^N \frac{1}{2}(\theta - y_i)^2.
$$

\medskip\pause

1. Find the minimizer $\theta^*$ and the sampled gradient. \pause
2. Complete the SGD update and plot its path. \pause
3. Compare $\gamma_k = 0.1$ with $\gamma_k = 1/k$. What changes?
:::::

::::: {.column width="47%"}
\footnotesize

```{r}
#| label: sgd-mean-exercise
#| eval: false

set.seed(1)

N <- 100
y <- rnorm(N, mean = 2)
steps <- 200
theta <- numeric(steps + 1)
theta[1] <- 0

for (k in seq_len(steps)) {
  i <- sample.int(N, 1)
  gamma <- 0.1

  theta[k + 1] <- NA # <1>
}

plot(0:steps, theta, type = "l")
abline(h = mean(y), lty = 2)
```

1. Complete the update.
:::::
:::

## Summary

Stochastic gradient descent is a popular method for large-scale optimization.

\medskip\pause

Convergence results are relatively weak, but for many applications this does not
matter.

\medskip\pause

In practice, most implementations use mini-batches.

## Next TIme

### Rcpp

We will look at how to use Rcpp to speed up R code.

\medskip\pause

Because SGD typically involves nested loops, these algorithms benefit greatly
from relying on C++.
