---
title: "Optimization"
---

{{< include _common.qmd >}}

## Recap: Monte Carlo Methods

### Standard Monte Carlo

Estimate $\E h(X)$ with $X \sim f$ by
$$
  \hat{\mu}_{\mathrm{MC}}
  = \frac{1}{N} \sum_{i=1}^N h(X_i) \quad \text{with } X_i \overset{iid}{\sim} f.
$$

. . .

### Importance Sampling

Use the "trick" that
$$
  \mu = \int h(x) f(x) \mathrm{d}x
      = \int h(x) \frac{f(x)}{g(x)} g(x) \ \mathrm{d}x
      = \int h(x) w^*(x) g(x) \ \mathrm{d}x
$$
and estimate
$$
  \hat{\mu}^*_{\mathrm{IS}}
  = \frac{1}{N} \sum_{i=1}^N h(X_i) w^*(X_i) \quad \text{with } X_i \overset{iid}{\sim} g.
$$

## Recap: Importance Sampling

### Normalized Weights

When $f$ is known only up to a normalizing constant, we can still use importance
sampling by estimating normalized weights.

. . .

### Picking a Good Importance Distribution

Pick an importance distribution $g$ that covers the important parts of $f$, with
heavy tails, and from which we can easily sample.

## Today

### Optimization

What is an optimization problem?

\medskip\pause

We learn about different kinds of optimization problems.

. . .

### Convex Optimization

Convex optimization problems are a special class of optimization problems that
are easier to solve.

. . .

### Algorithms

We learn about gradient descent and Newton's method.

# Optimization

## Optimization

The process of trying to optimize an objective function $H$.

\medskip\pause

Typically, we are interested in either **minimizing** or **maximizing** the
function.

\medskip\pause

Since minimizing $H$ is equivalent to maximizing $-H$, we will focus on
minimization.

. . .

### Examples {.example}

- Fitting a generalized linear model by minimizing the negative log-likelihood
  \pause
- Finding the shortest path in a network \pause
- Minimizing variance in a portfolio of financial assets

## Standard Form

In *standard form*, an optimization problem is written as
$$
  \begin{aligned}
     & \operatorname*{minimize}_\theta &  & H(\theta)                                  \\
     & \textrm{subject to}             &  & g_i(\theta) \leq 0, \quad i = 1, \ldots, m \\
     &                                 &  & h_j(\theta) = 0, \quad j = 1, \ldots, k.
  \end{aligned}
$$
where

- $H$ is the objective function,
- $g_i$ are inequality constraints, and
- $h_j$ are equality constraints.

. . .

### Solution

Let $C$ be the feasible set. An optimal point satisfies
$$
  \theta^* \in \argmin_{\theta \in C} H(\theta).
$$

## Examples of Optimization Problems

### Smoothing Splines

$$
  \operatorname*{minimize}_{\symbf{\beta}} \left( (\symbf{y} - \symbf{\Phi}\symbf{\beta})^T (\symbf{y} - \symbf{\Phi}\symbf{\beta}) + \lambda \symbf{\beta}^T \symbf{\Omega} \symbf{\beta} \right)
$$

. . .

### Finding the Hyperparameter for Smoothing Splines

$$
  \begin{aligned}
     & \operatorname*{minimize}_\lambda &  & \operatorname{CV}(\lambda) \\
     & \textrm{subject to}              &  & \lambda \geq 0,
  \end{aligned}
$$
where $\operatorname{CV}(\lambda)$ is the mean squared prediction error from
cross-validation.

## Examples of Optimization Problems

### Maximum Likelihood

$$
  \operatorname*{minimize}_{\theta} \big(-\ell(\theta)\big),
$$
where $\ell$ is the log-likelihood function.

. . .

### Minimum-Variance Portfolio

$$
  \begin{aligned}
     & \operatorname*{minimize}_w &  & w^T \Sigma w                     \\
     & \textrm{subject to}        &  & \bar{p}^T w \geq r_\mathrm{min}, \\
     &                            &  & \symbf{1}^T w = 1,
  \end{aligned}
$$
where $w$ are the asset weights, $\Sigma$ is the covariance matrix of asset
returns, $\bar{p}$ is the vector of expected returns, and $r_\mathrm{min}$ is
the minimum acceptable return.

## Example: Ordinary Least Squares (OLS) Regression

The problem is to
$$
  \begin{aligned}
     & \operatorname*{minimize}_{\beta \in \mathbb{R}^p} &  & H(\beta) = \frac{1}{2} \|y - X\beta\|_2^2.
  \end{aligned}
$$

. . .

### Solution

An optimal point satisfies
$$
  \beta^* \in \argmin_{\beta \in \mathbb{R}^p} H(\beta).
$$

## Types of Optimization Problems

### Convexity

- **Convex**
- Quasiconvex
- Nonconvex

. . .

### Smoothness

- **Smooth**
- Nonsmooth

. . .

### Constraints

- **Unconstrained**
- Constrained

## Optimality

For **unconstrained** minimization, suppose $H$ is twice continuously
differentiable.

. . .

### Necessary Conditions

If $\theta^*$ is a local minimum, then
$$
  \nabla H(\theta^*) = 0 \quad \text{and} \quad \nabla^2 H(\theta^*) \succeq 0.
$$
These conditions alone do not guarantee a local minimum.

. . .

### Sufficient Conditions

If
$$
  \nabla H(\theta^*) = 0 \quad \text{and} \quad \nabla^2 H(\theta^*) \succ 0,
$$
then $\theta^*$ is a **strict local minimum**.

## Convex Optimization Problems

Standard form:

$$
  \begin{aligned}
     & \operatorname*{minimize}_\theta &  & H(\theta)                                   \\
     & \textrm{subject to}             &  & g_i(\theta) \leq 0, \quad i = 1, \ldots, m, \\
     &                                 &  & a_j^T \theta = b_j, \quad j = 1, \ldots, k,
  \end{aligned}
$$

where $H$ and $g_i$ are convex.

## Convex Functions

### Jensen's Inequality

A function $H$ is convex iff its domain is convex and, for all
$\theta_1, \theta_2 \in \operatorname{dom} H$ and $\lambda \in [0, 1]$,
$$
  H(\lambda \theta_1 + (1 - \lambda) \theta_2) \leq \lambda H(\theta_1)
                                                    + (1 - \lambda) H(\theta_2).
$$

\pause

```{r}
#| label: jensens-inequality
#| echo: false
#| fig-width: 4
#| fig-height: 2
#| fig-cap: "Jensen's inequality for a convex function."

library(ggplot2)

# Convex function
f <- function(x) x^2

# Points for Jensen's inequality
x1 <- -0.5
x2 <- 1
lambda <- 0.3
x_mid <- lambda * x1 + (1 - lambda) * x2

df <- data.frame(
  x = c(seq(-2, 2, length.out = 100), x1, x2),
  y = c(f(seq(-2, 2, length.out = 100)), f(x1), f(x2)),
  type = c(rep("f", 100), rep("chord", 2))
)

pts <- data.frame(x = c(x1, x2), y = c(f(x1), f(x2)))

labels <- c(
  f = expression(H(theta)),
  chord = expression(
    lambda * H(theta[1]) + (1 - lambda) * H(theta[2])
  )
)

ggplot(df, aes(x, y)) +
  geom_line(aes(linetype = type, color = type)) +
  geom_point(data = pts, aes(x, y)) +
  scale_linetype_manual(
    values = c(f = "solid", chord = "dashed"),
    labels = labels
  ) +
  scale_color_manual(
    values = c(f = "black", chord = "steelblue"),
    labels = labels
  ) +
  geom_text(
    aes(x, y),
    data = pts,
    label = c("theta[1]", "theta[2]"),
    parse = TRUE,
    nudge_x = c(-0.3, 0.3)
  ) +
  labs(color = NULL, linetype = NULL) +
  labs(y = expression(H(theta)), x = expression(theta))
```

## First-Order Condition for Convexity

On an open convex domain, a **differentiable** function $H$ is convex iff
$$
  H(\theta_1) \geq H(\theta_2) + \nabla H(\theta_2)^T(\theta_1 - \theta_2)
$$
for all $\theta_1, \theta_2 \in \operatorname{dom} H$.

. . .

```{r}
#| label: first-order-condition
#| echo: false
#| fig-width: 4
#| fig-height: 2
#| fig-cap: "First-order condition for convexity."

library(ggplot2)
# Convex function
f <- function(x) x^2
f_prime <- function(x) 2 * x
# Points for tangent line
x0 <- 0.2
x_vals <- seq(-1, 2, length.out = 100)
df <- data.frame(
  x = c(x_vals, x0),
  y = c(f(x_vals), f(x0)),
  type = c(rep("f", 100), "tangent")
)
pts <- data.frame(x = x0, y = f(x0))

labels <- c(
  f = expression(H(theta)),
  tangent = expression(
    H(theta[2]) + nabla * H(theta[2])^T * (theta[1] - theta[2])
  )
)

ggplot(df, aes(x, y)) +
  geom_line(aes(linetype = type, color = type)) +
  geom_abline(
    slope = f_prime(x0),
    intercept = f(x0) - f_prime(x0) * x0,
    linetype = "dashed",
    color = "steelblue"
  ) +
  geom_point(data = pts, aes(x, y)) +
  geom_text(
    data = pts,
    aes(x, y),
    label = expression(theta[2]),
    nudge_y = 0.3
  ) +
  scale_linetype_manual(
    values = c(f = "solid", tangent = "dashed"),
    labels = labels
  ) +
  scale_color_manual(
    values = c(f = "black", tangent = "steelblue"),
    labels = labels
  ) +
  lims(y = c(-1, 4)) +
  labs(color = NULL, linetype = NULL) +
  labs(y = expression(H(theta)), x = expression(theta))
```

## Optimality for Convex Problems

Suppose $H$ is differentiable and convex, and $C$ is a convex feasible set. Any
local minimum is **global** (but not necessarily unique).

. . .

### First-Order Condition

A feasible point $\theta^* \in C$ is optimal iff
$$
  \nabla H(\theta^*)^T(\theta - \theta^*) \geq 0\quad\text{for all}\quad \theta
                                          \in C.
$$

. . .

### Unconstrained Problems

When $C = \mathbb{R}^p$, this simplifies to
$$
  \nabla H(\theta^*) = 0.
$$

## Quasiconvex Optimization Problems

### Standard Form

$$
  \begin{aligned}
     & \operatorname*{minimize}_\theta &  & H(\theta)                                   \\
     & \textrm{subject to}             &  & g_i(\theta) \leq 0, \quad i = 1, \ldots, m, \\
     &                                 &  & A^T \theta = b,
  \end{aligned}
$$
where $g_i$ are convex but $H$ is **quasiconvex**.

. . .

### Examples

- Minimizing negative quasiconcave utility (economics) \pause
- The floor function $H(\theta) = \lfloor \theta \rfloor$

. . .

### Algorithms

**Bisection** on objective values, solving a convex feasibility problem at each
step (not covered here).

## Nonconvex Optimization Problems

- Generally hard to solve.
- The objective or feasible set is nonconvex.
- Local minima are not necessarily global.

. . .

### Examples

- Deep learning
- Hyperparameter tuning (cross-validation)
- Combinatorial problems

. . .

### Algorithms

Gradient descent seek local solutions.

\medskip

**Stochastic algorithms** (next week) and **global optimization methods** are
also used.

##

```{r}
#| label: convexity-types
#| echo: false
#| warning: false
#| fig-height: 3.2
#| fig-width: 5.4
#| fig-cap: "$\\theta^2$ (convex), $\\sin(\\theta + \\pi)$ (nonconvex),
#|   and $\\sqrt{|\\theta|}$ (quasiconvex; derivative undefined at $\\theta = 0$)."

convex_function <- function(x) x^2
non_convex_function <- function(x) sin(x + pi)
quasi_convex_function <- function(x) sqrt(abs(x))

# Define the gradients
convex_gradient <- function(x) 2 * x
non_convex_gradient <- function(x) cos(x + pi)
quasi_convex_gradient <- function(x) {
  ifelse(x == 0, NA_real_, sign(x) / (2 * sqrt(abs(x))))
}

# Create a sequence of x values
x_vals <- seq(-2, 2, length.out = 201)

# Create data frames for plotting
convex_df <- data.frame(
  x = x_vals,
  y = convex_function(x_vals),
  gradient = convex_gradient(x_vals),
  type = "Convex"
)
non_convex_df <- data.frame(
  x = x_vals,
  y = non_convex_function(x_vals),
  gradient = non_convex_gradient(x_vals),
  type = "Nonconvex"
)
quasi_convex_df <- data.frame(
  x = x_vals,
  y = quasi_convex_function(x_vals),
  gradient = quasi_convex_gradient(x_vals),
  type = "Quasiconvex"
)

# Combine data frames
df <- bind_rows(convex_df, non_convex_df, quasi_convex_df)

# Function to add Jensen's inequality line segment
add_jensens_line <- function(df, func) {
  x1 <- -1.5
  x2 <- 0.1
  y1 <- func(x1)
  y2 <- func(x2)
  df |>
    mutate(
      jensens_y = ifelse(
        x >= x1 & x <= x2,
        y1 + (y2 - y1) * (x - x1) / (x2 - x1),
        NA
      )
    )
}

# Add Jensen's inequality line segments
convex_df <- add_jensens_line(convex_df, convex_function)
non_convex_df <- add_jensens_line(
  non_convex_df,
  non_convex_function
)
quasi_convex_df <- add_jensens_line(
  quasi_convex_df,
  quasi_convex_function
)

# Combine data frames with Jensen's line
df_with_jensens <- bind_rows(
  convex_df,
  non_convex_df,
  quasi_convex_df
)

# Plot the objective functions with Jensen's inequality line segments using facets
p1 <- ggplot(df_with_jensens, aes(x = x)) +
  geom_line(aes(y = y)) +
  geom_line(
    aes(y = jensens_y),
    linetype = "dashed",
    color = "steelblue4"
  ) +
  labs(y = expression(H(theta)), x = expression(theta)) +
  facet_wrap(~type, scales = "free_y")

# The quasiconvex example's derivative has separate branches at zero.
p2 <- ggplot(df, aes(x = x, y = gradient)) +
  geom_line(aes(
    group = ifelse(type == "Quasiconvex", sign(x), 0)
  )) +
  labs(y = expression(nabla * H(theta)), x = expression(theta)) +
  facet_wrap(~type, scales = "free_y")

# Print the plots
p1 / p2 +
  plot_layout(
    axis_title = "collect",
    guides = "collect",
    axes = "collect"
  )
```

# Solving Optimization Problems

## Types of Optimization Methods

### Direct Methods

Solve in finitely many operations in exact arithmetic.

\medskip\pause

**Examples:** solving a linear system of equations

. . .

### Iterative Optimization

Incrementally update the optimization variable until some convergence criterion
is met.

\medskip\pause

**Examples:** gradient descent, hill-climbing

## Running Example: OLS

```{r}
#| label: ols-example
#| echo: false

source(here::here("R", "gradient_descent.R"))
source(here::here("R", "lecture7-models.R"))

library(mvtnorm)
library(plot3D)

set.seed(24)

Sigma <- matrix(c(1, 0.8, 0.8, 1), 2, 2)
x <- rmvnorm(100, sigma = Sigma)

colnames(x) <- c("x1", "x2")

beta <- c(1, 2)

y <- x %*% beta

beta1_grid <- seq(-1, 4, length.out = 51)
beta2_grid <- seq(-1, 4, length.out = 51)

beta1_grid_persp <- seq(-1, 4, length.out = 21)
beta2_grid_persp <- seq(-1, 4, length.out = 21)

f_vals <- outer(
  beta1_grid,
  beta2_grid,
  Vectorize(function(b1, b2) ols_objective(c(b1, b2), x, y))
)

f_vals_persp <- outer(
  beta1_grid_persp,
  beta2_grid_persp,
  Vectorize(function(b1, b2) ols_objective(c(b1, b2), x, y))
)

colnames(f_vals) <- beta1_grid
rownames(f_vals) <- beta2_grid
colnames(f_vals_persp) <- beta1_grid_persp
rownames(f_vals_persp) <- beta2_grid_persp

optimum_df <- tibble(beta1 = beta[1], beta2 = beta[2])

beta_df <- tibble(
  beta1 = rep(beta1_grid, times = ncol(f_vals)),
  beta2 = rep(beta2_grid, each = nrow(f_vals)),
  f = as.vector(f_vals)
)

beta_df_persp <- tibble(
  beta1 = rep(beta1_grid_persp, times = ncol(f_vals_persp)),
  beta2 = rep(beta2_grid_persp, each = nrow(f_vals_persp)),
  f = as.vector(f_vals_persp)
)
```

We are going to use a simple example of OLS as a running example.

\medskip\pause

We will let $p = 2$, and generate some data from a multivariate normal
distribution:
$$
  \symbf{x} \sim \normal(
    \symbf{\mu} = \begin{bmatrix}
                    0 \\
                    0
                  \end{bmatrix},
    \symbf{\Sigma} = \begin{bmatrix}
                       1   & 0.8 \\
                       0.8 & 1
                     \end{bmatrix}
  )
$$

##

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: persp-ols
#| fig-width: 4
#| fig-height: 4
#| echo: false

persp3D(
  beta1_grid,
  beta2_grid,
  f_vals,
  theta = 30,
  phi = 30,
  expand = 0.5,
  border = "black",
  col = "steelblue1",
  xlab = "beta1",
  ylab = "beta2",
  zlab = "H"
)
```
:::::

::::: {.column width="47%"}
```{r}
#| label: contour-ols
#| fig-height: 2.8
#| echo: false

source(here::here("R", "gradient_descent.R"))

contour_scale <- scale_color_gradient(
  low = "grey50",
  high = "grey80",
  guide = NULL
)

step_size_illustration <- function(
  x,
  y,
  t = NULL,
  maxit = 35,
  line_search = FALSE,
  plot_convergence = TRUE,
  plot_approx = FALSE,
  method = c("gd", "newton"),
  iteration = NULL,
  show_next_step = FALSE
) {
  method <- match.arg(method)

  optimizer <- switch(method, gd = gd, newton = newton)

  res <- optimizer(
    x,
    y,
    t = t,
    maxit = maxit,
    line_search = line_search
  )

  L <- norm(crossprod(x), "2")

  if (is.null(t)) {
    if (method == "newton") {
      t <- 1
    } else {
      t <- 1 / L
    }
  }

  df <- t(res$betas)
  colnames(df) <- c("beta1", "beta2")

  approx_index <- if (is.null(iteration)) 1L else iteration + 1L
  approx_point <- df[approx_index, ]

  if (!is.null(iteration)) {
    df <- df[
      seq_len(approx_index + as.integer(show_next_step)),
      ,
      drop = FALSE
    ]
  }

  beta1_grid <- seq(-1, 4, length.out = 51)
  beta2_grid <- seq(-1, 4, length.out = 51)

  f_vals <- outer(
    beta1_grid,
    beta2_grid,
    Vectorize(function(b1, b2) ols_objective(c(b1, b2), x, y))
  )

  contour_breaks <- pretty(range(f_vals), n = 25)

  colnames(f_vals) <- beta1_grid
  rownames(f_vals) <- beta2_grid

  optimum_df <- tibble(beta1 = beta[1], beta2 = beta[2])

  beta_df <- tibble(
    beta1 = rep(beta1_grid, times = ncol(f_vals)),
    beta2 = rep(beta2_grid, each = nrow(f_vals)),
    f = as.vector(f_vals)
  )

  res <- ggplot(beta_df, aes(beta1, beta2)) +
    contour_scale +
    geom_point(
      data = optimum_df,
      color = "darkorange",
      size = 3,
      pch = 4
    ) +
    labs(x = expression(beta[1]), y = expression(beta[2])) +
    coord_fixed(
      xlim = c(-1, 4),
      ylim = c(-1, 4),
      expand = FALSE
    ) +
    geom_contour(
      aes(z = f, color = after_stat(level)),
      breaks = contour_breaks
    )

  if (plot_approx) {
    approx_vals <- outer(
      beta1_grid,
      beta2_grid,
      Vectorize(function(b1, b2) {
        if (method == "gd") {
          ols_gd_model(c(b1, b2), approx_point, x, y, t)
        } else {
          ols_newton_model(c(b1, b2), approx_point, x, y)
        }
      })
    )

    colnames(approx_vals) <- beta1_grid
    rownames(approx_vals) <- beta2_grid

    approx_df <- tibble(
      beta1 = rep(beta1_grid, times = ncol(approx_vals)),
      beta2 = rep(beta2_grid, each = nrow(approx_vals)),
      f = as.vector(approx_vals)
    )

    res <- res +
      geom_contour(
        data = approx_df,
        aes(beta1, beta2, z = f),
        color = "steelblue",
        linetype = "dashed",
        breaks = contour_breaks[
          seq(1, length(contour_breaks), by = 5)
        ]
      ) +
      annotate(
        "point",
        x = approx_point[1],
        y = approx_point[2],
        color = "steelblue",
        size = 2
      )
  }

  if (plot_convergence) {
    res <- res + geom_point(data = df, size = 0.5)

    if (nrow(df) > 1) {
      res <- res + geom_path(data = df)
    }

    if (show_next_step && !is.null(iteration)) {
      next_point <- df[nrow(df), ]
      res <- res +
        annotate(
          "segment",
          x = approx_point[1],
          y = approx_point[2],
          xend = next_point[1],
          yend = next_point[2],
          arrow = grid::arrow(
            length = grid::unit(0.06, "inches")
          )
        ) +
        annotate(
          "point",
          x = next_point[1],
          y = next_point[2],
          size = 2
        )
    }
  }

  res
}

step_size_illustration(x, y, plot_convergence = FALSE)
```
:::::
:::

## Direct Methods

Rounding error and conditioning affect numerical accuracy.

### Example: OLS

Solve
$$
  X^TX \beta = X^Ty
$$
for $\beta \in \mathbb{R}^p$, assuming $X$ has full column rank.

\medskip\pause

In R you call

```r
solve(crossprod(X), crossprod(X, y))
```

For ordinary numeric matrices, `solve()` uses **LU factorization** with partial
pivoting. Solving the least-squares problem by QR avoids forming $X^T X$.

## Iterative Methods

Update a current best guess iteratively until it is "good enough".

```{r}
#| label: order-information-setup
#| include: false

information_objective <- function(theta) {
  theta^4 / 4 - theta^2 / 2 + theta / 4 + 1
}

information_queries <- tibble(theta = c(1.25, 0, -1.25)) |>
  mutate(
    value = information_objective(theta),
    slope = theta^3 - theta + 1 / 4,
    curvature = 3 * theta^2 - 1
  )

information_curve <- tibble(
  theta = seq(-1.8, 1.8, length.out = 401)
) |>
  mutate(value = information_objective(theta))

information_models <- crossing(
  query = seq_len(nrow(information_queries)),
  offset = seq(-0.45, 0.45, length.out = 101)
) |>
  mutate(
    theta = information_queries$theta[query] + offset,
    tangent = information_queries$value[query] +
      information_queries$slope[query] * offset,
    quadratic = tangent +
      information_queries$curvature[query] * offset^2 / 2
  )

information_plot <- function(
  order = -1L,
  reference = FALSE,
  n_queries = nrow(information_queries)
) {
  p <- ggplot()
  if (order < 0 || reference) {
    p <- p +
      geom_line(
        data = information_curve,
        aes(theta, value),
        color = if (order < 0) "black" else "grey70"
      )
  }
  if (order > 0) {
    model <- if (order == 1) "tangent" else "quadratic"
    p <- p +
      geom_line(
        data = filter(information_models, query <= n_queries),
        aes(theta, .data[[model]], group = query),
        color = "darkorange"
      )
  }
  p +
    geom_point(
      data = head(information_queries, n_queries),
      aes(theta, value),
      color = "royalblue",
      size = 2
    ) +
    scale_x_continuous(
      breaks = information_queries$theta,
      labels = expression(theta[1], theta[2], theta[3])
    ) +
    scale_y_continuous(breaks = 0:2) +
    coord_cartesian(xlim = c(-1.8, 1.8), ylim = c(0, 2.5)) +
    labs(x = expression(theta), y = expression(H(theta)))
}
```

```{r}
#| label: order-objective
#| echo: false
#| dependson: order-information-setup
#| fig-width: 5
#| fig-height: 2.35
#| fig-show: animate

for (iteration in seq_len(nrow(information_queries))) {
  print(information_plot(n_queries = iteration))
}
```

\uncover<3->{What information do we get at each queried point?}

\pdfpcnote{
  - Reveal three improving queries, from right to left.\\
  - The full curve is shown for us; an optimizer queries individual points.\\
  - Keep these same three points in mind on the next slide.\\
}

## Optimization Methods by Order

Each order adds information at the same three points.

\medskip

::: {.columns}
::::: {.column width="31%"}
### Zeroth order

```{r}
#| label: order-values
#| echo: false
#| dependson: order-information-setup
#| fig-width: 1.7
#| fig-height: 2.1

information_plot(0)
```

Value: $H(\theta_i)$

\medskip

Grid search
:::::

\pause

::::: {.column width="31%"}
### First order

```{r}
#| label: order-tangents
#| echo: false
#| dependson: order-information-setup
#| fig-width: 1.7
#| fig-height: 2.1

information_plot(1)
```

Also slope: $H'(\theta_i)$

\medskip

Gradient descent
:::::

\pause

::::: {.column width="31%"}
### Second order

```{r}
#| label: order-quadratics
#| echo: false
#| dependson: order-information-setup
#| fig-width: 1.7
#| fig-height: 2.1

information_plot(2)
```

Also curvature: $H''(\theta_i)$

\medskip

Newton's method
:::::
:::

\pdfpcnote{
  - Same points and axes: values, then slopes, then curvature.\\
  - Lines and parabolas show local information, not prescribed update rules.\\
  - Zeroth-order methods can estimate slopes using extra function evaluations.\\
}

## Comparing with the Objective

Gray curves show the objective for reference.

\medskip

::: {.columns}
::::: {.column width="31%"}
### Zeroth order

```{r}
#| label: order-values-reference
#| echo: false
#| dependson: order-information-setup
#| fig-width: 1.7
#| fig-height: 2.1

information_plot(0, reference = TRUE)
```

Value: $H(\theta_i)$

\medskip

Grid search
:::::

::::: {.column width="31%"}
### First order

```{r}
#| label: order-tangents-reference
#| echo: false
#| dependson: order-information-setup
#| fig-width: 1.7
#| fig-height: 2.1

information_plot(1, reference = TRUE)
```

Also slope: $H'(\theta_i)$

\medskip

Gradient descent
:::::

::::: {.column width="31%"}
### Second order

```{r}
#| label: order-quadratics-reference
#| echo: false
#| dependson: order-information-setup
#| fig-width: 1.7
#| fig-height: 2.1

information_plot(2, reference = TRUE)
```

Also curvature: $H''(\theta_i)$

\medskip

Newton's method
:::::
:::

\pdfpcnote{
  - Compare each local description with the same objective.\\
  - Quadratics match curvature locally but can depart from the full curve.\\
}

## Gradient Descent

The quintessential first-order method!

\medskip\pause
::: {.columns}
::::: {.column width="47%"}
\begin{algorithm}[H]
  \caption{Gradient descent}
  \KwData{Step size $\gamma > 0$}
  \Repeat{convergence}{
    $\theta \gets \theta - \gamma \nabla H(\theta)$\;
  }
  \Return{$\theta$}
\end{algorithm}

\medskip

Converges to a **global minimum** if one exists, $H$ is convex on
$\mathbb{R}^p$, plus some other technical conditions.

### Step Size {.alert}

But how do we choose $\gamma$?
:::::

::::: {.column width="47%"}
```{r}
#| label: gd-illustration
#| echo: false

gd_illustration <- function(f, f_prime, x0, gamma, n_steps) {
  x_vals <- numeric(n_steps + 1)
  x_vals[1] <- x0

  for (i in 1:n_steps) {
    x_vals[i + 1] <- x_vals[i] - gamma * f_prime(x_vals[i])
  }

  df <- data.frame(
    x = seq(
      min(x_vals) - 0.5,
      max(x_vals) + 1,
      length.out = 100
    ),
    y = f(seq(
      min(x_vals) - 0.5,
      max(x_vals) + 1,
      length.out = 100
    ))
  )

  points_df <- data.frame(
    step = 0:n_steps,
    x = x_vals,
    y = f(x_vals)
  )

  ggplot(df, aes(x, y)) +
    geom_line() +
    geom_point(data = points_df, aes(x, y)) +
    geom_segment(
      data = points_df[-nrow(points_df), ],
      aes(
        x = x + 0.1 * f_prime(x),
        y = y + 0.1 * f_prime(x) * f_prime(x),
        xend = x - gamma * f_prime(x),
        yend = y - gamma * f_prime(x) * f_prime(x)
      ),
      arrow = arrow(length = unit(0.2, "cm")),
      color = "steelblue"
    ) + # Gamma scales the horizontal update, not the tangent arrow's length.
    annotate(
      "segment",
      x = x_vals[1],
      xend = x_vals[2],
      y = f(x0),
      yend = f(x0),
      arrow = arrow(length = unit(0.15, "cm")),
      color = "steelblue"
    ) +
    annotate(
      "segment",
      x = x_vals[2],
      xend = x_vals[2],
      y = f(x0),
      yend = f(x_vals[2]),
      linetype = "dotted",
      color = "steelblue"
    ) +
    annotate(
      "text",
      x = mean(x_vals[1:2]),
      y = f(x0),
      label = "-gamma * nabla * H(theta[0])",
      parse = TRUE,
      vjust = -0.7,
      size = 3,
      color = "steelblue"
    ) +
    geom_point(
      aes(x, y),
      data = tibble(x = 2, y = f(2)),
      color = "darkorange"
    ) +
    labs(y = expression(H(theta)), x = expression(theta))
}

gd_illustration(
  f = function(x) (x - 2)^2 + 1,
  f_prime = function(x) 2 * (x - 2),
  x0 = 0,
  gamma = 0.2,
  n_steps = 3
)
```
:::::
:::

--------------------------------------------------------------------------------

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: gd-illustration-small-step
#| echo: false
#| fig-width: 2.6
#| fig-height: 2.6
#| fig-cap: The step size is too small!

step_size_illustration(x, y, 0.0003)
```
:::::

\pause

::::: {.column width="47%"}
```{r}
#| label: gd-illustration-large-step
#| echo: false
#| fig-width: 2.6
#| fig-height: 2.6
#| fig-cap: The step size is too large!

step_size_illustration(x, y, 0.0115, maxit = 35)
```
:::::
:::

## Derivation of Update

Take the second-order Taylor expansion of $H$ around $\theta$:
$$
  H(\theta')
  \approx \hat{H}(\theta') = H(\theta) + \nabla H(\theta)^T (\theta' - \theta)
    + \frac{1}{2}(\theta' - \theta)^T \nabla^2 H(\theta) (\theta' - \theta).
$$

. . .

::: {.columns}
::::: {.column width="47%"}
Replace $\nabla^2 H(\theta)$ with $\frac{1}{\gamma} I$ and minimize with respect
to $\theta'$ to get
$$
  \theta_{n+1} = \theta_n - \gamma \nabla H(\theta_n).
$$

### Step Size {.alert}

But we still need to choose $\gamma$!
:::::

::::: {.column width="47%"}
```{r}
#| label: gd-taylor-illustration
#| echo: false

# show a illustration of this taylor expansion, and the fact that the
# gradient descent step is the minimizer of this taylor expansion
gd_taylor_illustration <- function(f, f_prime, x0, t) {
  x_vals <- seq(x0 - 0.5, x0 + 1.5, length.out = 100)
  df <- data.frame(
    x = x_vals,
    y = f(x_vals),
    taylor = f(x0) +
      f_prime(x0) * (x_vals - x0) +
      (x_vals - x0)^2 / (2 * t)
  )

  df_long <- df |>
    pivot_longer(
      cols = c("y", "taylor"),
      names_to = "type",
      values_to = "value"
    )

  taylor_min <- tibble(
    x = df$x[which.min(df$taylor)],
    y = min(df$taylor)
  )

  next_step <- tibble(
    x = c(x0, x0 - t * f_prime(x0)),
    y = c(f(x0), f(x0 - t * f_prime(x0)))
  )

  ggplot(df_long, aes(x)) +
    geom_line(aes(y = value, color = type, linetype = type)) +
    geom_point(
      aes(x, y),
      data = taylor_min,
      color = "steelblue"
    ) +
    geom_point(aes(x, y), data = next_step, color = "black") +
    geom_text(
      aes(x, y),
      data = next_step,
      nudge_x = -0.1,
      nudge_y = -0.35,
      label = c("theta[n]", "theta[n + 1]"),
      parse = TRUE
    ) +
    scale_color_manual(
      values = c(y = "black", taylor = "steelblue"),
      labels = c(
        y = expression(H(theta)),
        taylor = expression(hat(H)(theta))
      )
    ) +
    scale_linetype_manual(
      values = c(y = "solid", taylor = "dashed"),
      labels = c(
        y = expression(H(theta)),
        taylor = expression(hat(H)(theta))
      )
    ) +
    labs(
      y = expression(H(theta)),
      x = expression(theta),
      color = NULL,
      linetype = NULL
    ) +
    theme_void()
}

gd_taylor_illustration(
  f = function(x) (x - 2)^2 + 1,
  f_prime = function(x) 2 * (x - 2),
  x0 = 0,
  t = 0.2
)
```
:::::
:::

## The Descent Lemma (from Taylor's Theorem)

### Taylor's Theorem

Suppose $H$ is twice continuously differentiable on an open convex domain. For
any $\theta, \theta'$ in that domain, there is a $\zeta$ on the segment between
them such that
$$
  H(\theta')
  = H(\theta) + \nabla H(\theta)^T (\theta' - \theta)
    + \frac{1}{2}(\theta' - \theta)^T \nabla^2 H(\zeta) (\theta' - \theta).
$$

. . .

### Descent Lemma

Suppose we know that the Hessian is bounded above by $L I$, i.e.,
$$
  \nabla^2 H(\theta) \preceq L I \quad \text{for all } \theta.
$$

. . .

Then
$$
  H(\theta') \leq H(\theta) + \nabla H(\theta)^T (\theta' - \theta)
                  + \frac{L}{2} \|\theta' - \theta\|_2^2.
$$

## Lipschitz Constant

::: {.columns}
::::: {.column width="47%"}
On a convex domain, $L > 0$ is a **Lipschitz constant** of $\nabla H$ if, for
all $\theta, \theta'$,
$$
  \|\nabla H(\theta) - \nabla H(\theta')\|_2 \leq L \|\theta - \theta'\|_2.
$$

. . .

It measures how fast the gradient of $H$ can change, relative to the change in
$\theta$.

\pause

### Twice Differentiable Case

If $H$ is twice continuously differentiable on an open convex domain, this is
equivalent to
$$
  \|\nabla^2 H(\theta)\|_2 \leq L \quad \text{for all } \theta.
$$

For convex $H$, this reduces to $\nabla^2 H(\theta) \preceq L I$.
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: lipschitz-cones
#| echo: false
#| fig-show: animate
#| fig-height: 3.2
#| out-width: "\\linewidth"

lipschitz_illustration <- function(f_prime, L, x0, x_range) {
  # Include the anchor so the shaded cone meets the curve exactly there.
  x_vals <- sort(unique(c(
    seq(x_range[1], x_range[2], length.out = 401),
    x0
  )))

  df <- data.frame(x = x_vals, y = f_prime(x_vals))

  ggplot(df, aes(x, y)) +
    geom_ribbon(
      aes(
        ymin = f_prime(x0) - L * abs(x - x0),
        ymax = f_prime(x0) + L * abs(x - x0)
      ),
      fill = "grey50",
      alpha = 0.5
    ) +
    geom_line() +
    geom_point(
      aes(x, y),
      data = tibble(x = x0, y = f_prime(x0)),
      color = "darkorange",
      size = 2
    ) + # Fixed axes make the cone's motion visible without moving the curve.
    coord_cartesian(xlim = x_range, ylim = range(df$y)) +
    labs(y = expression(nabla * H(theta)), x = expression(theta))
}

smooth_gradient <- function(x) {
  ifelse(x < 0, 2 * x - 1, -1 / (1 + x))
}

piecewise_gradient <- function(x) {
  ifelse(
    x < 1,
    25 * x,
    ifelse(1 <= x & x < 2, x + 24, 25 * x - 24)
  )
}

smooth_anchors <- seq(-1.5, 1.5, length.out = 5)
piecewise_anchors <- seq(0, 3, length.out = 5)

for (i in seq_along(smooth_anchors)) {
  p1 <- lipschitz_illustration(
    f_prime = smooth_gradient,
    L = 2,
    x0 = smooth_anchors[i],
    x_range = c(-2, 2)
  ) +
    labs(title = expression(L == 2))

  p2 <- lipschitz_illustration(
    f_prime = piecewise_gradient,
    L = 25,
    x0 = piecewise_anchors[i],
    x_range = c(-0.5, 3.5)
  ) +
    labs(title = expression(L == 25))

  print(p1 / p2 + plot_layout(axis_titles = "collect"))
}
```
:::::
:::

## The Connection to Gradient Descent

If we apply the descent lemma with $\theta' = \theta - \gamma \nabla H(\theta)$,
we get
$$
  H(\theta - \gamma \nabla H(\theta))
  \leq H(\theta) - \gamma \|\nabla H(\theta)\|_2^2
       + \frac{L \gamma^2}{2} \|\nabla H(\theta)\|_2^2.
$$

. . .

This means that if $0 < \gamma \leq 1/L$, then
$$
  H(\theta - \gamma \nabla H(\theta))
  \leq H(\theta) - \frac{\gamma}{2} \|\nabla H(\theta)\|_2^2.
$$

. . .

Unless the gradient is zero, each step decreases the function value.

--------------------------------------------------------------------------------

```{r}
#| label: gd-illustration-right-step
#| echo: false
#| fig-width: 5
#| fig-height: 3.25
#| fig-cap: The step size is just right ($\gamma = 1/L$)!

step_size_illustration(x, y)
```

## Gradient Descent Step by Step

::: {.columns align=center}
::::: {.column width="47%"}
The dashed blue contours show the quadratic model at the current iterate.

\medskip

Minimize the model to take a step, then build a new model.
:::::

::::: {.column width="47%"}
```{r}
#| label: gd-quadratic-steps
#| echo: false
#| fig-show: animate
#| fig-height: 2.8
#| out-width: "\\linewidth"
#| fig-cap: "The gradient descent approximation."

for (iteration in 0:2) {
  for (show_next_step in c(FALSE, TRUE)) {
    stage <- if (show_next_step) {
      "Take the step"
    } else {
      "Build the model"
    }
    print(
      step_size_illustration(
        x,
        y,
        t = 0.001,
        maxit = 4,
        plot_approx = TRUE,
        iteration = iteration,
        show_next_step = show_next_step
      ) +
        labs(title = paste0("Step ", iteration + 1, ": ", stage))
    )
  }
}
```
:::::
:::

## Line Searching

In practice, we may not know $L$ or it may be difficult to compute.

. . .

### Exact Line Search

It may be tempting to try to solve
$$
  \gamma^* = \argmin_\gamma H(\theta - \gamma \nabla H(\theta)).
$$

. . .

But this is typically too expensive. Instead, we typically use an inexact line
search, like **backtracking line search**.

## Backtracking Line Search

::: {.columns}
::::: {.column width="53%"}
\begin{algorithm}[H]
  \caption{Backtracking line search}
  \KwData{Initial step size $\gamma_0 > 0$, parameters $c \in (0, 1/2]$ and $d \in (0, 1)$}
  $\gamma \gets \gamma_0$\;
  \While{$H(\theta - \gamma \nabla H(\theta)) > H(\theta) - c \gamma \|\nabla H(\theta)\|_2^2$}{
    $\gamma \gets d \gamma$\;
  }
  \Return{$\gamma$}
\end{algorithm}

\bigskip\pause

The **Armijo condition** requires a sufficient decrease in the function value.

\medskip\pause

Adapts to local curvature of the function.
:::::

::::: {.column width="41%"}
```{r}
#| label: backtracking-illustration
#| fig-height: 2.9
#| fig-width: 2.6
#| echo: false
#| fig-cap: An illustration of backtracking line search.

# show a illustration of backtracking line search
# that shows how the check work on a simple one-dimensional function
# with a tangent line and the next step
backtracking_illustration <- function() {
  H <- function(theta) (theta - 2)^2 + 1
  H_prime <- function(theta) 2 * (theta - 2)

  theta <- 0.5
  c_suff <- 0.3
  gamma <- seq(-0.2, 1, length.out = 102)

  df <- data.frame(
    gamma = gamma,
    f1 = H(theta - gamma * H_prime(theta)),
    f2 = H(theta) - gamma * H_prime(theta)^2,
    f3 = H(theta) - c_suff * gamma * H_prime(theta)^2
  )

  labels <- c(
    f1 = expression(H(theta - gamma * nabla * H(theta))),
    f2 = expression(H(theta) - gamma * nabla * H(theta)^2),
    f3 = expression(H(theta) - c * gamma * nabla * H(theta)^2)
  )

  df_long <- df |>
    pivot_longer(
      cols = c("f1", "f2", "f3"),
      names_to = "type",
      values_to = "y"
    ) |>
    filter(!(gamma < 0 & (type == "f2" | type == "f3")))

  ggplot(df_long, aes(gamma, y)) +
    geom_line(aes(color = type, linetype = type)) +
    labs(y = expression(H(gamma)), x = expression(gamma)) +
    scale_y_continuous(limits = c(-2, 5)) +
    scale_color_manual(
      values = c(
        f1 = "black",
        f2 = "darkorange",
        f3 = "steelblue"
      ),
      labels = labels
    ) +
    scale_linetype_manual(
      values = c(f1 = "solid", f2 = "dashed", f3 = "dashed"),
      labels = labels
    ) +
    labs(color = NULL, linetype = NULL) +
    guides(
      color = guide_legend(position = "inside"),
      linetype = guide_legend(position = "inside")
    ) +
    theme(legend.position.inside = c(0.6, 0.85))
}
backtracking_illustration()
```
:::::
:::

##

```{r}
#| label: gd-illustration-backtracking
#| echo: false
#| fig-width: 5
#| fig-height: 3.25
#| fig-cap: "Backtracking line search, starting each search at $\\gamma_0 = 1$."

step_size_illustration(x, y, t = 1, line_search = TRUE)
```

## Conditioning

::: {.columns}
::::: {.column width="46%"}
$L$ limits the safe step size.

\medskip\pause

For convex $H$, $L$ bounds the largest Hessian \alert{eigenvalue} and may be
pessimistic in other directions.

. . .

### Condition Number

For a positive-definite Hessian, its **condition number** is
$$
  \kappa = \frac{\lambda_{\max}}{\lambda_{\min}}.
$$

. . .

Large $\kappa$ means **ill-conditioned**; $\kappa$ near $1$ means
**well-conditioned**.

\medskip

If $\mu I \preceq \nabla^2 H(\theta) \preceq L I$ everywhere, with $\mu > 0$,
then $L/\mu$ bounds all local condition numbers.
:::::

::::: {.column width="46%"}
```{r}
#| label: gd-illustration-conditioning
#| echo: false
#| fig-height: 3.2

well_conditioned <- function(x) {
  0.5 * ((x[1] - beta[1])^2 + (x[2] - beta[2])^2)
}

ill_conditioned <- function(x) {
  0.5 * ((x[1] - beta[1])^2 + 5 * (x[2] - beta[2])^2)
}

conditioned_df <- function(f, beta1_grid, beta2_grid) {
  f_vals <- outer(
    beta1_grid,
    beta2_grid,
    Vectorize(function(b1, b2) f(c(b1, b2)))
  )

  colnames(f_vals) <- beta1_grid
  rownames(f_vals) <- beta2_grid

  tibble(
    beta1 = rep(beta1_grid, times = ncol(f_vals)),
    beta2 = rep(beta2_grid, each = nrow(f_vals)),
    f = as.vector(f_vals)
  )
}

well_df <- conditioned_df(
  well_conditioned,
  beta1_grid,
  beta2_grid
) |>
  mutate(type = "Well-conditioned (L = 1)")
ill_df <- conditioned_df(
  ill_conditioned,
  beta1_grid,
  beta2_grid
) |>
  mutate(type = "Ill-conditioned (L = 5)")

conditioned_df <- bind_rows(well_df, ill_df)

p1 <- ggplot(conditioned_df, aes(beta1, beta2)) +
  contour_scale +
  geom_point(
    data = optimum_df,
    color = "darkorange",
    size = 3,
    pch = 4
  ) +
  labs(x = expression(beta[1]), y = expression(beta[2])) +
  coord_fixed(
    xlim = c(-1, 4),
    ylim = c(0.5, 3.5),
    expand = FALSE
  ) +
  geom_contour(
    aes(z = f, color = after_stat(level)),
    bins = 35
  ) +
  facet_wrap(~type, ncol = 1)
p1
```
:::::
:::

## Convergence and Conditioning

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: gd-illustration-well-conditioned
#| fig-height: 2.5
#| echo: false
#| fig-cap: "Gradient descent on a well-conditioned problem."

x_well <- rmvnorm(100, sigma = diag(2))
y_well <- x_well %*% beta

step_size_illustration(x_well, y_well, plot_approx = TRUE)
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: gd-illustration-ill-conditioned
#| fig-height: 2.5
#| echo: false
#| fig-cap: "Gradient descent on an ill-conditioned problem."

x_ill <- rmvnorm(100, sigma = matrix(c(1, 0.8, 0.8, 1), 2, 2))
y_ill <- x_ill %*% beta

step_size_illustration(x_ill, y_ill, plot_approx = TRUE)
```
:::::
:::

\pdfpcnote{
  - How can we improve convergence in the ill-conditioned case?\\
}

## Newton's Method

Recall the second-order expansion of $H$ around $\theta$:
$$
  H(\theta')
  \approx \hat{H}(\theta') = H(\theta) + \nabla H(\theta)^T (\theta' - \theta)
    + \frac{1}{2}(\theta' - \theta)^T \nabla^2 H(\theta) (\theta' - \theta).
$$

. . .

::: {.columns}
::::: {.column width="47%"}
### Gradient Descent Update

Replace $\nabla^2 H(\theta)$ with $\frac{1}{\gamma} I$ and minimize.

. . .

### Newton Update

If $\nabla^2 H(\theta_n) \succ 0$, minimize $\hat{H}(\theta')$ directly:
$$
  \theta_{n+1} = \theta_n
                 - \big(\nabla^2 H(\theta_n)\big)^{-1} \nabla H(\theta_n).
$$

\medskip\pause

For OLS with $X$ of full column rank, the model is exact: one full Newton step
reaches the minimum.
:::::

::::: {.column width="47%"}
```{r}
#| label: newton-illustration
#| echo: false
#| fig-height: 2.4

step_size_illustration(
  x,
  y,
  plot_convergence = TRUE,
  method = "newton",
  plot_approx = TRUE,
  t = 1,
  maxit = 2
)
```
:::::
:::

## Newton's Method Beyond Quadratics

Consider the function we will use in the exercise:
$$
  H(\theta) = \log(1 + e^\theta) + \frac{1}{2}(\theta - 2)^2.
$$

. . .

```{r}
#| label: newton-nonquadratic
#| echo: false
#| fig-width: 4
#| fig-height: 2
#| fig-cap: "Newton's method on a non-quadratic function."

theta0 <- -2
theta_next <- theta0 -
  nonquadratic_gradient(theta0) / nonquadratic_hessian(theta0)

model_df <- tibble(
  theta = seq(-3, 3, length.out = 401),
  objective = nonquadratic_objective(theta),
  model = nonquadratic_model(theta, theta0)
) |>
  pivot_longer(
    cols = c(objective, model),
    names_to = "type",
    values_to = "value"
  )

step_df <- tibble(
  theta = c(theta0, theta_next),
  value = nonquadratic_model(theta, theta0),
  label = c("theta[n]", "theta[n + 1]")
)

ggplot(model_df, aes(theta, value)) +
  geom_line(aes(color = type, linetype = type)) +
  geom_point(data = step_df, color = "steelblue") +
  geom_text(
    data = step_df,
    aes(label = label),
    parse = TRUE,
    nudge_y = -0.7
  ) +
  scale_color_manual(
    values = c(objective = "black", model = "steelblue"),
    labels = c(
      objective = expression(H(theta)),
      model = expression(hat(H)(theta))
    )
  ) +
  scale_linetype_manual(
    values = c(objective = "solid", model = "dashed"),
    labels = c(
      objective = expression(H(theta)),
      model = expression(hat(H)(theta))
    )
  ) +
  labs(
    x = expression(theta),
    y = expression(H(theta)),
    color = NULL,
    linetype = NULL
  )
```

## Damped Newton

Assume $\nabla^2 H(\theta_n) \succ 0$. A full Newton step can still increase
$H$. **Damped Newton** uses
$$
  \theta_{n+1} = \theta_n - \gamma \nabla^2 H(\theta_n)^{-1} \nabla H(\theta_n).
$$
Backtrack from $\gamma = 1$ using an Armijo parameter $0 < c < 1/2$.

. . .

### Phases

- Far from the minimum, steps may be **damped** ($\gamma < 1$). \pause
- Near a minimizer with a positive-definite, locally Lipschitz Hessian, full
  steps are accepted thereafter: the **pure phase** has quadratic convergence.

\pdfpcnote{
  - Positive definiteness makes the Newton direction a descent direction.\\
  - Along $d$, Armijo checks $H(\theta + \gamma d) \leq H(\theta) + c\gamma\nabla H(\theta)^T d$.\\
  - An indefinite Hessian may need modification; backtracking alone is not enough.\\
}

##

```{r}
#| label: gd-newton-comparison
#| echo: false
#| fig-width: 5.4
#| fig-height: 3
#| fig-cap: "Gradient descent vs Newton's method on the ill-conditioned problem."

pl_gd <- step_size_illustration(
  x_ill,
  y_ill,
  method = "gd",
  plot_convergence = TRUE
) +
  ggtitle("Gradient Descent")
pl_newton <- step_size_illustration(
  x_ill,
  y_ill,
  method = "newton",
  t = 1,
  line_search = TRUE,
  plot_convergence = TRUE
) +
  ggtitle("Newton's Method")

pl_gd + pl_newton + plot_layout(guides = "collect")
```

## Comparing the Methods: Logistic Regression

For binary outcomes $y_i \in \{0, 1\}$, minimize
$$
  H(\beta)
  = \frac{1}{n} \sum_{i=1}^n \left[ \log\left( 1 + e^{x_i^\top\beta} \right) - y_i x_i^\top\beta \right]
    + \frac{\lambda}{2}\|\beta\|_2^2,
$$
with a small ridge penalty, $\lambda = 10^{-3}$.

\medskip

Unlike OLS, the Hessian changes with $\beta$.

\pause\medskip

GD uses a fixed step size $1/L$. Newton uses backtracking line search.

## Iterations and Numerical Work

```{r}
#| label: gd-newton-comparisons
#| include: false

source(here::here("R", "lecture7-comparison.R"))

comparison_basic <- compare_logistic_methods(logistic_comparison_problem(
  1500,
  300,
  design_condition = 5,
  seed = 71
))
comparison_gd <- compare_logistic_methods(logistic_comparison_problem(
  2000,
  400,
  design_condition = 2,
  seed = 72
))
comparison_newton <- compare_logistic_methods(logistic_comparison_problem(
  200,
  20,
  design_condition = 100,
  seed = 73
))

comparison_plots <- function(comparison) {
  curves <- comparison$history |>
    mutate(
      relative_gap = pmax(relative_gap, comparison$tolerance)
    )

  endpoints <- curves |>
    group_by(method) |>
    slice_tail(n = 1) |>
    ungroup()
  points <- bind_rows(
    filter(curves, method == "Newton"),
    filter(endpoints, method == "Gradient descent")
  )

  base_plot <- ggplot(
    curves,
    aes(y = relative_gap, color = method, linetype = method)
  ) +
    geom_hline(
      yintercept = comparison$tolerance,
      color = "grey60",
      linetype = "dotted"
    ) +
    geom_line() +
    geom_point(data = points, size = 1.5) +
    scale_x_continuous(
      transform = "log1p",
      breaks = c(0, 1, 10, 100, 1000),
      expand = expansion(mult = 0.03)
    ) +
    scale_y_log10(
      breaks = 10^seq(-8, 0, by = 2),
      labels = function(z) parse(text = paste0("10^", log10(z)))
    ) +
    scale_color_manual(
      values = c(
        "Gradient descent" = "royalblue",
        Newton = "darkorange"
      )
    ) +
    scale_linetype_manual(
      values = c(
        "Gradient descent" = "solid",
        Newton = "longdash"
      )
    ) +
    labs(
      y = "Relative objective gap",
      color = NULL,
      linetype = NULL
    ) +
    guides(
      color = guide_legend(position = "bottom", nrow = 1),
      linetype = guide_legend(position = "bottom", nrow = 1)
    )

  iteration_plot <- base_plot +
    aes(x = iteration) +
    labs(x = "Iterations, log(1 + x)")
  work_plot <- base_plot +
    aes(x = work) +
    labs(x = "Work (GD steps), log(1 + x)")

  list(iterations = iteration_plot, work = work_plot)
}
```

Both panels show the same runs on one simulated logistic regression problem:
1,500 observations and 300 coefficients.

::: {.columns}
::::: {.column width="48%"}
```{r}
#| label: gd-newton-basic-iterations
#| echo: false
#| fig-width: 2.6
#| fig-height: 2.25
#| out-width: "\\linewidth"
#| fig-cap: "Newton converges in a few iterations."

comparison_plots(comparison_basic)$iterations
```
:::::

. . .

::::: {.column width="48%"}
```{r}
#| label: gd-newton-basic-work
#| echo: false
#| fig-width: 2.6
#| fig-height: 2.25
#| out-width: "\\linewidth"
#| fig-cap: "But it takes more work per iteration."

comparison_plots(comparison_basic)$work
```
:::::
:::

\pdfpcnote{
  - Before revealing the work panel, ask: Which method looks faster?\\
  - Point out the dotted line: both methods reach the same accuracy target.\\
  - Reveal the work panel: same runs, only the horizontal axis changes.\\
  - One work unit is one GD step's estimated arithmetic cost, not a timing.\\
  - Newton takes fewer steps, but builds and factors a Hessian at each step.\\
  - GD wins on total work here. Fewer iterations need not mean less work.\\
}

## When Does Each Method Win?

::: {.columns}
::::: {.column width="48%"}
### Gradient Descent

2,000 observations, 400 coefficients. $\kappa(X^\top X) = 2$.

```{r}
#| label: gd-newton-gd-wins
#| echo: false
#| fig-width: 2.6
#| fig-height: 2.25
#| out-width: "\\linewidth"
#| fig-cap: "Large Hessians are costly."

comparison_plots(comparison_gd)$work
```
:::::

. . .

::::: {.column width="48%"}
### Newton's Method

200 observations, 20 coefficients. $\kappa(X^\top X) = 100$.

```{r}
#| label: gd-newton-newton-wins
#| echo: false
#| fig-width: 2.6
#| fig-height: 2.25
#| out-width: "\\linewidth"
#| fig-cap: "Curvature information pays off."

comparison_plots(comparison_newton)$work
```
:::::
:::

\pdfpcnote{
  - Ask: How should problem size and conditioning affect the winner?\\
  - Compare methods within each panel; each has its own GD work unit.\\
  - Left: good conditioning helps GD; a 400-by-400 Hessian makes Newton costly.\\
  - Right: Newton corrects for curvature with only a 20-by-20 Hessian.\\
  - The condition numbers describe the design Gram matrix; the Hessian varies.\\
  - Emphasize the tradeoff: cheap steps versus fewer, more expensive steps.\\
}

## Gradient Descent vs Newton's Method

Let $\theta^*$ be a minimizer and $e_n = \|\theta_n - \theta^*\|_2$.

::: {.columns}
::::: {.column width="47%"}
### Gradient Descent

- Cheap iterations: gradient evaluations
- Sensitive to conditioning
- **Sublinear** objective gap $O(1/n)$ for convex $H$
- **Linear** objective gap $O(q^n)$, $0 < q < 1$, if $H$ is strongly convex
- Assumes an $L$-Lipschitz gradient and $0 < \gamma \leq 1/L$
:::::

. . .

::::: {.column width="47%"}
### Newton's Method

- Costly iterations: Hessian and linear solve
- Uses curvature to rescale the step
- **Locally quadratic** with full steps: $e_{n+1} \leq C e_n^2$
- Reaches $e_n \leq \epsilon$ in $O(\log \log(1/\epsilon))$ local iterations
- Assumes a locally Lipschitz Hessian that is positive definite at $\theta^*$
:::::
:::

## Stopping Criteria

::: {.columns}
::::: {.column width="47%"}
### When to Stop?

Knowing when to stop is **important** but **hard**.

\medskip\pause

We want to stop when we are "close enough" to a minimizer, but we don't know the
minimizer in advance.
:::::

. . .

::::: {.column width="47%"}
### Common (ad hoc) choices

- Small gradient:
  $$
  \|\nabla H(\theta_n)\|_r \leq \epsilon
  $$
  \pause
- Small relative change in $\theta$:
  $$
  \frac{\|\theta_{n+1} - \theta_n\|_2}{\max\{1, \|\theta_n\|_2\}} \leq \epsilon
  $$
  \pause
- Small relative change in $H$:
  $$
  \frac{|H(\theta_{n+1}) - H(\theta_n)|}{\max\{1, |H(\theta_n)|\}} \leq \epsilon
  $$
:::::
:::

## Benchmarking: Which Method Is Faster?

```{r}
#| label: naive-optimization-benchmark
#| cache: false
#| fig-width: 5
#| fig-height: 1.8

naive_bench <- bench::mark(
  GD = gd(x, y, maxit = 20),
  Newton = newton(x, y, maxit = 20, line_search = FALSE),
  check = FALSE
)
plot(naive_bench)
```

--------------------------------------------------------------------------------

```{r}
#| label: optimization-benchmark-accuracy
#| include: false
#| cache: false

benchmark_target <- 1e-8
benchmark_paths <- list(
  GD = gd(x, y, maxit = 1000),
  Newton = newton(x, y, maxit = 20, line_search = FALSE)
)

# The noiseless OLS example has minimum zero, so the relative gap is known.
benchmark_gaps <- lapply(
  benchmark_paths,
  function(fit) fit$loss / fit$loss[1]
)
benchmark_maxit <- vapply(
  benchmark_gaps,
  function(gap) {
    reached <- which(gap <= benchmark_target)
    stopifnot(length(reached) > 0L)
    reached[1]
  },
  integer(1)
)
stopifnot(benchmark_gaps$GD[20] > benchmark_target)

# These implementations count the initial point in maxit.
benchmark_updates <- benchmark_maxit - 1L
accuracy_bench <- bench::mark(
  GD = gd(x, y, maxit = benchmark_maxit["GD"]),
  Newton = newton(
    x,
    y,
    maxit = benchmark_maxit["Newton"],
    line_search = FALSE
  ),
  check = FALSE
)

benchmark_time <- function(timing) {
  sprintf("%.3f ms", as.numeric(timing) * 1000)
}
```

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lrr}
  Comparison & GD & Newton \\
  \hline
  Median time, \texttt{maxit = 20}
    & `r benchmark_time(naive_bench$median[1])`
    & `r benchmark_time(naive_bench$median[2])` \\
  \uncover<2->{Relative objective gap}
    & \uncover<2->{`r format(benchmark_gaps$GD[20], scientific = TRUE, digits = 2)`}
    & \uncover<2->{`r format(benchmark_gaps$Newton[20], scientific = TRUE, digits = 2)`} \\
  \uncover<3->{Updates to reach a gap $\leq 10^{-8}$}
    & \uncover<3->{`r benchmark_updates["GD"]`}
    & \uncover<3->{`r benchmark_updates["Newton"]`} \\
  \uncover<3->{Median time for those updates}
    & \uncover<3->{`r benchmark_time(accuracy_bench$median[1])`}
    & \uncover<3->{`r benchmark_time(accuracy_bench$median[2])`} \\
\end{tabular}
\end{center}

\begin{overlayarea}{\textwidth}{1.2cm}
  \only<1|handout:0>{\textbf{GD is faster?}}
  \only<2|handout:0>{GD has not reached $10^{-8}$. Newton kept iterating after reaching it.}
  \only<3->{
    \textbf{Decide what counts as solving the problem before timing or plot the full trajectory.}
    \par\medskip
  }
\end{overlayarea}

\pdfpcnote{
  - Show the timings first and ask which method students would choose.\\
  - Reveal the gaps: maxit = 20 gives 19 updates, even after convergence.\\
  - One full Newton step solves this OLS problem up to numerical precision.\\
  - Reveal the common target: timing includes setup, but excludes calibration.\\
  - In practice, use a common accuracy check to decide when to stop.\\
  - check = FALSE skips output comparison; timings also depend on code and hardware.\\
}

## Exercise

Minimize
$$
  H(\theta) = \log(1 + e^\theta) + \frac{1}{2}(\theta - 2)^2.
$$

. . .

::: {.columns}
::::: {.column width="47%"}
### Steps

1. Derive $H'(\theta)$ and $H''(\theta)$.
2. Complete gradient descent with $g = H'(\theta)$ and $\gamma = 0.8$. Stop when
   $|g| \leq 10^{-8}$.
3. Replace the update with a Newton step. Which method needs fewer updates?
:::::

::::: {.column width="47%"}
```{r}
#| label: optimization-exercise-skeleton
#| eval: false

theta <- -2
path <- theta
for (i in seq_len(100)) {
  g <- ...
  if (...) {
    break
  }
  theta <- ...
  path <- c(path, theta)
}
plot(path, type = "b")
```
:::::
:::

\pdfpcnote{
  - Give students 15 minutes to derive the derivatives and complete the loop.\\
  - The path includes the starting point, so count length(path) - 1 updates.\\
  - Gradient descent needs seven updates; Newton needs four.\\
  - Ask why Newton needs fewer updates and what extra work each step requires.\\
}

## Summary

### Optimization

We talked about what an optimization problem is and what a solution looks like.

. . .

### Gradient Descent

The gradient descent method uses the gradient to take steps toward a minimizer.

. . .

### Newton's Method

The Newton method uses the Hessian to rescale the gradient and take steps toward
a minimizer.

## Next Time

### Testing

We will cover how to \alert{systematically} test your code using **testthat**.

. . .

### Debugging

We will cover how to fix bugs in your code using debugging tools in R.

\medskip\pause

This part will mostly be a hands-on session, so bring your laptop!

\lectureappendix

## Quasi-Newton Methods

Replace $\nabla^2 H(\theta)$ with an approximation $B_n$.

. . .

### BFGS

The update and direction calculation cost $O(p^2)$ per iteration, with $O(p^2)$
storage.

. . .

### L-BFGS

Store $m$ correction pairs: $O(mp)$ work per iteration and $O(mp)$ storage.

\medskip

These costs exclude objective evaluations, gradient evaluations, and line
search.
