---
title: "A Deeper Look at the EM Algorithm"
execute:
  cache: false
---

{{< include _common.qmd >}}

```{r}
#| label: setup
#| include: false

set.seed(1618)
library(patchwork)
library(testthat)
```

```{r}
#| label: moth_likelihood
#| echo: false

M <- function(y, group) as.vector(tapply(y, group, sum))

mult_likelihood <- function(
  x,
  group,
  pi_map,
  constraint = function(par) TRUE
) {
  function(par) {
    pr <- pi_map(par)
    if (!constraint(par) || any(pr > 1) || any(pr < 0)) {
      return(Inf)
    }
    -sum(x * log(M(pr, group)))
  }
}

pi_map <- function(p) {
  p[3] <- 1 - p[1] - p[2]
  c(
    p[1]^2,
    2 * p[1] * p[2],
    2 * p[1] * p[3],
    p[2]^2,
    2 * p[2] * p[3],
    p[3]^2
  )
}

e_step_mult <- function(theta, x, group) {
  p <- pi_map(theta)
  x[group] * p / M(p, group)[group]
}

m_step_mult <- function(y) {
  N <- sum(y)
  pC <- (y[1] + (y[2] + y[3]) / 2) / N
  pI <- (y[4] + (y[2] + y[5]) / 2) / N
  c(pC, pI)
}

constraint <- function(par) {
  par[1] <= 1 &&
    par[1] >= 0 &&
    par[2] <= 1 &&
    par[2] >= 0 &&
    1 - par[1] - par[2] >= 0
}

neg_loglik_obs <- mult_likelihood(
  x = c(85, 196, 341),
  group = c(1, 1, 1, 2, 2, 3),
  pi_map = pi_map,
  constraint = constraint
)

moth_optim <- optim(c(0.3, 0.3), neg_loglik_obs)
```

## Last Time

### Likelihood Optimization

We introduced the peppered moths problem and maximized the observed-data
likelihood using general-purpose and constrained optimization.

\medskip\pause

We talked about barrier methods for constrained optimization.

. . .

### The EM Algorithm

We introduced the EM algorithm for likelihood optimization in models with
missing data or latent variables.

## Recap: The EM Algorithm

The EM algorithm iteratively optimizes
$$
  \theta_{n+1} = \argmax_{\theta \in \Theta} Q(\theta \mid \theta_n),
$$
with
$$
  Q(\theta \mid \theta_n)
  = \E_{\theta_n}\big(\log p(X, Z \mid \theta) \mid X\big).
$$

\medskip\pause

Each iteration is guaranteed not to decrease the observed-data log-likelihood.

## Today

### More on the EM Algorithm

We tackle the peppered moths example with the EM algorithm.

. . .

### Fisher Information

After we have optimized the likelihood, we often want to estimate the variance
of our estimators. Fisher information helps us do this.

. . .

### Convergence and Generalized EM

We connect the EM convergence rate to missing information and introduce the
generalized EM algorithm.

## Recap: The Peppered Moths

::: {.columns}
::::: {.column width="47%"}
### Alleles

C, I, and T with frequencies $p_C$, $p_I$, and $p_T$, where
$$
  p_C + p_I + p_T = 1.
$$

### Genotypes

CC, CI, CT, II, IT, TT

### Phenotypes

Black, Mottled, Light-Colored
:::::

::::: {.column width="47%"}
```{r}
#| label: moth-genotype-phenotype
#| echo: false
#| fig-width: 2.8
#| fig-height: 1.6

alleles <- c("C", "I", "T")
moth_genotypes <- tidyr::expand_grid(
  row = alleles,
  column = alleles
) |>
  dplyr::mutate(
    genotype = paste0(pmin(row, column), pmax(row, column)),
    phenotype = case_when(
      row == "C" | column == "C" ~ "Black",
      row == "I" | column == "I" ~ "Mottled",
      .default = "Light-colored"
    )
  )

ggplot(moth_genotypes, aes(column, row, fill = phenotype)) +
  geom_tile(color = "white") +
  geom_text(
    aes(
      label = genotype,
      color = if_else(phenotype == "Black", "white", "black")
    ),
    size = 3
  ) +
  scale_x_discrete(
    limits = alleles,
    position = "top",
    expand = c(0, 0)
  ) +
  scale_y_discrete(limits = rev(alleles), expand = c(0, 0)) +
  scale_fill_manual(
    name = "Phenotype",
    values = c(
      "Black" = "grey20",
      "Mottled" = "darkorange",
      "Light-colored" = "grey90"
    ),
    breaks = c("Black", "Mottled", "Light-colored")
  ) +
  scale_color_identity() +
  coord_fixed() +
  labs(x = NULL, y = NULL)
```
:::::
:::

## Latent Variables

### Genotype Labels

Let each moth $i = 1,\ldots,N$ have an (unobserved) genotype label
$$
  Z_i \in \{1,2,3,4,5,6\} \quad\text{(CC, CI, CT, II, IT, TT)}.
$$

. . .

Via the Hardy--Weinberg principle, the **genotype** probabilities are
$$
  p = (p_C^2,\ 2p_Cp_I,\ 2p_Cp_T,\ p_I^2,\ 2p_Ip_T,\ p_T^2).
$$

. . .

### Genotype Counts

The genotype counts $Y_k$ are sufficient statistics for the complete data. They
are also latent, with
$$
  Y_k = \sum_{i=1}^N \mathbf{1}(Z_i = k),
  \quad Y = (Y_1,\ldots,Y_6), \quad \sum_k Y_k = N.
$$

## Observed Data

What we actually observe are the **phenotype** counts
$$
  X_j = \sum_{k \in A_j} Y_k,\quad X = M(Y),
$$
where the genotype categories are partitioned into
$$
  A_1 = \{1,2,3\}\ (\text{Black}),\quad A_2 = \{4,5\}\ (\text{Mottled}),
  \quad A_3 = \{6\}\ (\text{Light}).
$$

. . .

Here, $M : \mathbb{R}^6 \to \mathbb{R}^3$ is the cell-collapsing map
$$
  M(y) = (y_1 + y_2 + y_3,\ y_4 + y_5,\ y_6).
$$

. . .

### Multinomial Distribution

If $Y \sim \textrm{Mult}(N, p)$ with $p = (p_1, \ldots, p_6)$, then
$$
  X = M(Y) \sim \textrm{Mult}(N, M(p)).
$$

## Complete-Data Log-Likelihood

The complete-data log-likelihood, with $c(Y)$ independent of $p$, is
$$
  \ell(p \mid Y) = \sum_{k=1}^6 Y_k \log p_k + c(Y).
$$

. . .

### Optimization Variable

For our EM algorithm, we will optimize over the **allele frequencies**
$$
  \theta = (p_C, p_I), \qquad p_T = 1 - p_C - p_I.
$$

\medskip\pause

Write $\pi(\theta)$ for the genotype probabilities given above, with
$p_T = 1 - p_C - p_I$.

## Multinomial Conditional Distributions

Given phenotype counts $X$ and current allele frequencies $\theta'$, the
genotype counts $Y_{A_j} = (Y_k)_{k \in A_j}$ within phenotype $j$ satisfy
$$
  Y_{A_j} \mid X, \theta'
  \sim \textrm{Mult}\left( X_j, \left( \frac{\pi_k(\theta')}{M(\pi(\theta'))_j} \right)_{k \in A_j} \right).
$$

\medskip\pause

### Expected Genotype Counts

Hence for $k \in A_j$,
$$
  \E_{Y \mid X, \theta'}(Y_k) = \frac{X_j \pi_k(\theta')}{M(\pi(\theta'))_j}.
$$

\medskip

The expected counts divide each phenotype's observed count among its compatible
genotypes in proportion to their current probabilities.

## The Q Function

EM $Q$-function:
$$
  \begin{aligned}
    Q(\theta \mid \theta')
     & = \E_{Y \mid X, \theta'} \log p (Y \mid \theta)                                                        \\
    \pause
     & = \sum_{k=1}^6 \E_{Y \mid X, \theta'}(Y_k) \log \pi_k(\theta) + c(\theta')                             \\
    \pause
     & = \sum_{k=1}^6 \frac{X_{j(k)} \pi_k(\theta')}{M(\pi(\theta'))_{j(k)}} \log \pi_k(\theta) + c(\theta').
  \end{aligned}
$$

. . .

Here, $k \in A_{j(k)}$. The expected counts are computed at $\theta'$ and held
fixed while we maximize over $\theta$. The term
$c(\theta') = \E_{Y \mid X,\theta'}c(Y)$ does not affect this maximization.

\pdfpcnote{
  - The E-step computes expected genotype counts using the current allele frequencies.\\
}

## M-Step for the Multinomial Model

We need to maximize $Q(\theta \mid \theta')$ w.r.t. $\theta$.

\medskip\pause

With $y = (n_{CC}, n_{CI}, n_{CT}, n_{II}, n_{IT}, n_{TT})^T$ a complete
observation, it can be shown that the MLE is
$$
  \begin{aligned}
    \hat{p}_C & = \frac{n_{CC} + (n_{CI} + n_{CT}) / 2}{N} \\
    \hat{p}_I & = \frac{n_{II} + (n_{CI} + n_{IT}) / 2}{N}
  \end{aligned}
$$
where $N = \symbf{1}^T y$ is the total number of observations.

In the M-step, replace $y$ by its conditional expectation from the E-step.

## EM Algorithm Implementation

```{r}
#| label: em-mult

em_mult <- function(
  par,
  x,
  group,
  eps = 1e-12,
  maxit = 20,
  cb = NULL
) {
  for (i in seq_len(maxit)) {
    par0 <- par
    y <- e_step_mult(par, x, group)
    par <- m_step_mult(y)

    if (!is.null(cb)) {
      cb() # Record the current state for later inspection.
    }

    if (sum((par - par0)^2) <= eps * (sum(par^2) + eps)) {
      break
    }
  }
  par
}
```

## Peppered Moths EM Algorithm

The observed counts are 85 black, 196 mottled, and 341 light-colored moths.

```{r}
#| label: moth-em-data
#| results: "hold"

moth_x <- c(85, 196, 341)
moth_group <- c(1, 1, 1, 2, 2, 3)
```

. . .

::: {.columns}
::::: {.column width="47%"}
Then we can run the EM algorithm.

```{r}
#| label: fit-moth-em

em_mult(c(0.3, 0.3), moth_x, moth_group)
```
:::::

. . .

::::: {.column width="47%"}
This corresponds to the MLE found via `optim()`.

```{r}
#| label: moth-optim-estimate

moth_optim$par
```
:::::
:::

## Inside the EM Algorithm

The `tracer()` function in the `CSwR` package can be used to trace the
iterations of an algorithm.^[Use `Delta` to control verbosity. `0`: no output.]

. . .

```{r}
#| label: initialize-moth-tracer

library(CSwR)
em_tracer <- tracer("par", Delta = 0)
```

. . .

Each call to `tracer()` returns an S3 object of class `tracer`. The object has a
field `tracer`, which is a function that will record the current value of given
parameters each time it is called.

. . .

```{r}
#| label: run-moth-tracer

em_mult(c(0.3, 0.3), moth_x, moth_group, cb = em_tracer$tracer)
```

##

You can then summarize the collected values using `summary()`.

```{r}
#| label: summarize-moth-tracer

summary(em_tracer)
```

## Tracing the EM Algorithm

```{r}
#| label: em-tracer

em_tracer <- tracer(c("par0", "par"), Delta = 0)
p_hat <- em_mult(
  c(0.3, 0.3),
  moth_x,
  moth_group,
  eps = 0,
  cb = em_tracer$tracer
)
p_hat

em_trace <- summary(em_tracer)
```

##

```{r}
#| label: plot-curves
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "EM iterations for the peppered moths example. The orange cross marks
#|   the MLE. The contours show the negative Q function at each iteration, up to an
#|   additive constant. The black dot marks its minimum, the next EM iterate."

nll_pep <- Vectorize(function(p1, p2) neg_loglik_obs(c(p1, p2)))

em_trace <- transform(
  em_trace,
  n = seq_len(nrow(em_trace)),
  par_norm_diff = sqrt((par0.1 - par.1)^2 + (par0.2 - par.2)^2),
  nll = nll_pep(par.1, par.2)
)

group <- c(1, 1, 1, 2, 2, 3)

EStep <- function(theta) e_step_mult(theta, x, group)
MStep <- function(y) m_step_mult(y)

eps <- 1e-6
maxit <- 5

exp_vals <- matrix(NA, nrow = maxit, ncol = length(group))

par <- c(0.3, 0.3)

level_curves <- vector("list", length = maxit)

ll <- double(maxit)
x <- c(85, 196, 341)

n_mat <- 100
pC <- seq(0, 1, length.out = n_mat)
pI <- seq(0, 1, length.out = n_mat)

pars <- vector("list", length = maxit)

for (i in seq_len(maxit)) {
  par0 <- par
  y <- EStep(par)

  x_new <- M(y, group)
  z <- matrix(NA, n_mat, n_mat)

  x <- c(85, 196, 341)

  for (k in seq_len(n_mat)) {
    for (j in seq_len(n_mat)) {
      val <- -sum(y * log(pi_map(c(pC[k], pI[j]))))
      if (is.finite(val)) {
        z[k, j] <- val
      } else {
        z[k, j] <- -Inf
      }
    }
  }

  level_curves[[i]] <- z

  ll[i] <- -sum(y * log(pi_map(par)))

  exp_vals[i, ] <- y
  par <- MStep(y)
  pars[[i]] <- par
}

par_optim <- c(0.07084, 0.18874)

# Build long data frame for first 4 iterations
build_df <- function(z_mat, iter) {
  out <- data.frame(
    pC = rep(pC, times = length(pI)),
    pI = rep(pI, each = length(pC)),
    z = as.vector(z_mat),
    iter = iter
  )
  out
}

plot_iters <- 1:4
df_all <- do.call(
  rbind,
  lapply(plot_iters, function(i) build_df(level_curves[[i]], i))
)

# Points (current parameter in each iteration)
iter_points <- data.frame(
  pC = sapply(pars[plot_iters], `[[`, 1),
  pI = sapply(pars[plot_iters], `[[`, 2),
  iter = plot_iters
)

mle_point <- data.frame(pC = par_optim[1], pI = par_optim[2])

ggplot(df_all, aes(pC, pI)) +
  geom_contour(
    aes(z = z, color = after_stat(level)),
    breaks = seq(800, 3000, by = 200)
  ) +
  geom_point(
    data = mle_point,
    colour = "darkorange",
    size = 3,
    pch = 4
  ) +
  geom_point(data = iter_points, colour = "black", size = 2) +
  facet_wrap(
    ~iter,
    labeller = labeller(iter = function(i) paste0("k = ", i))
  ) +
  coord_equal() +
  guides(color = "none") +
  labs(x = expression(p[C]), y = expression(p[I])) +
  theme(panel.spacing = unit(0.8, "lines"))
```

##

```{r}
#| label: tracer-plot1
#| echo: false
#| fig-height: 2.7
#| fig-width: 3.5
#| fig-cap: "Parameter convergence of the EM algorithm applied to the peppered
#|   moths problem."

ggplot(em_trace, aes(n, par_norm_diff)) +
  geom_point() +
  geom_line() +
  labs(
    y = expression(group("||", theta[n] - theta[n - 1], "||"))
  ) +
  scale_y_log10()
```

##

```{r}
#| label: loglikelihood-convergence
#| warning: false
#| echo: false
#| fig-cap: "Negative log-likelihood gap relative to the best recorded iterate.
#|   An offset of 1e-14 keeps zero gaps visible on the logarithmic scale."
#| fig-width: 4
#| fig-height: 2.5

mutate(em_trace, diff_nll = 1e-14 + nll - min(nll)) |>
  ggplot(aes(n, diff_nll)) +
  labs(
    y = expression(
      max[k] * italic(l)(theta[k]) - italic(l)(theta[n])
    )
  ) +
  geom_point() +
  geom_line() +
  scale_y_log10()
```

## Multi-Start

A common problem in likelihood optimization is local maxima.

\medskip\pause

Trying a grid of starting values can help, but does not guarantee a global
maximum.^[The function `neg_loglik_obs()` computes the negative observed
log-likelihood.]

```{r}
#| label: moth-em-multistart

theta_grid <- expand.grid(
  pC = seq(0.01, 0.49, length.out = 5),
  pI = seq(0.01, 0.49, length.out = 5)
)

res <- apply(
  theta_grid,
  1,
  em_mult,
  x = moth_x,
  group = moth_group
)
neg_logliks <- apply(res, 2, neg_loglik_obs)

which.min(neg_logliks)
```

# Fisher Information

## The Fisher Information

When doing statistical inference, we need to measure the uncertainty of
estimates.

\medskip\pause

The Fisher information is a key quantity for this. It is defined as
$$
  \mathcal{I}(\theta) = -\E_\theta\left( D_\theta^2 \ell(\theta) \right),
$$
where $\ell(\theta) = \log p(X \mid \theta)$.

. . .

This is the expected negative Hessian of the log-likelihood.

### Complete, Observed, and Missing Information

In the presence of latent variables, we can define three types of Fisher
information:

- **Complete-data information** $\mathcal{I}_Y(\theta)$
- **Observed-data information** $\mathcal{I}_X(\theta)$
- **Missing information**
  $\mathcal{I}_{Y \mid X}(\theta) = \mathcal{I}_Y(\theta) - \mathcal{I}_X(\theta)$

```{r}
#| label: moth-probability-gradient
#| echo: false

pi_grad <- function(p) {
  matrix(
    c(
      2 * p[1],
      0,
      2 * p[2],
      2 * p[1],
      2 * p[3] - 2 * p[1],
      -2 * p[1],
      0,
      2 * p[2],
      -2 * p[2],
      2 * p[3] - 2 * p[2],
      -2 * p[3],
      -2 * p[3]
    ),
    ncol = 2,
    nrow = 6,
    byrow = TRUE
  )
}
```

## Observed Fisher Information

The observed information is the negative Hessian of the observed-data
log-likelihood:
$$
  \hat{i}_X(\theta) = -D_\theta^2 \ell(\theta).
$$

At the MLE, it estimates the expected Fisher information
$\mathcal{I}_X(\theta)$.

\medskip\pause

For the moth data, we already have the observed likelihood. We can differentiate
it directly, or recover the same information from the EM $Q$ function.

## Directly from the Observed Likelihood

Since `neg_loglik_obs` is the negative observed-data log-likelihood, its Hessian
at the MLE is the observed information.

```{r}
#| label: observed-information-direct

library(numDeriv)

theta_hat <- em_mult(
  c(0.3, 0.3),
  moth_x,
  moth_group,
  eps = 1e-20
)
iX_direct <- hessian(neg_loglik_obs, theta_hat)
iX_direct
```

\medskip\pause

The inverse of this matrix estimates the covariance matrix of $\hat{\theta}$.

## The Information Identity

Recall that $Q(\theta \mid \theta')$ averages the complete-data log-likelihood
over the missing genotypes. Define
$$
  H(\theta \mid \theta') = -\E_{Y \mid X,\theta'}\log p(Y \mid X,\theta).
$$
Then $\ell(\theta) = Q(\theta \mid \theta') + H(\theta \mid \theta')$.

\medskip\pause

At the MLE, this yields
$$
  \hat{i}_X
  = \underbrace{-D_\theta^2 Q(\hat{\theta} \mid \hat{\theta})}_{\hat{i}_Y}
    - \underbrace{D_\theta^2 H(\hat{\theta} \mid \hat{\theta})}_{\hat{i}_{Y \mid X}}.
$$

\medskip\pause

The missing-information term measures how much information we lose because the
genotypes are unobserved.

## Missing Information from $Q$

We can compute the missing-information term by differentiating $Q$ with respect
to both arguments:
$$
  \hat{i}_{Y \mid X} = D_{\theta'}D_\theta Q(\hat{\theta} \mid \hat{\theta}).
$$

\medskip\pause

The information identity therefore gives a second calculation:
$$
  \hat{i}_X = -D_\theta^2 Q(\hat{\theta} \mid \hat{\theta})
              - D_{\theta'}D_\theta Q(\hat{\theta} \mid \hat{\theta}).
$$

\medskip\pause

The first derivative treats the current parameter $\theta'$ as fixed; the mixed
derivative also accounts for how the E-step changes with $\theta'$.

## Implementation of $Q$

First we implement $Q$ as an R function, omitting $c(\theta')$ because it does
not affect derivatives with respect to the first argument.

```{r}
#| label: q-def

Q <- function(theta, theta0, x = c(85, 196, 341)) {
  theta[3] <- 1 - theta[1] - theta[2]
  theta0[3] <- 1 - theta0[1] - theta0[2]
  group <- c(1, 1, 1, 2, 2, 3)
  (x[group] *
    pi_map(theta0) /
    M(pi_map(theta0), group)[group]) %*%
    log(pi_map(theta))
}
```

## Gradient of $Q$

The mixed derivative requires the gradient of $Q$ with respect to its first
argument:

```{r}
#| label: gradq-def

Q_grad <- function(theta, theta0, x = c(85, 196, 341)) {
  theta[3] <- 1 - theta[1] - theta[2]
  group <- c(1, 1, 1, 2, 2, 3)
  (x[group] *
    pi_map(theta0) /
    (M(pi_map(theta0), group)[group] * pi_map(theta))) %*%
    pi_grad(theta)
}
```

. . .

`pi_grad()` is the Jacobian of the genotype probabilities w.r.t. allele
frequencies (defined in source code).

## Numerical Differentiation of $Q$

```{r}
#| label: q-hess

iY <- -hessian(Q, theta_hat, theta0 = theta_hat)
iY
```

. . .

```{r}
#| label: iyx-jacobian

iYX <- jacobian(
  function(theta) Q_grad(theta_hat, theta),
  theta_hat
)
iX_em <- iY - iYX
iX_em
```

```{r}
#| label: check-observed-information
#| include: false

test_that("Observed-information calculations agree", {
  expect_equal(iX_em, iX_direct, tolerance = 1e-5)
})
```

## Computational Cost

The direct route needs a numerical Hessian of the observed likelihood. The EM
route needs a numerical Hessian of $Q$ and a mixed derivative.

\medskip\pause

```{r}
#| label: benchmark-observed-information
#| echo: false

information_benchmark <- bench::mark(
  direct = hessian(neg_loglik_obs, theta_hat),
  via_em = {
    complete <- -hessian(Q, theta_hat, theta0 = theta_hat)
    missing <- jacobian(
      function(theta) Q_grad(theta_hat, theta),
      theta_hat
    )
    complete - missing
  },
  check = FALSE,
  iterations = 100
)
information_benchmark[, c("expression", "median")]
```

\medskip\pause

With $d$ parameters, both numerical Hessians require on the order of $d^2$
function evaluations. The EM route also differentiates the gradient of $Q$.
Their relative cost depends on how expensive those evaluations are.

##

```{r}
#| label: fisher-ellipses
#| warning: false
#| message: false
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "Approximate 95% confidence ellipse using observed information (solid),
#|   with a complete-data information ellipse (dashed) for comparison.
#|   The cross marks the MLE for the peppered moths example."

alpha <- 0.05

ellipse_points <- function(I, center, level = 0.95, n = 200) {
  Sigma <- solve(I)
  eig <- eigen(Sigma, symmetric = TRUE)
  A <- eig$vectors %*% diag(sqrt(pmax(eig$values, 0)))
  r <- sqrt(qchisq(level, df = 2))
  t <- seq(0, 2 * pi, length.out = n)
  pts <- t(sapply(
    t,
    function(phi) center + r * A %*% c(cos(phi), sin(phi))
  ))
  colnames(pts) <- c("pC", "pI")
  as.data.frame(pts)
}

clip_simplex <- function(df) {
  subset(df, pC >= 0 & pI >= 0 & pC + pI <= 1)
}

df_obs <- clip_simplex(transform(
  ellipse_points(iX_direct, theta_hat, 1 - alpha),
  Type = "Observed (i_X)"
))
df_comp <- clip_simplex(transform(
  ellipse_points(iY, theta_hat, 1 - alpha),
  Type = "Complete (i_Y)"
))

simplex_boundary <- data.frame(
  pC = c(0, 1, 0, 0),
  pI = c(0, 0, 1, 0)
)

theta_df <- data.frame(pC = theta_hat[1], pI = theta_hat[2])

ggplot() +
  geom_path(data = df_obs, aes(pC, pI, color = Type)) +
  geom_path(
    data = df_comp,
    aes(pC, pI, color = Type),
    linetype = 2
  ) +
  geom_point(
    data = theta_df,
    aes(pC, pI),
    color = "darkorange",
    pch = 4,
    size = 3
  ) +
  coord_equal() +
  lims(x = c(0, 0.15), y = c(0.1, 0.25)) +
  scale_color_manual(
    values = c(
      "Observed (i_X)" = "#1b6ca8",
      "Complete (i_Y)" = "#d95f02"
    ),
    breaks = c("Observed (i_X)", "Complete (i_Y)"),
    labels = c(expression(hat(i)[X]), expression(hat(i)[Y]))
  ) +
  labs(
    x = expression(p[C]),
    y = expression(p[I]),
    color = "Ellipse"
  )
```

## Further Reading

The course book covers two additional routes: Fisher's score identity and
differentiation of the EM mapping.

## Convergence of the EM Algorithm

We have already shown that the EM algorithm monotonically increases the observed
data log-likelihood:
$$
  \ell(\theta_{n+1}) \ge \ell(\theta_n).
$$

. . .

Parameter convergence requires additional assumptions. If the iterates converge
and the EM mapping $\Phi$ is continuous at their limit, then
$$
  \theta_{n+1} = \Phi(\theta_n), \quad \hat{\theta} = \Phi(\hat{\theta}).
$$

. . .

By the Taylor expansion of $\Phi$ around $\hat{\theta}$, we have
$$
  \theta_{n+1}
  - \hat{\theta} \approx D\Phi(\hat{\theta}) (\theta_n - \hat{\theta}).
$$

. . .

So the error at iteration $n + 1$ is approximately a linear transformation of
the error at iteration $n$.

\medskip\pause

And the rate of convergence is determined by the matrix $D\Phi(\hat{\theta})$.

## Convergence: Connection to Fisher Information

Let $\rho(A)$ be the spectral radius of a matrix $A$, i.e., the largest absolute
value of its eigenvalues.

\medskip\pause

This means that the convergence rate of the EM algorithm is governed by
$$
  \rho(D\Phi(\hat{\theta}))
  = \rho\left( \hat{i}_{Y \mid X} \left( \hat{i}_Y \right)^{-1} \right).
$$

\medskip\pause

If $\rho(D\Phi(\hat{\theta}))$ is close to 0, the EM algorithm converges
quickly, and if it is close to 1, the EM algorithm converges slowly.

. . .

### Relation to Missing Information

If the missing information is small **relative to the complete-data
information**, then $\rho(D\Phi(\hat{\theta}))$ is small and EM converges
quickly.

## Convergence Estimation

Near a regular optimum, EM typically converges linearly. If the dominant error
component is present, then asymptotically
$$
  \|\theta_{n+1} - \hat{\theta}\| \approx \rho \|\theta_n - \hat{\theta}\|,
  \qquad
  0 < \rho < 1.
$$

\medskip\pause

The rate can be estimated from successive parameter differences, which typically
decay at the same asymptotic rate:
$$
  \log \|\theta_n - \theta_{n-1}\| \approx n \log \rho + \text{constant}.
$$

. . .

```{r}
#| label: linear-convergence

log_lm <- lm(log(par_norm_diff) ~ n, data = em_trace)
exp(coefficients(log_lm)["n"])
```

\medskip\pause

It is small in this case, implying fast convergence.

## The Generalized EM Algorithm

::: {.columns}
::::: {.column width="47%"}
### Key Idea

Relax the M-step: instead of fully maximizing $Q(\theta \mid \theta_n)$, we just
require an **increase**:
$$
  Q(\theta_{n+1} \mid \theta_n) \ge Q(\theta_n \mid \theta_n).
$$

. . .

Full maximization may be **hard or computationally expensive**. GEM allows
**simpler updates** (e.g., a gradient step or a coordinate-wise update).

. . .
:::::

::::: {.column width="47%"}
### Properties

The E-step is unchanged, and the **log-likelihood is still nondecreasing**:
$$
  \ell(\theta_{n+1}) \ge \ell(\theta_n).
$$

\medskip\pause

Convergence to a **stationary point** requires additional regularity and
progress conditions; a nondecreasing likelihood alone is not enough.

. . .

### Trade-Off

Cheaper iterations may require more iterations to converge.
:::::
:::

## A Mixture of Two Gaussians

::: {.columns align="center"}
::::: {.column width="47%"}
### Goal

First fit a two-component Gaussian mixture model by ordinary EM.

. . .

### Model

We assume that the data $X_i \sim \operatorname{\mathcal{N}}(\mu_k, \sigma_k^2)$
with probability $\pi_k$, $k = 1, 2$.
:::::

::::: {.column width="47%"}
```{r}
#| label: gaussian-mixture-density
#| echo: false
#| fig-cap: "A mixture of two Gaussian distributions."

gausdens <- function(x, mu1, mu2, sigma1, sigma2, p) {
  (p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
    (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2) /
    sqrt(2 * pi)
}

xx <- seq(-3, 11, 0.1)

sigma1 <- 1
sigma2 <- 2
mu1 <- -0.5
mu2 <- 4
p <- 0.5

il_df <- data.frame(
  x = xx,
  y1 = dnorm(xx, mu1, sigma1) * p,
  y2 = dnorm(xx, mu2, sigma2) * (1 - p),
  ymix = gausdens(xx, mu1, mu2, sigma1, sigma2, p)
)

pl <- ggplot(il_df, aes(x, ymix)) +
  geom_line() +
  geom_line(
    aes(y = y1),
    linetype = "dashed",
    color = "steelblue"
  ) +
  geom_line(
    aes(y = y2),
    linetype = "dashed",
    color = "darkorange"
  ) +
  labs(y = "Density", x = "x")

pl
```
:::::
:::

## Finite Mixtures

In a finite mixture, the marginal density of $X$ is
$$
  p(x \mid \psi) = \sum_{k=1}^K \pi_k f_k(x \mid \psi_k)
$$
where $\pi_k \geq 0$, $\sum_k \pi_k = 1$, and $f_k$ are densities.

\medskip\pause

### Latent Variable

Assume each $X_i$ comes from one of the $K$ components, but we don't know which
one.

\medskip\pause

We introduce a latent variable $Z_i \in \{1, \ldots, K\}$ such that
$$
  \Pr(Z_i = k) = \pi_k.
$$

. . .

### Complete-Data Model

Then the complete-data model is
$$
  p(x_i, z_i \mid \psi) = \pi_{z_i} f_{z_i}(x_i \mid \psi_{z_i}).
$$

## Gaussian Mixture

For Gaussian mixtures, the components are Gaussian:
$$
  f_k(x \mid \mu_k, \sigma_k^2)
  = \frac{1}{\sqrt{2 \pi \sigma_k^2}} e^{-\frac{(x-\mu_k)^2}{2 \sigma_k^2}}.
$$
with $\psi_k = (\mu_k, \sigma_k^2)$.

\medskip\pause

### Complete-Data Log-Likelihood

Letting $\theta = (\pi_1, \ldots, \pi_K, \psi_1, \ldots, \psi_K)$, the
complete-data log-likelihood is
$$
  \begin{aligned}
    \ell_c (\theta) & = \sum_{i=1}^N \log p(x_i, z_i \mid \theta)                                                                     \\
    \pause
                    & = \sum_{i=1}^N \log \left( f(x_i \mid z_i, \theta) f(z_i \mid \theta) \right)                                   \\
    \pause
                    & = \sum_{i=1}^N \sum_{k=1}^K \symbf{1}(Z_i = k) \left( \log \pi_k + \log f_k(x_i \mid \mu_k, \sigma_k^2) \right)
  \end{aligned}
$$

## Simple Case: Fixed Component Variances, $K = 2$

::: {.columns align="center"}
::::: {.column width="47%"}
Let's simplify and assume $K = 2$ and that $\sigma_1$ and $\sigma_2$ are known.

\medskip\pause

Then the model has three unknown parameters:
$$
  \theta = (\pi, \mu_1, \mu_2).
$$
:::::

::::: {.column width="47%"}
\begin{tikzpicture}[
    node distance=1.6cm,
    latent/.style={circle,draw,minimum size=10mm,inner sep=0pt},
    observed/.style={circle,draw,fill=gray!20,minimum size=10mm,inner sep=0pt},
    param/.style={rectangle,draw,minimum size=7mm,inner sep=2pt},
  ]

  % Parameters for components
  \node[param] (pi) {$\pi$};
  \node[param, right=2.5cm of pi, yshift=1cm] (mu1) {$\mu_1$};
  \node[param, right=2.5cm of pi, yshift=-1cm] (mu2) {$\mu_2$};

  % Latent and observed variables
  \node[latent, below=1.8cm of pi] (z) {$Z_i$};
  \node[observed, below=1.8cm of z] (x) {$X_i$};

  % Arrows
  \draw[->] (pi) -- (z);
  \draw[->] (z) -- (x);
  \draw[->] (mu1) -- (x);
  \draw[->] (mu2) -- (x);

  % Plate for data points
  \node[plate={below:{$i = 1,\dots,N$}}, fit=(z)(x)] (plateN) {};
\end{tikzpicture}
:::::
:::

## The E-Step

The complete-data log-likelihood is now
$$
  \begin{aligned}
    \ell_c(\theta) & = \sum_{i=1}^N \symbf{1}(Z_i = 1) \left( \log(\pi) - \frac{(x_i - \mu_1)^2}{2 \sigma_1^2} \right)                       \\
                   & \phantom{ = {}} + \symbf{1}(Z_i = 2)\left( \log(1 - \pi) - \frac{(x_i - \mu_2)^2}{2 \sigma_2^2} \right) + \text{const}.
  \end{aligned}
$$

. . .

Taking expectations gives
$$
  Q(\theta \mid \theta')
  = \sum_{i=1}^N \tau_{i} \left( \log(\pi) - \frac{(x_i - \mu_1)^2}{2 \sigma_1^2} \right)
    + (1 - \tau_{i})\left( \log(1 - \pi) - \frac{(x_i - \mu_2)^2}{2 \sigma_2^2} \right)
$$
where $\tau_i = \Pr(Z_i = 1 \mid X_i = x_i, \theta')$ is the **responsibility**
of component 1 for observation $i$.

\medskip\pause

The responsibility tells us how responsible component 1 is for generating $x_i$.

## The Responsibilities

The conditional probability in a mixture model is generally
$$
  \Pr(Z = z \mid X = x)
  = \frac{\pi_z f_z(x \mid \psi_z)}{\sum_{k=1}^K \pi_k f_k(x \mid \psi_k)}.
$$

\medskip\pause

For the $K = 2$ Gaussian case, this gives us
$$
  \tau_i
  = \Pr(Z_i = 1 \mid X = x_i, \theta')
  = \frac{\pi'_1 e^{-\frac{(x_i-\mu_1')^2}{2 \sigma_1^2}}}{\left( \pi'_1 e^{-\frac{(x_i-\mu_1')^2}{2 \sigma_1^2}} + \frac{\sigma_1}{\sigma_2} (1 - \pi'_1) e^{-\frac{(x_i-\mu_2')^2}{2 \sigma_2^2}} \right)}.
$$

## The Q-Step

The MLEs of $\theta$ given $Q(\theta \mid \theta')$ are
$$
  \hat{\theta}
  = \left( \frac{1}{N} \sum_{i=1}^N \tau_i, \quad \frac{1}{\sum_{i=1}^N \tau_i} \sum_{i=1}^N \tau_i x_i, \quad \frac{1}{\sum_{i=1}^N (1 - \tau_i)} \sum_{i=1}^N (1 - \tau_i) x_i \right).
$$

. . .

::: {.columns}
::::: {.column width="47%"}
### Updating $\pi$

We update $\pi$ as
$$
  \hat{\pi} = \frac{1}{N} \sum_{i=1}^N \tau_i,
$$
which is the fraction of all data points "assigned" to component 1.
:::::

. . .

::::: {.column width="47%"}
### Updating $\mu_1$ and $\mu_2$

We update $\mu_1$ as
$$
  \hat{\mu}_1 = \frac{\sum_{i=1}^N \tau_i x_i}{\sum_{i=1}^N \tau_i},
$$
and similarly for $\mu_2$.

\medskip\pause

This is a weighted mean, weighted by how much each component is responsible for
each observation.
:::::
:::

## Example: Simulation

We start by setting the parameters.

```{r}
#| label: mixture-gaus-sim

sigma1 <- 1
sigma2 <- 2
mu1 <- -0.5
mu2 <- 4
p <- 0.5
```

. . .

Then we simulate data.

```{r}
#| label: simulate-gaussian-mixture

N <- 5000
z <- sample(
  c(TRUE, FALSE),
  N,
  replace = TRUE,
  prob = c(p, 1 - p)
)
x <- numeric(N)
N1 <- sum(z)
x[z] <- rnorm(N1, mu1, sigma1)
x[!z] <- rnorm(N - N1, mu2, sigma2)
```

##

```{r}
#| label: gaussian-mixture-histogram
#| echo: false
#| fig-cap: "Histogram of simulated data from a two-component Gaussian mixture model."
#| fig-width: 4
#| fig-height: 2.7

pl <- ggplot(data.frame(x), aes(x)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 40,
    fill = "lightgray"
  ) +
  geom_line(
    aes(x, gausdens(x, mu1, mu2, sigma1, sigma2, p)),
    color = "black",
    linewidth = 1
  ) +
  labs(y = "Density", x = "x")

pl
```

```{r}
#| label: gaus-mix-loglik
#| echo: false

neg_loglik <- function(par, x) {
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]

  if (p < 0 || p > 1) {
    return(Inf)
  }

  -sum(log(
    p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
      (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2
  ))
}
```

## General-Purpose Optimization

```{r}
#| label: gaus-mix-example

optim_par <- optim(c(0.5, -0.5, 4), neg_loglik, x = x)
optim_par[1:2]
```

. . .

But the starting value matters!

```{r}
#| label: gaus-mix-example-bad

optim(c(0.9, 3, 1), neg_loglik, x = x)[1:2]
```

## The EM Algorithm

```{r}
#| label: em-alg

em <- function(
  par,
  em_step,
  eps = 1e-12,
  maxit = 20,
  cb = NULL
) {
  for (i in seq_len(maxit)) {
    par0 <- par
    par <- em_step(par)

    if (!is.null(cb)) {
      cb() # Callback function (tracer)
    }

    if (sum((par - par0)^2) <= eps * (sum(par^2) + eps)) {
      break
    }
  }
  par
}
```

## Implementation

Next, we implement the E and M steps. Implementations are not shown here, but
can be found in the [source code](./slides/lecture11.qmd)

```{r}
#| label: source-em-mix
#| echo: false

library(numDeriv)

e_step_gauss <- function(par, x, sigma1 = 1, sigma2 = sigma1) {
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  a <- p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1
  b <- (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2
  a / (a + b)
}

m_step_gauss <- function(p_hat, x) {
  N <- length(x)
  N1 <- sum(p_hat)
  N2 <- N - N1
  c(N1 / N, sum(p_hat * x) / N1, sum((1 - p_hat) * x) / N2)
}

create_em_gauss_step <- function(
  x,
  sigma1 = 1,
  sigma2 = sigma1
) {
  force(x)
  force(sigma1)
  force(sigma2)

  function(par) {
    m_step_gauss(e_step_gauss(par, x, sigma1, sigma2), x)
  }
}

Q <- function(par, par_prime) {
  p_hat <- e_step_gauss(par_prime, x, sigma1, sigma2)
  p <- par[1]
  mu1 <- par[2]
  mu2 <- par[3]
  sum(
    p_hat * (log(p) - (x - mu1)^2 / (2 * sigma1^2)) +
      (1 - p_hat) * (log(1 - p) - (x - mu2)^2 / (2 * sigma2^2))
  )
}

grad1 <- function(par) {
  grad(function(par) neg_loglik(par, x), par)
}
grad2 <- function(par) -grad(Q, par, par_prime = par)

em_gauss_step <- create_em_gauss_step(x, sigma1, sigma2)
```

## Testing

```{r}
#| label: test-em-gauss-mix

em_par <- em(c(0.5, -0.5, 4), em_gauss_step)
```

. . .

```{r}
#| label: gaus-mix-example2

# testthat::test_that("EM and optim give similar results", {
#   testthat::expect_equal(em_par, optim_par$par, tolerance = 1e-3)
# })
```

. . .

```{r}
#| label: run-em-gauss-mix-again

em(c(0.9, 3, 1), em_gauss_step) # Starting value still matters
```

## Tracing Convergence

The `tracer` object from the **CSwR** package lets us evaluate arbitrary
expressions at each iteration of the EM algorithm via `expr`.

```{r}
#| label: tracing-convergence

library(CSwR)
em_tracer <- tracer(
  c("par0", "par", "nll"),
  Delta = 0, # No printing
  expr = quote({
    nll <- neg_loglik(par, x)
  })
)
```

. . .

```{r}
#| label: fit-gaussian-mixture-trace

par_hat <- em(c(0.2, 2, 2), em_gauss_step, cb = em_tracer$tracer)

em_trace <- summary(em_tracer)
```

##

::: {.columns}
::::: {.column width="47%"}
**CSwR** contains an `autoplot` S3 method for `trace` objects.

. . .

```{r}
#| label: gaussmix-convergence-plot
#| eval: false

autoplot(em_trace, y = nll - min(nll))
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: convergence-plots-2
#| warning: false
#| echo: false
#| ref-label: gaussmix-convergence-plot
#| fig-cap: "Convergence of the EM algorithm for Gaussian mixture model."
```
:::::
:::

##

```{r}
#| label: mixture-convergence
#| echo: false

gaussdens2 <- function(x, p, mu1, mu2) {
  (p * exp(-(x - mu1)^2 / (2 * sigma1^2)) / sigma1 +
    (1 - p) * exp(-(x - mu2)^2 / (2 * sigma2^2)) / sigma2) /
    sqrt(2 * pi)
}

em_gauss_illustration <- function(i, em_trace) {
  p_i <- em_trace$par0.1[i]
  mu1_i <- em_trace$par0.2[i]
  mu2_i <- em_trace$par0.3[i]

  pl +
    geom_line(
      aes(y = gaussdens2(x, p_i, mu1_i, mu2_i)),
      color = "darkorange",
      size = 1
    ) +
    lims(y = c(0, 0.25)) +
    labs(title = paste0("k = ", i))
}
```

```{r}
#| label: em-gauss-iterations
#| fig-width: 4.2
#| fig-height: 2.8
#| echo: false
#| fig-show: animate
#| fig-cap: "Iterations of the EM algorithm"

for (i in seq_len(20)) {
  print(em_gauss_illustration(i, em_trace))
}
```

## Benchmarking

You might be inclined to wonder which method is faster: EM or `optim()` with
BFGS?

\medskip\pause

```{r}
#| label: benchmark-em-gauss-mix

gauss_bm <- bench::mark(
  em = em(c(0.5, -0.5, 4), em_gauss_step, maxit = 100),
  bfgs = optim(
    c(0.5, -0.5, 4),
    neg_loglik,
    x = x,
    method = "BFGS"
  ),
  check = FALSE
)
```

##

```{r}
#| label: plot-gaussian-mixture-benchmark
#| fig-cap: "Benchmarking EM against BFGS for the Gaussian mixture model."
#| fig-width: 4.5

plot(gauss_bm)
```

## Convergence of EM and BFGS

```{r}
#| label: trace-gaussian-mixture-benchmark
#| include: false

start <- c(0.5, -0.5, 4)
start_nll <- neg_loglik(start, x)

em_benchmark_tracer <- tracer(
  "nll",
  Delta = 0,
  expr = quote(nll <- neg_loglik(par, x))
)
em(
  start,
  em_gauss_step,
  maxit = 100,
  cb = em_benchmark_tracer$tracer
)

optim_benchmark_tracer <- tracer("nll", Delta = 0)
traced_neg_loglik <- function(par, x) {
  nll <- neg_loglik(par, x)
  optim_benchmark_tracer$tracer()
  nll
}
optim(start, traced_neg_loglik, x = x, method = "BFGS")

trace_path <- function(tracer, method) {
  trace <- tracer$get(simplify = TRUE)
  data.frame(
    elapsed_ms = c(0, cumsum(trace[, ".time"])) * 1000,
    nll = cummin(c(start_nll, trace[, "nll"])),
    method = method
  )
}

convergence <- rbind(
  trace_path(em_benchmark_tracer, "EM"),
  trace_path(optim_benchmark_tracer, "BFGS")
)
```

```{r}
#| label: plot-gaussian-mixture-convergence
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "Best negative log-likelihood gap against elapsed time. The EM trace
#|   records iterations; the BFGS trace records objective evaluations."

convergence$gap <- convergence$nll - min(convergence$nll) + 1e-6

ggplot(convergence, aes(elapsed_ms, gap, color = method)) +
  geom_line() +
  scale_color_manual(
    values = c(EM = "darkorange", BFGS = "royalblue")
  ) +
  scale_y_log10() +
  labs(
    x = "Elapsed time (ms)",
    y = "Best negative log-likelihood gap",
    color = NULL
  )
```

## Label Switching

If we set $\sigma_1 = \sigma_2$, then the likelihood is invariant under
switching the labels of the components, i.e., $(p, \mu_1, \mu_2)$ and
$(1 - p, \mu_2, \mu_1)$ give the same likelihood.

\medskip\pause

So the EM algorithm may converge to either of the two symmetric modes.

```{r}
#| label: mixture-gaus-sim2
#| echo: false

sigma1 <- 1
sigma2 <- 1
mu1 <- -0.5
mu2 <- 4
p <- 0.8
N <- 5000
z <- sample(
  c(TRUE, FALSE),
  N,
  replace = TRUE,
  prob = c(p, 1 - p)
)
x <- numeric(N)
N1 <- sum(z)
x[z] <- rnorm(N1, mu1, sigma1)
x[!z] <- rnorm(N - N1, mu2, sigma2)

em_gauss_step_alt <- create_em_gauss_step(x, sigma1)
```

. . .

```{r}
#| label: fit-label-switched-mixture-first

em(c(0.2, 3, -0.5), em_gauss_step_alt)
```

. . .

```{r}
#| label: fit-label-switched-mixture-second

em(c(0.8, -3, -0.5), em_gauss_step_alt)
```

. . .

In this example, we generated data with $p = 0.8$ and $\sigma_1 = \sigma_2$.

## A Mixture with Covariate-Dependent Weights

We now observe a covariate $w_i$ alongside each $x_i$. Let the probability of
component 1 depend on it:
$$
  \pi_i = \Pr(Z_i = 1 \mid w_i) = \frac{1}{1 + e^{-(a+b w_i)}}.
$$

\medskip\pause

The component distributions are still Gaussian. To focus on the mixing weights,
we treat $\mu_1 = -0.5$, $\mu_2 = 4$, and $\sigma_1 = \sigma_2 = 1$ as known.

\medskip\pause

With a constant mixing weight, we had a closed-form M-step. Here, estimating
$(a,b)$ requires a logistic regression fit.

## E-Step with Varying Mixture Weights

Let $\phi_k(x)$ denote the density of the known Gaussian component $k$. At the
current parameters $(a_n,b_n)$, compute
$$
  \pi_i^{(n)} = \frac{1}{1 + e^{-(a_n+b_n w_i)}}
$$
and the responsibility
$$
  \tau_i
  = \Pr(Z_i = 1 \mid x_i,w_i,a_n,b_n)
  = \frac{\pi_i^{(n)} \phi_1(x_i)} {\pi_i^{(n)} \phi_1(x_i) + (1 - \pi_i^{(n)})\phi_2(x_i)}.
$$

\medskip\pause

As before, the E-step assigns each observation a fractional component label.

## A Generalized M-Step

Hold the responsibilities $\tau_i$ fixed. With $\eta_i = a + b w_i$, the part of
$Q$ that depends on $(a,b)$ is
$$
  Q(a,b \mid a_n,b_n)
  = \sum_{i=1}^N \left[ \tau_i \eta_i - \log(1 + e^{\eta_i}) \right]
    + \text{const}.
$$

\medskip\pause

There is no closed-form maximizer. Instead, take one gradient-ascent step:
$$
  \begin{pmatrix}a_{n+1}\\b_{n+1}\end{pmatrix}
  = \begin{pmatrix}a_n\\b_n\end{pmatrix}
  + t \sum_{i=1}^N (\tau_i-\pi_i^{(n)})
    \begin{pmatrix}1\\w_i\end{pmatrix}.
$$

\medskip\pause

Halve $t$ until $Q$ does not decrease. This is a GEM step, so the observed-data
log-likelihood does not decrease.

## Example: A Mixture with Varying Weights

We simulate a covariate and use it to choose the component for each observation.

```{r}
#| label: simulate-varying-mixture

w_gem <- runif(500, -2, 2)
z_gem <- rbinom(500, 1, plogis(-0.5 + 1.5 * w_gem))
x_gem <- rnorm(500, ifelse(z_gem == 1, -0.5, 4), 1)
```

. . .

```{r}
#| label: plot-varying-mixture
#| echo: false
#| fig-width: 4.5
#| fig-height: 2.5
#| fig-cap: "The fraction of observations near each Gaussian component changes
#|   with the covariate."

ggplot(data.frame(w = w_gem, x = x_gem), aes(w, x)) +
  geom_point(alpha = 0.45, color = "royalblue") +
  labs(x = expression(w), y = expression(x))
```

## E-Step and $Q$ in R

```{r}
#| label: varying-mixture-responsibilities

gate_responsibility <- function(coef) {
  p <- plogis(coef[1] + coef[2] * w_gem)
  f1 <- dnorm(x_gem, -0.5, 1)
  f2 <- dnorm(x_gem, 4, 1)
  p * f1 / (p * f1 + (1 - p) * f2)
}

gate_q <- function(coef, tau) {
  eta <- coef[1] + coef[2] * w_gem
  sum(tau * eta - pmax(eta, 0) - log1p(exp(-abs(eta))))
}
```

\medskip

The last line evaluates $\log(1 + e^\eta)$ without overflow.

## One GEM Step in R

```{r}
#| label: varying-mixture-gem-step

gem_gate_step <- function(coef) {
  tau <- gate_responsibility(coef)
  p <- plogis(coef[1] + coef[2] * w_gem)
  gradient <- c(sum(tau - p), sum(w_gem * (tau - p)))
  step <- 1 / length(x_gem)
  q_old <- gate_q(coef, tau)
  while (gate_q(coef + step * gradient, tau) < q_old) {
    step <- step / 2
  }
  coef + step * gradient
}
```

## GEM Increases the Observed Likelihood

Starting from $(a_0,b_0) = (0,0)$, we repeat the GEM step 40 times.

```{r}
#| label: varying-mixture-gem-trace
#| echo: false

gate_loglik <- function(coef) {
  p <- plogis(coef[1] + coef[2] * w_gem)
  sum(log(
    p * dnorm(x_gem, -0.5, 1) + (1 - p) * dnorm(x_gem, 4, 1)
  ))
}

gem_path <- matrix(NA_real_, nrow = 41, ncol = 2)
gem_path[1, ] <- c(0, 0)
for (j in seq_len(40)) {
  gem_path[j + 1, ] <- gem_gate_step(gem_path[j, ])
}
gem_loglik <- apply(gem_path, 1, gate_loglik)
```

. . .

```{r}
#| label: plot-varying-mixture-gem-trace
#| echo: false
#| fig-width: 4.5
#| fig-height: 2.5
#| fig-cap: "Observed-data log-likelihood gain over GEM iterations."

ggplot(
  data.frame(
    iteration = 0:40,
    gain = gem_loglik - gem_loglik[1]
  ),
  aes(iteration, gain)
) +
  geom_line(color = "darkorange") +
  geom_point(color = "darkorange") +
  labs(x = "Iteration", y = "Log-likelihood gain")
```

## Exercise

Implement the EM algorithm for a missing data problem:

```{r}
#| label: bernoulli-missing-data

x <- c(1, 0, 1, NA, 0, NA, 1, 1, 0)
```

Assume that $X_i \sim \text{Bernoulli}(\theta)$ i.i.d. and that missingness is
ignorable (missing at random, with separate parameters for the missingness
mechanism). The complete-data log-likelihood is
$$
  \ell_C(\theta)
  = \sum_{i=1}^N \left( X_i \log(\theta) + (1 - X_i) \log(1 - \theta) \right).
$$

. . .

### Step 1 (Derive)

- E-step: Compute
  $Q(\theta \mid \theta_n) = \E_{\theta_n}\left( \log p(X_\text{obs}, X_\text{miss} \mid \theta) \mid X_\text{obs} \right)$.
- M-step: Maximize $Q(\theta \mid \theta_n)$ over $\theta$.

. . .

### Step 2 (Implement)

Implement the EM algorithm and run it from two different starting points. Record
the path of $\theta_n$ and the log-likelihood values $\ell(\theta_n)$.

```{r}
#| label: em-bernoulli-missing-data
#| include: false

x <- c(1, 0, 1, NA, 0, NA, 1, 1, 0)

# EM for Bernoulli with missing entries (treated as latent)
em_bern_missing <- function(x, p0 = 0.5, maxit = 6) {
  obs <- !is.na(x)
  ll <- function(p) {
    sum(x[obs] * log(p) + (1 - x[obs]) * log(1 - p))
  }
  p <- p0
  ps <- numeric(maxit + 1)
  lls <- numeric(maxit + 1)

  for (t in 0:maxit) {
    ps[t + 1] <- p
    lls[t + 1] <- ll(p)
    if (t < maxit) {
      # E-step: expected value of each missing Xi is p
      k <- sum(!obs)
      p <- (sum(x[obs]) + k * p) / length(x) # M-step: mean of completed data
    }
  }
  list(
    p_final = p,
    path = ps,
    loglik = lls,
    loglik_diff = diff(lls)
  )
}

# Run from two starts
res1 <- em_bern_missing(x, p0 = 0.5)
res2 <- em_bern_missing(x, p0 = 0.1)

res1$path
res1$loglik_diff # ≥ 0 (monotone)
res2$path

# Closed-form fixed point: solves p = (sum(obs 1s) + 2 p)/9 -> p = 4/7
# Plug-in expected Fisher information for each sample size.
p_hat <- res1$p_final
n_obs <- sum(!is.na(x))
i_obs <- n_obs / (p_hat * (1 - p_hat)) # observed-data info
i_complete <- length(x) / (p_hat * (1 - p_hat)) # complete-data info
c(p_hat = p_hat, i_obs = i_obs, i_complete = i_complete)
```

## Summary

We applied the EM algorithm to the peppered moths example.

\medskip\pause

We computed the observed Fisher information directly and through the information
identity.

\medskip\pause

We connected the convergence rate to missing information and introduced the
generalized EM algorithm.
