---
title: "Parallelization and Scatterplot Smoothing"
---

{{< include _common.qmd >}}

## Today

### Parallelization

We use multiple cores to speed up computations.

. . .

### Scatterplot Smoothing

We will cover **nearest-neighbor smoothing** and **smoothing splines**.

# Parallelization

## Parallelization

Most processors have multiple cores.

\medskip\pause

Unless instructed otherwise, only a single core is used.

\medskip\pause

The computer doesn't automatically know that your computations are safe to do in
parallel.

## Embarrassingly Parallel Tasks

These tasks separate easily into independent subtasks.

\pause\medskip

### Examples {.example}

- Summing a vector (or matrix): `sum()`\pause
- Linear algebra operations: `%*%` (`crossprod()`)\pause
- Running iterations of a simulation\pause
- Cross-validation

## The foreach Package

::: {.columns}
::::: {.column width="47%"}
```{r}
#| label: foreach-load
#| warning: false
#| message: false

library(foreach)
```

```{r}
#| label: foreach-sqrt
#| eval: false

foreach(i = seq_len(3)) %dopar%
  {
    sqrt(i)
  }
```
:::::

. . .

::::: {.column width="47%"}
```{r}
#| label: foreach-sqrt-display
#| echo: false
#| ref-label: "foreach-sqrt"
```
:::::
:::

\pause\medskip

### Notes {.alert}

- Returns a list, so it is not really a `for` loop.\pause
- Nothing is actually parallel yet!\pause
- First, we need to register a **backend**.

## Backends

::: {.columns}
::::: {.column width="43%"}
Several backends are available as separate packages (**doParallel**, **doMC**,
**doFuture**). Load and **register** one before using `%dopar%`.

\pause

```{r}
#| label: doparallel

library(doParallel)

cl <- makeCluster(2)
registerDoParallel(cl)
```
:::::

. . .

::::: {.column width="51%"}
```{r}
#| label: doparallel-execute
#| message: false

foreach(i = seq_len(3)) %dopar%
  {
    paste("Hi from", Sys.getpid())
  }
```
:::::
:::

## Combining Results

`foreach()` returns a **list** by default.

\medskip\pause

If you want to reduce your result, use either the `.combine` argument or
manually reduce the resulting list.

. . .

```{r}
#| label: foreach-combine

x <- c(4, 1, 1e2, 2)

foreach(i = seq_len(4), .combine = c) %dopar%
  {
    log(x[[i]])
  }
```

## Caveats

### Realistic Expectations

In practice, twice as many cores $\neq$ twice the speed.

. . .

### Overhead

Parallelization comes with overhead from starting workers and communicating
between them.

. . .

### Thread Safety

Some operations are not thread-safe---for example, writing to the same file from
multiple workers.

. . .

### Redundancies

Parallelizing a task that is already parallelized may not help and may even
\alert{hurt} performance.

## Parallelization on One Computer

\begin{figure}
\begin{tikzpicture}[
    font=\small,
    box/.style={draw, rounded corners=2pt, minimum height=0.45cm,
      minimum width=1.7cm, inner sep=3pt},
    flow/.style={-{Stealth}, draw=RoyalBlue3}
  ]
  \draw[rounded corners=4pt, draw=black!50, fill=black!2]
    (-2, -1.35) rectangle (2, 3.25);
  \node at (0, 2.95) {One computer};
  \draw[rounded corners=2pt, draw=RoyalBlue3, fill=RoyalBlue3!5]
    (-1.35, -0.2) rectangle (1.35, 2.6);
  \node at (0, 2.3) {Processor};
  \node at (-4, 2.95) {Tasks};
  \node at (4, 2.95) {Results};

  \foreach \i in {1,...,4} {
    \node[box, draw=RoyalBlue3, fill=RoyalBlue3!10]
      (task-\i) at (-4, {2.35 - 0.55*\i}) {Task \i};
    \node[box, draw=RoyalBlue3, fill=RoyalBlue3!15]
      (core-\i) at (0, {2.35 - 0.55*\i}) {Core \i};
    \node[box, draw=DarkOrange3, fill=DarkOrange1!15]
      (result-\i) at (4, {2.35 - 0.55*\i}) {Result \i};
    \draw[flow] (task-\i) -- (core-\i);
    \draw[flow, draw=DarkOrange3] (core-\i) -- (result-\i);
  }

  \node[box, draw=black!50, fill=black!5, minimum width=2.7cm]
    (memory) at (0, -0.85) {Shared memory};
  \draw[{Stealth}-{Stealth}, draw=black!50] (0, -0.2) -- (memory.north);
  \end{tikzpicture}
  \caption{Independent tasks can run simultaneously on different cores of one processor.}
\end{figure}

\pdfpcnote{
  The cores share the memory of one computer.

  Most R code uses only one core unless we explicitly parallelize it.
}

## Parallelization on a Cluster

\begin{figure}
\begin{tikzpicture}[
    font=\small,
    box/.style={draw, rounded corners=2pt, align=center, inner sep=4pt},
    flow/.style={-{Stealth}, draw=RoyalBlue3}
  ]
  \node[box, draw=RoyalBlue3, fill=RoyalBlue3!10, minimum width=1.5cm]
    (jobs) at (-5.45, 0) {Job\\queue};
  \node[box, draw=RoyalBlue3, fill=RoyalBlue3!10, minimum height=0.9cm]
    (scheduler) at (-3, 0) {Scheduler};
  \node[box, draw=DarkOrange3, fill=DarkOrange1!15, minimum height=0.9cm]
    (results) at (5.25, 0) {Results};
  \draw[flow] (jobs) -- (scheduler);

  \foreach \i/\y in {1/1.7, 2/0, 3/-1.7} {
    \node[box, draw=black!50, fill=black!2,
      minimum width=4.3cm, minimum height=1.5cm]
      (node-\i) at (1.3, \y) {};
    \node at (1.3, {\y + 0.52}) {Node \i};
    \node[box, font=\scriptsize, draw=black!50, fill=black!5,
      minimum width=3.8cm, minimum height=0.34cm, inner sep=1pt]
      (memory-\i) at (1.3, {\y - 0.49}) {Shared memory};

    \foreach \j in {1,...,4} {
      \node[box, font=\scriptsize, draw=RoyalBlue3, fill=RoyalBlue3!15,
        minimum width=0.88cm, minimum height=0.42cm, inner sep=1pt]
        (core-\i-\j) at ({1.3 + (\j - 2.5)}, {\y + 0.02}) {Core \j};
      \draw[draw=black!50] (core-\i-\j.south)
        -- (core-\i-\j.south |- memory-\i.north);
    }

    \draw[flow] (scheduler.east) to[out=0, in=180] (node-\i.west);
    \draw[flow, draw=DarkOrange3]
      (node-\i.east) to[out=0, in=180] (results.west);
  }
  \end{tikzpicture}
  \caption{A scheduler assigns independent jobs to cores across several **nodes**.}
\end{figure}

\pdfpcnote{
  A node is one physical computer containing several cores and shared memory.

  Communication between nodes is usually more expensive than communication
  within a node.

  Job arrays are ideal for independent simulation scenarios or bootstrap
  batches.

  The program must not start more workers than the number of allocated cores.

  Have each job write to a separate output file.
}

# Scatterplot Smoothing

## Nuuk Temperature Data

```{r}
#| label: load-nuuk-data
#| echo: false
#| message: false

nuuk <- read_table(
  here::here("data", "nuuk.txt"),
  col_names = c("Year", 1:12),
  na = "-999",
  skip = 1
) |>
  gather(
    key = "Month",
    value = "Temperature",
    -Year,
    convert = TRUE
  ) |>
  mutate(Temperature = Temperature / 10) |>
  filter(Year > 1866)

nuuk_year <- group_by(nuuk, Year) |>
  summarise(
    Temperature = mean(Temperature),
    Median = median(Temperature),
    High = max(Temperature),
    Low = min(Temperature)
  )
N <- nrow(nuuk_year)
```

```{r}
#| label: nuuk-plot
#| fig-cap: Annual average temperature in Nuuk, Greenland
#| fig-width: 4

nuuk_plot <- ggplot(nuuk_year, aes(Year, Temperature)) +
  geom_point()
nuuk_plot
```

Data is available [here](../data/nuuk.txt).

## Nearest-Neighbor Estimation

Given data points $(x_1, y_1), \ldots, (x_N, y_N)$, the $k$-nearest-neighbor
estimator at $x_i$ is
$$
  \hat{f}_i = \frac{1}{k} \sum_{j \in \mathcal{N}_i} y_j
$$
where $\mathcal{N}_i$ is the set of indices for the $k$ nearest neighbors of
$x_i$.

\medskip\pause

This is an estimator of
$$
  f(x) = \E(Y \mid X = x).
$$

\pdfpcnote{
 N_i is the neighborhood of x_i.
}

\pdfpcnote{
  Show example with k = 3.
}

## Matrix Formulation

We can express all fitted values as a matrix-vector product:
$$
  \hat{\symbf{f}} = \symbf{S} \symbf{y}
$$
where $\symbf{y} = (y_1, \ldots, y_N)^T$ and $\symbf{S}$ is the **smoother
matrix**.

\pause

### Smoother Matrix

Each entry of $\symbf{S}$ is
$$
  S_{ij} = \frac{1}{k} \cdot \symbf{1}_{\{j \in \mathcal{N}_i\}}
$$
where $\symbf{1}_{\{j \in \mathcal{N}_i\}}$ is 1 if $j$ is a neighbor of $i$ and
0 otherwise.

\pdfpcnote{
  - Each fitted value is a weighted average of the y-values.\\
   -S is a relatively sparse N x N matrix.
}

## Smoother Matrix: An Example

Take $x_i = i$ for $i = 1, \ldots, 5$ and $k = 3$. Include each point itself
among its neighbors.

$$
  \symbf{S} = \frac{1}{3} \begin{pmatrix}
                            1                 & 1                 & 1                 & 0                 & 0                 \\
                            1                 & 1                 & 1                 & 0                 & 0                 \\
                            \alert < 2- > {0} & \alert < 2- > {1} & \alert < 2- > {1} & \alert < 2- > {1} & \alert < 2- > {0} \\
                            0                 & 0                 & 1                 & 1                 & 1                 \\
                            0                 & 0                 & 1                 & 1                 & 1
                          \end{pmatrix}.
$$

\pause

The third row selects and averages the neighbors of $x_3$:
$$
  \hat{f}_3 = \frac{y_2 + y_3 + y_4}{3}.
$$

\medskip\pause

### Sparsity

Our smoother has $N^2$ entries, but only $Nk$ nonzero values.

\pause

For $N = 10\,000$ and $k = 3$: **100 million entries, but only 30,000
nonzeros.**

\medskip\pause

We can take advantage of this!

\pdfpcnote{
  - Each row has three nonzero entries, each with weight 1/3.\\
  - At the endpoints, we still select the three closest points. This explains
    why the first two rows are identical, as are the last two rows.
}

## Sparse Matrices

A **sparse matrix** has mostly zero entries.

```{=latex}
\begin{center}
\begin{tikzpicture}[font=\normalsize]
  \foreach \row/\entries in {0/{4,0,0,0}, 1/{0,7,0,0},
                            2/{2,0,6,0}, 3/{0,9,0,5}} {
    \foreach \value [count=\col from 0] in \entries {
      \ifnum\value=0
        \node[text=black!35] at ({0.65*\col}, {-0.6*\row}) {$\value$};
      \else
        \node[text=RoyalBlue3] at ({0.65*\col}, {-0.6*\row}) {$\value$};
      \fi
    }
  }
  \draw (-0.2, 0.3) -- (-0.35, 0.3) -- (-0.35, -2.1) -- (-0.2, -2.1);
  \draw (2.15, 0.3) -- (2.3, 0.3) -- (2.3, -2.1) -- (2.15, -2.1);
  \node at (-0.9, -0.9) {$\symbf{A} =$};
  \node[anchor=west, align=left] at (3.3, 0.2)
    {\textbf{Dense storage}\\Store all 16 entries.};
  \uncover<2->{
    \node[anchor=west, align=left, text=RoyalBlue3] at (3.3, -2)
      {\textbf{Sparse storage}\\Store 6 nonzero values\\and their positions.};
  }
\end{tikzpicture}
\end{center}
```

\pause\pause

\pdfpcnote{
  Start with the matrix: which entries do we actually need to remember?
  On the first click, introduce sparse storage. Unstored entries are zero;
  this is an exact representation, not an approximation.

  On the second click, return to the smoother as motivation. It is sparse
  when k is small relative to N. Sparsity does not require a diagonal band
  or any other particular pattern.

  Positions also cost memory, so sparse storage need not save space for this
  tiny example. The benefit grows when the fraction of nonzeros is small.
}

## Sparse Storage: Column by Column

**Compressed sparse column (CSC):** read down each column, skipping zeros.

```{=latex}
\begin{center}
\begin{tikzpicture}[
    font=\small,
    cell/.style={draw=black!35, minimum width=0.72cm,
      minimum height=0.5cm, inner sep=0pt},
    flow/.style={-{Stealth}, draw=DarkOrange3}
  ]
  \fill[DarkOrange1!15] (0.39, 0.3) rectangle (0.91, -2.1);
  \node[text=DarkOrange3, font=\scriptsize] at (0.65, 0.6) {column 2};
  \node[text=black!60, font=\scriptsize] at (-0.85, 0.6) {row};
  \foreach \row/\entries in {0/{4,0,0,0}, 1/{0,7,0,0},
                            2/{2,0,6,0}, 3/{0,9,0,5}} {
    \node[text=black!60, font=\scriptsize]
      at (-0.85, {-0.6*\row}) {$\row$};
    \foreach \value [count=\col from 0] in \entries {
      \ifnum\value=0
        \node[text=black!35] at ({0.65*\col}, {-0.6*\row}) {$\value$};
      \else
        \ifnum\col=1
          \node[text=DarkOrange3] at ({0.65*\col}, {-0.6*\row}) {$\value$};
        \else
          \node[text=RoyalBlue3] at ({0.65*\col}, {-0.6*\row}) {$\value$};
        \fi
      \fi
    }
  }
  \draw (-0.2, 0.3) -- (-0.35, 0.3) -- (-0.35, -2.1) -- (-0.2, -2.1);
  \draw (2.15, 0.3) -- (2.3, 0.3) -- (2.3, -2.1) -- (2.15, -2.1);

  \uncover<2->{
    \fill[DarkOrange1!15] (6.78, 0.25) rectangle (8.22, -1.05);
    \node[anchor=east, font=\scriptsize, text=black!60] at (5.1, 0.6) {offset};
    \node[anchor=east] at (5.1, 0) {Values};
    \node[anchor=east] at (5.1, -0.8) {Row indices};
    \foreach \offset/\value/\row in {0/4/0, 1/2/2, 2/7/1, 3/9/3, 4/6/2, 5/5/3} {
      \node[font=\scriptsize, text=black!60]
        at ({5.7 + 0.72*\offset}, 0.6) {$\offset$};
      \node[cell] at ({5.7 + 0.72*\offset}, 0) {$\value$};
      \node[cell] at ({5.7 + 0.72*\offset}, -0.8) {$\row$};
    }
    \draw[flow] (0.91, 0.3) to[out=35, in=150] (7.14, 0.85);
  }

  \uncover<3->{
    \node[anchor=east] at (5.1, -2.5) {Column pointers};
    \foreach \col/\offset in {0/0, 1/2, 2/4, 3/5, 4/6} {
      \ifnum\col=1
        \node[cell, fill=DarkOrange1!15] (pointer-\col)
          at ({5.7 + 0.72*\col}, -2.5) {$\offset$};
      \else
        \ifnum\col=2
          \node[cell, fill=DarkOrange1!15] (pointer-\col)
            at ({5.7 + 0.72*\col}, -2.5) {$\offset$};
        \else
          \node[cell] (pointer-\col)
            at ({5.7 + 0.72*\col}, -2.5) {$\offset$};
        \fi
      \fi
    }
    \draw[flow] (pointer-1.north) -- (6.78, -1.1);
    \draw[flow] (pointer-2.north) -- (8.22, -1.1);
    \node[anchor=west, align=left, text=DarkOrange3] at (3.3, -3.4)
      {Column 2 starts at offset 2 and stops before offset 4.};
  }
\end{tikzpicture}
\end{center}
```

\pause\pause

Row indices and offsets start at zero. Unstored entries are zero.

\pdfpcnote{
  First trace down the highlighted column: skip the zeros, keep 7 and 9.

  On the first click, show the values and row indices. The pair (7, 1) means
  value 7 in row index 1. Ask where the 9 belongs before reading its row index.
  The offsets above the arrays identify positions in storage, not matrix rows.

  On the second click, show the pointers. Column 1 occupies offsets 0 and 1;
  column 2 occupies offsets 2 and 3. Its pointers are 2 and 4, so it has
  4 minus 2 equals 2 stored entries. No column index is needed for each value.
  The final pointer, 6, is the total number of stored entries. An empty column
  would have equal start and end pointers.

  These arrays are the x, i, and p slots of a dgCMatrix. R's sparseMatrix
  constructor accepts one-based indices, but stores zero-based indices.

  For the smoother, CSC stores Nk values, Nk row indices, and N + 1 pointers:
  O(Nk + N) = O(Nk) storage instead of O(N squared).
}

## Sparse Matrices in R

Construct the smoother matrix directly from its nonzero entries.

::: {.columns}
::::: {.column width="56%"}
```{r}
#| label: sparse-smoother-example

S_example <- Matrix::sparseMatrix(
  i = rep(1:5, each = 3),
  j = c(1:3, 1:3, 2:4, 3:5, 3:5),
  x = 1 / 3,
  dims = c(5, 5)
)
```
:::::

. . .

::::: {.column width="40%"}
```{r}
#| label: sparse-smoother-display
#| echo: false

Matrix::printSpMatrix(3 * S_example)
```
:::::
:::

\bigskip

The display is scaled by 3. Dots represent zeros that are not stored.

\pause\medskip

The usual `%*%` operator also works with sparse matrices, using only the stored
entries when multiplying by a vector.

\pdfpcnote{
  The i and j arguments give row and column indices. The x argument gives the
  corresponding values; a single value is repeated for every listed position.

  A base R matrix containing many zeros still uses dense storage. Constructing
  the sparse matrix directly avoids allocating that dense matrix first.
}

## Linear Smoothers

Smoothers of the form $\hat{\symbf{f}} = \symbf{S} \symbf{y}$ are called *linear
smoothers*.

\medskip

Here, linear means linear in $\symbf{y}$, with the $x$-values and $k$ fixed.

\pause\medskip

The $k$-nearest-neighbor smoother is a simple example of a linear smoother that
works for $x$-values in any metric space.

\pause\medskip

The representation of a linear smoother as a matrix-vector product,
$$
  \symbf{S} \symbf{y}
$$
is theoretically useful, but often not the best way to actually compute
$\hat{\symbf{f}}$.

## Running Mean

When $x_i \in \mathbb{R}$, we can sort the data according to the $x$-values and
then use a *symmetric* neighbor definition:
$$
  \mathcal{N}_i
  = \left\{ i - \frac{k - 1}{2}, i - \frac{k - 1}{2} + 1, \ldots, i - 1 , i, i + 1, \ldots, i + \frac{k - 1}{2} \right\}
$$
for odd $k$.

\pause\medskip

This is a version of nearest-neighbor smoothing that assumes ordered and
symmetric neighborhoods.

\pause

### Simplifies Computations

We don't need to keep track of metric comparisons. Only the order matters.

\pdfpcnote{
- Symmetric neighborhoods do not take distance into account.

- We only need to sort.

- Only the order matters.
}

## Running Mean: An Illustration

```{r}
#| label: running-mean-illustration
#| echo: false
#| warning: false
#| fig-show: animate
#| fig-width: 5.5
#| fig-height: 2.8
#| fig-cap: Illustration of the running mean with $k = 3$. The orange points are
#|   the three nearest neighbors of the blue point.

running_mean_data <- tibble(
  x = seq_len(6),
  y = c(1, 7, 4, 10, 4, 7)
) |>
  mutate(pair = sprintf("(%d, %d)", x, y))

running_mean_centers <- 2:5
running_mean_fits <- tibble(
  x = running_mean_centers,
  y = map_dbl(
    running_mean_centers,
    \(i) mean(running_mean_data$y[(i - 1):(i + 1)])
  )
)

running_mean_frame <- function(step) {
  center <- running_mean_centers[step]
  neighbors <- (center - 1):(center + 1)
  current_y <- running_mean_fits$y[step]
  window_values <- running_mean_data$y[neighbors]
  displayed_fits <- running_mean_fits[seq_len(step), ]
  equation <- sprintf(
    paste0(
      "paste(hat(f)[%d] == frac(y[%d] + y[%d] + y[%d], 3), ",
      "' = ', frac(%d + %d + %d, 3), ' = ', %d)"
    ),
    center,
    neighbors[1],
    neighbors[2],
    neighbors[3],
    window_values[1],
    window_values[2],
    window_values[3],
    current_y
  )

  running_mean_data |>
    mutate(in_window = x %in% neighbors) |>
    ggplot(aes(x, y)) +
    geom_line(color = "grey80") +
    geom_point(aes(color = in_window), size = 3) +
    geom_text(
      aes(label = pair, color = in_window),
      nudge_y = 0.8,
      show.legend = FALSE
    ) +
    geom_line(
      aes(x, y),
      data = displayed_fits,
      color = "royalblue"
    ) +
    geom_point(
      aes(x, y),
      data = displayed_fits,
      color = "royalblue",
      size = 2.5
    ) +
    annotate(
      "point",
      x = center,
      y = current_y,
      color = "royalblue",
      size = 4
    ) +
    annotate(
      "segment",
      x = neighbors[1],
      xend = neighbors[3],
      y = 0,
      yend = 0,
      color = "darkorange"
    ) +
    annotate(
      "segment",
      x = c(neighbors[1], neighbors[3]),
      xend = c(neighbors[1], neighbors[3]),
      y = 0,
      yend = 0.35,
      color = "darkorange"
    ) +
    annotate(
      "text",
      x = center,
      y = -0.55,
      label = sprintf("N[%d]", center),
      parse = TRUE,
      color = "darkorange"
    ) +
    annotate(
      "text",
      x = 2,
      y = 10,
      label = equation,
      parse = TRUE,
      color = "royalblue"
    ) +
    scale_color_manual(
      values = c(`FALSE` = "grey60", `TRUE` = "darkorange"),
      guide = "none"
    ) +
    scale_x_continuous(breaks = running_mean_data$x) +
    coord_cartesian(
      xlim = c(0.65, 6.35),
      ylim = c(-1.3, 12.7),
      expand = FALSE
    ) +
    labs(x = expression(x[i]), y = expression(y[i]))
}

for (step in seq_along(running_mean_centers)) {
  print(running_mean_frame(step))
}
```

\pdfpcnote{
  Each click shifts the three-point neighborhood one position to the right.

  Only the first and last observations in the neighborhood change.

  This sets up the efficient update on the following slides.
}

## Running Mean (Naive Implementation)

Assume $y$ is sorted and $k$ is odd.

```{r}
#| label: runmean-simple

run_mean_naive <- function(y, k) {
  N <- length(y)
  m <- (k - 1) / 2
  f_hat <- rep(NA, N)

  for (i in (m + 1):(N - m)) {
    f_hat[i] <- mean(y[(i - m):(i + m)])
  }

  f_hat
}
```

\pdfpcnote{
- We can speed this up by replacing the mean computation.
}

## Efficient Running Mean

We can achieve a more efficient implementation by noting that
$$
  \hat{f}_{i+1} = \hat{f}_{i} - \frac{y_{i-(k-1)/2}}{k}
                  + \frac{y_{i+(k+1)/2}}{k}.
$$

. . .

```{r}
#| label: run_mean_def

run_mean <- function(y, k) {
  N <- length(y)
  m <- floor((k - 1) / 2)
  k <- 2 * m + 1 # Ensures k is odd and m = (k - 1) / 2
  y <- y / k
  f_hat <- rep(NA, N)
  f_hat[m + 1] <- sum(y[1:k])

  for (i in (m + 1):(N - m - 1)) {
    f_hat[i + 1] <- f_hat[i] - y[i - m] + y[i + 1 + m]
  }
  f_hat
}
```

\pdfpcnote{
  - A sliding window of size k is used to compute the running mean.\\
  - This ensures that k is odd.\\
  - The complexity decreases from O(Nk) to O(N).
}

## Visualization

```{r}
#| label: nuuk-NN-plot2
#| echo: false
#| warning: false
#| fig-width: 4
#| fig-height: 2.2
#| fig-cap: Running mean smoother with $k = 11$.

f_hat <- run_mean(nuuk_year$Temperature, 11)
nuuk_plot + geom_line(aes(y = f_hat), color = "blue")
```
. . .

For a response vector `y`, R provides the same running mean:

```{r}
#| label: run-mean-filter
#| eval: false

stats::filter(y, rep(1 / 11, 11))
```

\pdfpcnote{
  This takes five observations to the right and five to the left.

  Warn about the name clash with dplyr::filter().

  `rep(1 / 11, 11)` contains the moving-average coefficients.

  For this example, set `y <- nuuk_year$Temperature` and
  `f_hat_filter <- stats::filter(y, rep(1 / 11, 11))`.
  Inspect the boundary values with `f_hat_filter[c(1:10, 137:147)]`.
  The first and last five fitted values are NA.

  Check agreement with `all.equal(f_hat, as.numeric(f_hat_filter))`,
  which returns TRUE.
}

##

```{r}
#| label: runMeanBench
#| echo: false
#| warning: false
#| message: false
#| fig-width: 4
#| fig-height: 2.6
#| fig-cap: "Dense matrix-vector multiplication is $O(N^2)$."

nn_bench <- bench::press(N = c(512, 1024, 2048, 4096), {
  k <- 11
  m <- (k - 1) / 2
  y <- rnorm(N)
  S <- matrix(0, N - 2 * m, N)
  for (i in seq_len(nrow(S))) {
    S[i, i:(i + k - 1)] <- 1 / k
  }
  S_sparse <- Matrix::sparseMatrix(
    i = rep(seq_len(nrow(S)), each = k),
    j = rep(seq_len(nrow(S)), each = k) +
      rep(0:(k - 1), times = nrow(S)),
    x = 1 / k,
    dims = dim(S)
  )
  bench::mark(
    dense = S %*% y,
    sparse = S_sparse %*% y,
    running = run_mean(y, k = k),
    filter = stats::filter(y, rep(1 / k, k)),
    check = FALSE
  )
})

mutate(nn_bench, method = as.character(expression)) |>
  ggplot(aes(N, median, color = method)) +
  geom_point() +
  geom_line() +
  labs(y = "time (s)")
```

\medskip\pause

For fixed $k$, sparse multiplication and the other two algorithms are $O(N)$.

\pdfpcnote{
  Both matrix representations are constructed before timing. The benchmark
  measures applying an already constructed smoother to a vector.

  The matrix products return only the interior fitted values; the other methods
  also return missing values at the boundaries.
}

## Three Ways to Compute a Running Mean

  | Method                | Storage for $\symbf{S}$ | Work for all fitted values |
  | :-------------------- | :---------------------- | :------------------------- |
  | Dense matrix          | $O(N^2)$                | $O(N^2)$                   |
  | Sparse matrix         | $O(Nk)$                 | $O(Nk)$                    |
  | Sliding-window update | No matrix               | $O(N)$                     |

\pause\medskip

Sparse multiplication skips the zero weights.

\pause\medskip

The sliding-window update also reuses the overlap between neighborhoods.

\pdfpcnote{
  These costs assume the observations are already ordered and k is the window
  size. Matrix construction and finding neighborhoods are not included.

  All methods still need storage for the data and fitted values. The table
  compares the additional storage used to represent the smoother matrix.

  With k fixed, both sparse multiplication and the sliding-window update are
  linear in N. Varying k reveals their different dependence on window size.
}

## Bias--Variance Trade-Off

There is a bias-variance trade-off in nearest-neighbor smoothing with respect to
$k$.

\medskip\pause

If the data are i.i.d. with $\V(Y \mid X) = \sigma^2$ and
$f(x) = \E(Y \mid X = x)$,

\pause

$$
  \begin{aligned}
    \mathrm{MSE}(x) & = \E(f(x) - \hat{f}(x))^2                                                                                                                              \\
    \pause
                    & = \V\big(f(x) - \hat{f}(x)\big) + \big(\E(f(x) - \hat{f}(x))\big)^2                                                                                    \\
    \pause
                    & = \underbrace{\frac{\sigma^2}{k}}_{\text{variance}} + \bigg( \underbrace{f(x) - \frac{1}{k} \sum_{l \in \mathcal{N}(x)} f(x_l)}_{\text{bias}}\bigg)^2.
  \end{aligned}
$$

. . .

### Bias-Variance Trade-Off

- Small $k$ gives large variance and small bias (if $f$ is smooth).\pause
- Large $k$ gives small variance and potentially large bias (if $f$ is not
  constant).

## How to Choose $k$?

We know there is a bias-variance trade-off in $k$---so how do we choose $k$?

\pause

### Leave-One-Out Cross-Validation (LOOCV)

Leave out one observation at a time, predict using the remaining data, and
measure the prediction error.

\pause\medskip

::: {.columns}
::::: {.column width="47%"}
The general LOOCV formula is
$$
  \mathrm{LOOCV} = \frac{1}{N} \sum_{i=1}^N (y_i - \hat{f}^{-i}_i)^2.
$$

\pause

Keeping the neighborhoods fixed, we can express the leave-one-out fitted values
as
$$
  \hat{f}^{-i}_i = \sum_{j \neq i} \frac{S_{ij}y_j}{1 - S_{ii}},
$$
:::::

. . .

::::: {.column width="47%"}
This leads to the simplified formula
$$
  \mathrm{LOOCV}
  = \frac{1}{N} \sum_{i=1}^N \left( \frac{y_i - \hat{f}_i}{1 - S_{ii}} \right)^2.
$$
:::::
:::

\pdfpcnote{
For the running mean, we can, in principle, compute this without leaving out any
observations.
}

## LOOCV for Nearest Neighbors

For nearest neighbors, $S_{ii} = \frac{1}{k}$ for all $i$, so
$$
  \mathrm{LOOCV}
  = \frac{1}{N} \sum_{i=1}^N \left( \frac{y_i - \hat{f}_i}{1 - 1/k} \right)^2.
$$

. . .

### Implementation

```{r}
#| label: LOOCV-runMean

loocv <- function(k, y) {
  f_hat <- run_mean(y, k)
  mean(((y - f_hat) / (1 - 1 / k))^2, na.rm = TRUE)
}
```

. . .

### Note {.alert}

The implementation ignores the missing fitted values introduced by our boundary
handling. It uses `mean()` rather than `sum()` so that the criterion is averaged
over the remaining observations.

\pdfpcnote{
- For nearest neighbors, $S_{ii} = \frac{1}{k}$ for all $i$.

- **Derive the formula on the blackboard!**
}

--------------------------------------------------------------------------------

```{r}
#| label: nuuk-running-loocv

k <- seq(3, 40, 2)
cv_error <- sapply(
  k,
  function(kk) loocv(kk, nuuk_year$Temperature)
)
k_opt <- k[which.min(cv_error)]
```

\medskip\pause

```{r}
#| label: nuuk-running-loocv-plot
#| fig-width: 5
#| fig-height: 2
#| echo: false
#| fig-cap: LOOCV error for nearest-neighbor smoothing as a function of $k$.

ggplot(tibble(k, cv_error), aes(k, cv_error)) +
  geom_vline(xintercept = k_opt, color = "dark orange") +
  geom_line() +
  geom_point() +
  labs(y = "LOOCV Error", x = expression(k))
```

##

```{r}
#| label: nuuk-NN-plot3
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 2.5
#| fig-cap: Nearest-neighbor smoother with LOOCV-optimal $k=15$.

nuuk_plot +
  geom_line(
    aes(y = run_mean(nuuk_year$Temperature, k_opt)),
    color = "blue"
  )
```

## Wiggliness

The running mean is wiggly!

\pause\medskip

We can use kernel smoothing (as we did last week) to fix this.

\pause\medskip

But another idea is to use \alert{smoothing splines} instead.

# Smoothing Splines

## Smoothing Splines

We want a smooth function $f$ that balances goodness of fit and wiggliness.

\medskip\pause

To do this, we minimize an objective function that penalizes wiggliness:
$$
  L(f) = \sum_{i=1}^N (y_i - f(x_i))^2
         + \lambda \underbrace{\int f''(z)^2 \ \mathrm{d}z}_{\|f''\|_2^2}
$$
where $\lambda \geq 0$ is a tuning parameter that controls the trade-off.

\pause\medskip

The minimizer is a **natural cubic spline** with **knots** at the data points
$x_i$.

\medskip\pause

That is, it has the representation
$$
  f = \sum_j \beta_j \varphi_j,
$$
where $\{\varphi_j\}$ is a basis for the $N$-dimensional space of such splines.

\medskip\pause

The problem reduces to finding the coefficients $\symbf{\beta}$.

\pdfpcnote{
Without the penalty term, any interpolating function would work well.

The penalty term penalizes solutions whose second derivatives have large norms.
}

## Loss Function

In vector notation,
$$
  \symbf{f} = \symbf{\Phi}\symbf{\beta}
$$
where $\symbf{\Phi}_{ij} = \varphi_j(x_i)$. Then \pause
$$
  \begin{aligned}
    L(\symbf{f}) & = (\symbf{y} - \symbf{f})^T (\symbf{y} - \symbf{f}) + \lambda \|f''\|_2^2                                                                  \\
                 & = (\symbf{y} - \symbf{\Phi}\symbf{\beta})^T (\symbf{y} - \symbf{\Phi}\symbf{\beta}) + \lambda \symbf{\beta}^T \symbf{\Omega} \symbf{\beta}
  \end{aligned}
$$
with
$$
  \symbf{\Omega}_{jk} = \int \varphi_j''(z) \varphi_k''(z) \ \mathrm{d}z.
$$

\pause

### Solution

The solution is a *linear smoother* with smoother matrix $\symbf{S}_{\lambda}$
and penalty matrix $\symbf{\Omega}$.

\pdfpcnote{
  This is similar to ridge (Tikhonov) regression: OLS plus a squared $\ell_2$
  penalty.

  We can write the penalty term in the quadratic form
  $\lambda \symbf{\beta}^T \symbf{\Omega} \symbf{\beta}$.

  Derive the minimizer on the blackboard.
}

## From Basis Functions to a Design Matrix

To compute $\symbf{\Phi}$, we need concrete basis functions $\varphi_j$.

\pause\medskip

**Cubic B-splines** are piecewise cubic basis functions with local support.
Their weighted sum gives the spline:
$$
  f(x) = \sum_j \beta_j \varphi_j(x).
$$

\pause\medskip

Evaluating the basis at the observations gives
$\symbf{\Phi}_{ij} = \varphi_j(x_i)$: one column per basis function.

\medskip

In R, `splineDesign(knots, x)` computes this matrix.

\pause\medskip

Natural cubic splines additionally satisfy $f''(a) = f''(b) = 0$ at the
boundaries.

\pdfpcnote{
  The next slide illustrates an ordinary cubic B-spline basis with equally
  spaced knots. This spans a larger space than the natural cubic splines
  with the same knots. Natural boundary conditions restrict the coefficients
  of their weighted sum.

  The smoothing-spline result uses knots at the observations. The small,
  equally spaced set in the illustration makes the basis functions easier
  to see.
}

## A Cubic B-Spline Basis in R

Illustration: an ordinary cubic basis with equally spaced knots.

```{r}
#| label: splines
#| message: false
#| warning: false

knots <- c(0, 0, 0, seq(0, 1, 0.2), 1, 1, 1)
xx <- seq(0, 1, 0.005)
b_splines <- splines::splineDesign(knots, xx)
```

. . .

```{r}
#| label: splines-plot
#| echo: false
#| fig-width: 5
#| fig-height: 2
#| fig-cap: Cubic B-spline basis functions with knots at 0, 0.2, 0.4, 0.6, 0.8,
#|   and 1.

df <- data.frame(
  x = rep(xx, times = ncol(b_splines)),
  y = as.vector(b_splines),
  group = as.character(rep(
    seq_len(ncol(b_splines)),
    each = length(xx)
  ))
)

knot_points <- c(0, 0.2, 0.4, 0.6, 0.8, 1)
spline_colors <- setNames(
  RColorBrewer::brewer.pal(ncol(b_splines), "Dark2"),
  seq_len(ncol(b_splines))
)

ggplot(df, aes(x = x, y = y, group = group, color = group)) +
  scale_x_continuous(breaks = knot_points, minor_breaks = NULL) +
  scale_color_manual(values = spline_colors, guide = "none") +
  geom_line() +
  labs(x = expression(x), y = expression(varphi[j](x)))
```

\pdfpcnote{
  Each column of b\_splines contains one basis function evaluated on xx,
  a grid used for plotting. Evaluating at the observed predictor values
  instead would give the design matrix from the loss function.

  splineDesign defaults to order four, which means cubic polynomials.
  Each endpoint occurs four times in the knot vector, including its entry
  in seq. This constructs the full cubic basis on the interval; it does
  not impose natural boundary conditions.

  No response values or fitted coefficients are involved yet. The curves
  show the basis functions that we will combine to construct a fit.
}

```{r}
#| label: spline-illustration-setup
#| echo: false

pen_mat <- function(inner_knots) {
  knots <- sort(c(rep(range(inner_knots), 3), inner_knots))
  d <- diff(inner_knots)
  g_ab <- splines::splineDesign(knots, inner_knots, derivs = 2)
  knots_mid <- inner_knots[-length(inner_knots)] + d / 2
  g_ab_mid <- splines::splineDesign(knots, knots_mid, derivs = 2)
  g_a <- g_ab[-nrow(g_ab), ]
  g_b <- g_ab[-1, ]
  (crossprod(d * g_a, g_a) +
    4 * crossprod(d * g_ab_mid, g_ab_mid) +
    crossprod(d * g_b, g_b)) /
    6
}

spline_illustration <- local({
  data <- running_mean_data |>
    transmute(x = (x - min(x)) / diff(range(x)), y)
  Phi <- splines::splineDesign(knots, data$x)
  omega <- pen_mat(knot_points)
  lambdas <- 10^seq(-4, 0)
  coefficients <- vapply(
    lambdas,
    function(lambda) {
      drop(solve(
        crossprod(Phi) + lambda * omega,
        crossprod(Phi, data$y)
      ))
    },
    numeric(ncol(Phi))
  )

  basis <- tibble(
    x = rep(xx, times = ncol(b_splines)),
    y = as.vector(b_splines),
    group = rep(names(spline_colors), each = length(xx))
  )
  fits <- lapply(seq_along(lambdas), function(i) {
    components <- sweep(b_splines, 2, coefficients[, i], `*`)
    list(
      components = mutate(basis, y = as.vector(components)),
      curve = tibble(x = xx, y = rowSums(components))
    )
  })

  list(
    data = data,
    basis = basis,
    lambdas = lambdas,
    coefficients = coefficients,
    fits = fits
  )
})

spline_illustration_frame <- function(stage = 3, fit_index = 2) {
  fit <- spline_illustration$fits[[fit_index]]
  basis_plot <- ggplot(
    spline_illustration$basis,
    aes(x, y, color = group)
  ) +
    geom_line() +
    scale_color_manual(values = spline_colors, guide = "none") +
    scale_x_continuous(breaks = c(0, 0.5, 1)) +
    coord_cartesian(ylim = c(0, 1.05)) +
    labs(
      title = "Basis functions",
      x = expression(x),
      y = expression(varphi[j](x))
    )

  component_plot <- ggplot(
    fit$components,
    aes(x, y, color = group)
  ) +
    geom_blank() +
    scale_color_manual(values = spline_colors, guide = "none") +
    scale_x_continuous(breaks = c(0, 0.5, 1)) +
    coord_cartesian(ylim = c(0, 10.5)) +
    labs(
      title = "Weighted functions",
      x = expression(x),
      y = expression(hat(beta)[j] * varphi[j](x))
    )

  fit_plot <- ggplot(spline_illustration$data, aes(x, y)) +
    geom_point(color = "grey45", size = 1.8) +
    scale_x_continuous(breaks = c(0, 0.5, 1)) +
    coord_cartesian(ylim = c(0, 10.5)) +
    labs(
      title = "Fitted curve",
      x = expression(x),
      y = expression(y)
    )

  if (stage >= 2) {
    component_plot <- component_plot + geom_line()
  }
  if (stage >= 3) {
    fit_plot <- fit_plot +
      geom_line(data = fit$curve, color = "royalblue")
  }
  if (stage >= 4) {
    point_basis <- filter(
      spline_illustration$basis,
      near(x, 0.4),
      y > 1e-8
    )
    point_components <- filter(
      fit$components,
      near(x, 0.4),
      abs(y) > 1e-8
    )
    point_fit <- filter(fit$curve, near(x, 0.4))
    basis_plot <- basis_plot +
      geom_vline(
        xintercept = 0.4,
        color = "grey65",
        linetype = "dashed"
      ) +
      geom_point(data = point_basis, size = 2)
    component_plot <- component_plot +
      geom_vline(
        xintercept = 0.4,
        color = "grey65",
        linetype = "dashed"
      ) +
      geom_point(data = point_components, size = 2)
    fit_plot <- fit_plot +
      geom_vline(
        xintercept = 0.4,
        color = "grey65",
        linetype = "dashed"
      ) +
      geom_point(
        data = point_fit,
        color = "royalblue",
        size = 2.5
      )
  }

  basis_plot + component_plot + fit_plot + plot_layout(nrow = 1)
}
```

## From Basis Functions to a Fitted Curve

```{r}
#| label: spline-basis-illustration
#| echo: false
#| fig-show: animate
#| fig-width: 5.5
#| fig-height: 2.8
#| fig-cap: The six observations from the running-mean example, with $x$
#|   rescaled to $[0, 1]$. Here $\lambda = 0.001$.

point_components <- spline_illustration$fits[[2]]$components |>
  filter(near(x, 0.4), abs(y) > 1e-8)
point_equation <- sprintf(
  "At x = 0.4: %s = %.1f (rounded)",
  paste(sprintf("%.1f", point_components$y), collapse = " + "),
  sum(point_components$y)
)
spline_stages <- c(
  "Start with the basis functions",
  "Multiply by the fitted coefficients",
  "Add the weighted functions",
  point_equation
)

for (stage in seq_along(spline_stages)) {
  print(
    spline_illustration_frame(stage) +
      plot_annotation(title = spline_stages[stage])
  )
}
```

\pdfpcnote{
  The colors identify the same eight basis functions as in the preceding
  slide. The left panel stays fixed throughout.

  First show the basis and the observations. On the first click, multiply
  each basis function by its fitted coefficient. The coefficients have
  already been estimated jointly by minimizing the penalized loss.
  They are not the observed responses, and they need not be positive.

  On the second click, add the colored contributions vertically at each x.
  Their sum is the blue fitted curve. This is the operation Phi times beta.

  On the third click, focus on x = 0.4. Only three basis functions are
  nonzero at this knot. Add their weighted values to obtain about 6.2,
  the blue point. Its observed response is 4.

  Here every observation is a knot. With positive lambda, minimizing the
  penalized loss over the full cubic basis gives a natural smoothing spline.
}

## The Penalty Changes the Coefficients

```{r}
#| label: spline-penalty-illustration
#| echo: false
#| fig-show: animate
#| fig-width: 5.5
#| fig-height: 2.8
#| fig-cap: The basis and observations stay fixed. Increasing $\lambda$
#|   penalizes curvature more strongly, giving a smoother fitted curve.

for (i in seq_along(spline_illustration$lambdas)) {
  print(
    spline_illustration_frame(fit_index = i) +
      plot_annotation(
        title = bquote(
          lambda == 10^.(log10(spline_illustration$lambdas[i]))
        )
      )
  )
}
```

\pdfpcnote{
  Each click increases lambda by a factor of ten. The left panel never
  changes: the basis determines which curves we can represent.

  The coefficients are re-estimated for each lambda. Follow the changing
  contributions in the middle and their sum on the right. All axes stay
  fixed across the animation.

  The penalty controls the curvature of the sum. Individual coefficients
  need not shrink toward zero as lambda increases. Straight lines have
  zero curvature, so the fit approaches the least-squares line.

  The next slide shows the penalty matrix used to compute these fits.
}

## Penalty Matrix $\symbf{\Omega}$

```{r}
#| label: Omega-Simpson
#| echo: false
#| fig-cap: The penalty matrix $\symbf{\Omega}$ for a cubic spline with knots at
#|   0, 0.2, ..., 0.8, 1.
#| fig-width: 5
#| fig-height: 3

library(Matrix)
omega <- pen_mat(seq(0, 1, 0.1))

omega_mat <- as.matrix(omega)

n_row <- nrow(omega_mat)
n_col <- ncol(omega_mat)

omega_df <- expand.grid(
  Row = seq_len(n_row),
  Col = seq_len(n_col)
)
omega_df$Value <- as.vector(omega_mat)

ggplot(omega_df, aes(x = Col, y = Row, fill = Value)) +
  geom_tile() +
  scale_y_reverse() +
  scale_fill_gradient2() +
  coord_fixed()
```

## Sparsity in Spline Computations

B-spline basis functions have **local support**: each is nonzero only on a few
neighboring knot intervals. This makes $\symbf{\Phi}$ and $\symbf{\Omega}$
sparse.

\pause\medskip

The coefficients solve
$$
  (\symbf{\Phi}^T \symbf{\Phi} + \lambda \symbf{\Omega}) \hat{\symbf{\beta}}
  = \symbf{\Phi}^T \symbf{y}.
$$
With the basis ordered by knot location, this system is **banded**: its nonzero
entries lie near the diagonal.

\pause\medskip

Solve for $\hat{\symbf{\beta}}$, then compute
$\hat{\symbf{f}} = \symbf{\Phi}\hat{\symbf{\beta}}$. The full smoother matrix
$\symbf{S}_{\lambda}$ is generally dense.

\pause\medskip

In R, `splineDesign(..., sparse = TRUE)` creates a sparse design matrix.

\pdfpcnote{
  At any x, at most four cubic B-spline basis functions are nonzero.

  Basis functions with disjoint supports also have zero entries in the penalty
  matrix. Its band structure is visible in the preceding figure.

  Sparse structure must be reflected in the storage to obtain the computational
  benefits. The following implementation uses dense storage to show the algebra.

  The inverse of a sparse matrix need not be sparse. Solving the system avoids
  explicitly constructing the inverse and the full smoother matrix.
}

## Fitting a Smoothing Spline

We implement the matrix algebra directly to compute
$\symbf{S}_{\lambda} \symbf{y}$.

```{r}
#| label: nuuk-smooth-spline

inner_knots <- nuuk_year$Year

# Note that the order does not matter.
knots <- c(rep(range(inner_knots), 3), inner_knots)

Phi <- splines::splineDesign(knots, inner_knots)
```

. . .

```{r}
#| label: nuuk-smooth-spline2

# Omega can be reused because it depends only on the knots.
omega <- pen_mat(inner_knots)
smoother <- function(lambda) {
  Phi %*%
    solve(
      crossprod(Phi) + lambda * omega, # Phi^T Phi + lambda Omega
      crossprod(Phi, nuuk_year$Temperature) # Phi^T y
    )
}
```

##

```{r}
#| label: nuuk-smoother-plot
#| fig-width: 5
#| fig-height: 2.5

nuuk_plot +
  geom_line(aes(y = smoother(10)), color = "blue") +
  geom_line(aes(y = smoother(1000)), color = "red") +
  geom_line(aes(y = smoother(100000)), color = "orange")
```

## Generalized Cross-Validation (GCV)

For smoothing splines, $S_{ii}$ varies with $i$, which makes LOOCV expensive to
compute.

\medskip\pause

### Idea

Replace $S_{ii}$ in LOOCV with $\mathrm{df} / N$, where
$\mathrm{df} = \operatorname{trace}(\symbf{S}) = \sum_{i=1}^N S_{ii}$. This
gives the **generalized** cross-validation criterion
$$
  \mathrm{GCV}
  = \frac{1}{N} \sum_{i=1}^N \left( \frac{y_i - \hat{f}_i}{1 - \mathrm{df} / N} \right)^2.
$$

. . .

### Implementation

```{r}
#| label: gcv-smooth-spline

gcv <- function(lambda, y) {
  S <- Phi %*% solve(crossprod(Phi) + lambda * omega, t(Phi))
  df <- sum(diag(S))
  mean(((y - S %*% y) / (1 - df / length(y)))^2, na.rm = TRUE)
}
```

\pdfpcnote{
  - The df approximation relies on S_ii being nearly constant.\\
  - When S_ii is constant, GCV coincides with the LOOCV formula.
}

## GCV-Optimal $\lambda$

Apply `gcv()` across a grid of $\lambda$-values and choose the $\lambda$ that
minimizes GCV.

```{r}
#| label: nuuk-spline-gcv

lambda <- seq(50, 250, 2)
gcv <- sapply(lambda, gcv, y = nuuk_year$Temperature)
lambda_opt <- lambda[which.min(gcv)]
```

```{r}
#| label: gc-plot
#| echo: false
#| fig-width: 5
#| fig-height: 2

tibble(lambda, gcv) |>
  ggplot(aes(lambda, gcv)) +
  geom_vline(xintercept = lambda_opt, color = "dark orange") +
  geom_line() +
  labs(y = "GCV Error", x = expression(lambda))
```

## GCV-Optimal Smoothing Spline

```{r}
#| label: nuuk-spline-opt
#| fig-width: 5
#| fig-height: 2.5

smooth_opt <- smoother(lambda_opt)
nuuk_plot + geom_line(aes(y = smooth_opt), color = "blue")
```

## Using `smooth.spline()`

```{r}
#| label: nuuk-spline-opt2
#| fig-width: 5
#| fig-height: 2.5

smooth <- smooth.spline(nuuk_year$Year, nuuk_year$Temperature)
nuuk_plot + geom_line(aes(y = smooth$y), color = "blue")
```

\pdfpcnote{
- Disable the fast heuristic to obtain results similar to ours.
}

## Efficient Computations

When $N$ is large, computing $\symbf{S}_{\lambda} \symbf{y}$ directly is
expensive.

\medskip\pause

In practice, we therefore often use fewer basis functions than data points,
$p < N$.

\pause

### Singular Value Decomposition

Using the singular value decomposition
$$
  \Phi = \symbf{U} D \symbf{V}^T
$$
\pause we can rewrite the smoother matrix as
$$
  \symbf{S}_{\lambda}
  = \widetilde{\symbf{U}} (I + \lambda \Gamma)^{-1} \widetilde{\symbf{U}}^T
$$
where $\widetilde{\symbf{U}} = \symbf{U} \symbf{W}$ and \pause
$$
  D^{-1} \symbf{V}^T \symbf{\Omega} \symbf{V} D^{-1}
  = \symbf{W} \Gamma \symbf{W}^T.
$$

\medskip\pause

The columns of $\widetilde{\symbf{U}}$ form the **Demmler-Reinsch basis**.

\pdfpcnote{
  This can be done whether we use all or a subset of the basis functions.

  The smoother matrix S_\lambda is diagonalized.

  Derive $\symbf{S}_\lambda$ on blackboard.
}

# Exercises

## Exercises

### Exercise 1: Parallel Square

Use the `foreach` and `doParallel` packages to compute the squares of a vector
in parallel. Benchmark the parallel version against a regular `for` loop using
`bench::mark()`.

```{r}
#| label: parallel-square
#| echo: false
#| eval: false

library(foreach)
library(doParallel)
library(bench)

x <- rnorm(10000)

loop_square <- function(x) {
  n <- length(x)
  out <- numeric(n)

  for (i in seq_len(n)) {
    out[i] <- x[i]^2
  }

  out
}

parallel_square <- function(x) {
  out <- foreach(i = seq_along(x), .combine = c) %dopar%
    {
      x[i]^2
    }
  out
}

cl <- makeCluster(detectCores() / 2)
registerDoParallel(cl)

bench::mark(
  loop = loop_square(x),
  parallel = parallel_square(x),
  check = FALSE
)
```

. . .

### Exercise 2: Smoothing Air Quality Data

Apply `stats::filter()` and `smooth.spline()` to the `airquality$Temp` data. Try
different window sizes for the moving average and different `spar` values for
the spline. Plot and compare the results.

## Summary

### Parallelization

- Parallelization delegates work across your CPU cores to speed up
  computations.\pause
- Use `foreach` and `%dopar%` to parallelize computations.

. . .

### Smoothing Splines

- A smoothing spline is a natural cubic spline that minimizes a penalized least
  squares criterion.\pause
- The penalty term controls the trade-off between goodness of fit and
  wiggliness.

## Next Time

### Pseudo-Random Numbers

How do you simulate random numbers in R?

. . .

### Rejection Sampling

General and useful method for sampling from a target distribution

\appendix

# Appendix

## Interpretation

Project the data $\symbf{y}$ onto the Demmler-Reinsch basis (the columns of
$\widetilde{\symbf{U}}$) to obtain the coefficients
$\hat{\symbf{\beta}} = \widetilde{\symbf{U}}^T \symbf{y}$.

\pause\medskip

Each coefficient is shrunk according to the corresponding eigenvalue $\gamma_i$
and smoothing parameter $\lambda$:
$$
  \hat{\symbf{\beta}}_i(\lambda) = \frac{\hat{\beta}_i}{1 + \lambda \gamma_i}.
$$

\pause\medskip

The final smoothed values are reconstructed by combining the shrunk coefficients
with the basis functions:
$$
  \hat{\symbf{f}} = \widetilde{\symbf{U}} \hat{\symbf{\beta}}(\lambda).
$$

\pdfpcnote{
  The $\hat\beta_i$ are the original coefficients when $\lambda = 0$.
}

## The Demmler-Reinsch Basis (Columns of $\widetilde{\symbf{U}}$)

```{r}
#| label: spline-diagonalization
#| echo: false
#| fig-width: 5.5
#| fig-height: 3

inner_knots <- seq(1867, 2013, length.out = 18)
Phi <- splines::splineDesign(
  c(rep(range(inner_knots), 3), inner_knots),
  nuuk_year$Year
)
omega <- pen_mat(inner_knots)
Phi_svd <- svd(Phi)
omega_tilde <- t(
  t(crossprod(Phi_svd$v, omega %*% Phi_svd$v)) / Phi_svd$d
) /
  Phi_svd$d

Omega_tilde_svd <- svd(omega_tilde)
U_tilde <- Phi_svd$u %*% Omega_tilde_svd$u

colnames(U_tilde) <- paste0(rep("u", 20), 1:20)
bind_cols(select(nuuk_year, Year), as_tibble(U_tilde)) |>
  gather(
    key = "term",
    value = "value",
    -Year,
    factor_key = TRUE
  ) |>
  ggplot(aes(Year, value, color = term)) +
  geom_line() +
  facet_wrap(~term) +
  theme(legend.position = "none") +
  ylab("")
```

##

```{r}
#| label: spline-gamma
#| echo: false
#| fig-width: 5.5
#| fig-height: 3
#| fig-cap: The eigenvalues $\gamma_i$ of the penalty matrix in the
#|   Demmler-Reinsch basis.

library(patchwork)

d <- Omega_tilde_svd$d

eigen_data <- tibble(i = 1:20, Eigenvalues = d)

p1 <- ggplot(eigen_data, aes(i, Eigenvalues)) +
  geom_point() +
  labs(y = expression(gamma[i]))

p2 <- p1 + scale_y_log10()

p1 / p2 + plot_layout(axes = "collect")
```
