---
title: "Standard Deviation with Rcpp: Exercise Solution"
format: html
execute:
  cache: false
---

Solution to the exercise in [Lecture 12](../slides/lecture12.pdf).

```{r}
#| label: rcpp-sd-setup
#| include: false

library(Rcpp)
```

## Implementation

R's `sd()` computes the sample standard deviation,
$$
  s = \sqrt{\frac{1}{n - 1}\sum_{i=1}^n (x_i - \bar{x})^2},
  \qquad
  \bar{x} = \frac{1}{n}\sum_{i=1}^n x_i.
$$
The denominator is $n - 1$, rather than $n$. When $n < 2$, the sample standard
deviation is undefined, and `sd()` returns `NA`.

We use two loops: the first computes the mean, and the second sums the squared
deviations. Computing deviations from the mean avoids subtracting two large,
nearly equal sums, as in $\sum_i x_i^2 - n\bar{x}^2$. This matters when the
observations have a large mean but little variation.

```{Rcpp}
//| label: rcpp-sd-solution

#include <Rcpp.h>
#include <cmath>

// [[Rcpp::export]]
double
cpp_sd(const Rcpp::NumericVector& x)
{
  const R_xlen_t n = x.size();
  if (n < 2) {
    return NA_REAL;
  }

  double mean = 0.0;
  for (R_xlen_t i = 0; i < n; ++i) {
    if (Rcpp::NumericVector::is_na(x[i])) {
      return NA_REAL;
    }
    mean += x[i] / n;
  }

  double sum_squared = 0.0;
  for (R_xlen_t i = 0; i < n; ++i) {
    const double deviation = x[i] - mean;
    sum_squared += deviation * deviation;
  }

  return std::sqrt(sum_squared / (n - 1));
}
```

The Rcpp chunk compiles the function and makes `cpp_sd()` available in R.
Alternatively, save the code in a `.cpp` file and load it with
`Rcpp::sourceCpp()`. The `const` reference helps prevent accidental modification
of the input. We return `NA_REAL` for short vectors and missing values, matching
the default behavior of `sd()`. This function does not implement `na.rm = TRUE`.

## Check against R

We check ordinary data, constant data, large values with small differences,
missing values, and the two short-vector cases. Floating-point results can
differ slightly, so we use `all.equal()` rather than requiring exact equality.

```{r}
#| label: rcpp-sd-checks

set.seed(1)
inputs <- list(
  ordinary = c(-3, 0, 2, 8),
  random = rnorm(1000),
  constant = rep(2, 10),
  large_mean = 1e10 + c(-2, -1, 0, 1, 2),
  missing = c(1, NA_real_, 3),
  nan = c(1, NaN, 3),
  empty = numeric(),
  singleton = 1
)

for (x in inputs) {
  stopifnot(isTRUE(all.equal(cpp_sd(x), sd(x))))
}

data.frame(
  input = names(inputs),
  R = vapply(inputs, sd, numeric(1)),
  Rcpp = vapply(inputs, cpp_sd, numeric(1)),
  row.names = NULL
)
```

The empty vector and singleton both return `NA`, and constant data return zero.
All comparisons above must pass for the document to render successfully.

## Benchmark

We compile the function and generate the input before benchmarking, so the
timings measure calls to the functions. `bench::mark()` also checks that their
results agree.

```{r}
#| label: rcpp-sd-benchmark

set.seed(2)
x <- rnorm(1e5)

timings <- bench::mark(
  R = sd(x),
  Rcpp = cpp_sd(x),
  min_iterations = 20
)
timings[, c("expression", "median", "itr/sec", "mem_alloc")]
```

```{r}
#| label: rcpp-sd-speedup
#| include: false

speedup <- as.numeric(timings$median[1] / timings$median[2])
comparison <- if (speedup >= 1) {
  sprintf("runs %.2f times as fast as", speedup)
} else {
  sprintf("takes %.2f times as long per call as", 1 / speedup)
}
```

The median reports the typical time per call; `itr/sec` reports calls per
second. A smaller median indicates a faster implementation. The memory column
measures allocations tracked by R, so it does not include every C++ allocation.

In this run, the Rcpp implementation `r comparison` R's `sd()`, based on the
median times.

The Rcpp version is not guaranteed to win: R's `sd()` already uses compiled code
to compute the variance. Results depend on the input size, compiler, and
machine. Try changing the vector length to see when the cost of calling C++
matters relative to the work inside the function.
