---
title: "Optimization: Exercise Solution"
format: html
execute:
  cache: false
---

Solution to the exercise in [Lecture 7](../slides/lecture7.pdf).

## Derivatives

For $H(\theta) = \log(1 + e^\theta) + \frac{1}{2}(\theta - 2)^2$,
$$
  H'(\theta) = \frac{e^\theta}{1 + e^\theta} + \theta - 2,
  \qquad
  H''(\theta) = \frac{e^\theta}{(1 + e^\theta)^2} + 1.
$$

```{r}
#| label: derivatives

H_prime <- function(theta) {
  exp(theta) / (1 + exp(theta)) + theta - 2
}
H_double_prime <- function(theta) {
  exp(theta) / (1 + exp(theta))^2 + 1
}
```

## Gradient descent

Fill in the gradient, stopping test, and update:

```{r}
#| label: gradient-descent
#| fig-width: 5
#| fig-height: 2.5

theta <- -2
path <- theta
for (i in seq_len(100)) {
  g <- H_prime(theta)
  if (abs(g) <= 1e-8) {
    break
  }
  theta <- theta - 0.8 * g
  path <- c(path, theta)
}
plot(path, type = "b")
```

## Newton's method

Restart from `theta <- -2` and replace the update with

```{r}
#| label: newton-update
#| eval: false

theta <- theta - g / H_double_prime(theta)
```

Gradient descent takes **7 updates**, whereas Newton takes **4**. Count updates
with `length(path) - 1`, since the path includes the starting point. Both reach
$\theta^* \approx 1.22675$; Newton uses the local curvature to converge faster
near the minimum.
