---
title: "Absolute-Value Constraints: Exercise Solution"
format: html
execute:
  cache: false
---

Solution to the exercise in [Lecture 9](../slides/lecture9.pdf).

We want to solve
$$
  \begin{aligned}
     & \operatorname*{minimize}_{\theta \in \mathbb{R}}
     &                                                  & H(\theta) = \frac{1}{2}(\theta - 2)^2 \\
     & \text{subject to}                                &                                       & |\theta| \leq 1.
  \end{aligned}
$$

## Standard form

The constraint $|\theta| \leq 1$ is equivalent to $-1 \leq \theta \leq 1$. In
standard form, we write both inequalities as $g_i(\theta) \leq 0$:
$$
  \begin{aligned}
     & \operatorname*{minimize}_{\theta \in \mathbb{R}}
     &                                                  & \frac{1}{2}(\theta - 2)^2 \\
     & \text{subject to}                                &                           & g_1(\theta) = \theta - 1 \leq 0,  \\
     &                                                  &                           & g_2(\theta) = -\theta - 1 \leq 0.
  \end{aligned}
$$

## Constraints for `constrOptim()`

`constrOptim()` expects constraints of the form `ui %*% theta >= ci`. Writing
our constraints as $\theta \geq -1$ and $-\theta \geq -1$ gives
$$
  \underbrace{\begin{bmatrix} 1 \\ -1 \end{bmatrix}}_{\texttt{ui}} \theta
  \succeq
  \underbrace{\begin{bmatrix} -1 \\ -1 \end{bmatrix}}_{\texttt{ci}},
$$
where $\succeq$ means that each component on the left is at least the
corresponding component on the right. Each row of `ui` represents one
constraint, and its single column corresponds to our one parameter.

```{r}
#| label: constraints

ui <- rbind(1, -1)
ci <- c(-1, -1)
```

## Numerical solution

The derivative is $H'(\theta) = \theta - 2$. We supply both the objective and
its gradient, and start at $\theta_0 = 0.5$. The starting point must be strictly
feasible: `ui %*% theta0 > ci`, or $-1 < \theta_0 < 1$.

```{r}
#| label: constrained-optimization

H <- function(theta) 0.5 * (theta - 2)^2
gradient <- function(theta) theta - 2

theta0 <- 0.5
fit <- constrOptim(
  theta = theta0,
  f = H,
  grad = gradient,
  ui = ui,
  ci = ci
)

fit[c("par", "value", "convergence")]
```

The estimated parameter is close to $1$, and the objective value is close to
$1/2$. A convergence code of zero indicates successful convergence. The barrier
method approaches the boundary from inside the feasible interval, so the
numerical estimate need not equal $1$ exactly.

## Checking the solution

The unconstrained minimum is at $\theta = 2$, which is outside $[-1, 1]$.
Throughout the feasible interval, $H'(\theta) = \theta - 2 < 0$, so the
objective decreases as $\theta$ increases. The constrained minimum is therefore
at the right endpoint:
$$
  \theta^* = 1, \qquad H(\theta^*) = \frac{1}{2}.
$$
